---
title: "Pipeline walkthrough"
execute:
  eval: false
---

This article shows the package-level workflow. The examples are intentionally
not evaluated during site builds because fitting and nested validation are
long-running operations.

## Train a deployment kit

```{r}
library(PAGe)

historical <- load_flu_hist()
m0 <- build_m0(historical)
m1 <- build_m1(historical, m0)
m2 <- train_m2(historical, m0, m1, best_spec = NULL)
kit <- assemble_kit(m0, m1, m2)

saveRDS(kit, "m2_production.rds")
```

Retuning is optional when using the production defaults:

```{r}
m0_tuning <- tune_m0(historical)
m1_tuning <- tune_m1(historical, m0, m1)
```

## Run the weekly pipeline

Current-season data must use the same field definitions and feature scales as
the training data. Given a trained kit, run all stages sequentially:

```{r}
kit <- load_prospective_kit("m2_production.rds")
current <- getCurrentD(season = "2025-26")
forecast <- run_pipeline(kit, current)
plot_forecast(forecast, history = historical)
```

For operational systems, archive the issue date, input snapshot, kit version,
manual overrides, and returned M0/M1/M2 state. This makes every forecast
reproducible and prevents revised surveillance data from silently changing
historical evaluations.
