---
title: "Pipeline walkthrough"
execute:
  eval: false
---

The examples are not executed during package builds because fitting and nested
validation are long-running.

## Prepare data and refresh

```{r}
library(PAGe)

historical <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()
training <- train_pipeline(historical, mode = "refresh")
kit <- training$kit
```

PAGe does not bundle surveillance observations. The default prospective
holdout is 2025-26, which remains outside every fit until promotion.

## Replay and promote

```{r}
candidate <- replay_season_holdout(kit, historical, season = "2025-26")
incumbent <- replay_season_holdout(incumbent_kit, historical, season = "2025-26")
promotion <- check_promotion(candidate$metrics, incumbent$metrics)
next_training <- train_pipeline(historical, mode = "refresh", promotion = promotion)
```

## Retune and select

```{r}
retuned <- train_pipeline(
  historical,
  mode = "retune",
  previous_results = prior_m2_results,
  selection_method = "min_nll", # or "one_se" / "pareto"
  racing = FALSE
)
```

The adaptive grid uses compatible prior results and expands around promising
boundaries. Optional racing requires a callback and cannot replace final full
nested-LOSO evaluation.

## Forecast

```{r}
current <- prepare_surveillance_data(current_csv, season = "2026-27")
forecast <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(forecast, history = historical)
```

Frozen-GAM prediction is the default; weekly refitting is an explicit legacy
compatibility option.
