---
title: "Pipeline walkthrough"
execute:
  eval: false
---

::: {.callout-caution}
The API snippets below illustrate model development only. An in-memory
`check_promotion()` result is not immutable release evidence. For a real
release, use `docs/deployment-workflow.qmd` and preserve the acceptance,
fixed-spec refresh, and deployment manifests before loading the promoted kit.
:::

The examples are not executed during package builds because fitting and nested
validation are long-running.

## Prepare data and declare season sets

```{r}
library(PAGe)

historical <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()
selection <- validate_season_selection(
  historical,
  training_seasons = development_seasons,
  exclude_seasons = excluded_seasons,
  holdout_seasons = "2025-26",
  application_seasons = "2026-27"
)
```

PAGe does not bundle surveillance observations. The four season sets are
explicit and disjoint; 2025-26 remains outside every development fit.

## Guard M0, M1, and M2 independently

```{r}
m0_tuning <- tune_m0(historical, selection = selection)
validate_m0_tuning(m0_tuning)
m0 <- fit_m0(historical, selection, m0_tuning$best_params) |>
  freeze_m0(m0_tuning)

m1_tuning <- tune_m1(historical, m0 = m0, selection = selection)
validate_m1_tuning(m1_tuning)
m1_config <- as.list(m1_tuning$best[1, , drop = FALSE])
m1 <- fit_m1(historical, selection, m0, m1_config) |>
  freeze_m1(m1_tuning)

m2_tuning <- tune_m2(historical, selection, m0, m1, m2_grid)
validate_m2_tuning(m2_tuning)
m2 <- fit_m2(historical, selection, m0, m1, m2_tuning$best_spec) |>
  freeze_m2(m2_tuning)

kit <- assemble_kit(m0, m1, m2)
validate_page_kit(kit)
```

Tuning, fitting, and freezing are distinct states. M1 requires a frozen,
selection-matched M0; M2 requires the exact frozen M0/M1 chain; governed kit
assembly rejects drafts and identity mismatches.

For a locked-spec compatibility refresh:

```{r}
training <- train_pipeline(historical, mode = "refresh")
kit <- training$kit
```

`train_pipeline()` has not yet been refactored to compose the guarded lifecycle
above.

## Replay and promote

Governed holdout evaluation uses
`scripts/acceptance/replay_2025_26.R`, not a bare in-memory
`check_promotion()` report. The acceptance script writes a private decision
bundle and disclosure-safe manifest that bind the authorized data, candidate,
and incumbent by SHA-256.

After a passing decision, run `season2526/run_retrain_venkata.R` with those
saved artifacts and both kit paths. It verifies the hashes before constructing
the short-lived promotion evidence accepted by `train_pipeline()`. R classes
can be forged, so the retained and verified artifacts are the production
safety boundary.

## Retune and select

```{r}
retuned <- train_pipeline(
  historical,
  mode = "retune",
  previous_results = prior_m2_results,
  selection_method = "min_nll", # or "one_se" / "pareto"
  racing = FALSE
)
```

The adaptive grid uses compatible prior results and expands around promising
boundaries. Optional racing requires a callback and cannot replace final full
nested-LOSO evaluation.

### Tuning tips

A winning parameter should usually be bracketed by tested values. When it sits
at a lower or upper grid edge, retain the current candidates and add one valid
level beyond that edge using adjacent observed spacing. Rerun the same complete
LOSO folds and selection rule. Start with one-factor local neighbors; add
targeted interactions only when fold results suggest them.

Boundary values can be legitimate. Zero may mean an optional M2 smooth is off,
and probability, count, window, and basis-size parameters have hard valid
domains. Record each boundary as expanded, accepted as a null/constraint, or
unresolved at the computational cap.

For the default M1 grid, lower-edge selections at `k_ref=25` and
`slope_weight=8` suggest controlled next values of `20` and `4`, respectively,
if they remain selected after complete validation. For M2,
`plan_m2_grid(previous_results, max_finalists, max_specs)` retains the
incumbent and finalists, adds local neighbors, and expands valid winning
boundaries one adjacent step.

Stop when material axes are bracketed, a boundary has a documented
null/constraint interpretation, improvement is flat relative to fold
uncertainty, or the predeclared cap is reached. Finish all expansion before
opening the prospective holdout.

## Forecast

```{r}
kit <- load_promoted_kit(
  kit_path =
    "/secure/PAGe/deployment-registry/<deployment-id>/promoted_kit.rds",
  deployment_manifest_path =
    "results/deployment-audit/<deployment-id>/deployment_manifest.json"
)
current <- prepare_surveillance_data(current_csv, season = "2026-27")
forecast <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(forecast, history = historical)
```

Frozen-GAM prediction is the default; weekly refitting is an explicit legacy
compatibility option.
