---
title: "Pipeline walkthrough"
execute:
  eval: false
---

::: {.callout-caution}
The API snippets below illustrate model development only. An in-memory
`check_promotion()` result is not immutable release evidence. For a real
release, use `docs/deployment-workflow.qmd` and preserve the acceptance,
fixed-spec refresh, and deployment manifests before loading the promoted kit.
:::

The examples are not executed during package builds because fitting and nested
validation are long-running.

## Prepare data and refresh

```{r}
library(PAGe)

historical <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()
training <- train_pipeline(historical, mode = "refresh")
kit <- training$kit
```

PAGe does not bundle surveillance observations. The default prospective
holdout is 2025-26, which remains outside every fit until promotion.

## Replay and promote

Governed holdout evaluation uses
`scripts/acceptance/replay_2025_26.R`, not a bare in-memory
`check_promotion()` report. The acceptance script writes a private decision
bundle and disclosure-safe manifest that bind the authorized data, candidate,
and incumbent by SHA-256.

After a passing decision, run `season2526/run_retrain_venkata.R` with those
saved artifacts and both kit paths. It verifies the hashes before constructing
the short-lived promotion evidence accepted by `train_pipeline()`. R classes
can be forged, so the retained and verified artifacts are the production
safety boundary.

## Retune and select

```{r}
retuned <- train_pipeline(
  historical,
  mode = "retune",
  previous_results = prior_m2_results,
  selection_method = "min_nll", # or "one_se" / "pareto"
  racing = FALSE
)
```

The adaptive grid uses compatible prior results and expands around promising
boundaries. Optional racing requires a callback and cannot replace final full
nested-LOSO evaluation.

## Forecast

```{r}
kit <- load_promoted_kit(
  kit_path =
    "/secure/PAGe/deployment-registry/<deployment-id>/promoted_kit.rds",
  deployment_manifest_path =
    "results/deployment-audit/<deployment-id>/deployment_manifest.json"
)
current <- prepare_surveillance_data(current_csv, season = "2026-27")
forecast <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(forecast, history = historical)
```

Frozen-GAM prediction is the default; weekly refitting is an explicit legacy
compatibility option.
