---
title: ""
format:
  html:
    page-layout: full
    grid:
      body-width: 1200px
editor: visual
---

```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(mgcv)
library(gamm4)
library(gratia)
library(ggplot2)
library(plotly)
library(kableExtra)
library(purrr)
library(glue)
library(quartabs)
library(htmltools)
library(data.table)
library(ISOweek)
library(MMWRweek)
source("helpers.R")
startWeek = 27 ##week number to start the surveillance season
season = "2025-26"
grid = data.frame(weekF = seq(1,52, by = 1)) ## grid for predictions
min_window = 10 ## minimum time window to the start/end of the data for ignition point detection
df = 10 ## degrees of freedom for the GAM smoothing

#####season ignition parameters
p_thresh = 0.01 #at least 1% to reach ignition
k = 0.002 # min slope for ignition at least 0.1% increase in positivity




```

## Data processing

```{r}
#| warning: false


#fluWeek = read.csv("allweeks.csv") %>% select(week = fluweek,year, date = week_start_date) %>% 
#  mutate(week = as.integer(week), year = as.integer(year), date = as.Date(date,"%d-%b-%y")) %>% distinct()
  

theDs = read.csv("positive.csv") %>% left_join(read.csv("test.csv")) %>% setNames(c("week","y","season","N")) %>% 
  mutate(y = as.integer(gsub(",", "", y)), N = as.integer(gsub(",", "", N)),  neg = N-y, season = as.factor(season)) %>%  
  filter(!season %in% c("2022-23", "2023-24"))

#OLIS data
# theDD = read.csv("Influenza_OLIS.csv") %>%rename(week = Surveillance.week) %>%
#                                       mutate(date = as.Date(Week.start.date, format="%d-%b-%y"),
#                                              N = as.integer(Total.number.of.tests),
#                                              y = as.integer(Total.number.of.positive.tests),
#                                              neg = N -y,
#                                              season = as.factor(Surveillance.period), 
#                                              date = as.character(date)) %>% 
#                                       select(c(date,names(theDs)))


theDD = read.csv("weeklyTotal.csv") %>% left_join(read.csv("weeklyPos.csv")) %>% 
                                            rename(week = Surveillance.week) %>%
                                             mutate(N = as.integer(Total.number.of.tests),
                                             y = as.integer(Total.number.of.positive.tests),
                                             neg = N -y,
                                             season = as.factor(Surveillance.period)) %>% 
                                      dplyr::select(names(theDs))
  


theD = rbind(theDs, theDD %>% filter(season %in% c("2022-23", "2023-24", "2024-25"))) %>%
  mutate(season =  droplevels(season)) %>%
              group_by(season) %>% 
              mutate( p = y/N,
                      start_year = as.integer(substr(season, 1, 4)),
                      mmwr_year  = ifelse(week >= 35L, start_year, start_year + 1L),
                      Rdate      = MMWRweek2Date(mmwr_year, week, 1L),
                      nW_true    = n_weeks_in_start_year(start_year),
                      weekS      = ((week - 35L) %% nW_true) + 1L,
                      weekF      = ((week - startWeek) %% nW_true) + 1L,
                      cYear      = as.factor(lubridate::year(Rdate))
                     )  %>%  ungroup() %>% arrange(season, weekS, Rdate) %>% rename(date = Rdate)






#current "season", with around 8 weeks from last season
currentSeason = theDD %>% filter(season %in% c("2024-25", "2025-26")) %>%
  mutate(season =  droplevels(season)) %>%
              group_by(season) %>% 
              mutate( p = y/N,
                      start_year = as.integer(substr(season, 1, 4)),
                      mmwr_year  = ifelse(week >= 35L, start_year, start_year + 1L),
                      Rdate      = MMWRweek2Date(mmwr_year, week, 1L),
                      nW_true    = n_weeks_in_start_year(start_year),
                      weekS      = ((week - 35L) %% nW_true) + 1L,
                      weekF      = ((week - startWeek) %% nW_true) + 1L,
                      cYear      = as.factor(lubridate::year(Rdate))
                     )  %>%  ungroup() %>% arrange(season, weekS, Rdate) %>% 
                  filter(week >= startWeek, Rdate > as.Date("2025-06-20")) %>% rename(date = Rdate)
 

save(theD, currentSeason, file = "flu.RData")
```

### Data overview

::: panel-tabset
### Raw data by calender week and year

```{r}
ggplotly(ggplot(theD, aes(x = week, y = p, group = cYear, color = cYear)) + geom_point() + geom_line())
```

### by flu week and flu season

```{r}
ggplotly(ggplot(theD, aes(x = weekS, y = p, group = season, color = season)) + geom_point() + geom_line() +
  xlab("Season Week: aligned to start at 35"))
```

### Starting in `r startWeek`

Create new week indicator using `r startWeek` as week 1

```{r}

p1 = ggplotly(ggplot(theD, aes(x = weekF, y = p, group = season, color = season)) + geom_point() + geom_line() +xlab("Week starting in July"))
p1
```
:::

## Ignition detection

### Smoothing by season

Apply gam for smoothing, for each season, find 1st and 2nd derivatives

```{r}
#| warning: false

gam_fit <- pred <- d1 <- d2 <-as.list(unique(theD$season)) %>% setNames(unique(theD$season)) # create empty list for gam fits
predD = theD %>%  add_column(fit = NA, se.fit = NA)

for (s in unique(theD$season)) {
  gam_fit[[s]] <- gam(cbind(y, neg) ~ s(weekF, k = df, bs = "ps"), data = predD %>% filter(season == s), family = "binomial", method = "REML")
  pred[[s]] <- predict(gam_fit[[s]], newdata = predD%>% filter(season == s), type = "response", se.fit = T)
  predD$fit[predD$season == s] = pred[[s]]$fit
  predD$se.fit[predD$season == s] = pred[[s]]$se.fit
  
  d1[[s]] <- gratia::derivatives(gam_fit[[s]], intervel = "simultaneous", order = 1, data = predD %>% filter(season ==s)) %>% select(.derivative, .lower_ci, .upper_ci) %>% rename(d1 = .derivative, d1.lower = .lower_ci, d1.upper = .upper_ci)
  
  d2[[s]] <- gratia::derivatives(gam_fit[[s]], intervel = "simultaneous", order = 2, data =  predD %>% filter(season ==s)) %>% select(.derivative, .lower_ci, .upper_ci) %>% rename(d2 = .derivative, d2.lower = .lower_ci, d2.upper = .upper_ci)
  
}

predD = predD %>% mutate(binom.se = sqrt(fit*(1 - fit) / N),
                         total.se = sqrt(binom.se^2 + se.fit^2),
                         low = fit - 1.96*total.se,
                         high = fit + 1.96*total.se) %>% mutate(low = ifelse(low<0, 0, low), 
                                                                season = as.factor(as.character(season)))


    
```

### Model results

::: panel-tabset
### Observed vs smoothed

```{r}
 ggplotly(ggplot(predD, aes(x = weekF, y = p, group = season, color = season)) + geom_point() +
            xlab("Week") + 
            geom_line(aes(y = fit), linewidth = 1) +
            geom_ribbon(aes(x = weekF, ymin = low, ymax = high), alpha = 0.2) +
            ylab("Proportion of positive tests") )
```

### Model summary

```{r}
for (s in unique(theD$season)) {
  cat("----------------------------------------------------------------------\n")
  print(paste("Season:", s))
  print(summary(gam_fit[[s]]))
}
```

### GAM plot

```{r}
for (s in unique(theD$season)) {
  print(paste("Season:", s))
  print(gratia::draw(gam_fit[[s]], residuals = TRUE, rug = TRUE) + ggtitle(paste("GAM fit for season", s)))
}
```
:::

### Derivatives

Estimate derivatives of the GAM fit, 1st and 2nd order, on the response scale

```{r}

allD <- pmap(
  .l = list(split(predD, predD$season), d1,d2),
  .f = bind_cols
) %>% lapply( \(x) {
  x %>% mutate(d1_resp = fit * (1 - fit) * d1, 
               d2_resp = fit * (1 - fit) * d2+ fit * (1 - fit) * (1 - 2*fit) * d1^2,
               d1_lo_resp = fit * (1 - fit) * d1.lower,
               d1_up_resp = fit * (1 - fit) * d1.upper,
               d2_lo_resp = fit * (1 - fit) * d2.lower +fit * (1 - fit) * (1 - 2*fit) * (d1.lower^2),
               d2_up_resp = fit * (1 - fit) * d2.upper + fit * (1 - fit) * (1 - 2*fit) * (d1.upper^2),
               slope_rank  = rank(desc(d1_resp)),
               acc_rank = rank(desc(d2_resp)),
               signs1 = sign(d1_resp),
               signs2 = sign(d2_resp),
               concaveDown = c(NA,(signs2[-1] == -1 & signs2[-length(signs2)] == 1)), #this is the first negative after the zero crossing
               concaveUp = c(NA,(signs2[-1] == 1 & signs2[-length(signs2)] == -1)),
               increase = c(NA,(signs1[-1] == -1 & signs1[-length(signs1)] == 1)), 
               decrease = c(NA,(signs1[-1] == 1 & signs1[-length(signs1)] == -1)))%>% select(-c(signs1, signs2)) 
  
})  %>% lapply(flagIgnition)  


plots = allD
for (s in names(allD)) {
toPlot = allD[[s]]
plots[[s]] = ggplotly(ggplot(toPlot %>% filter(flag_window), aes(x = weekF, y = p)) + geom_point(alpha = 0.5)+
           geom_line(aes(y = fit),linewidth = 2)  +geom_hline(yintercept = 0) +
           geom_line(aes(y = d1_resp), color = "red", linewidth = 1) +
           geom_line(aes(y = d2_resp), color = "blue", linewidth = 1) +
           geom_vline(data = toPlot %>% filter(flag), aes(xintercept = weekF), color = "green", alpha = 0.5) +   
           xlab("Week") + ylab("Derivative and the GAM fit") + ggtitle(paste("Season", s)))
}
```

### Ignition detection {.tabset}

In order to be an ignition point leading to the start of the season, candidate points must

-   At least `r min_window` weeks after the start of the season, and before end of the season
-   Have a positivity of at least `r p_thresh * 100`%
-   Have a positive first derivative bigger than `r k * 100`% indicating an upward trend
-   Have a positive second derivative indicating an acceleration
-   If multiple ignition points are detected, the earliest one is selected

```{r gen-tabs,  results='asis'}

df1 <- tibble(
  name   = names(plots),
  widget = map(plots, htmltools::div))

df1 |> render_tabset(
  tabset_vars = name,
  output_vars = widget
)
```

## Inflection point detection

### Align seasons at ignition point

::: panel-tabset
### After

```{r}
#| echo: false
theD = allD %>% rbindlist() %>% arrange(season, weekF)  %>% mutate(anchorWeek = median(iWeek),offset = iWeek- anchorWeek, newWeek = ((weekF - offset - 1) %% 52) + 1)

ggplotly(ggplot(theD, aes(x = newWeek, y = p, group = season, color = season),alpha = 0.5) + geom_point() + geom_line()  +
                      geom_vline(xintercept = theD$anchorWeek[1], linetype = "dashed") +
                      geom_hline(yintercept = c(p_thresh,0.05)) +
                      xlab(paste("Week (aligned ignition at", theD$anchorWeek[1], ")")))
```

### Before

```{r}
#| echo: false
p1
```
:::

### Data smoothing

::: panel-tabset
### plot

```{r}
#| warning: false
grid = data.frame(newWeek = seq(1,52, by = 1)) 
mod2 <- gamm4(
  cbind(y, neg) ~ 
    s(newWeek,k=df),      # one global smooth
    random = ~(1 | season),
  data   = theD,
  family = binomial(),
  method = "REML"
)

pred = predict(mod2$gam, newdata = grid %>% add_column(season = "fit"), type = "response", se.fit = T)

predD = grid %>% add_column(N = tapply(theD$N, theD$newWeek, median),
                           fit = pred$fit,
                           se.fit = pred$se.fit,
                           binom.se = sqrt(fit*(1 - fit) / N),
                           total.se = sqrt(binom.se^2 + pred$se.fit^2),
                           low = pred$fit - 1.96*total.se,
                           high = pred$fit + 1.96*total.se,
                           p = 0) %>% mutate(low = ifelse(low<0, 0, low), season = "fit")

ggplotly(ggplot(theD, aes(x = newWeek, y = p, group = season, color = season)) + 
            geom_point(alpha = 0.5)  + geom_line(alpha = 0.5)+
            xlab("Week aligned") + 
            geom_line(data = predD, aes(x = newWeek, y = fit), color = "black", linewidth = 1.5) +
            geom_ribbon(data = predD, aes(x = newWeek, ymin = low, ymax = high), alpha = 0.2) +
            geom_vline(xintercept = theD$anchorWeek[1], linetype = "dashed") +
            ylab("Proportion of positive tests") + ggtitle("GAM fit with common trend"))

```

### Model summary

```{r}
summary(mod2)
```
:::

### Derivatives

```{r}
mod2 = mod2$gam

seasons = levels(theD$season)
d1 <- gratia::derivatives(mod2, intervel = "simultaneous", order = 1, data = predD %>% mutate(season = seasons[1]), select = "s(newWeek)") %>% select(.derivative, .lower_ci, .upper_ci) %>% rename(d1 = .derivative, d1.lower = .lower_ci, d1.upper = .upper_ci)%>% 
     add_column(fit = predD$fit, season = seasons[1]) %>%
     mutate(d1_resp = fit * (1 - fit) * d1, 
            d1_lo_resp = fit * (1 - fit) * d1.lower,
            d1_up_resp = fit * (1 - fit) * d1.upper)%>% select(d1_resp, d1_lo_resp, d1_up_resp)
  
d2 <- gratia::derivatives(mod2, intervel = "simultaneous", order = 2, data = predD %>% mutate(season = seasons[1]), select = "s(newWeek)") %>% select(.derivative, .lower_ci, .upper_ci) %>% rename(d2 = .derivative, d2.lower = .lower_ci, d2.upper = .upper_ci) %>% 
     add_column(fit = predD$fit, season = seasons[1]) %>%
     mutate(d2_resp = fit * (1 - fit) * d2, 
            d2_lo_resp = fit * (1 - fit) * d2.lower,
            d2_up_resp = fit * (1 - fit) * d2.upper) %>% select(d2_resp, d2_lo_resp, d2_up_resp)

predD <- predD %>% cbind(d1,d2) %>%
               mutate(
               signs1 = sign(d1_resp),
               signs2 = sign(d2_resp),
               concaveDown = c(NA,(signs2[-1] == -1 & signs2[-length(signs2)] == 1)), #this is the first negative after the zero crossing
               concaveUp = c(NA,(signs2[-1] == 1 & signs2[-length(signs2)] == -1)),
               decrease = c(NA,(signs1[-1] == -1 & signs1[-length(signs1)] == 1)), 
               increase = c(NA,(signs1[-1] == 1 & signs1[-length(signs1)] == -1)),
               inflection =  (d2_lo_resp < 0 & d1_lo_resp ==max(d1_lo_resp)),
               peak = (fit==max(fit)))


ggplotly(ggplot(theD, aes(x = newWeek, y = fit, group = season, color = season),alpha = 0.5) + 
                      geom_point(aes(y=p),alpha = 0.5) + geom_line(alpha = 0.5)  +
                      xlab(paste("Week (aligned ignition at", theD$anchorWeek[1], ")")) +
                      geom_line(data = predD, aes(y = fit), color = "black", linewidth = 2) +
                      geom_line(data = predD, aes(y = d1_resp), color = "red", linewidth = 0.5, alpha = 0.5) +
                      geom_line(data = predD, aes(y = d2_resp), color = "blue", linewidth = 0.5, alpha = 0.5) +
                      geom_vline(xintercept = theD$anchorWeek[1], linetype = "dashed") +
                      geom_hline(yintercept = c(0,0.05), alpha = c(1,0.5), linewidth = c(0.7,0.5)) +
                      geom_vline(xintercept = predD %>% filter(inflection) %>%.$newWeek, color = "blue") +
                      geom_vline(xintercept = predD %>% filter(peak)%>%.$newWeek, color = "red") 
         
         
         )
```

## Alignment

### Estimated reference curve

```{r}

weeks = 1:52
grid <- tibble(newWeek = weeks)
g_ref_fun <- (function(gam_obj, grid) {
  eta_hat <- drop(predict(gam_obj, newdata = grid, type = "link"))
  splinefun(grid$newWeek, eta_hat, method = "natural")  # rule=2 extrapolates
})(mod2$gam, grid)

ref_df <- tibble(
  newWeek = weeks,
  p_gamm  = plogis(g_ref_fun(weeks))
)

ggplotly(ggplot() +
  geom_line(data = ref_df, aes(newWeek, p_gamm), color = "steelblue") +
  labs(title = "Estimated reference curve",
       x = "Week", y = "Probability") + theme_minimal(12))

```

### Alignment hyperparameters and forecast

```{r}
g_ref_safe <- function(u) g_ref_fun(pmin(pmax(u, 1), 52))

# 3) Mean & SE from the fitted GAM at arbitrary u
g_ref_mu_se <- (function(gam_obj) {
  function(u) {
    nd <- data.frame(newWeek = u)
    pr <- predict(gam_obj, newdata = nd, type = "link", se.fit = TRUE)
    list(mu = drop(pr$fit), se = drop(pr$se.fit))
  }
})(mod2)


#=========================================================learn about shifts and streches
theD = theD  %>% mutate(newWeek = weekF)
hyper <- learn_alignment_hyperparams(theD %>% mutate(newWeek = weekF), g_ref_fun)
#hyper$TAU_BOUNDS; hyper$DELTA_BOUNDS; hyper$WEEK_THRESHOLD_DELTA; hyper$LAMBDA_DELTA
currentSeason$newWeek = currentSeason$weekF 

##check dilation
scale_diag <- check_scale_identifiability(
  currentD  = currentSeason,
  g_ref_fun = g_ref_fun,
  hyper     = hyper
)
print(scale_diag)



res <- align_forecast_pipeline_dilate(
  currentSeason, g_ref_fun, hyper = hyper,
  allow_scale =NULL,   # auto: FALSE early, TRUE later  scale_diag$allow_scale_rec
  use_weights = TRUE,
  level = 0.95
)

cat(sprintf(" taû=%.2f, deltâ=%.3f%s | Peak week ~ %.1f (95%% CI [%.1f, %.1f]) | Peak p=%.3f\n",
           
            res$tau, res$delta,
            if (is.na(res$fallback_reason)) "" else " [τ-only fallback]",
            res$peak$t_peak, res$peak$t_peak_ci[1], res$peak$t_peak_ci[2],
            res$peak$p_peak))

p= ggplot() +
  geom_ribbon(
    data = subset(res$pred_df, kind == "forecast"),
    aes(x = newWeek, ymin = p_lo, ymax = p_hi),
    fill = "steelblue", alpha = 0.20
  ) +
  geom_line(
    data = subset(res$pred_df, kind == "forecast"),
    aes(x = newWeek, y = p_hat),
    color = "steelblue", linewidth = 1.2
  ) +
  geom_line(
    data = theD %>% filter(weekF >8),
    aes(x = newWeek, y = fit, group = season),
    color = "grey70", linewidth = 0.6, alpha = 0.4
  ) +
  geom_point(
    data = subset(res$pred_df, kind == "observed"),
    aes(x = newWeek, y = p_hat),
    color = "tomato", size = 2
  ) +
  geom_vline(xintercept = res$last_obs, linetype = 2, color = "grey50") +
  { if (!any(is.na(res$peak$t_peak))) geom_vline(xintercept = res$peak$t_peak, color = "purple", linetype = 2) } +
  labs(
    title = sprintf("shift= %.2f, dilation= %.3f",
                    res$tau, res$delta),
    subtitle = if (!any(is.na(res$peak$t_peak))) {
      sprintf("Peak ~ week %.1f ",
              res$peak$t_peak)
    } else "Peak CI unavailable yet (early/unstable δ).",
    x = "Week of year", y = "Probability (positivity)"
  ) +
  theme_minimal(base_size = 13)

ggplotly(p)


xD = res$pred_df %>% left_join(currentSeason %>% select(date, newWeek = weekF)) %>%
                     add_column(season = season) %>%     
                     mutate( date = as.Date(date),
                             start_year = as.integer(substr(season, 1, 4)),
                             nW_true    = n_weeks_in_start_year(start_year),
                             week = ((newWeek + startWeek - 2L) %% nW_true) + 1L,
                             mmwr_year  = ifelse(week >= 35L, start_year, start_year + 1L),
                             Rdate      = MMWRweek2Date(mmwr_year, week, 1L),
                             date = as.Date(ifelse(is.na(date), Rdate, date)))
                  

xD = xD %>% left_join(ref_df)

ggplotly(xD %>% ggplot(aes(x = date, y = p_hat)) + geom_line()  +
  geom_point(data = xD %>% filter(kind=="observed")) +     
    geom_ribbon(aes(ymin = p_lo, ymax = p_hi), alpha = 0.20) +
      geom_line(aes(y= p_gamm), color = "steelblue") +
    geom_text(data = xD %>% filter(newWeek %%2 ==0),aes(y = 0.1,x= date,label = newWeek), size = 3, angle = 90) + 
  scale_x_date(
    breaks = xD$date,
    labels = xD$week
  ) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)))


```
