---
title: "Canonical PAGe v3: Ignition, Peak Timing, and A/B Forecasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Canonical PAGe v3: Ignition, Peak Timing, and A/B Forecasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PAGe ships the audited weekF12 canonical v3 shadow model as package data. The model bundle contains the frozen M0-A, M1-A, M2-A, M1-B, and M2-B artifacts and a hash manifest tied to release:

`5472d08992b5a9da40a9419b75c7427847ff7b1999070041d5e38b0a71da853b`

No repository-level `source()` calls or external model directories are required after package installation.

## Inspect the bundled model identity

```{r models, eval=FALSE}
library(PAGe)
models <- page_v3_models()
attr(models, "release_id")
attr(models, "manifest")
```

Every packaged runtime file is SHA-256 verified before use. The manifest also records the canonical source-artifact SHA-256 values. M2-A and M2-B are deterministic runtime-only projections: they preserve the frozen coefficients/contracts (and the learned normalized B shape template required for posterior-C2) while omitting historical target-count frames. The runtime remains `shadow_only` and `production_eligible = FALSE`.

## Run from an OLIS snapshot

```{r olis, eval=FALSE}
result <- page_v3_forecast(
  "path/to/hist_olis.RData",
  season = "2026-27"
)

result
```

The returned object contains the full operational state:

```{r inspect, eval=FALSE}
result$monitoring$A$m0   # ignition decision and crossing week
result$monitoring$A$m1   # peak mean, 90% interval, weeks-to-peak, probabilities
result$monitoring$B$m1   # B activity/timing state
result$forecasts         # A/B +1/+2 forecasts and routes
result$provenance        # release/model/input identities
```

## Run from a typed A/B panel

The package also accepts a data frame with one aligned row per week and these required columns:

`season`, `weekF`, `y_A`, `N_A`, `p_A`, `y_B`, `N_B`, `p_B`.

```{r panel, eval=FALSE}
result <- page_v3_forecast(
  typed_ab_panel,
  season = "2026-27",
  origin_weekF = 12
)
```

Dates are optional for model execution but recommended for audit/reporting.

## How routing works

At weekF12 and later, the runtime requires the exact latest three consecutive observations through the requested origin.

Influenza A:

- M0-A determines ignition;
- M1-A estimates peak timing after ignition;
- M2-A +1 and +2 remain exact A1 state forecasts by frozen policy.

Influenza B:

- B+1 is exact B1 state;
- B+2 uses posterior-C2 only when causal B timing is available;
- otherwise B+2 is exact B1 fallback.

Before weekF12, `page_v3_forecast()` returns a valid object with `issued = FALSE` and does not silently turn early diagnostics into issued forecasts.

## Weekly operation

A minimal weekly R job is simply:

```{r weekly, eval=FALSE}
library(PAGe)

result <- page_v3_forecast(
  "/secure/path/hist_olis.RData",
  season = "2026-27"
)

saveRDS(result, file = sprintf(
  "results/page-v3-%s-weekF%02d.rds",
  result$season, result$origin_weekF
))
```

Scheduling (Posit Connect, cron, CI, or another orchestrator) is intentionally separate from the forecasting package. The model, validation, ignition/peak monitoring, and forecast routing all remain inside PAGe.
