---
title: "Getting started with PAGe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PAGe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**PAGe** (Phase-Aligned Gated Epidemic Forecasting) produces 1--2 week
ahead forecasts of seasonal respiratory-virus percentage positivity
from surveillance data. The pipeline is three stages run in sequence
each week:

* **M0** -- ignition detection (when the season has started)
* **M1** -- alignment to a learned reference curve via shift `tau` and
  optional dilation `delta`
* **M2** -- binomial GAM forecast with adaptive Holt EMA bias correction

## Data and training

```{r}
library(PAGe)

# PAGe does not bundle surveillance observations.
allD <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()

# Offline locked-spec refresh. The 2025-26 holdout is excluded by default.
training <- train_pipeline(allD, mode = "refresh")
kit <- training$kit
```

Use `mode = "retune"` for full M0/M1/M2 tuning. Its adaptive M2 grid is
informed by compatible prior results and supports `"min_nll"`, `"one_se"`,
and `"pareto"` selection. Optional conservative racing only removes clear
losers; finalists still receive full nested-LOSO evaluation.

## Holdout replay and promotion

```{r}
candidate <- replay_season_holdout(kit, allD, season = "2025-26")
incumbent <- replay_season_holdout(incumbent_kit, allD, season = "2025-26")
promotion <- check_promotion(candidate$metrics, incumbent$metrics)

# Passing all gates permits inclusion in training for 2026-27.
next_training <- train_pipeline(allD, mode = "refresh", promotion = promotion)
```

The default gates require 2% NLL improvement, at most 5% horizon-MAE
degradation, and at most 10% phase-MAE degradation.

## Weekly forecasting

Given `kit` and the current season's observed weeks so far, produce a
forecast with:

```{r}
current <- prepare_surveillance_data(current_csv, season = "2026-27")
res <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(res, history = allD)
```

Frozen-GAM prediction is the default validated deployment path. Weekly
refitting is retained only as an explicit compatibility option.

## Where to read more

* [Pipeline overview](pipeline-overview.html) -- architecture and notation.
* [Pipeline walkthrough](pipeline-walkthrough.html) -- an end-to-end training
  and deployment walkthrough.
* `?build_m0`, `?build_m1`, `?build_m2`, `?assemble_kit`,
  `?run_pipeline` -- reference pages for the high-level API.
