---
title: "Getting started with PAGe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PAGe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**PAGe** (Phase-Aligned Gated Epidemic Forecasting) produces 1--2 week
ahead forecasts of seasonal respiratory-virus percentage positivity
from surveillance data. The pipeline is three stages run in sequence
each week:

* **M0** — ignition detection (when the season has started)
* **M1** — alignment to a learned reference curve via shift `tau` and
  optional dilation `delta`
* **M2** — binomial GAM forecast with adaptive Holt EMA bias correction

## Training a kit

```{r}
library(PAGe)

# Historical surveillance data shipped with the package.
allD <- load_flu_hist()

# M0: ignition detection.
m0 <- build_m0(allD)

# M1: reference curve + alignment hyperparameters.
m1 <- build_m1(allD, m0)

# M2: forecast GAM (uses the pre-tuned production spec).
m2 <- train_m2(allD, m0, m1, best_spec = NULL)

# Bundle everything a deployment needs into one list.
kit <- assemble_kit(m0, m1, m2)
```

## Weekly forecasting

Given `kit` and the current season's observed weeks so far, produce a
forecast with:

```{r}
current <- getCurrentD(season = "2025-26")
res <- run_pipeline(kit, current)
plot_forecast(res, history = allD)
```

## Where to read more

* [Pipeline overview](pipeline-overview.html) — architecture and notation.
* [Pipeline walkthrough](pipeline-walkthrough.html) — an end-to-end training
  and deployment walkthrough.
* `?build_m0`, `?build_m1`, `?build_m2`, `?assemble_kit`,
  `?run_pipeline` — reference pages for the high-level API.
