---
title: "PAGe v1, v2, and v3: What Can Be Compared Fairly?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PAGe v1, v2, and v3: What Can Be Compared Fairly?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PAGe evolved in stages, and the evaluation ledgers changed with the model. A single pooled "v1 vs v2 vs v3" score would mix different targets, origins, pathogens, and model-routing decisions. This vignette therefore reports only comparisons supported by matched or explicitly chronological evidence.

```{r metrics}
library(PAGe)
metrics <- page_version_metrics()
metrics[, c(
  "comparison_scope", "component", "horizon", "metric", "unit",
  "n_rows", "n_seasons", "v1_legacy", "v2", "v3",
  "relative_gain_v1_to_v2", "relative_gain_v2_to_v3"
)]
```

## M1-A peak timing: v1/legacy versus v2

The strict common ledger contains 81 matched origin/truth rows across 10 seasons. Mean absolute peak-timing error is approximately:

- v1/legacy: 1.375 weeks;
- v2: 1.093 weeks;
- relative reduction: 20.6%.

The same v2 timing family feeds the canonical v3 A stack, but this exact historical ledger is not an independent v3 re-score. For that reason the v3 metric cell is intentionally `NA` rather than copied from v2.

The underlying v2 ledger also has mean 90% interval coverage of approximately 0.802 and mean interval width of about 4.01 weeks. Those uncertainty metrics describe that M1-v2 evaluation ledger, not a separate v3 experiment.

## M2-A: v1/legacy versus v2 on matched targets

For 10 historical seasons whose target counts are identical across the two systems, the matched-target benchmark gives:

```{r a-table}
metrics[metrics$component == "M2_A",
        c("horizon", "n_rows", "n_seasons", "v1_legacy", "v2",
          "relative_gain_v1_to_v2", "caveat")]
```

At +1 week, MAE falls from about 8.41 percentage points to 1.78 percentage points. At +2 weeks, MAE falls from about 9.21 to 2.41 percentage points.

This benchmark matches target rows and target counts, but historical training snapshots are not perfectly vintage-matched. It should therefore be interpreted as strong stage-specific evidence rather than a fully prospective head-to-head trial.

Canonical v3 preserves the governed A1 state route. It was not independently re-scored on this exact v1/legacy comparison ledger, so v3 remains `NA` in this table.

## M2-B +2: the direct v2 to v3 improvement

The clearest direct v2-to-v3 performance comparison is Influenza B +2 forecasting. v2's exact B1 state forecast is the baseline. v3 uses posterior-C2 when causal B timing is available and otherwise falls back to the exact B1 route.

Across the seven-season chronological B benchmark:

```{r b-table}
metrics[metrics$component == "M2_B",
        c("comparison_scope", "horizon", "n_rows", "n_seasons", "v2", "v3",
          "relative_gain_v2_to_v3", "caveat")]
```

Overall +2 MAE improves from about 0.515 percentage points to 0.436 percentage points, a 15.3% reduction.

On the timing-active subset, 89 rows across five seasons, MAE improves from about 0.736 to 0.552 percentage points, a 24.9% reduction. All five timing-active seasons improved on +2 MAE in that benchmark.

## What is intentionally not claimed

PAGe does not currently publish a single universal v1/v2/v3 score. In particular:

- there is no verified common-ledger M0 performance comparison suitable for a v1/v2/v3 accuracy claim;
- the v1/legacy M2 benchmark is A-only;
- v3's main new historical accuracy evidence is on B +2 routing;
- inherited/frozen A components are not assigned v3 metrics merely because their lineage descends from v2;
- prospective 2026-27 v2/v3 paired accuracy should be accumulated only after each forecast target is observed.

The full table, evidence paths, and caveats are available programmatically:

```{r all-metrics}
page_version_metrics()
```
