---
title: "Train PAGe from Seasonal Surveillance Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Train PAGe from Seasonal Surveillance Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PAGe training starts with the surveillance record and an explicit expert review of epidemic ignition. The package does not silently substitute its historical default ignition labels when you use the workflow in this vignette.

## 1. Prepare seasonal data

Use one row per season/week with positive counts, test counts, and positivity:

```{r schema, eval=FALSE}
seasonal_data <- data.frame(
  season = c("2022-23", "2022-23", "2023-24", "2023-24"),
  weekF = c(1, 2, 1, 2),
  y = c(5, 8, 4, 7),
  N = c(1000, 1000, 950, 960)
)
seasonal_data$p <- seasonal_data$y / seasonal_data$N
```

`prepare_surveillance_data()` validates and canonicalizes the input. Real model training should use complete historical seasons rather than this tiny schema illustration.

## One-call interactive workflow

If you want PAGe to lead the workflow, omit `labels` and `ignition_weeks`. `page_train_workflow()` will plot each season first, prompt for the expert decimal ignition week, validate the complete label set, and only then start training:

```{r one-call, eval=FALSE}
fit <- page_train_workflow(
  seasonal_data,
  annotator = "your-name",
  interactive = TRUE,
  mode = "refresh",
  prospective_holdout = NULL
)
```

Use the two-step workflow below when you want to review/save the labels separately before starting a potentially expensive training run.

## 2. Review each season and label ignition

In an interactive R session, call:

```{r interactive-label, eval=FALSE}
library(PAGe)

labels <- page_label_ignitions(
  seasonal_data,
  annotator = "your-name",
  interactive = TRUE
)
```

For each season PAGe first opens the same Plotly review produced by `review_expert_ignition()`. After you inspect the curve, the console asks for the ignition week on PAGe's continuous week coordinate. For example, `18.5` means halfway through the interval represented by integer week 18.

The decimal annotation is retained for provenance. The current M0/M1 training contract uses the containing integer week, so `18.5` becomes integer training anchor 18. PAGe makes this conversion explicitly with `floor()`.

For a reproducible script or CI job, provide the reviewed values directly:

```{r reproducible-label, eval=FALSE}
labels <- page_label_ignitions(
  seasonal_data,
  ignition_weeks = c(
    "2018-19" = 19.4,
    "2019-20" = 22.1,
    "2022-23" = 15.6,
    "2023-24" = 20.2,
    "2024-25" = 23.0,
    "2025-26" = 19.3
  ),
  annotator = "your-name",
  interactive = FALSE
)

labels$table
labels$manual_labels
```

Expert input is limited to ignition timing in this workflow. PAGe does not ask the user to hand-label forecast targets or manufacture future peak information.

## 3. Train a model

`page_train_workflow()` reuses the package's governed `train_pipeline()` implementation. It does not create a second training engine.

A fixed-specification refresh is the fastest path when the model structure has already been selected:

```{r refresh, eval=FALSE}
fit <- page_train_workflow(
  seasonal_data,
  labels = labels,
  mode = "refresh",
  prospective_holdout = NULL,
  n_cores = 4
)

fit
```

For model development, keep a prospective season out of all fitting, tuning, **and expert ignition review**. Let the workflow label only the trainable seasons, or supply ignition weeks that omit the holdout:

```{r holdout, eval=FALSE}
fit <- page_train_workflow(
  seasonal_data,
  ignition_weeks = c(
    "2018-19" = 19.4,
    "2019-20" = 22.1,
    "2022-23" = 15.6,
    "2023-24" = 20.2,
    "2024-25" = 23.0
  ),
  annotator = "your-name",
  interactive = FALSE,
  mode = "retune",
  prospective_holdout = "2025-26",
  checkpoint_dir = "checkpoints/page-retune",
  n_cores = 8
)
```

`page_train_workflow()` excludes unreleased holdout and explicitly excluded seasons before its interactive review step. If a pre-existing label set contains those seasons, their labels are retained in the returned provenance object but stripped before `train_pipeline()` is called.

A governed retune is substantially more expensive because it performs leakage-safe leave-one-season-out model selection. Retuning and holdout release should follow the promotion/evaluation workflow documented elsewhere in PAGe; do not inspect the final holdout to choose hyperparameters and then report it as prospective evidence.

## 4. Inspect and save the frozen kit

The training workflow returns the original expert label set, the complete `page_training_result`, and its frozen deployment kit:

```{r inspect, eval=FALSE}
fit$labels$table
fit$training_result
validate_page_kit(fit$kit, mode = "frozen")

page_save_kit(fit, "artifacts/my-page-kit.rds")
```

The saved kit is self-contained for frozen forecasting. `page_load_kit()` validates it again when read.

```{r load, eval=FALSE}
kit <- page_load_kit("artifacts/my-page-kit.rds")
```

Continue with the deployment vignette to forecast new weekly observations.
