---
title: "2025-26 Holdout Decision and Post-Promotion Refit"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
execute:
  eval: false
---

This is a two-phase, private-data workflow. The 2025-26 season is an untouched
prospective holdout until a separately saved decision bundle passes the locked
promotion gates. It must not be used for retrospective tuning, replay after a
refit, or other in-sample performance claims.

## Phase 1: untouched holdout replay and promotion evidence

Run the acceptance workflow against the locked incumbent kit and authorized
2025-26 surveillance data. It produces a private
`page_holdout_decision_bundle` RDS and a separate disclosure-safe promotion
manifest only after the holdout replay and canonical `check_promotion()`
decision are complete. The bundle contains a run identifier, creation time,
season binding, canonical locked-threshold report, and SHA-256 identities for
the candidate, incumbent, and authorized data. The manifest repeats those
identities and records the SHA-256 hash of the immutable decision bundle.

The bundle is the authorization input for Phase 2; a bare report, custom
thresholds, malformed report, failed gates, a different season, or mismatched
authorized-data hash is rejected before training.

```{r decision-bundle-schema, eval=true}
# This compact verification is disclosure-safe and needs no private artifact.
expected_bundle_fields <- c(
  "schema", "schema_version", "run_id", "created_at", "holdout_season",
  "report", "source_artifact_hashes"
)
stopifnot(identical(expected_bundle_fields[[3]], "holdout_season"))
expected_bundle_fields
```

Keep the resulting bundle and its referenced private inputs immutable. The
release decision demonstrates performance only for the pre-refit locked model.

## Phase 2: post-decision refit

Only after Phase 1 passes, use the governed high-level API through the opt-in
script below. It calls `train_pipeline(mode = "refresh",
prospective_holdout = "2025-26", promotion = evidence)` with artifact-bound
evidence created by `verify_promotion_evidence()` and the exact promoted
candidate M0/M1 settings and M2 specification. A bare report cannot release
the holdout. Phase 2 does not retune, search a grid, or choose a new
specification after seeing the holdout.

First validate the inputs without starting a training job:

```{bash, eval=false}
Rscript season2526/run_retrain_venkata.R \
  --data=/secure/PAGe/authorized_combined_surveillance.csv \
  --promotion-bundle=/secure/PAGe/holdout_decision_2025-26.rds \
  --promotion-manifest=/secure/PAGe/holdout_promotion_manifest_2025-26.rds \
  --candidate-kit=/secure/PAGe/promoted_candidate_kit.rds \
  --incumbent-kit=/secure/PAGe/incumbent_pre_holdout_kit.rds \
  --output-dir=/secure/PAGe/private-refits \
  --manifest-dir=results/post-promotion-manifests \
  --preflight-only
```

To perform the actual private refit, remove `--preflight-only`. The script does
not download surveillance data and accepts only an explicitly authorized
combined data file (CSV or RDS), an explicit decision bundle, its separately
saved promotion manifest, and the exact candidate and incumbent kits. Their
SHA-256 hashes, training seasons, and the candidate specification identity
must match the acceptance evidence.

Before writing any artifact, the script asserts that the holdout status is
`released`, that `2025-26` appears in
`retuned$kit$m2_production$training_seasons`, and that a selected specification
identity is present. A failed assertion writes neither private model nor
manifest.

Successful output uses a spec- and SHA-256-based immutable filename such as
`post_promotion_refit_<spec>_data-<hash>_promotion-<hash>.rds`. The full refit
remains in `--output-dir`; the separate `--manifest-dir` receives a
disclosure-safe aggregate manifest in both RDS and Markdown, listing training
seasons and source hashes.
Existing output names are never overwritten.

```{r manifest-contract, eval=true}
# A rendered document may verify the public manifest schema without exposing
# row-level surveillance data or reading any private artifact.
manifest_provenance_fields <- c(
  "spec_id", "training_seasons", "source_artifact_hashes"
)
stopifnot(length(manifest_provenance_fields) == 3L)
manifest_provenance_fields
```

The source document is authoritative. Any pre-existing rendered HTML is stale
until deliberately rendered in an environment authorized to access only the
intended disclosure-safe outputs.
