---
title: "PAGe Pipeline Retrain Reproduction & Walk-Forward Evaluation"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
execute:
  eval: false
---

This document provides the exact code used to reproduce the full PAGe pipeline retrain (incorporating the 2025-26 season) and retrospectively evaluate the final model on each season in a walk-forward manner.

## Setup and Data Preparation

```{r setup, message=FALSE, warning=FALSE}
library(PAGe)
library(dplyr)
library(ggplot2)
library(purrr)
library(future)

# 1. Load historical data
source("scripts/fresh_run/00_shared.R")
histD <- load_allD(exclude = c(), include_holdout = TRUE) %>% 
  filter(season != "2025-26")

# 2. Extract current season (2025-26)
currD <- getCurrentD(season = "2025-26") %>%
  filter(season == "2025-26")

# 3. Combine and prepare canonical dataset
all_raw <- bind_rows(histD, currD)
allD <- prepare_surveillance_data(all_raw)

# Setup multisession parallelization
n_cores <- parallel::detectCores() - 1L
plan(multisession, workers = n_cores)
```

## Section 1: M0 - Ignition Detection Tuning

M0 identifies the start of the exponential growth phase. We retune the ignition parameters using a grid search over the newly assembled dataset.

```{r m0-tuning}
# Tune M0 over the default grid
m0_tuning <- tune_m0(allD, n_cores = n_cores, verbose = TRUE)
m0_best <- m0_tuning$best_params

cat("Best M0 parameters:\n")
print(m0_best)
```

## Section 2: M1 - Alignment Tuning

M1 aligns the current season to the reference median curve and historical templates. 

```{r m1-tuning}
# Build M1 reference curve
m1_built <- build_m1(allD, m0 = m0_tuning)

# Retune M1 hyperparameters
m1_tuning <- tune_m1(allD, m0 = m0_tuning, m1 = m1_built, n_cores = n_cores, verbose = TRUE)
m1_best <- m1_tuning$best_params

cat("Best M1 parameters:\n")
print(m1_best)
```

## Section 3: M2 - Forecast Model Construction

M2 builds the final GAM utilizing the M1 aligned projections as covariates.

```{r m2-tuning}
# Build and evaluate the M2 specification grid using Nested LOSO
m2_built <- build_m2(
  allD, 
  m0 = m0_tuning, 
  m1 = m1_built, 
  loso_seasons = "all", 
  n_cores = n_cores, 
  verbose = TRUE
)
m2_best_spec <- m2_built$best_spec

cat("Best M2 nested-LOSO NLL:", m2_built$summary$bernoulli_nll[1], "\n")
print(m2_best_spec)

# Train the final production M2 model on all data
m2_trained <- train_m2(
  allD, 
  m0 = m0_tuning, 
  m1 = m1_built, 
  best_spec = m2_best_spec, 
  verbose = TRUE
)

# Assemble and save the final production kit
kit <- assemble_kit(m2_trained)
saveRDS(kit, "data/retuned_pipeline_kit.rds")
```

*(Note: The above three sections can also be run in a single pass using `train_pipeline(allD, mode = "retune")` as we did on the remote worker).*

## Walk-forward Retrospective Evaluation

We now use the fully retuned kit to perform a retrospective walk-forward evaluation on every season. This simulates what the prospective pipeline would have forecasted at each weekly step during the given season.

```{r evaluate-walkforward, fig.width=10, fig.height=6}
# Load the final kit
kit <- readRDS("data/retuned_pipeline_kit.rds")

# Identify all valid seasons to evaluate
eval_seasons <- sort(setdiff(unique(allD$season), c("2011-12", "2015-16", "2020-21", "2021-22")))

# Run walkforward evaluation for each season
walkforward_results <- map(eval_seasons, function(s) {
  message("Replaying season: ", s)
  replay_season_holdout(kit, allD, season = s)
})
names(walkforward_results) <- eval_seasons

# Plotting function for retrospective M2 performance
plot_retrospective_forecast <- function(season_name, replay_out) {
  preds <- replay_out$predictions %>% filter(h == 2) # Plot 2-week ahead forecasts
  obs <- replay_out$history %>% filter(season == season_name) %>% mutate(p_obs = y / N)
  
  ggplot() +
    geom_point(data = obs, aes(x = weekF, y = p_obs), color = "black", alpha = 0.5) +
    geom_line(data = preds, aes(x = target_weekF, y = m2_p), color = "firebrick", size = 1) +
    geom_ribbon(data = preds, aes(x = target_weekF, ymin = m2_lo, ymax = m2_hi), fill = "firebrick", alpha = 0.2) +
    labs(
      title = paste("Walk-forward 2-Week Ahead Forecast:", season_name),
      x = "Season Week (weekF)",
      y = "Test Positivity"
    ) +
    theme_minimal()
}

# Generate and display plots
for (s in eval_seasons) {
  print(plot_retrospective_forecast(s, walkforward_results[[s]]))
}
```
