---
title: "Reference curve estimation for flu A"
format:
  html:
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(gamm4)
library(gratia)
library(plotly)
library(data.table)
library(MMWRweek)
devtools::load_all("../PAGe", quiet = TRUE)

# ---- Settings ----
startWeek   <- 27     # MMWR week number that defines weekF = 1
min_window  <- 10     # ignition search window at both ends
df          <- 20     # k for the common-curve smooth (increased from 10; reference needs flexibility too)
p_thresh    <- 0.01
k1          <- 0.4
k2          <- -0.01
manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
) -1

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Build the aligned historical dataset used downstream by the rest of the pipeline.
- Estimate the common reference curve and per-season FS curves.
- Inspect aligned peaks, derivatives, and the saved reference artifacts.
:::

## Notation and data structure

Let $s = 1,\dots,S$ index influenza seasons and let $w$ denote the within-season flu-week number. For each season–week $(s,w)$, define:

- $x_{s,w}$: number of influenza-positive tests (code column `y`)
- $n_{s,w}$: number of tests performed (code column `N`)
- $p_{s,w} = x_{s,w} / n_{s,w}$: observed weekly positivity (code column `p`)
- $\pi_{s,w}$: underlying probability of positivity (the quantity modelled/forecast)

In the dataset we keep several week indices:

- `week`: MMWR **epidemiologic week-of-year** from surveillance data.
- `weekS`: **surveillance-week index** (season starts at MMWR week 35).
- `weekF`: $w_s^f$, **flu-week index** (season starts at MMWR week `startWeek`; here 27).
- `newWeek`: $w^a_{s,w}$, **aligned week index** after shifting weeks so ignition lines up across seasons (defined later).


```{r}
#| label: load-data

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season,
    week,
    year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg = N - y,
    date = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))
```


## Season-level smoothing and derivatives

For each season $s$, a binomial GAM with P-spline smooth (REML) is fitted to the raw counts, with observations between the ignition week and the observed peak upweighted 3× (soft exponential decay after peak). This makes the smooth most faithful in the rising-limb-to-peak region that matters for alignment and forecasting.

We first try `k = 10` to inspect EDF, then increase to `k = 20` since all seasons push against the basis limit.

```{r}
#| label: derivatives
k_smooth <- 20L   # increased from 10 after EDF inspection showed all seasons near limit

res <- estimateDerivs(
  allD,
  k                 = k_smooth,
  peak_weight_boost = 3,
  peak_weight_decay = 0.3,
  ignition_weeks    = manual_labels
)

# EDF per season
edf_tbl <- purrr::map_dfr(names(res$models), function(s) {
  g <- res$models[[s]]
  tibble::tibble(
    season = s,
    k      = k_smooth,
    edf    = round(sum(g$edf), 2),
    k_used_pct = round(100 * sum(g$edf) / (k_smooth - 1), 1)
  )
}) %>% dplyr::arrange(season)

knitr::kable(edf_tbl,
  col.names = c("Season", "k", "EDF", "EDF / (k-1) %"),
  caption   = "Effective degrees of freedom per season. EDF / (k-1) % near 100 means the smooth is still at its flexibility limit.")
```

The **effective degrees of freedom (EDF)** tells how much of the `k - 1` available flexibility REML actually used. With `k = 10` all seasons reached ~100% — the smooth was constrained. Increasing to `k = 20` gives REML room to choose a more faithful peak shape without forcing extra curvature where it isn't needed.

From the fitted smooth, first and second derivatives on the probability scale are computed via `gratia::derivatives()`:

$$
d1_{s,w} = \frac{d}{dw}\tilde p_s(w), \qquad
d2_{s,w} = \frac{d^2}{dw^2}\tilde p_s(w).
$$

These are used as features for ignition detection (slope and curvature signals) and downstream in $M_2$ as real-time covariates.

### Per-season observed vs fitted peaks

Each tab shows one season. Points = raw observed positivity $p = y/N$. Blue line = weighted-smooth fit from `estimateDerivs()` (3× weight on ignition-to-peak). Shaded band = 95% CI.

- **Red dashed**: observed peak (`argmax(p)`)
- **Blue solid**: smoothed peak (`argmax(fit)`)

```{r}
#| label: peak-comparison
peak_comp <- res$data %>%
  group_by(season) %>%
  summarise(
    obs_peak    = weekF[which.max(replace(p,   is.na(p),   -Inf))],
    smooth_peak = weekF[which.max(replace(fit, is.na(fit), -Inf))],
    .groups = "drop"
  ) %>%
  left_join(
    tibble::enframe(manual_labels, name = "season", value = "iWeek"),
    by = "season"
  )

peak_comp %>%
  mutate(delta = smooth_peak - obs_peak) %>%
  knitr::kable(caption = "Observed vs smoothed peak week per season (delta = smooth − obs; iWeek = manual ignition week)")
```

::: panel-tabset

```{r}
#| label: season-peak-tabs
#| results: asis
#| echo: false
#| warning: false

for (s in sort(unique(res$data$season))) {
  cat(sprintf("\n### %s\n\n", s))

  sdat  <- dplyr::filter(res$data, season == s)
  speak <- dplyr::filter(peak_comp, season == s)

  p <- ggplot(sdat, aes(x = weekF)) +
    geom_ribbon(aes(ymin = fit_low, ymax = fit_high),
                fill = "#1e5f8e", alpha = 0.15) +
    geom_line(aes(y = fit), colour = "#1e5f8e", linewidth = 1.2) +
    geom_point(aes(y = p), colour = "grey40", size = 1.8, alpha = 0.8) +
    geom_vline(xintercept = speak$obs_peak,
               colour = "red", linetype = "dashed", linewidth = 0.9) +
    geom_vline(xintercept = speak$smooth_peak,
               colour = "#1e5f8e", linetype = "solid", linewidth = 0.9) +
    geom_vline(xintercept = speak$iWeek,
               colour = "darkorange", linetype = "dotted", linewidth = 1.0) +
    annotate("text", x = speak$obs_peak,    y = Inf, vjust = 1.5, hjust = -0.1,
             label = paste0("obs=", speak$obs_peak),    colour = "red",      size = 3) +
    annotate("text", x = speak$smooth_peak, y = Inf, vjust = 1.5, hjust =  1.1,
             label = paste0("fit=", speak$smooth_peak), colour = "#1e5f8e", size = 3) +
    annotate("text", x = speak$iWeek, y = 0, vjust = -0.4, hjust = 1.1,
             label = paste0("ign=", speak$iWeek), colour = "darkorange", size = 3) +
    labs(title = s, x = "weekF", y = "Positivity") +
    theme_minimal()

  print(htmltools::tagList(ggplotly(p)))
  cat("\n")
}
```

:::

## Ignition week definition

**For all historical seasons used here, ignition weeks are manually verified labels** supplied via `manual_labels`. The algorithmic gate conditions ($\tilde p > 0.01$, slope $> 0.4$, curvature $> -0.01$, etc.) are bypassed when a season appears in `manual_labels` — `flagIgnition()` uses the supplied week directly. Manual labels are used for reference curve estimation because they give a cleaner, bias-free alignment baseline; the algorithmic detector is trained and evaluated separately in $M_0$.

The `manual_labels` used here:

```{r}
#| echo: false
knitr::kable(
  tibble::enframe(manual_labels, name = "season", value = "iWeek"),
  caption = "Manually verified ignition weeks (weekF)"
)
```

```{r}
#| label: ignition
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = p_thresh,
    k1            = k1,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = min_window,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = k2,
    manual_labels = manual_labels
  ))

theD <- bind_rows(map(outs, "data"))
ignD <- bind_rows(map(outs, "ignition"))

ggplotly(plotSeasonCurves(bind_rows(map(outs, "data")), x="weekF"))
```

## Align seasons 

We align seasons so ignition weeks coincide. Define:

- `iWeek`: $w_s^{(\mathrm{ign})}$, ignition week within each season.
- $w^{an}$ (`anchorWeek`): the **median** ignition week across seasons, i.e. $w^{an} = \mathrm{median}_s\, w_s^{(\mathrm{ign})}$.
- `newWeek`: $w^a_{s,w}$, aligned within-season week index after shifting by $w^{an} - w_s^{(\mathrm{ign})}$

```{r}
#| label: align-ignition
alignedD <- alignIgnition(outs) %>%
  mutate(phase = if_else(weekF < iWeek, 0L, 1L))
anchorWeek <- attr(alignedD, "anchorWeek")
anchorWeek
```

## Common curve estimation (template)

Using aligned time `newWeek`, we estimate a population-level reference curve $g_{\mathrm{ref}}(w^a)$ that represents the typical seasonal trajectory of flu positivity. This curve serves two purposes:

1. **Alignment**: its peak location anchors the aligned time axis for peak-week estimation.
2. **$M_2$ covariate**: $g_{\mathrm{ref}}(w^a)$ enters the forecasting model as a template predictor encoding "where in the typical season are we."

### Model specification

We use a **factor-smooth interaction** GAM on the logit-transformed per-season smoothed positivity $\tilde p_{s,w}$ (output of `estimateDerivs()`):

$$
\mathrm{logit}(\tilde p_{s,w}) = f_s(w^a_{s,w}) + \varepsilon_{s,w}, \qquad f_s \sim \mathcal{GP}(0, \sigma^2_f),
$$

where each season $s$ gets its own cyclic smooth $f_s(\cdot)$ via `s(newWeek, season, bs = "fs", xt = list(bs = "cc"))`. The **factor-smooth (`bs = "fs"`)** basis ties all per-season curves together through a shared smoothness penalty $\sigma^2_f$ estimated by REML: seasons with more typical shapes are shrunk toward the group, while outlier seasons (e.g. 2017--18 with its low, late peak) are allowed to deviate.

### Weighting

Observations are weighted to emphasize the **ignition-to-peak region** that matters most for alignment and forecasting:

- **Pre-season** (phase = 0, before ignition): weight = `trough_weight` (0.1). Deep-trough logit values ($\approx -7$) would otherwise dominate and flatten the peak.
- **Ignition to peak**: weight = `peak_weight_boost` (3.0). The rising limb through peak is the critical region.
- **Post-peak**: soft exponential decay from `peak_weight_boost` back to 1.0 ($\text{half-life} \approx 2$ weeks at `decay = 0.3`).

### Population curve extraction

The population-level reference curve is obtained by taking the **pointwise median of logit-scale predictions across all season levels**:

$$
\eta_{\mathrm{ref}}(w) = \mathrm{median}_s\, \hat f_s(w), \qquad
g_{\mathrm{ref}}(w) = \mathrm{logistic}\bigl(\eta_{\mathrm{ref}}(w)\bigr).
$$

The median is preferred over the mean because it is robust to outlier seasons (e.g. 2017--18 with its low, late peak) and preserves the sharper peak shape that the majority of seasons exhibit. Since each $\hat f_s$ is already a smooth cyclic spline, the pointwise median of $S \approx 10$ smooth curves is itself nearly smooth — no re-smoothing is needed.

This approach is more robust than fitting a single global smooth with random intercepts (which pools levels but not shapes), because it lets the population curve inherit the typical *shape* across seasons.

```{r}
#| label: common-curve

ex <- c("2011-12","2020-21","2021-22","2025-26")

ref <- estimateRef(alignedD, exSeason = ex, k = df, method = "fs",
                   trough_weight = 0.1, peak_weight_boost = 3, agg = "median")

# --- Per-season fitted curves from the fs model (stored in ref$eta_mat) ---
eta_mat <- ref$eta_mat
grid <- data.frame(newWeek = seq_len(52))

# Per-season curves on probability scale (for plotting)
fs_season_df <- purrr::imap_dfr(as.data.frame(eta_mat), function(eta, s) {
  tibble::tibble(newWeek = grid$newWeek, season = s, fit_fs = plogis(eta))
})

# Mean and median population curves (logit scale -> plogis)
eta_mean   <- rowMeans(eta_mat)
eta_median <- apply(eta_mat, 1, median)

pop_curves <- tibble::tibble(
  newWeek    = grid$newWeek,
  mean_p     = plogis(eta_mean),
  median_p   = plogis(eta_median)
)

# Peak stats
pk_mean_wk <- pop_curves$newWeek[which.max(pop_curves$mean_p)]
pk_med_wk  <- pop_curves$newWeek[which.max(pop_curves$median_p)]

# Per-season peak stats in aligned space (from raw data)
season_peaks <- alignedD %>%
  filter(!season %in% ex) %>%
  group_by(season) %>%
  summarise(
    obs_peak_nw    = newWeek[which.max(replace(p,   is.na(p),   -Inf))],
    smooth_peak_nw = newWeek[which.max(replace(fit, is.na(fit), -Inf))],
    .groups = "drop"
  )

peak_tbl <- tibble::tibble(
  ` ` = c("**FS median (chosen)**", "FS mean",
          "Data median (obs p)", "Data median (smooth fit)",
          "Data mean (obs p)", "Data mean (smooth fit)"),
  `Peak week` = c(pk_med_wk, pk_mean_wk,
    round(median(season_peaks$obs_peak_nw), 1),
    round(median(season_peaks$smooth_peak_nw), 1),
    round(mean(season_peaks$obs_peak_nw), 1),
    round(mean(season_peaks$smooth_peak_nw), 1)),
  `Peak p` = c(round(max(pop_curves$median_p), 4),
               round(max(pop_curves$mean_p), 4),
               NA, NA, NA, NA)
)

knitr::kable(peak_tbl, caption = "Reference curve peak week vs per-season peak statistics in aligned space")
```

::: panel-tabset

### Reference curve (raw data bg)

```{r}
pop_long <- pop_curves %>%
  pivot_longer(c(mean_p, median_p), names_to = "agg", values_to = "p_ref") %>%
  mutate(agg = recode(agg, mean_p = "Mean", median_p = "Median (chosen)"))

agg_cols <- c("Median (chosen)" = "darkorange", "Mean" = "purple")

ggplotly(
  ggplot() +
    geom_line(data = ref$dat, aes(x = newWeek, y = p, group = season),
              colour = "grey70", linewidth = 0.5, alpha = 0.3) +
    geom_line(data = pop_long,
              aes(x = newWeek, y = p_ref, colour = agg), linewidth = 1.3) +
    scale_colour_manual(values = agg_cols) +
    geom_vline(xintercept = anchorWeek, linetype = "dashed") +
    labs(x = "newWeek", y = "Positivity", colour = NULL,
         title = "FS reference curve: median (chosen) vs mean") +
    theme_minimal() + theme(legend.position = "top")
)
```

### FS per-season curves

```{r}
ggplotly(
  ggplot() +
    geom_line(data = fs_season_df,
              aes(x = newWeek, y = fit_fs, group = season),
              colour = "grey60", linewidth = 0.6, alpha = 0.5) +
    geom_line(data = pop_long,
              aes(x = newWeek, y = p_ref, colour = agg), linewidth = 1.4) +
    scale_colour_manual(values = agg_cols) +
    geom_vline(xintercept = anchorWeek, linetype = "dashed") +
    labs(x = "newWeek", y = "Positivity", colour = NULL,
         title = "Per-season FS fitted curves (grey) with median & mean") +
    theme_minimal() + theme(legend.position = "top")
)
```

### Derivatives

```{r}
ggplotly(
  ref$pred_df %>%
    select(newWeek, fit, d1_eta, d2_eta, d1_p, d2_p) %>%
    pivot_longer(-newWeek) %>%
    ggplot(aes(x = newWeek, y = value, group = name, colour = name)) +
    geom_line(linewidth = 1) + theme(legend.title = element_blank()) +
    labs(x = "newWeek", y = "value")
)
```

:::

### Save the reference curve and derivatives for package
```{r, eval = F}
g_ref_fun    <- ref$g_ref_fun
g_ref_mu_se  <- ref$g_ref_mu_se
ref_df       <- ref$ref_df
hyper <- learn_alignment_hyperparams(ref$dat %>% mutate(newWeek = weekF), g_ref_fun)
remove_global_functions()

save.image("ref_curve.RData")
```
