---
title: "PAGe: Phase-Aligned Gated Epidemic Forecasting"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/retro_estimation.R")
source("../R/ignitionTraining.R")
source("../R/prospective_training.R")
source("../R/prospective_running.R")
source("../R/module_training.R")

# ---- Settings ----
startWeek   <- 27     # MMWR week number that defines weekF = 1
min_window  <- 10     # ignition search window at both ends
df          <- 10     # k for the common-curve smooth
p_thresh    <- 0.01
k1          <- 0.4
k2          <- -0.01
manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}


# NOTE: replace with your real file path or a project-relative path
allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season,
    week,
    year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg = N - y,
    date = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))





```

# Notations 

Let $s = 1,\dots,S$ index influenza seasons and let $w$ denote the within-season flu-week number. For each season–week $(s,w)$, define:

- $x_{s,w}$: number of influenza-positive tests (code column `y`)  
- $n_{s,w}$: number of tests performed (code column `N`)  
- $p_{s,w} = x_{s,w} / n_{s,w}$: observed weekly positivity (code column `p`)  
- $\pi_{s,w}$: underlying probability of positivity (the quantity modeled/forecast)

In the data we keep several week indices:

- `week`: MMWR **epidemiologic week-of-year** from surveillance data.
- `weekS`: **surveillance-week index** (season starts at MMWR week 35), for interpretation only.
- `weekF`: $w_s^f$, **flu-week index** (flu season starts at MMWR week `startWeek`; here 27).
- `iWeek`: $w_s^{(\mathrm{ign})}$,  ignition week for season $s$ (user defined or estimated)
- `newWeek`: $w^a_{s,w}$ **aligned week index** after shifting $w^f_{s,w}$ so $w_s^{(\mathrm{ign})}$ lines up across seasons

Let $P(s,w)\in\{0,1\}$ be a phase indicator, with
$$
P(s,w)=
\begin{cases}
0, & \text{pre-ignition (pre-season)}\\
1, & \text{post-ignition (in-season)}
\end{cases}
$$




We define the framework as a mixture of two components: a “pre-ignition” component $M_0$, and a “post-ignition” component $M_1$, where


- **$M_0$ (Ignition detector):** using only information available up to week $w$, estimate whether the season has ignited (i.e., whether $P(s,w)=1$).
- **$M_1$ (Forecast model):** once ignition is detected, forecast h weeks ahead using the aligned template:

$$
\pi_{s,w+h} \;=\; \Pr\!\big(\text{positive at } w+h \mid \, P(s,w)=1,\, M_1\big),
$$


Primary forecasting targets: one- and two-week-ahead probabilities $\pi_{s,w+1}$ and $\pi_{s,w+2}$.


The remaining of this document is divided into three parts: 

- Estimation of ignition weeks and the common template curve using historical data
- Training of $M_0$ and $M_1$  using the ignition week label

$$
P(s,w)=\mathbb{1}\{w^f_{s,w}\ge w_s^{(\mathrm{ign})}\},
$$

<!-- We train a collection gating rules $M_0$ derived from historical data to produce a decision $\widehat P(s,w)\in\{0,1\}$ which is monitored weekly, once ignition is detected ($\widehat P(s,w)=1$), historical data is aligned and used to train the forecast model $M_1$. -->

- Application of $M_0$ and $M_1$ in real time, where $P_(s,w)$ is estimated on the fly and $M_1$ is calibrated in real time

In addition,  an R package **`flualign`** that implements the framework, including retrospective estimation, training and prospective running functions are demostrated as along the way

# Retrospective template estimation

This section explains how to estimate ignition weeks and the common template curve from historical data, which are used as training labels for $M_0$ and covariates for $M_1$.

## Season-level smoothing and derivatives estimation

For each season $s$, smooth positivity over $w$ to obtain a denoised curve $\tilde p_s(w)$ (stored as `fit` in `estimateDerivs()` output). From the fitted smooth compute:

$$
d1_{s,w} = \frac{d}{dw}\tilde p_s(w), \qquad
d2_{s,w} = \frac{d^2}{dw^2}\tilde p_s(w).
$$

**Implementation note.** Derivatives are computed from the fitted smooth (not finite differences of raw data), which is typically more stable. For ignition detection we work with derivatives on the **link (logit) scale** as returned by `estimateDerivs()` (code columns `d1` and `d2`).

```{r}
#| label: derivatives
res <- estimateDerivs(allD, k = 10)
```

::: panel-tabset


### Smoothed curves

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = p, group = season, colour = season)) +
    geom_point() +
    geom_line(aes(y = fit), linewidth = 1) +
    geom_ribbon(aes(ymin = fit_low, ymax = fit_high), alpha = 0.2) +
    labs(x = "weekF", y = "Positivity")
)
```

### First derivative

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = d1, group = season, colour = season)) +
    geom_line(linewidth = 1) +
    labs(x = "weekF", y = "d1 (derivative)")
)
```

### Second derivative

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = d2, group = season, colour = season)) +
    geom_line(linewidth = 1) +
    labs(x = "weekF", y = "d2 (second derivative)")
)
```

### Raw data snapshot

```{r}
#| echo: false
allD %>% head(20) %>% gt()
```

:::

## Retrospective ignition detection {#sec-ignition-detection}

Define the ignition week $w_s^{(\mathrm{ign})}$ as the earliest week satisfying fixed criteria such as

-   $\tilde p_s(w) > p_t$ (code: smoothed mean `fit` vs `p_thresh`),
-   $d1_{s,w} > k1$ (code: link-scale derivative `d1` vs `k1`),
-   $d2_{s,w} > k2$ (code: link-scale derivative `d2` vs `d2_relax = k2`).

Additional parameters (such as minimum and maximum weeks in which ignition is allowed) are controlled by the arguments to `flagIgnition()`.

```{r}
#| label: ignition
#| echo: TRUE
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = p_thresh,
    k1            = k1,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = min_window,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = k2,
    manual_labels = manual_labels-1
  ))
```


```{r}
#| echo: FALSE
theD <- bind_rows(map(outs, "data"))
ignD <- bind_rows(map(outs, "ignition"))

ggplotly(plotSeasonCurves(bind_rows(map(outs, "data")), x="weekF"))
```





## Retrospective template curve estimation 

Using aligned time `newWeek`, estimate a common **template** curve on the link scale via:

$$
\mathrm{logit}\left(\pi_{s,w}\right) = b_s + f\!\left(w^a_{s,w}\right),
$$

where $w^a_{s,w}$ is the aligned week index (`newWeek`), $b_s$ is a season random intercept, and $f(\cdot)$ is a cyclic smooth (`s(newWeek, bs="cc", k=df)` inside `estimateRef()`).

```{r}
#| label: common-curve
#| echo: TRUE
alignedD <- alignIgnition(outs)
anchorWeek <- attr(alignedD, "anchorWeek")

ref <- estimateRef(
  alignedD  = alignedD,
  exSeason  = c("2011-12","2020-21","2021-22","2025-26"),
  k         = df,     
  n_weeks   = 52
)
```

```{r}
#| echo: FALSE

template_df <- ref$pred_df
g_ref_fun    <- ref$g_ref_fun
g_ref_mu_se  <- ref$g_ref_mu_se


ggplotly(
  ggplot() +
    geom_line(data = template_df, aes(x = newWeek, y = fit), linewidth = 1.2) +
    geom_ribbon(data = template_df, aes(x = newWeek, ymin = low, ymax = high), alpha = 0.2) +
    geom_line(data = ref$dat, aes(x = newWeek, y = p, group = season),
              colour = "grey70", linewidth = 0.6, alpha = 0.35) +
    geom_vline(xintercept = anchorWeek, linetype = "dashed") +
    labs(x = "newWeek", y = "Positivity")
)

```

The estimated common curve $f(\cdot)$ serves as a **template** for forecasting: once we detect ignition, we can use the template shape (plus derivatives) to help forecast the future trajectory.It is also used in peak timing forecast by continuously aligning with real-time observed data, allowing the template to stretch and shift as needed to match the observed curve. 




# Prospective 2-stage model training

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```


## $M_0$ — prospective ignition detection (offline tuning, online use)

We pose ignition detection as a **prospective** decision problem. For each season $s$ and week $w$, the detector uses information available up to week $w$ to decide whether ignition has occurred, and returns the **earliest** ignition week.

### Ignition gates

We restrict ignition to an eligible within-season week window:

$$
g_{\mathrm{win}}(s,w)=\mathbf{1}\{w_{\min}\le w^f_{s,w}\le w_{\max}\}.
$$


```{r, eval = F, echo = F, include = F}
ign_fit <- fitIgnition(
  dat = alignedD,
  event_k = 1L,
  lead = 1L,
  A_pre = 6L,
  B_post = 6L,
  k_week = 6L,
  k_p = 8L,
  k_fs = 4L,
  fit_base = TRUE,
  verbose = FALSE
)
```

We define four epidemiological gates used in voting. A fifth signal, the velocity gate (`cond_dp`), is computed and shown in the detection table for diagnostics but does not contribute to the vote.

::: panel-tabset

### Rolling-sum evidence gate

*Gate: `cond_sum` · Signal: `p_sumK` = $S_K(s,w)$ · Parameters: `K_sum` = $K_{\mathrm{sum}}$, `p_sum_thr` = $\tau_{\mathrm{sum}}$*

**Purpose.** A single elevated week could be a sampling artefact. Requiring the *sum* of positivity over several consecutive weeks to exceed a threshold filters out isolated spikes and demands that elevated signal be sustained across multiple observations before contributing a vote.

Let $K_{\mathrm{sum}}\ge 1$ and define a short rolling sum of weekly positivity:

$$
S_{K}(s,w)=\sum_{j=0}^{K_{\mathrm{sum}}-1} p_{s,w-j}.
$$

The rolling-sum gate is

$$
g_{\mathrm{sum}}(s,w)=
\mathbf{1}\left\{S_{K}(s,w)\ge \tau_{\mathrm{sum}}\right\}.
$$

### Smoothed positivity level gate

*Gate: `cond_p` · Signal: `p_sm` = $\tilde{p}_{s,w}$ · Parameters: `L` = $L$, `p_thr` = $\tau_p$*

**Purpose.** The raw weekly positivity $p_{s,w}$ is noisy. Smoothing over $L$ weeks suppresses week-to-week noise and confirms that the *current level* of positivity is meaningfully elevated rather than a transient fluctuation.

Let $\tilde p_{s,w}$ be a short rolling mean of weekly positivity with window length $L$:

$$
\tilde p_{s,w}=
\frac{1}{L}\sum_{j=0}^{L-1} p_{s,w-j},
\qquad L \in \{2,3\}.
$$

The level gate is

$$
g_{p}(s,w)=
\mathbf{1}\left\{\tilde p_{s,w}\ge \tau_{p}\right\}.
$$

### Cumulative prevalence gate

*Gate: `cond_prev` · Signal: `prev` = $\mathrm{Prv}_{s,w}$ · Parameter: `prev_thr` = $\tau_{\mathrm{prev}}$*

**Purpose.** A short burst of positive tests in a small sample can transiently elevate weekly positivity without representing a true population-level signal. The cumulative prevalence gate integrates evidence *across the entire season so far*, making it resistant to brief noise episodes: the season-wide test burden must reveal a meaningfully infected fraction of specimens.

Define cumulative prevalence evidence:

$$
\mathrm{Prv}_{s,w}=
\frac{\sum_{u \le w} y_{s,u}}{\sum_{u \le w} N_{s,u}}.
$$

The prevalence gate is

$$
g_{\mathrm{prev}}(s,w)=
\mathbf{1}\left\{\mathrm{Prv}_{s,w}\ge \tau_{\mathrm{prev}}\right\}.
$$

### Trend confirmation gate

*Gate: `cond_inc` · Signal: `dp` = $\Delta\tilde{p}_{s,w} = \tilde{p}_{s,w}-\tilde{p}_{s,w-1}$ · Parameters: `n_consec` = $n_{\mathrm{consec}}$, `eps` = $\epsilon$*

**Purpose.** Elevated positivity that is already declining is not ignition — it could be a receding false alarm. This gate requires that the smoothed signal has not been *sustainedly falling* over recent weeks, confirming that the elevated level is part of an ongoing rise rather than a peak already past. The noise tolerance $\epsilon$ allows one minor dip without disqualifying a genuine uptrend.

Let $k=n_{\mathrm{consec}}-1$ and $\epsilon \ge 0$ be a tolerance. We declare a sustained upward trend if at least $k-1$ of the last $k$ differences of `dp` exceed $-\epsilon$:

$$
g_{\mathrm{inc}}(s,w)=
\mathbf{1}\left\{
\sum_{j=0}^{k-1}\mathbf{1}\{\Delta \tilde p_{s,w-j} > -\epsilon\}\ \ge\ k-1
\right\}.
$$

Note that `cond_inc` checks *local direction* (week-to-week signs) but not magnitude. It passes as soon as positivity stops falling — even if it has been flat for weeks at a moderate level.

### Velocity gate (display only)

*Gate: `cond_dp` · Signal: `p_sm` lagged difference over `K_dp` weeks = $\tilde{p}_{s,w}-\tilde{p}_{s,w-K_{\mathrm{dp}}}$ · Parameters: `K_dp` = $K_{\mathrm{dp}}$, `dp_thr` = $\tau_{\mathrm{dp}}$*

**Note.** This gate is computed and included in the per-season detection table for diagnostic purposes, but it does **not** contribute to the $N_{\mathrm{req}}$-of-4 vote.

**Purpose.** The four voting gates can all be satisfied by a flat, moderately elevated signal. The velocity gate demands a *net displacement* of at least $\tau_{\mathrm{dp}}$ over the past $K_{\mathrm{dp}}$ weeks, distinguishing a genuinely rising epidemic from a flat signal sitting at a moderate level.

Let $K_{\mathrm{dp}}\ge 1$ be a look-back lag. The velocity gate is

$$
g_{\mathrm{dp}}(s,w)=
\mathbf{1}\left\{
\tilde p_{s,w}-\tilde p_{s,w-K_{\mathrm{dp}}}\ge \tau_{\mathrm{dp}}
\right\}.
$$

Unlike `cond_inc`, which checks whether each step is non-negative, `cond_dp` measures the *total rise* over the window.

:::

### Ignition rule: $N_{\mathrm{req}}$-of-4 voting

Define the vote count over the four epidemiological gates:

$$
H(s,w)=
g_{\mathrm{sum}}(s,w)+
g_{p}(s,w)+
g_{\mathrm{prev}}(s,w)+
g_{\mathrm{inc}}(s,w).
$$

Ignition is declared at the earliest week within the eligible window where all $N_{\mathrm{req}}$ gates agree:

$$
\hat w_s^{(\mathrm{ign})}=
\min\left\{
w:\
g_{\mathrm{win}}(s,w)=1,\
H(s,w)\ge N_{\mathrm{req}}
\right\},
$$

with $\hat w_s^{(\mathrm{ign})}=\mathrm{NA}$ if no week satisfies the rule. With $N_{\mathrm{req}}=4$, all four gates must agree simultaneously, ensuring that no single gate can trigger detection on its own.

---

### Leave One Season Out (LOSO) training of thresholds

To assess how well the ignition detector transfers to an unseen season, we use **leave-one-season-out (LOSO)** cross-validation. For each season $s$, we tune the rule thresholds using all seasons except $s$, then run the detector **prospectively** on the held-out season to obtain $\hat w_s^{(\mathrm{ign})}$. We summarize performance by aggregating the per-season ignition timing errors across folds, and select hyperparameters that minimize the chosen LOSO objective. This procedure reduces overfitting to idiosyncratic seasons and better reflects real deployment, where ignition must be detected for a new season without access to its future trajectory.


Let $w_s^{(\mathrm{ign})}$ be the ignition week for season $s$, $\theta$ denote the full set of thresholds and rule hyperparameters. 

$$
\theta = (\tau_{\mathrm{sum}},K_{\mathrm{sum}},\tau_p,\tau_{\mathrm{prev}}, n_{\mathrm{consec}}, L, \epsilon, N_{\mathrm{req}}, w_{\min}, w_{\max})
$$

Detection error for season $s$ under parameter set $\theta$ is

$$
\mathrm{diff}_s(\theta)=\hat w_s^{(\mathrm{ign})}(\theta)-w_s^{(\mathrm{ign})}.
$$

If no week fires within $[w_{\min}, w_{\max}]$, the detector assigns $\hat w_s^{(\mathrm{ign})} = w_{\max}$ as a fallback, so misses never occur.

Detection one week early is treated as exact via an adjusted error with a $-1$ wiggle room:

$$
\mathrm{adj\_diff}_s = \max(\mathrm{diff}_s,\,0)+\max(-1-\mathrm{diff}_s,\,0),
$$

so $\mathrm{adj\_diff}=0$ when $\mathrm{diff}\in\{-1,0\}$, and errors grow symmetrically beyond that window.

Per-season loss:

$$
\ell(\mathrm{adj\_diff}) = \mathrm{adj\_diff}
+ \gamma\max(\mathrm{adj\_diff}-2,\,0),
\qquad \gamma \ge 0,
$$

where the $\gamma$ term penalizes adjusted errors exceeding 2 weeks. Late and missing detections are not penalized separately.

A global tuning objective is

$$
\mathrm{Score}(\theta)=
\sum_s \ell(\mathrm{adj\_diff}_s(\theta))
+\lambda\max_s \mathrm{adj\_diff}_s(\theta).
$$

**Parameter selection.** Among all grid candidates, the best $\hat\theta$ is selected by the following lexicographic criterion (in priority order):

1. Minimize $n_{>2}(\theta)$: number of seasons with $\mathrm{adj\_diff}>2$.
2. Minimize $\max_s\,\mathrm{adj\_diff}_s(\theta)$.
3. Minimize $\mathrm{Score}(\theta)$ (final tiebreaker).

The tuned $\hat\theta$ is then applied prospectively in real time to determine ignition timing and trigger downstream forecasting.



```{r, eval = T}
if(file.exists("../data/stage1_tuning.rds")) {
  tuned <- readRDS("../data/stage1_tuning.rds")
} else { 
 
grid_loso <- data.table::CJ(
  # classifier gate excluded from vote (use_cls=FALSE); cls_thr kept for display only
  cls_thr   = 0.26,                            # fixed (display only)
  use_cls   = FALSE,                           # 4-gate voting: cond_sum, cond_p, cond_prev, cond_inc
  p_thr     = c(0.002, 0.003, 0.004, 0.005),  # shifted left: prev best was at lower boundary
  prev_thr  = c(0.001, 0.002, 0.003),          # shifted left: prev best was at lower boundary
  n_consec  = 5L,                              # fixed: stable optimum across two runs
  L         = 2L,                              # fixed at best
  eps       = 0,                               # fixed at best
  K_sum     = 5L,                              # fixed at best
  p_sum_thr = c(0.050, 0.055, 0.060),          # shifted right: prev best was at upper boundary
  N_req     = 4L,                              # require all 4 gates
  w_min     = 13L,
  w_max     = 26L,
  K_dp      = 3L,                              # fixed at best
  dp_thr    = 0.01,                            # fixed at best
  sorted    = FALSE
)
  
tuned<- loso_M0v2(
  dat  = alignedD,
  grid = as.data.frame(grid_loso),   # <-- the small LOSO grid we built
  score_col = "p_cls_p",
  drop_seasons = c("2015-16"),       # remove from LOSO folds
  exSeason_tune = NULL,              # optional extra exclusions inside tuning

  fit_args = list(
    fit_base = TRUE, fit_slope = FALSE, fit_fs = FALSE,
    event_k = 1L, lead = 1L, A_pre = 6L, B_post = 6L,
    k_week = 6L, k_p = 8L, k_fs = 4L,
    select = FALSE, verbose = FALSE
  ),

  tune_args = list(
    miss_penalty = 0,
    lambda = 20,
    kappa = 0,
    gamma = 25,
    gamma_late = 0,
    iWeek = TRUE,
    ncores = 10L,
    verbose = FALSE,
    progress_every = 200L
  ),

  verbose = TRUE
)

  saveRDS(tuned, "../data/stage1_tuning.rds")
}


tuned$compare %>%  gt()

```





::: panel-tabset

### LOSO results
```{r, eval = T}
tuned$runtime
keep_params <- c("p_thr","prev_thr","p_sum_thr","eps","dp_thr",
                 "n_consec","L","K_sum","N_req","w_min","w_max","K_dp")
bp_display  <- tuned$best_params[names(tuned$best_params) %in% keep_params]
data.frame(Parameter = names(unlist(bp_display)),
           Value     = unlist(bp_display), row.names = NULL) |>
  gt() |>
  tab_header(title = "Best M0 parameters (LOSO tuned)")

det_all <- detect_ignition_from_tuning(tuned, alignedD)
det_all$compare |>
  gt() |>
  tab_header(title = "Ignition detection: predicted vs truth") |>
  cols_label(iWeek_hat = "Detected week", iWeek_true = "True week", diff = "Diff") |>
  tab_style(style = cell_fill(color = "#fff3cd"),
            locations = cells_body(rows = abs(diff) > 2)) |>
  tab_footnote(
    footnote = "True week is the retrospective label from flagIgnition() fitted to the full available season data (smoothed positivity + derivative criteria). It is algorithm-defined, not an externally validated ground truth. For the current season (2025-26) this label is provisional and uses all data available at render time.",
    locations = cells_column_labels(columns = iWeek_true)
  )
```


```{r, eval = T, results = 'asis', echo = FALSE}
seasons_in_data <- sort(unique(det_all$data$season))
for (s in seasons_in_data) {
  cat(paste0("\n\n### ", s, "\n\n"))
  cat(gt::as_raw_html(plot_season_detection_table(det_all, s)))
  cat("\n")
}
```


:::

## $M_1$ - Prospective forecast training (offline training, online use)

After retrospective ignition labeling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week  $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$,and $w^{(\mathrm{ign})}_s$ be the ignition week for season $s$. We model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_w>0\}\;
  S_{2}^{h}\!\left(w^a_{s,w};\; k_w\right) \\[2pt]
&+ \mathbb{I}\{k_s>0\}\;
  S_{3,s}^{h}\!\left(w^a_{s,w};\;S, k_s\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_1>0\}\;
  S_{6}^{h}\!\left(d1_{s,w};\; k_1\right) \\[2pt]
&+ \mathbb{I}\{k_2>0\}\;
  S_{7}^{h}\!\left(d2_{s,w};\; k_2\right).
\end{aligned}
$$ {#eq-mod1}



### Model training (offline)
For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.


Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$  
  - Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
  - **Hyperparams:** none.  
  - **Toggle:** always on.  
  - **R:** `-1 + lead`.

###  $b_s$  
  - Season-level baseline shift shared across weeks/leads.  
  - **Hyperparams:** none. 
  - **Toggle:** always on.
  - **R:** `s(season, bs="re")`.  
  - **New-season prediction:** typically excluded (see “Prediction note” below).

### $z_{s,w}(\delta,K)$  

- Template backbone covariate: injects the aligned template shape (shifted by $\delta$) and ramps it in after ignition over $K \ne 0$ weeks.  

  $$
  z_{s,w}(\delta,K)=\omega(t_{s,w};K)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  \qquad
  t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s,
  $$
 where
  $$
  \omega(t;K)=\min\!\left(1,\max\!\left(0,\dfrac{t}{K}\right)\right)
  $$
- **Hyperparams:** 
  - shift $\delta$ (integer weeks)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning “no ramp”), 
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$), 
  - smooth basis $k_f$ **only if** `T="S"`.  
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.  
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`  
  - `T="O"`: `offset(logit_f_eff)`  
  - `T="N"`: (omit template term)


### $S_{2}^{h}(w^a_{s,w};k_w)$  
  - Global aligned-time correction, allows the model to “self-align” in real time with observed data
  - **Hyperparams:** $k_w$ (basis dimension).  
  - **Toggle:** on if $k_w>0$; off if $k_w=0$.  
  - **R:** `s(newWeek, by=lead, bs="ts", k=k_w)`.

### $S^{h}_{3,s}(w^a_{s,w};S,k_s)$  
  - Season-specific deviation over aligned time, allow each season deviate smoothly from the global aligned-time pattern.  
  - **Hyperparams:** $k_s$ (basis dimension).  
  - **Toggle:** on if $k_s>0$; off if $k_s=0$.  
  - **R:** `s(newWeek, by=interaction(season,lead), bs="fs", k=k_s, xt=...)`.  
  - **New-season prediction:** typically excluded because a brand-new season has an unseen factor level.

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts: $\alpha_{\text{state}}=1$ gives the raw logit (no smoothing); small $\alpha_{\text{state}}$ gives a long memory that changes slowly.

  - **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay, tuned jointly with $\lambda_w$) and $k_e$ (smooth basis dimension).
  - **Toggle:** on if $k_e>0$; off if $k_e=0$.
  - **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
  - **R (model):** `s(z_ema, by=lead, bs=”ts”, k=k_e)`.

### $S_{6}^{h}(d1_{s,w};k_1)$  
  - Prospective derivative (slope), captures local growth/decline momentum on the logit scale, estimated in real time  
  - **Hyperparams:** $k_1$ (basis dimension).  
  - **Toggle:** on if $k_1>0$; off if $k_1=0$.  
  - **R (feature):** `d1_now`.  
  - **R (model):** `s(d1_now, by=lead, bs="ts", k=k_1)`.

###  $S_{7}^{h}(d2_{s,w};k_2)$
  - Prospective derivative (curvature), captures acceleration/curvature on the logit scale (often the noisiest term), estimated in real time
  - **Hyperparams:** $k_2$ (basis dimension).
  - **Toggle:** on if $k_2>0$; off if $k_2=0$ (recommended if noisy).
  - **R (feature):** `d2_now`.
  - **R (model):** `s(d2_now, by=lead, bs="ts", k=k_2)`.

### $\lambda_w$

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition (pre-buffer rows have $t_{s,w}<0$). Normalising by $\bar{w}_s$ keeps the weighted log-likelihood on the same scale across $\lambda_w$ values. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Interpretation.** Larger $\lambda_w$ concentrates the fit on the first few post-ignition weeks and down-weights the tail of the season. This is appropriate because forecasts are consumed immediately after ignition; tail accuracy matters less in practice. The trade-off is that a very large $\lambda_w$ may overfit the early-season pattern across training seasons.

**Tuning.** $\lambda_w$ is treated as a hyperparameter and selected jointly with $\alpha_{\text{state}}$ by LOSO. However, evaluation during LOSO always uses **uniform weights** over the first `eval_window` post-ignition weeks, so all specs are compared on the same objective regardless of their $\lambda_w$ value — $\lambda_w$ governs what the model learns, not what it is judged on.

**R implementation:** weights are stored as column `.w` in `d_train` and passed to `mgcv::bam(weights = .w, ...)`. The tuned value is stored in the spec as `spec$lambda_w` and applied identically during the weekly refit.

### Late-season floor (`w_floor`, `t_floor_start`)

An aggressive $\lambda_w$ (e.g. 0.20) reduces the effective weight on late-season observations to near zero ($e^{-0.2 \times 30} \approx 0.002$). While this is intentional for the training objective, it leaves the smooth terms unconstrained in the late-season covariate range, producing numerically unstable extrapolation well beyond the epidemic peak.

To prevent this without distorting the pre-peak fit, we apply a **two-phase weight floor**:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) is the week beyond which the floor activates, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. Setting $t_{\text{floor}} = 14$ ensures the floor is inactive throughout the entire pre-peak window (historical median peak at $t \approx 9$, maximum at $t = 12$), so the tuned early-window behaviour is completely preserved. The floor is a **fitting-time regularisation only** — it is not a tuning hyperparameter and does not affect the LOSO objective.

**Practical effect:** observations at $t_{s,w} > 14$ receive at least 5% of the peak-week weight. This is enough to anchor the smooth terms in the late-season covariate range (preventing wild extrapolation) while keeping late-season predictions operationally meaningless weeks from influencing the peak-window fit. As a side effect, the REML optimisation converges significantly faster because the information matrix is no longer near-singular at late covariate values.

**R implementation:** `spec$w_floor` and `spec$t_floor_start` are read by both `train_stage2_joint_prepped()` (prospective pipeline) and `train_stage2_joint_m1()` (module tuning pipeline) and applied identically during the weekly refit.

:::

### Weekly refit (online use)

Let $s^*$ denote the current live season and $\mathcal{D}_{\text{hist}}$ the full historical dataset of all past seasons. At each week $w \geq w^{(\mathrm{ign})}_{s^*}$ after ignition detection, $M_1$ is refit on the augmented dataset

$$
\mathcal{D}_w = \mathcal{D}_{\text{hist}} \;\cup\; \bigl\{\,(s^*,\,w',\,x_{s^*,w'},\,n_{s^*,w'}) : w' \leq w\,\bigr\}.
$$

This allows the season random effect $b_{s^*}$ and the factor-smooth $S^h_{3,s^*}$ to be estimated from current-season observations, so both the level and shape of the forecast are corrected as data accumulate. All hyperparameters (including $\lambda_w$, $\alpha_{\text{state}}$, and basis dimensions $k_*$) are fixed at their LOSO-tuned values; only the GAM coefficients are updated each week.

::: panel-tabset

### Prospective derivatives estimation

For each season $s$ and each week $w$, we estimate derivatives on the **link scale** using only the observed history up to week $w$.
At week $w$, let $p_{s,w}=x_{s,w}/n_{s,w}$ be observed positivity (code: `p`). We work on the logit scale:

$$
z_{s,w} = \mathrm{logit}\!\left(\mathrm{clamp}(p_{s,w},\varepsilon,1-\varepsilon)\right) =  a + b\,u + c\,u^2
$$

To estimate derivatives at the boundary without peeking ahead, we fit a **local quadratic** using only the most recent $k$ points up to week $w$, time is centered at the evaluation week so that the first- and second-order derivatives at the boundary correspond directly to the linear and quadratic coefficients of the local polynomial fit.

- window: $\{w-k+1,\dots,w\}$ (clipped to available weeks)
- centered time: $u = w' - w$ so $u=0$ at the boundary week $w$
- fit: $z \approx a + b\,u + c\,u^2$

Then the prospective derivatives (link scale) at week $w$ are:

$$
d1_{s,w} = \left.\frac{dz}{dw}\right|_{w} = b,
\qquad
d2_{s,w} = \left.\frac{d^2 z}{dw^2}\right|_{w} = 2c.
$$

This estimator is **prospective-safe** because it only uses weeks $\le w$, and it is stable at the boundary because the polynomial is centered at the evaluation point.



```{r, eval = T}
#| label: compute-prospective-derivs
#| fig-height: 10
#| fig-width: 12
#| echo: TRUE


### Prospective Derivatives by season, using alignedD, i.e. real ignition week
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)
alignedD_long <- alignedD_prosp %>%select(season, weekF, d1_link, d2_link) %>%pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line() +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "w (within-season week)", y = "Derivative on logit scale", color = "")
)
```

### Hyperparameters tuning

We tune hyperparameters of $M_1$ using **leave-one-season-out (LOSO)** cross-validation.  
For each candidate spec (from `expand_grid_specs()`), we repeatedly:

1. Hold out one season $s$.
2. Fit the joint model on all remaining seasons.
3. Predict the held-out season **as a new season** by excluding season-dependent terms (e.g., `s(season)` and the `fs` term when present).
4. Score predictions using a **weighted** binomial negative log-likelihood (NLL).


- **$\delta$ (template shift):** integer weeks (typically $\{-3,\ldots,3\}$). Set `delta=NA` to disable the template covariate at the feature level.
- **`T` (template entry mode):** one of `{"S","O","N"}` (smooth, offset, none). *(In code: `template_mode ∈ {"smooth","offset","none"}`.)*
- **$Kr$ (ramp length):** integer with $Kr \ge 1$. Use $Kr=1$ for “no ramp” (immediate template effect). Set `Kr=NA` to disable the template covariate at the feature level.
- **$k_f$ (template smooth basis):** integer $\ge 1$ used only when `T="S"` (typical 4–12).
- **$k_w,k_s,k_e,k_1,k_2$ (basis dimensions):** integers $\ge 0$ with the on/off rule: **0 disables the term**; typical values 4–12.
- **$k_n$:** always fixed at 0 (logN term dropped; structurally redundant in binomial model).
- **$\alpha_{\text{state}}$ (EWMA decay):** real number in $(0,1)$ (grid $\{0.15, 0.25, 0.35\}$), tuned jointly with $\lambda_w$; only relevant when $k_e>0$.
- **$\lambda_w$ (time-decay weight):** non-negative real (grid $\{0, 0.05, 0.10, 0.20\}$); $\lambda_w=0$ is standard MLE.
- **$Kb$** integer $\ge 0$ (commonly 1–2), defines $M_1$ training start as `weekF >= iWeek_used - Kb`.

**Primary selection metric: mean NLL over the evaluation window**

Let $W_{\mathrm{eval}}$ denote the **evaluation window** — the number of post-ignition origin weeks used to score each held-out season. We set $W_{\mathrm{eval}} = 12$, covering the first twelve weeks after ignition detection. This value is **not tuned**; it is a fixed design choice calibrated to the observed ignition-to-peak distribution across historical seasons. Across ten evaluable seasons (excluding pandemic/anomalous years), the median time from ignition to peak positivity is **8.5 weeks** and the maximum is **12 weeks** (2018-19), so $W_{\mathrm{eval}} = 12$ ensures that the tuning objective covers the approach to peak — the period of greatest operational interest — in all historical seasons. Using a fixed window also ensures that all candidate specs are evaluated on exactly the same set of rows regardless of season length or tail behaviour.

The evaluation set for season $s$ is

$$
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\},
\qquad t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})},
$$

and the LOSO objective is the mean NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i).
$$

Evaluation uses **uniform weights** even when $\lambda_w > 0$ for training — $\lambda_w$ controls what the model learns, not what it is judged on. We select the best spec by minimizing $\overline{\mathrm{NLL}}$.

Secondary diagnostics: Brier and RMSE(p)

We also report weighted calibration/accuracy diagnostics:
$$
\mathrm{Brier} = \frac{\sum_i w_i(\hat p_i - x_i/n_i)^2}{\sum_i w_i},
\qquad
\mathrm{RMSE}(p)=\sqrt{\mathrm{Brier}}.
$$
```{r, eval = F}
# ---- Spec grid (shared by both tuning runs) ----
# Run this block + the refit-LOSO block below on the server (eval=F here).
#
# Grid reduction history (top-20% frequency analysis across runs):
#   k_e=4   dropped  — k_e=6 wins 100% of top-20%
#   k_1=6   dropped  — k_1=4 wins 100% of top-20%
#   k_s=6   dropped  — k_s=4 wins 100% of top-20% (eval_window=12 run)
#
# w_floor now tuned in the grid (previously hard-coded post-hoc to 0.05).
# lambda_w densified around the refit-LOSO optimum (~0.20).
nc <- min(parallel::detectCores() - 1L, 32L)

spec_grid <- expand_grid_specs(
  delta_grid  = -2:2,              # full range; -2 is weaker but kept for coverage
  Kr_grid     = 2:6,               # ramp length 2–6 weeks
  Kb_grid     = c(1L, 2L),        # training buffer around ignition
  T_grid      = "S",
  k_f_grid    = c(6L, 8L),        # template smooth df
  alpha_state = c(0.15, 0.25, 0.35),
  k_w_grid    = c(0L, 4L),        # week smooth on/off
  k_e_grid    = 6L,               # z_ema: 6 always wins
  k_n_grid    = 0L,               # logN: dropped (never meaningful)
  k_1_grid    = 4L,               # d1 slope: 4 always wins
  k_2_grid    = 0L,               # d2 curvature: dropped
  k_s_grid    = 4L,               # fs season-shape: 4 always wins
  w_floor_grid = c(0, 0.03, 0.05, 0.08),  # late-season weight floor (tuned here)
  verbose     = TRUE
)
# Expected grid size:
# delta(5) x Kr(5) x Kb(2) x k_f(2) x alpha(3) x k_w(2) x w_floor(4) = 2,400 base specs
# x lambda_w(7) in refit-LOSO tuning call = 16,800 total evaluations
```

```{r, eval = F}
# ---- Refit-LOSO tuning (correct objective for prospective deployment) ----
if (file.exists("../data/stage2_tuning_refit.rds")) {
  tuned2_refit <- readRDS("../data/stage2_tuning_refit.rds")
} else {
  tuned2_refit <- tune_stage2_loso_spec_grid_parallel(
    alignedD_prosp = alignedD_prosp,
    template_df    = template_df,
    spec_grid      = spec_grid,
    seasons        = NULL,
    eval_window    = 12L,
    refit_loso     = TRUE,
    lambda_w_grid  = c(0, 0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
    workers        = nc,
    chunk_size     = 8L,
    nthreads       = 1L,
    verbose        = TRUE
  )
  saveRDS(tuned2_refit, "../data/stage2_tuning_refit.rds")
}

ggplotly(plot_tune_stage2_heatmap(tuned2_refit, metric="mean_nll", normalize01=TRUE,
                          norm_scope="facet", agg=TRUE, base_size=10))
```

```{r, eval = F}
# ---- Standard LOSO tuning (optional — for historical in-sample plot only) ----
# lambda_w=0 dominates here; run only if you need a refreshed stage2_tuning.rds.
tuned2 <- tune_stage2_loso_spec_grid_parallel(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec_grid      = spec_grid,
  seasons        = NULL,
  eval_window    = 12L,
  refit_loso     = FALSE,
  lambda_w_grid  = c(0, 0.05, 0.10, 0.20),
  workers        = nc,
  chunk_size     = 8L,
  nthreads       = 1L,
  verbose        = TRUE
)
saveRDS(tuned2, "../data/stage2_tuning.rds")
```



### Traning results (true ingnition week)

```{r, eval = F, echo = T}
spec_stage2_best <- stage2_spec_from_tuning(tuned2)
# Drop non-scalar slots (formula, best_row, exclude_newseason) for display
skip_slots <- c("best_row", "formula", "exclude_newseason")
spec_display <- spec_stage2_best[!names(spec_stage2_best) %in% skip_slots]
data.frame(
  Specification = names(spec_display),
  Value = sapply(spec_display, function(x) {
    paste(as.character(x), collapse = ", ")
  }),
  row.names = NULL
) |> gt::gt() |> gt::tab_header(title = "Best M1 spec")



joint_out <- train_stage2_joint(
  dat         = alignedD_prosp,
  template_df = template_df,
  spec        = spec_stage2_best
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```


### Fit Summary
```{r, eval = F, echo = T}
summary(joint_out$fit)
```

### Refit-LOSO: best spec and fit by season

```{r refit-loso-fit, eval = F, echo = T, fig.height = 14, fig.width = 14}
tuned2_refit <- readRDS("../data/stage2_tuning_refit.rds")
spec_refit_best <- stage2_spec_from_tuning(tuned2_refit)
spec_refit_best$w_floor <- 0.05   # floor weight for late-season constraint

skip_slots <- c("best_row", "formula", "exclude_newseason")
spec_display_refit <- spec_refit_best[!names(spec_refit_best) %in% skip_slots]
data.frame(
  Specification = names(spec_display_refit),
  Value = sapply(spec_display_refit, function(x) paste(as.character(x), collapse = ", ")),
  row.names = NULL
) |> gt::gt() |> gt::tab_header(title = "Best M1 spec (refit-LOSO)")
```

```{r refit-loso-plot, eval = F, echo = T, fig.height = 14, fig.width = 14}
joint_out_refit <- train_stage2_joint(
  dat         = alignedD_prosp,
  template_df = template_df,
  spec        = spec_refit_best
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out_refit,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```
:::


# Prospective deployment and forecasting

Data from the 2025-26 season are used prospectively. Each week after ignition detection, we refit the Stage-2 GAM on all historical seasons **plus the current season's data observed so far**, using time-decay weights ($\lambda_w = 0.20$) that down-weight older observations. Season-specific terms ($b_s$, factor-smooth $S_3$) are excluded from predictions because 2025-26 is an unseen factor level in the base model; the correction comes through $z_\text{ema}$ and the updated weighted fit.

## Pre-trained model

All trained components are bundled into a single `kit` object. The refit-LOSO spec (`lambda_w = 0.20`, `w_floor = 0.05`) is used as the fixed hyperparameter set; only the GAM coefficients are updated each week.

| Variable | Source | Description |
|---|---|---|
| `template_df` | `ref$pred_df` | Smoothed reference curve (`newWeek`, `fit`). Drives $x^\delta_{s,w}$ and the ramp weight. |
| `ign_fit` | `fitIgnition()` | Stage-1 model object; `make_prospective_kit` extracts the GAM classifier automatically. |
| `tuned$best_params` | `loso_M0v2()` | LOSO-tuned Stage-1 threshold parameters. |
| `joint_out_refit` | `train_stage2_joint()` | Stage-2 training output using refit-LOSO best spec (includes `lambda_w`, `w_floor`). |

```{r prosp-kit, eval=F, echo=T}
kit <- make_prospective_kit(
  template_df   = template_df,
  ign_fit       = ign_fit,
  params_stage1 = tuned$best_params,
  joint_out     = joint_out_refit    # refit-LOSO spec: lambda_w=0.20, w_floor=0.05
)
```

## Current season snapshot

2025-26 is excluded from training data and is loaded here directly from the raw CSV.

```{r prosp-current, eval=F, echo=T}
nW_2025       <- get_full_cycle_weeks(2025)   # 53 MMWR weeks in 2025
currentSeason <- read.csv("../data/flu_testing_data.csv", stringsAsFactors = FALSE) %>%
  filter(season == "2025-26") %>%
  select(season, week, year, start_year = seasonstart,
         date = week_start_date, y = pos_flua, N = test_flu) %>%
  mutate(
    date  = as.Date(date),
    neg   = N - y,
    p     = y / N,
    weekF = ((week - startWeek) %% nW_2025) + 1L
  )
```

::: panel-tabset

### Plot

```{r, echo=F}
ggplotly(
  ggplot(currentSeason, aes(x = weekF, y = p)) +
    geom_point() + geom_line() +
    labs(x = "weekF", y = "Positivity (p)")
)
```

### Data

```{r, echo=F}
DT::datatable(currentSeason, options = list(pageLength = 25, scrollX = TRUE))
```

:::

## $M_0$: Walk-forward ignition detection (from weekF = 5)

`run_ignition_weekly` steps through each week from `start_week` and applies the Stage-1 detector with only data available at that week.

```{r prosp-ignition, eval=F, echo=T}
ign_out <- run_ignition_weekly(
  currentSeason  = currentSeason,
  ign_fit_or_gam = kit$stage1$gam_cls,
  params         = kit$stage1$params,
  start_week     = 5L
)
plots_ign        <- plot_ignition_weekly_snapshots(
  ign_out, currentSeason, facet = FALSE, start_week = 5L, maxWeek = 20
)
iWeek_hat_locked <- ign_out$iWeek_hat_locked
```

::: panel-tabset

```{r, results="asis", echo=F, eval=F}
for (nm in names(plots_ign)) {
  cat(knitr::knit_child(
    text = c(
      paste0("### ", nm), "",
      "```{r echo=FALSE}",
      sprintf("plotly::ggplotly(plots_ign[[%s]])", deparse(nm)),
      "```", ""
    ),
    envir = environment(), quiet = TRUE
  ), sep = "\n")
}
```

### Detection table

```{r, echo=F, eval=F}
local({
  df <- ign_out$df
  fmt_val <- function(val, flag) {
    v <- ifelse(is.na(val), "", sprintf("%.4f", as.numeric(val)))
    ifelse(flag %in% TRUE,
           paste0('<span style="color:red;font-weight:bold">', v, "</span>"), v)
  }
  disp <- data.frame(
    Week      = df$weekF,
    p         = round(df$p_now, 4),
    Sum5w     = fmt_val(df$cum_p_now, df$ok_cum_p    %in% TRUE),
    Prev      = fmt_val(df$prev_now,  df$ok_prev     %in% TRUE),
    Inc       = {
      v    <- ifelse(is.na(df$n_hit_now), "", as.character(as.integer(df$n_hit_now)))
      flag <- df$ok_n_consec %in% TRUE
      ifelse(flag, paste0('<span style="color:red;font-weight:bold">', v, "</span>"), v)
    },
    iWeek_hat = df$iWeek_hat_dynamic,
    stringsAsFactors = FALSE
  )
  highlight_rows <- which(df$ignite_ok_now %in% TRUE)
  DT::datatable(
    disp, rownames = FALSE, escape = FALSE,
    options = list(pageLength = 30, scrollX = TRUE, dom = "t")
  ) |>
    DT::formatStyle(
      columns         = "Week",
      target          = "row",
      backgroundColor = DT::styleRow(rows   = highlight_rows,
                                     values = rep("#fff3cd", length(highlight_rows)))
    )
})
```

:::

```{r, echo=F}
cat("Locked ignition week:", iWeek_hat_locked, "\n")
```

## Ignition week

Override manually if needed, or leave `NULL` to use the auto-detected week.

```{r prosp-iweek, eval=F, echo=T}
override_iWeek <- NULL   # set to e.g. 17L to force a specific week
iWeek_used <- if (!is.null(override_iWeek)) as.integer(override_iWeek) else iWeek_hat_locked
```

If `iWeek_used` is `NA` (ignition not yet detected), **stop here**.

## $M_1$: Walk-forward weekly refit forecast

Starting from the locked ignition week, the Stage-2 GAM is refit at each observed week using all historical seasons plus 2025-26 data up to that week. Season-specific terms are excluded during prediction (2025-26 is unseen at the base-model level); the real-time signal enters through $z_\text{ema}$ and the time-decay-weighted fit.

```{r prosp-walkforward, eval=F, echo=T}
if (is.na(iWeek_used)) {
  message("Ignition not detected (iWeek_used is NA). Skipping Stage-2 walk-forward.")
} else {
  n_weeks_season <- get_full_cycle_weeks(2025)

  # Build pseudo-prospective snapshots once.
  # Derivatives in each snapshot are prospective-safe (add_prospective_derivs_link
  # only uses data up to each snapshot's week).
  pp_full <- build_stage2_pseudo_prospective_list(
    currentSeason = currentSeason,
    kit           = kit,
    iWeek_hat     = iWeek_used,
    n_weeks       = n_weeks_season
  )

  # asof-week index for each pre-built snapshot.
  # The snapshot names are "newWeek=X_asofWeekF=Y"; extract Y.
  # Fallback: derive from toFit column if name pattern doesn't match.
  snap_wks <- {
    nm_raw <- names(pp_full$df)
    parsed <- suppressWarnings(as.integer(sub(".*_asofWeekF=(\\d+).*", "\\1", nm_raw)))
    if (any(is.na(parsed))) {
      vapply(pp_full$df, function(d) {
        fit_idx <- !is.na(d$toFit) & d$toFit == 1L
        if (any(fit_idx)) max(as.integer(d$weekF[fit_idx]), na.rm = TRUE) else NA_integer_
      }, integer(1L))
    } else parsed
  }

  # Walk-forward weeks: observed weeks at or after ignition that have a snapshot
  obs_weeks  <- sort(currentSeason$weekF[is.finite(currentSeason$p)])
  walk_weeks <- intersect(obs_weeks[obs_weeks >= iWeek_used], snap_wks)

  # ── Diagnostic: test a single prediction before the full loop ──────────────
  w_test   <- walk_weeks[1]
  joint_test <- suppressWarnings(refit_stage2_weekly(
    current_obs  = currentSeason[currentSeason$weekF <= w_test, ],
    iWeek_used   = iWeek_used,
    hist_data    = alignedD_prosp,
    template_df  = kit$stage2$template_df,
    spec         = kit$stage2$spec_stage2,
    season_label = "2025-26",
    verbose      = FALSE
  ))
  snap_test <- list(meta = pp_full$meta, df = pp_full$df[which(snap_wks == w_test)])
  ppp_test  <- stage2_predict_series(snap_test, joint_test$fit,
                                     which = "all", exclude_season_re = TRUE, interval = "ci")
  diag_df <- ppp_test[[1]]
  message("=== DIAG w=", w_test, " ===")
  message("lead levels in model: ", paste(levels(joint_test$fit$model$lead), collapse=", "))
  message("lead values in snapshot: ", paste(unique(as.character(snap_test$df[[1]]$lead)), collapse=", "))
  message("toFit==1 rows in snapshot: ", sum(!is.na(snap_test$df[[1]]$toFit) & snap_test$df[[1]]$toFit == 1L))
  message("p_hat_h1 non-NA: ", sum(!is.na(diag_df$p_hat_h1)))
  message("p_hat_h2 non-NA: ", sum(!is.na(diag_df$p_hat_h2)))
  message("cols in result: ", paste(names(diag_df), collapse=", "))
  # ────────────────────────────────────────────────────────────────────────────

  walk_preds <- vector("list", length(walk_weeks))
  names(walk_preds) <- paste0("w", walk_weeks)

  for (i in seq_along(walk_weeks)) {
    w <- walk_weeks[i]

    # Refit Stage-2 on historical + 2025-26 up to week w
    joint_w <- suppressWarnings(refit_stage2_weekly(
      current_obs  = currentSeason[currentSeason$weekF <= w, ],
      iWeek_used   = iWeek_used,
      hist_data    = alignedD_prosp,
      template_df  = kit$stage2$template_df,
      spec         = kit$stage2$spec_stage2,
      season_label = "2025-26",
      verbose      = FALSE
    ))

    # Select the pre-built snapshot for this week
    snap_idx <- which(snap_wks == w)
    pp_w     <- list(meta = pp_full$meta, df = pp_full$df[snap_idx])

    # Exclude season-specific terms: 2025-26 is not a level in the base model factor
    ppp_w <- stage2_predict_series(
      pp_w, joint_w$fit,
      which             = "all",
      exclude           = kit$stage2$exclude_terms,
      exclude_season_re = TRUE,
      interval          = "ci"
    )
    walk_preds[[i]] <- ppp_w[[1]]
  }
  walk_preds <- Filter(Negate(is.null), walk_preds)
}
```

```{r prosp-walkforward-plots, eval=F, echo=F}
if (!is.na(iWeek_used) && length(walk_preds) > 0) {
  plots_wf <- plot_stage2(
    walk_preds,
    ign_week = iWeek_used,
    facet     = FALSE,
    interval  = "ci",
    show_ref  = TRUE,
    show_pi   = TRUE
  )
}
```

::: panel-tabset

```{r, results="asis", echo=F, eval=F}
if (!is.na(iWeek_used) && exists("plots_wf") && length(plots_wf) > 0) {
  for (nm in names(plots_wf)) {
    wk       <- suppressWarnings(as.integer(sub("^w", "", nm)))
    post_ign <- if (is.finite(wk) && is.finite(iWeek_used)) wk - as.integer(iWeek_used) else NA
    tab_label <- if (is.finite(wk))
      sprintf("wk %d (+%d post-ign)", wk, post_ign) else nm

    cat(knitr::knit_child(
      text = c(
        paste0("### ", tab_label), "",
        "```{r echo=FALSE, fig.height=5, fig.width=9}",
        sprintf("plotly::ggplotly(plots_wf[[%s]])", deparse(nm)),
        "```", ""
      ),
      envir = environment(), quiet = TRUE
    ), sep = "\n")
  }
}
```

:::



# Summary
**Implementation map (functions / key objects).**

- Retrospective smoothing + derivatives: `estimateDerivs()` produces `fit`, `d1`, `d2` on the `weekF` scale.
- Retrospective ignition labeling + alignment (for template building): `flagIgnition()` (threshold rule) → `alignIgnition()` adds `iWeek`, `phase`, and aligned `newWeek` (attribute `anchorWeek`).
- Common template curve: `estimateRef()` fits the pooled cyclic GAMM and returns the template `pred_df` (columns `newWeek`, `fit`, `low`, `high`).
- Prospective derivatives (no look-ahead): `add_prospective_derivs_link()` adds `d1_link`, `d2_link` week-by-week.
- Stage 1 ignition detector (prospective gate): `fitIgnition()` → `tuneIgnitionGrid_M0v2()` / `loso_M0v2()` → `detectIgnitionBySeason_M0v2()`.
- Stage 2 forecast model: `tune_stage2_loso_shift_template()` → `prep_stage2_joint()` → `train_stage2_joint()` (Stage-2 design columns include `logit_f_eff`, `z_ema`, `logN_now`, `d1_now`, `d2_now`).




## Retro-estimation
- Identify training seasons
- Ignition specification for each season
  - Manual specification or estimation 
- Align Seasons curves
- Reference curve estimation

## Prospective training

- Ignition model specification and training
  - hyperparameter grid tuning using **true ignition week** for alignment and labeling
  - Apply ignition model on historical training data
- Forecast model specification and training
  - prospective derivative estimation
  - hyperparameter LOSO tuning 
  - Train final model on all data using **true ignition week** for alignment and labeling

## Prospective forecasting for new season

- Pull current season data (up to current week)
- Apply ignition model week by week until detection
- Once ignition is detected,prepare data for model forecast
  - Align with reference curve and
  - Prepare design matrix using tunned hyperparameters
- Make forecasts using the pre-trained forecast model, excluding season random effect for the new season
