---
title: "PAGe: Phase-Aligned Gated Epidemic Forecasting"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(gamm4)
library(gratia)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/retro_estimation.R")
source("../R/ignitionTraining.R")
source("../R/prospective_training.R")
source("../R/prospective_running.R")

# ---- Settings ----
startWeek   <- 27     # MMWR week number that defines weekF = 1
min_window  <- 10     # ignition search window at both ends
df          <- 10     # k for the common-curve smooth
p_thresh    <- 0.01
k1          <- 0.4
k2          <- -0.01

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}


# NOTE: replace with your real file path or a project-relative path
allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season,
    week,
    year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg = N - y,
    date = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))





```

# Notations 

Let $s = 1,\dots,S$ index influenza seasons and let $w$ denote the within-season flu-week number. For each season–week $(s,w)$, define:

- $x_{s,w}$: number of influenza-positive tests (code column `y`)  
- $n_{s,w}$: number of tests performed (code column `N`)  
- $p_{s,w} = x_{s,w} / n_{s,w}$: observed weekly positivity (code column `p`)  
- $\pi_{s,w}$: underlying probability of positivity (the quantity modeled/forecast)

In the data we keep several week indices:

- `week`: MMWR **epidemiologic week-of-year** from surveillance data.
- `weekS`: **surveillance-week index** (season starts at MMWR week 35), for interpretation only.
- `weekF`: $w_s^f$, **flu-week index** (flu season starts at MMWR week `startWeek`; here 27).
- `iWeek`: $w_s^{(\mathrm{ign})}$,  ignition week for season $s$ (user defined or estimated)
- `newWeek`: $w^a_{s,w}$ **aligned week index** after shifting $w^f_{s,w}$ so $w_s^{(\mathrm{ign})}$ lines up across seasons

Let $P(s,w)\in\{0,1\}$ be a phase indicator, with
$$
P(s,w)=
\begin{cases}
0, & \text{pre-ignition (pre-season)}\\
1, & \text{post-ignition (in-season)}
\end{cases}
$$




We define the framework as a mixture of two components: a “pre-ignition” component $M_0$, and a “post-ignition” component $M_1$, where


- **$M_0$ (Ignition detector):** using only information available up to week $w$, estimate whether the season has ignited (i.e., whether $P(s,w)=1$).
- **$M_1$ (Forecast model):** once ignition is detected, forecast h weeks ahead using the aligned template:

$$
\pi_{s,w+h} \;=\; \Pr\!\big(\text{positive at } w+h \mid \, P(s,w)=1,\, M_1\big),
$$


Primary forecasting targets: one- and two-week-ahead probabilities $\pi_{s,w+1}$ and $\pi_{s,w+2}$.


The remaining of this document is divided into three parts: 

- Estimation of ignition weeks and the common template curve using historical data
- Training of $M_0$ and $M_1$  using the ignition week label

$$
P(s,w)=\mathbb{1}\{w^f_{s,w}\ge w_s^{(\mathrm{ign})}\},
$$

<!-- We train a collection gating rules $M_0$ derived from historical data to produce a decision $\widehat P(s,w)\in\{0,1\}$ which is monitored weekly, once ignition is detected ($\widehat P(s,w)=1$), historical data is aligned and used to train the forecast model $M_1$. -->

- Application of $M_0$ and $M_1$ in real time, where $P_(s,w)$ is estimated on the fly and $M_1$ is calibrated in real time

In addition,  an R package **`flualign`** that implements the framework, including retrospective estimation, training and prospective running functions are demostrated as along the way

# Retrospective template estimation

This section explains how to estimate ignition weeks and the common template curve from historical data, which are used as training labels for $M_0$ and covariates for $M_1$.

## Season-level smoothing and derivatives estimation

For each season $s$, smooth positivity over $w$ to obtain a denoised curve $\tilde p_s(w)$ (stored as `fit` in `estimateDerivs()` output). From the fitted smooth compute:

$$
d1_{s,w} = \frac{d}{dw}\tilde p_s(w), \qquad
d2_{s,w} = \frac{d^2}{dw^2}\tilde p_s(w).
$$

**Implementation note.** Derivatives are computed from the fitted smooth (not finite differences of raw data), which is typically more stable. For ignition detection we work with derivatives on the **link (logit) scale** as returned by `estimateDerivs()` (code columns `d1` and `d2`).

```{r}
#| label: derivatives
res <- estimateDerivs(allD, k = 10)
```

::: panel-tabset


### Smoothed curves

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = p, group = season, colour = season)) +
    geom_point() +
    geom_line(aes(y = fit), linewidth = 1) +
    geom_ribbon(aes(ymin = fit_low, ymax = fit_high), alpha = 0.2) +
    labs(x = "weekF", y = "Positivity")
)
```

### First derivative

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = d1, group = season, colour = season)) +
    geom_line(linewidth = 1) +
    labs(x = "weekF", y = "d1 (derivative)")
)
```

### Second derivative

```{r}
#| echo: false
ggplotly(
  ggplot(res$data, aes(x = weekF, y = d2, group = season, colour = season)) +
    geom_line(linewidth = 1) +
    labs(x = "weekF", y = "d2 (second derivative)")
)
```

### Raw data snapshot

```{r}
#| echo: false
allD %>% head(20) %>% gt()
```

:::

## Retrospective ignition detection {#sec-ignition-detection}

Define the ignition week $w_s^{(\mathrm{ign})}$ as the earliest week satisfying fixed criteria such as

-   $\tilde p_s(w) > p_t$ (code: smoothed mean `fit` vs `p_thresh`),
-   $d1_{s,w} > k1$ (code: link-scale derivative `d1` vs `k1`),
-   $d2_{s,w} > k2$ (code: link-scale derivative `d2` vs `d2_relax = k2`).

Additional parameters (such as minimum and maximum weeks in which ignition is allowed) are controlled by the arguments to `flagIgnition()`.

```{r}
#| label: ignition
#| echo: TRUE
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df         = .x,
    p_thresh   = p_thresh,
    k1         = k1,
    k_c        = 0.01,
    n_consec   = 2,
    min_window = min_window,
    w_min      = 21,
    w_max      = 21,
    d2_relax   = k2
  ))
```


```{r}
#| echo: FALSE
theD <- bind_rows(map(outs, "data"))
ignD <- bind_rows(map(outs, "ignition"))

ggplotly(plotSeasonCurves(bind_rows(map(outs, "data")), x="weekF"))
```





## Retrospective template curve estimation 

Using aligned time `newWeek`, estimate a common **template** curve on the link scale via:

$$
\mathrm{logit}\left(\pi_{s,w}\right) = b_s + f\!\left(w^a_{s,w}\right),
$$

where $w^a_{s,w}$ is the aligned week index (`newWeek`), $b_s$ is a season random intercept, and $f(\cdot)$ is a cyclic smooth (`s(newWeek, bs="cc", k=df)` inside `estimateRef()`).

```{r}
#| label: common-curve
#| echo: TRUE
alignedD <- alignIgnition(outs)
anchorWeek <- attr(alignedD, "anchorWeek")

ref <- estimateRef(
  alignedD  = alignedD,
  exSeason  = c("2011-12","2020-21","2021-22","2025-26"),
  k         = df,     
  n_weeks   = 52
)
```

```{r}
#| echo: FALSE

template_df <- ref$pred_df
g_ref_fun    <- ref$g_ref_fun
g_ref_mu_se  <- ref$g_ref_mu_se


ggplotly(
  ggplot() +
    geom_line(data = template_df, aes(x = newWeek, y = fit), linewidth = 1.2) +
    geom_ribbon(data = template_df, aes(x = newWeek, ymin = low, ymax = high), alpha = 0.2) +
    geom_line(data = ref$dat, aes(x = newWeek, y = p, group = season),
              colour = "grey70", linewidth = 0.6, alpha = 0.35) +
    geom_vline(xintercept = anchorWeek, linetype = "dashed") +
    labs(x = "newWeek", y = "Positivity")
)

```

The estimated common curve $f(\cdot)$ serves as a **template** for forecasting: once we detect ignition, we can use the template shape (plus derivatives) to help forecast the future trajectory.It is also used in peak timing forecast by continuously aligning with real-time observed data, allowing the template to stretch and shift as needed to match the observed curve. 




# Prospective 2-stage model training



## $M_0$ — prospective ignition detection (offline tuning, online use)

We pose ignition detection as a **prospective** decision problem. For each season $s$ and week $w$, the detector uses information available up to week $w$ to decide whether ignition has occurred, and returns the **earliest** ignition week.

### Ignition gates

We restrict ignition to an eligible within-season week window:

$$
g_{\mathrm{win}}(s,w)=\mathbf{1}\{w_{\min}\le w^f_{s,w}\le w_{\max}\}.
$$


```{r, eval = T, echo = F, include = F}
ign_fit <- fitIgnition(
  dat = alignedD,
  event_k = 1L,
  lead = 1L,
  A_pre = 6L,
  B_post = 6L,
  k_week = 6L,
  k_p = 8L,
  k_fs = 4L,
  fit_base = TRUE,
  verbose = FALSE
)
```

We define four epidemiological gates:

::: panel-tabset

### Rolling-sum evidence gate

Let $K_{\mathrm{sum}}\ge 1$ and define a short rolling sum of weekly positivity:

$$
S_{K}(s,w)=\sum_{j=0}^{K_{\mathrm{sum}}-1} p_{s,w-j}.
$$

The rolling-sum gate is

$$
g_{\mathrm{sum}}(s,w)=
\mathbf{1}\left\{S_{K}(s,w)\ge \tau_{\mathrm{sum}}\right\}.
$$

### Smoothed positivity level gate

Let $\tilde p_{s,w}$ be a short rolling mean of weekly positivity with window length $L$:

$$
\tilde p_{s,w}=
\frac{1}{L}\sum_{j=0}^{L-1} p_{s,w-j},
\qquad L \in \{2,3\}.
$$

The level gate is

$$
g_{p}(s,w)=
\mathbf{1}\left\{\tilde p_{s,w}\ge \tau_{p}\right\}.
$$

### Cumulative prevalence gate

Define cumulative prevalence evidence:

$$
\mathrm{Prv}_{s,w}=
\frac{\sum_{u \le w} y_{s,u}}{\sum_{u \le w} N_{s,u}}.
$$

The prevalence gate is

$$
g_{\mathrm{prev}}(s,w)=
\mathbf{1}\left\{\mathrm{Prv}_{s,w}\ge \tau_{\mathrm{prev}}\right\}.
$$

### Trend confirmation gate (noise-tolerant)

Let $k=n_{\mathrm{consec}}-1$ and $\epsilon \ge 0$ be a tolerance. Define the week-to-week change in smoothed positivity:

$$
\Delta \tilde p_{s,w}=\tilde p_{s,w}-\tilde p_{s,w-1}.
$$

We declare a sustained upward trend if at least $k-1$ of the last $k$ differences exceed $-\epsilon$:

$$
g_{\mathrm{inc}}(s,w)=
\mathbf{1}\left\{
\sum_{j=0}^{k-1}\mathbf{1}\{\Delta \tilde p_{s,w-j} > -\epsilon\}\ \ge\ k-1
\right\}.
$$

This allows at most one small dip (up to $\epsilon$) over the last $k$ differences.

:::

### Ignition rule: $N_{\mathrm{req}}$-of-4 voting

Define the vote count over the four epidemiological gates:

$$
H(s,w)=
g_{\mathrm{sum}}(s,w)+
g_{p}(s,w)+
g_{\mathrm{prev}}(s,w)+
g_{\mathrm{inc}}(s,w).
$$

Ignition is declared at the earliest week within the eligible window where all $N_{\mathrm{req}}=4$ gates agree:

$$
\hat w_s^{(\mathrm{ign})}=
\min\left\{
w:\
g_{\mathrm{win}}(s,w)=1,\
H(s,w)\ge N_{\mathrm{req}}
\right\},
$$

with $\hat w_s^{(\mathrm{ign})}=\mathrm{NA}$ if no week satisfies the rule. Requiring all four gates simultaneously prevents false triggers from transient positivity spikes: a week with elevated rolling sum but no sustained rising trend (or vice versa) will not trigger ignition.

---

### Leave One Season Out (LOSO) training of thresholds

To assess how well the ignition detector transfers to an unseen season, we use **leave-one-season-out (LOSO)** cross-validation. For each season $s$, we tune the rule thresholds using all seasons except $s$, then run the detector **prospectively** on the held-out season to obtain $\hat w_s^{(\mathrm{ign})}$. We summarize performance by aggregating the per-season ignition timing errors across folds, and select hyperparameters that minimize the chosen LOSO objective. This procedure reduces overfitting to idiosyncratic seasons and better reflects real deployment, where ignition must be detected for a new season without access to its future trajectory.


Let $w_s^{(\mathrm{ign})}$ be the ignition week for season $s$, $\theta$ denote the full set of thresholds and rule hyperparameters. 

$$
\theta = (\tau_{\mathrm{sum}},K_{\mathrm{sum}},\tau_p,\tau_{\mathrm{prev}}, n_{\mathrm{consec}}, L, \epsilon, N_{\mathrm{req}}, w_{\min}, w_{\max})
$$

Detection error for season $s$ under parameter set $\theta$ is

$$
\mathrm{diff}_s(\theta)=\hat w_s^{(\mathrm{ign})}(\theta)-w_s^{(\mathrm{ign})}.
$$

A late-aware per-season loss is

$$
\ell(\mathrm{diff}) = |\mathrm{diff}|
+ \kappa\max(\mathrm{diff},0)
+ \gamma\max(|\mathrm{diff}|-2,\,0)
+ \gamma_{\mathrm{late}}\max(\mathrm{diff}-2,\,0),
\qquad \kappa,\gamma,\gamma_{\mathrm{late}} \ge 0,
$$

where the $\gamma$ term penalizes any error exceeding 2 weeks, and $\gamma_{\mathrm{late}}$ adds an additional penalty for being late by more than 2 weeks.

A global tuning objective is

$$
\mathrm{Score}(\theta)=
\sum_s \ell(\mathrm{diff}_s(\theta))
+\lambda\max_s |\mathrm{diff}_s(\theta)|
+\mathrm{miss\_penalty}\cdot n_{\mathrm{miss}}(\theta),
$$

where $n_{\mathrm{miss}}(\theta)$ counts seasons with $\hat w_s^{(\mathrm{ign})}(\theta)=\mathrm{NA}$.

**Parameter selection.** Among all grid candidates, the best $\hat\theta$ is selected by the following lexicographic criterion (in priority order):

1. Minimize $n_{>2}(\theta)$: number of seasons with $|\mathrm{diff}|>2$.
2. Minimize $n_{\mathrm{late},>2}(\theta)$: number of seasons with $\mathrm{diff}>2$.
3. Minimize $\max_s|\mathrm{diff}_s(\theta)|$.
4. Minimize $n_{\mathrm{miss}}(\theta)$.
5. Minimize $\mathrm{Score}(\theta)$ (final tiebreaker).

The tuned $\hat\theta$ is then applied prospectively in real time to determine ignition timing and trigger downstream forecasting.




::: panel-tabset

### LOSO tuning results
```{r, eval = T}
if(file.exists("../data/stage1_tuning.rds")) { 
  tuned <- readRDS("../data/stage1_tuning.rds")
} else { 
 
grid_loso <- data.table::CJ(
  # classifier gate excluded from vote (use_cls=FALSE); cls_thr kept for display only
  cls_thr   = 0.26,                     # fixed (display only)
  use_cls   = FALSE,                    # 4-gate voting: cond_sum, cond_p, cond_prev, cond_inc
  p_thr     = c(0.008, 0.009, 0.010),  # smoothed-level threshold
  prev_thr  = c(0.005, 0.006, 0.007),  # cumulative prevalence threshold
  n_consec  = c(3L, 4L, 5L),           # trend gate window
  L         = 2L,                       # fixed at best
  eps       = 0,                        # fixed at best
  K_sum     = 5L,                       # fixed at best
  p_sum_thr = c(0.045, 0.050, 0.055),  # rolling-sum threshold (N_req=4 reinstates ~0.050)
  N_req     = 4L,                       # require all 4 epi gates (prevents false positives)
  w_min     = 13L,
  w_max     = 30L,
  sorted    = FALSE
)
  
tuned<- loso_M0v2(
  dat  = alignedD,
  grid = as.data.frame(grid_loso),   # <-- the small LOSO grid we built
  score_col = "p_cls_p",
  drop_seasons = c("2015-16"),       # remove from LOSO folds
  exSeason_tune = NULL,              # optional extra exclusions inside tuning

  fit_args = list(
    fit_base = TRUE, fit_slope = FALSE, fit_fs = FALSE,
    event_k = 1L, lead = 1L, A_pre = 6L, B_post = 6L,
    k_week = 6L, k_p = 8L, k_fs = 4L,
    select = FALSE, verbose = FALSE
  ),

  tune_args = list(
    miss_penalty = 20,
    lambda = 20,
    kappa = 2,
    gamma = 25,
    gamma_late = 25,
    iWeek = TRUE,
    ncores = 10L,
    verbose = FALSE,
    progress_every = 200L
  ),

  verbose = TRUE
)

  saveRDS(tuned, "../data/stage1_tuning.rds") 
} 

tuned$runtime
data.frame(Parameter = names(unlist(tuned$best_params)),
           Value     = unlist(tuned$best_params), row.names = NULL) |>
  gt() |>
  tab_header(title = "Best M0 parameters (LOSO tuned)")
# Note: tuned$best_params is the median (continuous params) / mode (integer params)
# of fold-specific best parameters across all LOSO folds.

```


### LOSO tuned ignition detection
```{r, eval = T}
det_all <- detect_ignition_from_tuning(tuned, alignedD)
det_all$compare |>
  gt() |>
  tab_header(title = "Ignition detection: predicted vs truth") |>
  cols_label(iWeek_hat = "Detected week", iWeek_true = "True week", diff = "Diff") |>
  tab_style(style = cell_fill(color = "#fff3cd"),
            locations = cells_body(rows = abs(diff) > 2)) |>
  tab_footnote(
    footnote = "True week is the retrospective label from flagIgnition() fitted to the full available season data (smoothed positivity + derivative criteria). It is algorithm-defined, not an externally validated ground truth. For the current season (2025-26) this label is provisional and uses all data available at render time.",
    locations = cells_column_labels(columns = iWeek_true)
  )
```


### Plot

```{r, eval = T}

plot_ignition_detect_vs_truth(det_all, x_max = 30, y_max = 0.3, ncol = 4)
```

### Per-season detection

```{r}
#| results: asis
#| echo: false
seasons_in_data <- sort(unique(det_all$data$season))
for (s in seasons_in_data) {
  cat(knitr::knit_child(
    text = c(
      paste0("#### ", s),
      "",
      "```{r echo=FALSE}",
      sprintf("plot_season_detection_table(det_all, %s)", deparse(s)),
      "```",
      ""
    ),
    envir = environment(),
    quiet = TRUE
  ), sep = "\n")
}
```
:::




## $M_1$ - Prospective forecast training (offline training, online use)

After retrospective ignition labeling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week  $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$,and $w^{(\mathrm{ign})}_s$ be the ignition week for season $s$. We model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_w>0\}\;
  S_{2}^{h}\!\left(w^a_{s,w};\; k_w\right) \\[2pt]
&+ \mathbb{I}\{k_s>0\}\;
  S_{3,s}^{h}\!\left(w^a_{s,w};\;S, k_s\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_n>0\}\;
  S_{5}^{h}\!\left(\log n_{s,w};\; k_n\right) \\[2pt]
&+ \mathbb{I}\{k_1>0\}\;
  S_{6}^{h}\!\left(d1_{s,w};\; k_1\right) \\[2pt]
&+ \mathbb{I}\{k_2>0\}\;
  S_{7}^{h}\!\left(d2_{s,w};\; k_2\right).
\end{aligned}
$$ {#eq-mod1}

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term. The model is fit retrospectively using the aligned historical data, and then applied prospectively in real time after ignition is detected.


Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$  
  - Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
  - **Hyperparams:** none.  
  - **Toggle:** always on.  
  - **R:** `-1 + lead`.

###  $b_s$  
  - Season-level baseline shift shared across weeks/leads.  
  - **Hyperparams:** none. 
  - **Toggle:** always on.
  - **R:** `s(season, bs="re")`.  
  - **New-season prediction:** typically excluded (see “Prediction note” below).

### $z_{s,w}(\delta,K)$  

- Template backbone covariate: injects the aligned template shape (shifted by $\delta$) and ramps it in after ignition over $K \ne 0$ weeks.  

  $$
  z_{s,w}(\delta,K)=\omega(t_{s,w};K)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  \qquad
  t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s,
  $$
 where
  $$
  \omega(t;K)=\min\!\left(1,\max\!\left(0,\dfrac{t}{K}\right)\right)
  $$
- **Hyperparams:** 
  - shift $\delta$ (integer weeks)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning “no ramp”), 
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$), 
  - smooth basis $k_f$ **only if** `T="S"`.  
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.  
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`  
  - `T="O"`: `offset(logit_f_eff)`  
  - `T="N"`: (omit template term)


### $S_{2}^{h}(w^a_{s,w};k_w)$  
  - Global aligned-time correction, allows the model to “self-align” in real time with observed data
  - **Hyperparams:** $k_w$ (basis dimension).  
  - **Toggle:** on if $k_w>0$; off if $k_w=0$.  
  - **R:** `s(newWeek, by=lead, bs="ts", k=k_w)`.

### $S^{h}_{3,s}(w^a_{s,w};S,k_s)$  
  - Season-specific deviation over aligned time, allow each season deviate smoothly from the global aligned-time pattern.  
  - **Hyperparams:** $k_s$ (basis dimension).  
  - **Toggle:** on if $k_s>0$; off if $k_s=0$.  
  - **R:** `s(newWeek, by=interaction(season,lead), bs="fs", k=k_s, xt=...)`.  
  - **New-season prediction:** typically excluded because a brand-new season has an unseen factor level.

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$  
  - EMA state (current level memory), stabilizes the “current level” of activity using an EWMA on the logit scale, estimated in real time 
  - **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay) and $k_e$ (basis dimension).  
  - **Toggle:** on if $k_e>0$; off if $k_e=0$.  
  - **R (feature):** `z_ema` computed in `prep_stage2_joint()` using `alpha_state`.  
  - **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{5}^{h}(\log n_{s,w};k_n)$  
  - Testing volume effect , allows dependence on current testing intensity.  
  - **Hyperparams:** $k_n$ (basis dimension).  
  - **Toggle:** on if $k_n>0$; off if $k_n=0$.  
  - **R (feature):** `logN_now`.  
  - **R (model):** `s(logN_now, by=lead, bs="ts", k=k_n)`.

### $S_{6}^{h}(d1_{s,w};k_1)$  
  - Prospective derivative (slope), captures local growth/decline momentum on the logit scale, estimated in real time  
  - **Hyperparams:** $k_1$ (basis dimension).  
  - **Toggle:** on if $k_1>0$; off if $k_1=0$.  
  - **R (feature):** `d1_now`.  
  - **R (model):** `s(d1_now, by=lead, bs="ts", k=k_1)`.

###  $S_{7}^{h}(d2_{s,w};k_2)$  
  - Prospective derivative (curvature), captures acceleration/curvature on the logit scale (often the noisiest term), estimated in real time
  - **Hyperparams:** $k_2$ (basis dimension).  
  - **Toggle:** on if $k_2>0$; off if $k_2=0$ (recommended if noisy).  
  - **R (feature):** `d2_now`.  
  - **R (model):** `s(d2_now, by=lead, bs="ts", k=k_2)`.



:::

::: panel-tabset

### Prospective derivatives estimation

For each season $s$ and each week $w$, we estimate derivatives on the **link scale** using only the observed history up to week $w$.
At week $w$, let $p_{s,w}=x_{s,w}/n_{s,w}$ be observed positivity (code: `p`). We work on the logit scale:

$$
z_{s,w} = \mathrm{logit}\!\left(\mathrm{clamp}(p_{s,w},\varepsilon,1-\varepsilon)\right) =  a + b\,u + c\,u^2
$$

To estimate derivatives at the boundary without peeking ahead, we fit a **local quadratic** using only the most recent $k$ points up to week $w$, time is centered at the evaluation week so that the first- and second-order derivatives at the boundary correspond directly to the linear and quadratic coefficients of the local polynomial fit.

- window: $\{w-k+1,\dots,w\}$ (clipped to available weeks)
- centered time: $u = w' - w$ so $u=0$ at the boundary week $w$
- fit: $z \approx a + b\,u + c\,u^2$

Then the prospective derivatives (link scale) at week $w$ are:

$$
d1_{s,w} = \left.\frac{dz}{dw}\right|_{w} = b,
\qquad
d2_{s,w} = \left.\frac{d^2 z}{dw^2}\right|_{w} = 2c.
$$

This estimator is **prospective-safe** because it only uses weeks $\le w$, and it is stable at the boundary because the polynomial is centered at the evaluation point.



```{r}
#| label: compute-prospective-derivs
#| fig-height: 10
#| fig-width: 12
#| echo: TRUE


### Prospective Derivatives by season, using alignedD, i.e. real ignition week
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)
alignedD_long <- alignedD_prosp %>%select(season, weekF, d1_link, d2_link) %>%pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line() +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "w (within-season week)", y = "Derivative on logit scale", color = "")
)
```

### Hyperparameters tuning

We tune hyperparameters of $M_1$ using **leave-one-season-out (LOSO)** cross-validation.  
For each candidate spec (from `expand_grid_specs()`), we repeatedly:

1. Hold out one season $s$.
2. Fit the joint model on all remaining seasons.
3. Predict the held-out season **as a new season** by excluding season-dependent terms (e.g., `s(season)` and the `fs` term when present).
4. Score predictions using a **weighted** binomial negative log-likelihood (NLL).


- **$\delta$ (template shift):** integer weeks (typically $\{-3,\ldots,3\}$). Set `delta=NA` to disable the template covariate at the feature level.
- **`T` (template entry mode):** one of `{"S","O","N"}` (smooth, offset, none). *(In code: `template_mode ∈ {"smooth","offset","none"}`.)*
- **$Kr$ (ramp length):** integer with $Kr \ge 1$. Use $Kr=1$ for “no ramp” (immediate template effect). Set `Kr=NA` to disable the template covariate at the feature level.
- **$k_f$ (template smooth basis):** integer $\ge 1$ used only when `T="S"` (typical 4–12).
- **$k_w,k_s,k_e,k_n,k_1,k_2$ (basis dimensions):** integers $\ge 0$ with the on/off rule: **0 disables the term**; typical values 4–12.
- **$\alpha_{\text{state}}$ (EWMA decay):** real number in $(0,1)$ (typical grid 0.15–0.50),only relevant when $k_e>0$.
- **$Kb$** integer $\ge 0$ (commonly 0–2), defines $M_1$ training start as `weekF >= iWeek_used - Kb`.

**Primary selection metric: NLL**


Define target-week time since ignition:
$$
t_{\text{target}} = (w + h) - w_s^{(\mathrm{ign})}.
$$

We emphasize the first $k_t$ weeks after ignition (on the **target week**) by using weights:
$$
w(t_{\text{target}})=
\begin{cases}
w_{\text{early}}, & 0 \le t_{\text{target}} \le k_t,\\
1, & t_{\text{target}} > k_t.
\end{cases}
$$

The **weighted NLL**:
$$
\mathrm{NLL} =
-\sum_i w_i \log \Pr\!\left(x_i \mid n_i,\hat p_i\right),
\qquad
\mathrm{mean\_nll} = \frac{\mathrm{NLL}}{\sum_i w_i}.
$$

We select the best spec by **minimizing** `mean_nll`.

Secondary diagnostics: Brier and RMSE(p)

We also report weighted calibration/accuracy diagnostics:
$$
\mathrm{Brier} = \frac{\sum_i w_i(\hat p_i - x_i/n_i)^2}{\sum_i w_i},
\qquad
\mathrm{RMSE}(p)=\sqrt{\mathrm{Brier}}.
$$
```{r, eval = T}

if (file.exists("../data/stage2_tuning.rds")) {
  tuned2 <- readRDS("../data/stage2_tuning.rds")
} else {
  

spec_grid <- expand_grid_specs(
  delta_grid = -3:3,
  Kr_grid    = 1:4,
  Kb_grid    = c(0L, 1L, 2L),
  T_grid     = c("O","S"),
  k_f_grid   = c(6L, 8L, 10L),
  alpha_state = c(0.25),
  k_w_grid = c(6L, 8L),
  k_e_grid = c(0L, 6L),
  k_n_grid = c(0L, 6L),
  k_1_grid = c(4L, 6L),
  k_2_grid = c(0L),
  k_s_grid = c(4L,6L,10L),
  verbose = TRUE
)
  
  
tuned2 <- tune_stage2_loso_spec_grid_parallel(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec_grid      = spec_grid,
  ignD           = ignD,
  k_t            = 10L,
  w_early        = 2,
  workers        = 10L,
  strategy       = "auto",   # Windows -> multisession, Linux -> multicore
  chunk_size     = 4L,       # reduce I/O per worker; 4–16 is typical
  nthreads       = 1L,       # keep 1 when parallel
  verbose        = TRUE
)
  saveRDS(tuned2, "../data/stage2_tuning.rds")
}



ggplotly(plot_tune_stage2_heatmap(tuned2, metric="mean_nll", normalize01=TRUE,
                          norm_scope="facet", agg=TRUE, base_size=10))
```



### Traning results (true ingnition week)

```{r, eval = T, echo = T}
spec_stage2_best <- stage2_spec_from_tuning(tuned2)
spec_display <- within(spec_stage2_best, rm(best_row))
data.frame(
  Specification = names(spec_display),
  Value = sapply(spec_display, function(x) {
    if (inherits(x, "formula")) paste(deparse(x), collapse = " ") else paste(as.character(x), collapse = ", ")
  }),
  row.names = NULL
) |> gt() |> tab_header(title = "Best M1 spec")



joint_out <- train_stage2_joint(
  dat         = alignedD_prosp,
  template_df = template_df,
  spec        = spec_stage2_best
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```


### Fit Summary
```{r, eval = T, echo = T}
summary(joint_out$fit)
```
:::


### Offline calibration for prospective probability accuracy

::: panel-tabset

### Description
The Stage-2 model is trained on historical seasons and produces a linear predictor $\eta^{h}_{s,w}$ (equivalently a raw probability $\hat\pi^{h}_{s,w}=\mathrm{logit}^{-1}(\eta^{h}_{s,w})$). However, strong in-sample likelihood fit does not guarantee that $\hat\pi^{h}_{s,w}$ is well calibrated when applied to a **new season**. In particular, predicted probabilities can be systematically biased or overly confident even when point forecasts appear reasonable.

To improve new-season probability accuracy, we learn an **offline calibration map** from **out-of-sample** predictions. For each horizon $h\in\{1,2\}$, we apply Platt scaling on the logit scale:
$$
\mathrm{logit}\!\left(\pi^{h,\mathrm{cal}}_{s,w}\right)
\;=\;
a_h \;+\; b_h\,\mathrm{logit}\!\left(\hat\pi^{h}_{s,w}\right),
$$
where $a_h$ adjusts an overall bias and $b_h$ rescales confidence (values $b_h<1$ shrink extreme probabilities toward $0.5$; values $b_h>1$ sharpen them).

### $(a_h,b_h)$ training using LOSO 

A key point is that $(a_h,b_h)$ are *not* estimated from in-sample fitted values. Instead, we construct a calibration training set using leave-one-season-out (LOSO) cross-fitting:

1. Hold out season $s$.
2. Fit the Stage-2 model on seasons $\{s'\neq s\}$ (with Stage-2 hyperparameters fixed).
3. Predict $\hat\pi^{h}_{s,w}$ for all weeks $w$ in the held-out season and both horizons $h$.
4. Pair each prediction with the observed outcome $(x_{s,w+h},n_{s,w+h})$.

Stacking the held-out predictions across seasons yields a single out-of-sample dataset
$$
\mathcal{D}_{\mathrm{cal}}^{\mathrm{LOSO}}
=
\bigcup_{s\in\mathcal{S}}
\Big\{(s,w,h,x_{s,w+h},n_{s,w+h},\hat\pi^{h}_{s,w})\Big\},
$$
on which $(a_h,b_h)$ are estimated. Because each $\hat\pi^{h}_{s,w}$ is generated by a Stage-2 model trained **without** season $s$, this calibration map directly targets the prospective setting where the season is unknown at training time.

Offline calibration $(a_h,b_h)$ is estimated once from $\mathcal{D}_{\mathrm{cal}}^{\mathrm{LOSO}}$ and then held fixed during deployment, providing an immediate adjustment to probability scale even before substantial current-season outcomes are available.


```{r}

if(file.exists("../data/stage2_calib_loso.rds")) {
  d_calib_oos <- readRDS("../data/stage2_calib_loso.rds")
} else {
d_calib_oos <- make_calib_train_loso(joint_out, parallel = TRUE, workers = 12)
saveRDS(d_calib_oos, "../data/stage2_calib_loso.rds")
}


# 2) train calibrator
cal_off <- train_calib_platt(d_calib_oos)
plot_fit_obs_cal_by_season(joint_out = joint_out, cal = cal_off, interactive = TRUE)

```
:::


### Online calibration for real time adaption

::: panel-tabset

### Description

In prospective deployment, the current season can be systematically higher or lower than historical seasons even after phase alignment and gating. Rather than refitting the full Stage-2 model online, we introduce a lightweight **logit-scale intercept shift** that adapts as new data arrive.

Let $\eta^{h,\mathrm{cal}}_{s,w} = a_h + b_h\,\eta^{h}_{s,w}$ denote the offline-calibrated linear predictor (equivalently, $\eta^{h,\mathrm{cal}}_{s,w} = \mathrm{logit}(\pi^{h,\mathrm{cal}}_{s,w})$), where $\eta^{h}_{s,w}$ is the raw Stage-2 linear predictor from Eq. (@eq-mod1) evaluated prospectively (e.g., excluding season-specific effects). We define the online-adjusted predictor
$$
\mathrm{logit}\!\left(\tilde\pi^{h}_{s,w}\right)
\;=\;
\eta^{h,\mathrm{cal}}_{s,w}
\;+\;
c_{s,w},
$$
where $c_{s,w}\in\mathbb{R}$ is a scalar estimated at within-season week $w$ for season $s$. The same $c_{s,w}$ is applied to both horizons $h\in\{1,2\}$, so the updater corrects a **level shift** on top of the already-calibrated predictor, without altering Stage-2 shape or lead-specific dynamics.

At each week $w$ (once outcomes become available), we re-estimate $c_{s,w}$ using an expanding window of origins $u\le w$, starting at $w_s^{(\mathrm{ign})}-k_{\mathrm{pre}}$ to allow up to $k_{\mathrm{pre}}\ge 0$ pre-ignition weeks. Treating $\eta^{h,\mathrm{cal}}_{s,u}$ as a fixed offset, we estimate $c_{s,w}$ by maximizing a penalized binomial log-likelihood over all outcomes already observed by week $w$:
$$
\hat c_{s,w}
=
\arg\max_{c\in\mathbb{R}}
\left\{
\sum_{u\,=\,w_s^{(\mathrm{ign})}-k_{\mathrm{pre}}}^{w}
\ \sum_{h\in\{1,2\}}
\mathbf{1}\{u+h\le w\}\;
\log \Pr\!\Big(x_{s,u+h}\,\big|\, n_{s,u+h},\ \mathrm{logit}^{-1}\!\left(\eta^{h,\mathrm{cal}}_{s,u}+c\right)\Big)
\;-\;
\lambda_c \, c^2
\right\},
$$
where $x_{s,u+h}$ and $n_{s,u+h}$ are the observed positives and totals at week $u+h$, and $\lambda_c\ge 0$ shrinks $c$ toward $0$ for stability (especially early in the season). The Stage-2 model and offline calibration map $(a_h,b_h)$ are both fixed during deployment.

### Penalty $\lambda_c$ tuning

We choose $\lambda_c$ offline via LOSO by *replaying* the online updating procedure under the same information constraints as deployment. The LOSO out-of-sample predictions $\mathcal{D}_{\mathrm{cal}}^{\mathrm{LOSO}}$ and offline calibration map $(a_h,b_h)$ are pre-computed by `make_calib_train_loso` and `train_calib_platt` respectively; `tune_lambda_c` only performs the simulation below. For each candidate $\lambda_c\in\Lambda$ and each held-out season $s$:

1. Form the calibrated linear predictor for season $s$ using the pre-computed offline calibration map,
   $$
   \eta^{h,\mathrm{cal}}_{s,w}
   \;=\;
   a_h + b_h\,\mathrm{logit}\!\left(\hat\pi^{h}_{s,w}\right),
   $$
   where $\hat\pi^{h}_{s,w}$ are the LOSO out-of-sample predictions from $\mathcal{D}_{\mathrm{cal}}^{\mathrm{LOSO}}$.

2. For each week $w$ in season $s$, estimate $c_{s,w}$ using only outcomes already observed by week $w$ (i.e., include an origin $u$ only if $u+h\le w$):
   $$
   \hat c_{s,w}(\lambda_c)
   =
   \arg\max_{c\in\mathbb{R}}
   \left[
   \sum_{u=w_s^{(\mathrm{ign})}-k_{\mathrm{pre}}}^{w}
   \ \sum_{h\in\{1,2\}}
   \mathbf{1}\{u+h\le w\}\;
   \log \Pr\!\Big(x_{s,u+h}\,\big|\,
   n_{s,u+h},\ \mathrm{logit}^{-1}(\eta^{h,\mathrm{cal}}_{s,u}+c)\Big)
   \;-\;
   \lambda_c\,c^2
   \right].
   $$

3. Score the resulting forecasts
   $$
   \tilde\pi^{h}_{s,w}(\lambda_c)
   =
   \mathrm{logit}^{-1}\!\left(\eta^{h,\mathrm{cal}}_{s,w}+\hat c_{s,w}(\lambda_c)\right)
   $$
   using binomial negative log-likelihood aggregated over $w$ and $h$.

We repeat across all seasons and select the $\lambda_c$ that minimizes the average held-out score.

```{r, eval = T}

res_lambda <- tune_lambda_c(
  d_calib_oos = d_calib_oos,
  cal_off     = cal_off,
  ign_df      = ign_fit$iWeek_by_season,
  time_col    = "newWeek",
  k_pre       = spec_stage2_best$pre_buffer,
  lambda_grid = c(500, 800, 1000, 1250, 1600, 2000), 
  parallel    = TRUE,
  workers     = 12
)

res_lambda$best$lambda_c

plot_fit_obs_cal_online_by_season(
  joint_out    = joint_out,
  cal          = cal_off,
  lambda_c     = res_lambda$best$lambda_c,   # your best lambda
  k_pre        =  spec_stage2_best$pre_buffer,  # whatever you selected already
  interactive  = TRUE
)

```
:::

# Prospective deployment and forecasting 

Data from 2025-26 season are used prospectively, taking each week start at a given week as a snapshot, we produce a forecast at each week and evaluate how the forecast evolves as more data become available and the season unfolds. The workflow is as follows:with the following workflow:

In deployment, we apply these layers sequentially: 

- Ignition detection ($M_0$) produces an estimated ignition week $w^{(\mathrm{ign})}_s$ for the current season $s$.
- Forecast base model ($M_1$) produces $\hat\pi^{h}_{s,w}$ 
- Offline calibration maps it to $\pi^{h,\mathrm{cal}}_{s,w}$
- Online updater optionally adds a week-specific intercept shift $c_{s,w}$ to correct a current-season level offset.



## Pre-trained model needed

To simplify prospective deployment, all trained components and fixed settings are bundled into a single object `kit` via `make_prospective_kit()`. The table below describes each variable passed into that function.

| Variable | Source | Description |
|---|---|---|
| `template_df` | `ref$pred_df` | Smoothed reference positivity curve over aligned week index `newWeek`. Used to construct the template-shift covariate $x^{\delta}_{s,w}$ and ramp weight in Stage-2. |
| `ign_fit` | `fitIgnition()` | Full Stage-1 model object. Contains the fitted GAM classifier (`ign_fit$fits$base$gam`) and the scored training data. `make_prospective_kit` extracts the classifier automatically. |
| `tuned$best_params` | `loso_M0v2()` | LOSO-tuned threshold parameters for the Stage-1 ignition detector — positivity thresholds (`p_thr`, `prev_thr`, `p_sum_thr`), trend requirement (`n_consec`), minimum vote count (`N_req`), and allowable ignition window (`w_min`, `w_max`). |
| `joint_out` | `train_stage2_joint()` | Joint Stage-2 training output. Contains the tuned hyperparameter spec (`joint_out$spec`), the final fitted model (`joint_out$fit`), and the training design table (`joint_out$train_data`) used for safe prospective prediction. |
| `cal_off` | `train_calib_platt()` | Offline Platt calibration map trained on LOSO out-of-sample predictions. For each horizon $h$ it applies $\mathrm{logit}(\pi^{h,\mathrm{cal}}) = a_h + b_h\,\mathrm{logit}(\hat\pi^h)$ to correct bias and rescale confidence. Held fixed during deployment. |
| `res_lambda$best$lambda_c` | `tune_lambda_c()` | Ridge penalty $\lambda_c \ge 0$ for online intercept updating. `res_lambda` is the output of `tune_lambda_c()`; `$best$lambda_c` extracts the cross-validated optimal scalar. Larger values yield more conservative within-season updates. |
| `spec_stage2_best$pre_buffer` | `stage2_spec_from_tuning()` | Number of pre-ignition weeks $k_{\mathrm{pre}} \ge 0$ included in the online update window. Extracted from the best Stage-2 specification object. |

In deployment, Stage-2 produces $\hat\pi^{h}_{s,w}$, offline calibration maps it to $\pi^{h,\mathrm{cal}}_{s,w}$, and the online updater optionally applies a season-specific intercept shift so that
$$
\mathrm{logit}\!\left(\tilde\pi^{h}_{s,w}\right)
=
\mathrm{logit}\!\left(\pi^{h,\mathrm{cal}}_{s,w}\right)
+
c_{s,w}.
$$
```{r}

kit <- make_prospective_kit(
  template_df   = template_df,
  ign_fit       = ign_fit,               # extracts gam_cls automatically
  params_stage1 = tuned$best_params,     # tuned Stage-1 threshold parameters
  joint_out     = joint_out,             # extracts spec, fit, train_data automatically
  cal_off       = cal_off,               # can be NULL if not trained yet
  lambda_c      = res_lambda$best$lambda_c,
  k_pre         = spec_stage2_best$pre_buffer
)


saveRDS(kit, file = "prospective_kit.rds")
```

## Current season snapshot 

`currentSeason` should include only weeks observed so far and contain at least:
`season, weekF, y, N, neg, p` (if `p` is missing, your Stage-1 code can compute `p=y/N`).

```{r}
#| label: prospective-current-season
#| eval: true
#| echo: true

currentSeason <- getCurrentD() |>dplyr::select(-newWeek, -season)
```

::: panel-tabset

### Plot

```{r}
#| echo: false
ggplotly(
  ggplot(currentSeason, aes(x = weekF, y = p)) +
    geom_point() +
    geom_line() +
    labs(x = "weekF", y = "Positivity (p)")
)
```

### Data

```{r}
#| echo: false
DT::datatable(currentSeason, options = list(pageLength = 25, scrollX = TRUE))
```

:::

## $M_0$: Ignition week estimation 

This function takes the current season, starting at `start_week` (weekF),  it returns a detection result for each week after.


```{r}
#| label: prospective-stage1
#| eval: true
#| echo: true

ign_out <- run_ignition_weekly(
  currentSeason  = currentSeason,
  ign_fit_or_gam = kit$stage1$gam_cls,
  params         = kit$stage1$params,
  start_week     = 5L
)
plots <- plot_ignition_weekly_snapshots(
  ign_out, currentSeason,
  facet = FALSE, start_week = 5L, maxWeek = 20
)
iWeek_hat_locked <- ign_out$iWeek_hat_locked
```

::: panel-tabset

```{r}
#| results: asis
#| echo: false

for (nm in names(plots)) {
  cat(knitr::knit_child(
    text = c(
      paste0("### ", nm),
      "",
      "```{r echo=FALSE}",
      sprintf("plotly::ggplotly(plots[[%s]])", deparse(nm)),
      "```",
      ""
    ),
    envir = environment(),
    quiet = TRUE
  ), sep = "\n")
}
```

### Detection table

```{r}
#| echo: false
DT::datatable(ign_out$df, options = list(pageLength = 25, scrollX = TRUE))
```

:::

```{r}
#| echo: false
cat("Locked ignition week:", iWeek_hat_locked)
```



<!-- ## Ignition week estimation -->

<!-- Use the locked week by default, or optionally override manually if disagree with the detection results, or for sensitivity analysis -->

<!-- ```{r} -->
<!-- #| label: prospective-iweek-used -->
<!-- #| eval: true -->
<!-- #| echo: true -->

<!-- override_iWeek <- 17  # set to e.g. 18L to force, otherwise keep NULL -->
<!-- iWeek_used <- if (!is.null(override_iWeek)) as.integer(override_iWeek) else iWeek_hat_locked -->
<!-- ``` -->

<!-- If ignition has not triggered yet (`iWeek_used` is `NA`), **stop here**. -->

## $M_1$ real time forecast




### Forcasting model data prep

<!-- Treating each week since ignition as "current" week, this block prepare the data for forecasting by building the design matrix with all features needed used in stage 2 model, including the template covariate with the tuned shift and ramp-in, and the prospective derivatives computed in a real-time safe way.  -->


<!-- ```{r} -->
<!-- #| label: prospective-stage2 -->
<!-- #| eval: true -->
<!-- #| echo: true -->

<!-- if (is.na(iWeek_used)) { -->
<!--   message("Ignition not detected yet (iWeek_used is NA). Stop after Stage 1.") -->
<!-- } else { -->
<!--   pp <- build_stage2_pseudo_prospective_list( -->
<!--     currentSeason = currentSeason, -->
<!--     template_df   = template_df, -->
<!--     best_mean_nll = best_mean_nll, -->
<!--     iWeek_hat     = iWeek_used , -->
<!--     align         = TRUE, -->
<!--      n_weeks =   get_full_cycle_weeks(2025), -->
<!--     anchorWeek    = anchorWeek, -->
<!--     pre_buffer    = spec_stage2$pre_buffer, -->
<!--     deriv_k       = 5L -->
<!--   ) -->

<!-- } -->


<!-- knitr::kable(unlist(pp$meta)) -->
<!-- ``` -->


### Forecast:

<!-- Note that for a brand-new season we exclude season-level terms (at minimum `s(season)`, and also `s(newWeek,season)` if you fit the optional `bs="fs"` term). In code we use `exclude_stage2 <- spec_stage2$exclude_newseason`. -->

<!-- ```{r, eval=T} -->
<!-- if (!is.na(iWeek_used)) { -->
<!--   ppp <- stage2_predict_series(pp, stage2_fit, which = "all", ref_col = "template_fit_shift", exclude_season_re = TRUE) -->
<!--   allP <- ggplotly(plot_stage2(ppp, ign_week = ign_out$iWeek_hat_locked, interval = "none")) -->
<!--   plots <-plot_stage2(ppp, ign_week = ign_out$iWeek_hat_locked, interval = "pi", facet = FALSE) -->
<!-- } -->

<!-- #allP -->
<!-- ``` -->


<!-- ::: panel-tabset -->


<!-- ```{r} -->
<!-- #| results: asis -->
<!-- #| echo: false -->
<!-- library(knitr) -->
<!-- library(plotly) -->

<!-- for (nm in names(plots)) { -->
<!--   cat(knitr::knit_child( -->
<!--     text = c( -->
<!--       paste0("### ", nm), -->
<!--       "", -->
<!--       "```{r echo=FALSE}", -->
<!--       sprintf("plotly::ggplotly(plots[[%s]])", deparse(nm)), -->
<!--       "```", -->
<!--       "" -->
<!--     ), -->
<!--     envir = environment(), -->
<!--     quiet = TRUE -->
<!--   ), sep = "\n") -->
<!-- } -->
<!-- ``` -->

<!-- ::: -->


# Summary
**Implementation map (functions / key objects).**

- Retrospective smoothing + derivatives: `estimateDerivs()` produces `fit`, `d1`, `d2` on the `weekF` scale.
- Retrospective ignition labeling + alignment (for template building): `flagIgnition()` (threshold rule) → `alignIgnition()` adds `iWeek`, `phase`, and aligned `newWeek` (attribute `anchorWeek`).
- Common template curve: `estimateRef()` fits the pooled cyclic GAMM and returns the template `pred_df` (columns `newWeek`, `fit`, `low`, `high`).
- Prospective derivatives (no look-ahead): `add_prospective_derivs_link()` adds `d1_link`, `d2_link` week-by-week.
- Stage 1 ignition detector (prospective gate): `fitIgnition()` → `tuneIgnitionGrid_M0v2()` / `loso_M0v2()` → `detectIgnitionBySeason_M0v2()`.
- Stage 2 forecast model: `tune_stage2_loso_shift_template()` → `prep_stage2_joint()` → `train_stage2_joint()` (Stage-2 design columns include `logit_f_eff`, `z_ema`, `logN_now`, `d1_now`, `d2_now`).




## Retro-estimation
- Identify training seasons
- Ignition specification for each season
  - Manual specification or estimation 
- Align Seasons curves
- Reference curve estimation

## Prospective training

- Ignition model specification and training
  - hyperparameter grid tuning using **true ignition week** for alignment and labeling
  - Apply ignition model on historical training data
- Forecast model specification and training
  - prospective derivative estimation
  - hyperparameter LOSO tuning 
  - Train final model on all data using **true ignition week** for alignment and labeling

## Prospective forecasting for new season

- Pull current season data (up to current week)
- Apply ignition model week by week until detection
- Once ignition is detected,prepare data for model forecast
  - Align with reference curve and
  - Prepare design matrix using tunned hyperparameters
- Make forecasts using the pre-trained forecast model, excluding season random effect for the new season
