---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/m0_retro.R")
source("../R/m0_training.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m2_spec_grid.R")
source("../R/m2_nested_loso.R")
# Pull internal helpers missing from R/ mirror
.plot_stage2_joint_fit_prosp <- flualign:::.plot_stage2_joint_fit_prosp

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
hyper       <- ref_cache$hyper
if (is.null(hyper)) stop("hyper not found in ref_production.rds — re-run docs/run.qmd first")
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned   <- readRDS("../data/stage1_tuning.rds")
params  <- tuned$best_params  # M0 detection params; used by m1_walkforward_multi
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Define the `$M_2$` model and its feature set.
- Explain the weekly refit logic used after ignition.
- Tune the `$M_2$` hyperparameters with nested LOSO.
- Inspect the fit-back of the production model on historical seasons.

Inputs assumed here:

- `alignedD` — historical data aligned using manual ignition labels
- `template_df` — reference curve from `estimateRef()`
- `det_all` — per-season ignition detections from tuned `$M_0$`
:::

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_w>0\}\;
  S_{2}^{h}\!\left(w^a_{s,w};\; k_w\right) \\[2pt]
&+ \mathbb{I}\{k_s>0\}\;
  S_{3,s}^{h}\!\left(w^a_{s,w};\;S, k_s\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_1>0\}\;
  S_{6}^{h}\!\left(d1_{s,w};\; k_1\right) \\[2pt]
&+ \mathbb{I}\{k_2>0\}\;
  S_{7}^{h}\!\left(d2_{s,w};\; k_2\right).
\end{aligned}
$$ {#eq-mod1}

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,Kr)$

- Template backbone covariate: injects the reference curve shape into $M_2$'s linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §$M_1$ training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,Kr)=\omega(t_{s,w};Kr)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;Kr)=\min\!\left(1,\max\!\left(0,\dfrac{t}{Kr}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $Kr$ weeks (with $Kr=1$ meaning no ramp).

- **Notation warning — $\delta$ here is not $M_1$'s dilation $\delta_w$.** In $M_1$, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time $M_2$ sees it. The $\delta$ in $M_2$ is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if $M_1$'s alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that $M_1$'s alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{2}^{h}(w^a_{s,w};k_w)$
- Global aligned-time correction, allows the model to "self-align" in real time with observed data
- **Hyperparams:** $k_w$ (basis dimension).
- **Toggle:** on if $k_w>0$; off if $k_w=0$.
- **R:** `s(newWeek, by=lead, bs="ts", k=k_w)`.

### $S^{h}_{3,s}(w^a_{s,w};S,k_s)$
- Season-specific deviation over aligned time, allows each season to deviate smoothly from the global aligned-time pattern.
- **Hyperparams:** $k_s$ (basis dimension).
- **Toggle:** on if $k_s>0$; off if $k_s=0$.
- **R:** `s(newWeek, by=interaction(season,lead), bs="fs", k=k_s, xt=...)`.
- **New-season prediction:** typically excluded because a brand-new season has an unseen factor level.

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay, tuned jointly with $\lambda_w$) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{6}^{h}(d1_{s,w};k_1)$
- Prospective derivative (slope), captures local growth/decline momentum on the logit scale, estimated in real time
- **Hyperparams:** $k_1$ (basis dimension).
- **Toggle:** on if $k_1>0$; off if $k_1=0$.
- **R (feature):** `d1_now`.
- **R (model):** `s(d1_now, by=lead, bs="ts", k=k_1)`.

### $S_{7}^{h}(d2_{s,w};k_2)$
- Prospective derivative (curvature), captures acceleration/curvature on the logit scale
- **Hyperparams:** $k_2$ (basis dimension).
- **Toggle:** on if $k_2>0$; off if $k_2=0$ (recommended if noisy).
- **R (feature):** `d2_now`.
- **R (model):** `s(d2_now, by=lead, bs="ts", k=k_2)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Tuning.** $\lambda_w$ is treated as a hyperparameter and selected jointly with $\alpha_{\text{state}}$ by LOSO. Evaluation during LOSO always uses **uniform weights** over the first `eval_window` post-ignition weeks — $\lambda_w$ governs what the model learns, not what it is judged on.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Online weekly refit

Let $s^*$ denote the current live season and $\mathcal{D}_{\text{hist}}$ the full historical dataset. At each week $w \geq w^{(\mathrm{ign})}_{s^*}$ after ignition detection, $M_2$ is refit on the augmented dataset

$$
\mathcal{D}_w = \mathcal{D}_{\text{hist}} \;\cup\; \bigl\{\,(s^*,\,w',\,x_{s^*,w'},\,n_{s^*,w'}) : w' \leq w\,\bigr\}.
$$

This allows the season random effect $b_{s^*}$ and the factor-smooth $S^h_{3,s^*}$ to be estimated from current-season observations, so both the level and shape of the forecast are corrected as data accumulate. All hyperparameters are fixed at their LOSO-tuned values; only the GAM coefficients are updated each week.

### Prospective derivatives

For each season $s$ and week $w$, derivatives are estimated on the **link scale** using only the observed history up to week $w$. We fit a **local quadratic** using only the most recent $k$ points up to week $w$, centred at $w$:

$$
z_{s,w} = \mathrm{logit}(\mathrm{clamp}(p_{s,w})) \approx a + b\,u + c\,u^2,
\qquad u = w' - w
$$

The prospective derivatives at week $w$ are:

$$
d1_{s,w} = b, \qquad d2_{s,w} = 2c.
$$

This estimator is **prospective-safe** — it only uses weeks $\le w$.

```{r compute-prospective-derivs}
#| fig-height: 10
#| fig-width: 12
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)

alignedD_long <- alignedD_prosp %>%
  select(season, weekF, d1_link, d2_link) %>%
  pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_line() +
    facet_wrap(~ season, scales = "free_y") +
    labs(x = "weekF", y = "Derivative on logit scale", color = "")
)
```

## Hyperparameter tuning

$M_2$ hyperparameters are selected by **nested leave-one-season-out (LOSO) cross-validation**. Standard LOSO leaks alignment information: it uses a single reference curve fitted on all seasons, including the held-out one. Nested LOSO fixes this by re-estimating the reference curve and all $M_1$ predictions per fold.

**Protocol per fold:**

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold template.
3. Run $M_1$ walk-forward on training seasons → $M_1$ stacking features.
4. Train $M_2$ with the per-fold template and $M_1$ features.
5. Run $M_1$ walk-forward on $s^*$ → $M_1$ test features.
6. Evaluate $M_2$ on $s^*$ using uniform-weight NLL over the first 12 post-ignition weeks.

Steps 1–3 (Phase 1) are pre-computed once per fold and cached; Phase 2 (steps 4–6) runs in minutes regardless of grid size.

### Tuning parameters

The table below lists every hyperparameter, its role in the model, and the values searched. Parameters marked *fixed* were held constant throughout based on prior analysis or implementation constraints.

```{r tuning-param-table}
#| echo: false
tibble::tribble(
  ~Parameter,  ~Description,                                                                                                                                                    ~Type,          ~`Grid / value`,
  "δ (delta)", "Integer-week nudge to the template covariate on the aligned axis. Compensates for residual timing offset after $M_1$ alignment. δ=0 means no nudge.",             "Integer",      "Fixed: 0",
  "Kr",        "Template ramp length in weeks. Kr=1 means full template weight from week 1 of ignition; Kr>1 blends in gradually over Kr weeks.",                              "Integer ≥ 1",  "{1, 2, 3}",
  "k_f",       "Basis dimension for the template smooth $S_1^h$. Controls how nonlinearly the model maps the template covariate to the response. k_f=2 is near-linear.",       "Integer ≥ 2",  "{2, 3, 4}",
  "k_e",       "Basis dimension for the EMA-state smooth $S_4^h$. Controls nonlinearity of the current-level correction.",                                                     "Integer ≥ 0",  "{3, 4, 6}",
  "α_state",   "EWMA smoothing decay for the current-level covariate z_ema. Lower α = slower adaptation, longer effective memory (≈ 1/α weeks).",                             "Real ∈ (0,1)", "{0.03, 0.05, 0.07, 0.10, 0.15, 0.20, 0.25}",
  "k_s",       "Basis dimension for per-season deviation smooth $S_3^h$. k_s=0 disables season-specific shape correction.",                                                    "Integer ≥ 0",  "Fixed: 0",
  "λ_w",       "Time-decay weight exponent. Down-weights late-season training rows. λ_w=0 is standard uniform MLE.",                                                           "Real ≥ 0",     "Fixed: 0",
  "k_1",       "Basis dimension for slope smooth $S_6^h$ (prospective first derivative d1).",                                                                                  "Integer ≥ 0",  "Fixed: 4",
  "k_2",       "Basis dimension for curvature smooth $S_7^h$ (prospective second derivative d2). k_2=0 disables.",                                                             "Integer ≥ 0",  "Fixed: 0",
  "k_w",       "Basis dimension for global aligned-time smooth $S_2^h$. k_w=0 disables.",                                                                                      "Integer ≥ 0",  "Fixed: 0"
) |>
  gt() |>
  tab_header(title = "M_2 tuning parameters") |>
  cols_width(Parameter ~ px(90), Description ~ px(460),
             Type ~ px(110), `Grid / value` ~ px(240)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#f0f4f8"),
            locations = cells_body(rows = c(2, 3, 4, 5)))
```

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the full approach to peak (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean binomial NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** regardless of $\lambda_w$.

Predicted probabilities $\hat{p}_i$ pass through the **soft positivity ceiling** before NLL is computed, matching the transformation applied at deployment time. The ceiling is a tanh squash above the 95th-percentile of training positivity ($p_{\text{knee}}$), asymptoting toward $p_{\text{ceil}} = p_{\max}^{(\text{train})} + 0.5(p_{\max}^{(\text{train})} - p_{\text{knee}})$. Because historical seasons never exceed $\sim$34% positivity the cap is near-identity within the training distribution and has negligible effect on NLL or spec ranking. Its purpose is to ensure that the model evaluated during training is identical to the model deployed prospectively.

### Tuning call

The nested LOSO is implemented as a two-phase script (`scripts/run_nested_loso_v4.R`). Phase 1 pre-computes $M_1$ predictions once per fold; Phase 2 loops over all specs.

```{r tuning-call}
#| eval: false

# --- Phase 1: build fold + pre-compute $M_1$ (one-time per held-out season) ---
fold <- nested_loso_build_fold(
  allD = allD, test_season = test_s, exclude_seasons = "2015-16",
  k_ref = 25L, ref_method = "fs", manual_labels = manual_labels
)
m1_train <- m1_walkforward_multi(
  allD = allD, ref = fold$ref, hyper = fold$hyper, params = params,
  seasons = fold$train_seasons,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L, parallel = TRUE
)
m1_test <- m1_walkforward_predictions(
  seasonD = allD[allD$season == test_s, ],
  ref = fold$ref, hyper = fold$hyper, params = params,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L
)

# --- Phase 2: for each candidate spec, train M2 and score held-out season ---
# v4 grid: Kr x k_f x k_e x alpha_state (189 specs, fixed delta=0, k_s=0, lambda_w=0)
grid <- tidyr::crossing(
  Kr          = c(1L, 2L, 3L),
  k_f         = c(2L, 3L, 4L),
  k_e         = c(3L, 4L, 6L),
  alpha_state = c(0.03, 0.05, 0.07, 0.10, 0.15, 0.20, 0.25)
)
spec <- stage2_make_spec(
  delta = 0L, Kr = Kr, T = "S", k_f = k_f, k_e = k_e,
  k_1 = 4L, k_2 = 0L, k_w = 0L, k_s = 0L,
  alpha_state = alpha_state, lambda_w = 0, w_floor = 0.05
)

m2      <- nested_loso_m2_train(fold = fold, m1_train_preds = m1_train, spec = spec)
eval_out <- nested_loso_m2_eval(
  allD = allD, fold = fold, m2_fit = m2, m1_test_preds = m1_test,
  spec = spec, eval_window = 12L, k_deriv = 10L,
  manual_labels = manual_labels, flag_args = flag_args
)
```

### Full grid results

```{r load-production-results}
#| eval: true

# Prefer v8 (weekly refit + soft cap + expanded k_f grid + no clamping in eval).
# Fall back through v7 → v5 → v4 for older environments.
if (file.exists("../data/nested_loso_v8_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v8_production.rds")
  cat("v8 (weekly refit, k_f expanded) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v7_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v7_production.rds")
  cat("v7 (weekly refit, M1-based logit_f_eff) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v5_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v5_production.rds")
  cat("v5 (weekly refit) grid:", nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
} else {
  prod_results_v4 <- readRDS("../data/nested_loso_v4_production.rds")
  cat("v4 grid (frozen-fit eval):", nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
}
cat("Best:", prod_results_v4$best_spec_id,
    "| mean_nll =", round(prod_results_v4$summary$mean_nll[1], 4), "\n")
```

### Best spec {#sec-best-spec-interp}

```{r load-best-spec}
#| include: false
best_spec_nested <- prod_results_v4$best_spec
```

The tuned spec is `Kr=1, k_f=7, k_e=2, alpha_state=0.02` with `delta=0, k_s=0, lambda_w=0`.
Selected by v8 nested LOSO (756 specs; weekly refit + soft cap; M1-based `logit_f_eff`; no clamping in eval).

**`delta = 0` — no template shift.**
$M_1$'s alignment already places the current season correctly on the reference curve. No residual systematic timing offset remains, validating multi-template alignment with `k_ref=25`.

**`Kr = 1` — no ramp.**
The template covariate is applied at full weight from the first week after ignition. Once ignition fires the reference curve shape is immediately a reliable anchor — a gradual blend adds no benefit.

**`k_f = 7` — flexible template smooth.**
The logit-scale mapping from the template covariate $z_{s,w}$ uses 7 knots, capturing nonlinear relationships between the reference curve position and realised positivity. This is notably more flexible than the v4 spec (`k_f=2`) and reflects the fact that weekly refit (which incorporates current-season data at each step) can make effective use of a higher-capacity template basis without overfitting. The boundary check confirms `k_f=7` is interior: `k_f=8` is consistently worse.

**`k_e = 2` — minimal EWMA smooth.**
The EMA state `z_ema` enters with a 2-knot (near-linear) smooth. Under weekly refit, the current season's own trajectory is directly observed, so the model needs less nonlinear correction from the EMA. A simpler correction prevents the EMA term from conflicting with the rich template smooth.

**`alpha_state = 0.02` — very slow EMA, ≈50-week memory.**
The effective lookback is $1/\alpha = 50$ weeks — essentially a seasonal-mean tracker rather than a reactive signal. Combined with `k_e=2`, the EMA provides a stable baseline offset, not a dynamic correction. The boundary check confirms `alpha=0.02` is interior: `alpha=0.01` appears in the top 30 but with worse NLL.

**`k_s = 0` — no per-season deviation.**
Allowing each season its own weekly shape hurts out-of-sample. The shared reference curve captures enough cross-season variation.

**`lambda_w = 0` — uniform training weights.**
Down-weighting late-season observations provides no benefit. The model learns equally from all post-ignition weeks.

**Summary.** The v8 tuned $M_2$ uses a **flexible template smooth with a near-linear EMA correction** under weekly refit semantics. The shift toward higher `k_f` relative to v4 is a direct consequence of evaluating specs under the weekly refit path (which makes the template the primary signal) and including the soft cap in the LOSO objective (which prevents the stiff-template shortcut that v4 exploited).


## Fit-back on historical data

Refit $M_2$ on all historical seasons using the best nested-LOSO spec and the production reference curve.

**M1 walk-forward predictions for historical seasons** are computed here and saved with the production kit. They are passed to both `nested_loso_refit_best()` (frozen GAM) and `refit_stage2_weekly()` at deployment so that `logit_f_eff` is M1-based in all three contexts — frozen training, weekly refit, and LOSO evaluation — ensuring no inconsistency. There is no data leakage: the predictions are walk-forward (each season uses only past data) and the current season is never in `m1_train_preds`.

```{r compute-m1-production-preds}
#| eval: true
#| message: false
# Production M1 predictions for ALL training seasons (no leakage — walk-forward).
# These replace template-based logit_f_eff in the frozen GAM and in every
# weekly refit at deployment.  Save them in m2_production.rds so
# load_prospective_kit() can thread them through refit_stage2_weekly().
EXCLUDE_SEAS <- "2015-16"  # Same exclusion used throughout LOSO evaluation.
m1_preds_production <- m1_walkforward_multi(
  allD               = allD,
  ref                = ref,
  hyper              = hyper,
  params             = params,
  seasons            = sort(setdiff(unique(allD$season), EXCLUDE_SEAS)),
  temperature        = 0.25,
  rise_weight        = 1.0,
  trough_weight      = 0.1,
  peak_decay         = 0.3,
  slope_weight       = 0.5,
  slope_window       = 4L,
  dynamic_temp       = TRUE,
  dynamic_temp_pivot = 10L,
  parallel           = TRUE,
  verbose            = FALSE
)
message("m1_preds_production: ", nrow(m1_preds_production), " rows across ",
        length(unique(m1_preds_production$season)), " seasons")
```

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
joint_out_nested <- nested_loso_refit_best(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec           = best_spec_nested,
  m1_preds       = m1_preds_production
)

p_fitback <- plot_stage2_joint_fit_by_season(
  out_m1     = joint_out_nested,
  feat_full  = joint_out_nested$train_data,
  dat_raw    = alignedD_prosp,
  ign_hat_df = det_all$compare
)

ggsave("figures/fit_back_by_season.png", p_fitback,
       width = 14, height = 14, dpi = 150)

ggplotly(p_fitback)
```

```{r save-m2-production}
#| eval: true
# Save the production M2 object for use by load_prospective_kit().
#
# Contents:
#   fit            — frozen production GAM (trained with M1-based logit_f_eff)
#   spec           — best nested-LOSO spec
#   template_df    — reference curve (newWeek, fit) for refit_stage2_weekly()
#   m1_train_preds — M1 walk-forward predictions for all historical training
#                    seasons.  load_prospective_kit() reads this and passes it
#                    to refit_stage2_weekly() so that logit_f_eff at deployment
#                    matches the feature space used to train the frozen GAM.
#                    No data leakage: predictions are walk-forward and the
#                    current season is never included.
m2_production <- list(
  fit            = joint_out_nested$fit,
  spec           = best_spec_nested,
  template_df    = template_df[, c("newWeek", "fit")],
  m1_train_preds = m1_preds_production
)
saveRDS(m2_production, "../data/m2_production.rds")
message("Saved m2_production.rds")
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```
