---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/retro_estimation.R")
source("../R/ignitionTraining.R")
source("../R/prospective_training.R")
source("../R/prospective_running.R")
source("../R/module_training.R")
source("../R/nested_loso.R")

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned <- readRDS("../data/stage1_tuning.rds")
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

This document covers the $M_2$ positivity forecast model of the PAGe pipeline. $M_2$ is trained offline using all historical seasons and refitted online each week after ignition is detected. It takes alignment outputs from §1–6 of the [pipeline overview](pipeline_overview.html) as inputs.

**Prerequisites:**

- `alignedD` — historical data aligned using manual ignition labels (§1.2, reference curve estimation)
- `template_df` — reference curve from `estimateRef()` (§1.2)
- `det_all` — per-season ignition detections from tuned $M_0$ (§1.1)

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_w>0\}\;
  S_{2}^{h}\!\left(w^a_{s,w};\; k_w\right) \\[2pt]
&+ \mathbb{I}\{k_s>0\}\;
  S_{3,s}^{h}\!\left(w^a_{s,w};\;S, k_s\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_1>0\}\;
  S_{6}^{h}\!\left(d1_{s,w};\; k_1\right) \\[2pt]
&+ \mathbb{I}\{k_2>0\}\;
  S_{7}^{h}\!\left(d2_{s,w};\; k_2\right).
\end{aligned}
$$ {#eq-mod1}

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,K)$

- Template backbone covariate: injects the reference curve shape into M2's linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §M1 training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,K)=\omega(t_{s,w};K)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;K)=\min\!\left(1,\max\!\left(0,\dfrac{t}{K}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $K$ weeks (with $K \le 1$ meaning no ramp).

- **Notation warning — $\delta$ here is not M1's dilation $\delta_w$.** In M1, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time M2 sees it. The $\delta$ in M2 is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if M1's alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that M1's alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{2}^{h}(w^a_{s,w};k_w)$
- Global aligned-time correction, allows the model to "self-align" in real time with observed data
- **Hyperparams:** $k_w$ (basis dimension).
- **Toggle:** on if $k_w>0$; off if $k_w=0$.
- **R:** `s(newWeek, by=lead, bs="ts", k=k_w)`.

### $S^{h}_{3,s}(w^a_{s,w};S,k_s)$
- Season-specific deviation over aligned time, allows each season to deviate smoothly from the global aligned-time pattern.
- **Hyperparams:** $k_s$ (basis dimension).
- **Toggle:** on if $k_s>0$; off if $k_s=0$.
- **R:** `s(newWeek, by=interaction(season,lead), bs="fs", k=k_s, xt=...)`.
- **New-season prediction:** typically excluded because a brand-new season has an unseen factor level.

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay, tuned jointly with $\lambda_w$) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{6}^{h}(d1_{s,w};k_1)$
- Prospective derivative (slope), captures local growth/decline momentum on the logit scale, estimated in real time
- **Hyperparams:** $k_1$ (basis dimension).
- **Toggle:** on if $k_1>0$; off if $k_1=0$.
- **R (feature):** `d1_now`.
- **R (model):** `s(d1_now, by=lead, bs="ts", k=k_1)`.

### $S_{7}^{h}(d2_{s,w};k_2)$
- Prospective derivative (curvature), captures acceleration/curvature on the logit scale
- **Hyperparams:** $k_2$ (basis dimension).
- **Toggle:** on if $k_2>0$; off if $k_2=0$ (recommended if noisy).
- **R (feature):** `d2_now`.
- **R (model):** `s(d2_now, by=lead, bs="ts", k=k_2)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Tuning.** $\lambda_w$ is treated as a hyperparameter and selected jointly with $\alpha_{\text{state}}$ by LOSO. Evaluation during LOSO always uses **uniform weights** over the first `eval_window` post-ignition weeks — $\lambda_w$ governs what the model learns, not what it is judged on.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Online weekly refit

Let $s^*$ denote the current live season and $\mathcal{D}_{\text{hist}}$ the full historical dataset. At each week $w \geq w^{(\mathrm{ign})}_{s^*}$ after ignition detection, $M_2$ is refit on the augmented dataset

$$
\mathcal{D}_w = \mathcal{D}_{\text{hist}} \;\cup\; \bigl\{\,(s^*,\,w',\,x_{s^*,w'},\,n_{s^*,w'}) : w' \leq w\,\bigr\}.
$$

This allows the season random effect $b_{s^*}$ and the factor-smooth $S^h_{3,s^*}$ to be estimated from current-season observations, so both the level and shape of the forecast are corrected as data accumulate. All hyperparameters are fixed at their LOSO-tuned values; only the GAM coefficients are updated each week.

### Prospective derivatives

For each season $s$ and week $w$, derivatives are estimated on the **link scale** using only the observed history up to week $w$. We fit a **local quadratic** using only the most recent $k$ points up to week $w$, centred at $w$:

$$
z_{s,w} = \mathrm{logit}(\mathrm{clamp}(p_{s,w})) \approx a + b\,u + c\,u^2,
\qquad u = w' - w
$$

The prospective derivatives at week $w$ are:

$$
d1_{s,w} = b, \qquad d2_{s,w} = 2c.
$$

This estimator is **prospective-safe** — it only uses weeks $\le w$.

```{r compute-prospective-derivs}
#| fig-height: 10
#| fig-width: 12
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)

alignedD_long <- alignedD_prosp %>%
  select(season, weekF, d1_link, d2_link) %>%
  pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_line() +
    facet_wrap(~ season, scales = "free_y") +
    labs(x = "weekF", y = "Derivative on logit scale", color = "")
)
```

## Hyperparameter tuning

We tune hyperparameters of $M_2$ using **leave-one-season-out (LOSO)** cross-validation. For each candidate spec (from `expand_grid_specs()`), we repeatedly:

1. Hold out one season $s$.
2. Fit the joint model on all remaining seasons.
3. Predict the held-out season **as a new season** (excluding season-dependent terms).
4. Score predictions using binomial NLL.

### Tuning grid

- **$\delta$ (template shift):** integer weeks (typically $\{-3,\ldots,3\}$). Set `delta=NA` to disable the template covariate.
- **`T` (template entry mode):** one of `{"S","O","N"}` (smooth, offset, none).
- **$Kr$ (ramp length):** integer with $Kr \ge 1$. Use $Kr=1$ for "no ramp".
- **$k_f, k_w, k_s, k_e, k_1, k_2$ (basis dimensions):** integers $\ge 0$; 0 disables the term.
- **$\alpha_{\text{state}}$ (EWMA decay):** real in $(0,1)$, tuned jointly with $\lambda_w$.
- **$\lambda_w$ (time-decay weight):** non-negative real; $\lambda_w=0$ is standard MLE.
- **$Kb$:** integer $\ge 0$, defines training start as `weekF >= iWeek_used - Kb`.

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the approach to peak in all historical seasons (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** even when $\lambda_w > 0$ for training.

### Tuning code

```{r spec-grid}
#| eval: false
nc <- min(parallel::detectCores() - 1L, 32L)

spec_grid <- expand_grid_specs(
  delta_grid   = -2:2,
  Kr_grid      = 2:6,
  Kb_grid      = c(1L, 2L),
  T_grid       = "S",
  k_f_grid     = c(6L, 8L),
  alpha_state  = c(0.15, 0.25, 0.35),
  k_w_grid     = c(0L, 4L),
  k_e_grid     = 6L,
  k_n_grid     = 0L,
  k_1_grid     = 4L,
  k_2_grid     = 0L,
  k_s_grid     = 4L,
  w_floor_grid = c(0, 0.03, 0.05, 0.08),
  verbose      = TRUE
)
```

```{r refit-loso}
#| eval: true
if (file.exists("../data/stage2_tuning_refit.rds")) {
  tuned2_refit <- readRDS("../data/stage2_tuning_refit.rds")
} else {
  tuned2_refit <- tune_stage2_loso_spec_grid_parallel(
    alignedD_prosp = alignedD_prosp,
    template_df    = template_df,
    spec_grid      = spec_grid,
    seasons        = NULL,
    eval_window    = 12L,
    refit_loso     = TRUE,
    lambda_w_grid  = c(0, 0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
    workers        = nc,
    chunk_size     = 8L,
    nthreads       = 1L,
    verbose        = TRUE
  )
  saveRDS(tuned2_refit, "../data/stage2_tuning_refit.rds")
}

ggplotly(plot_tune_stage2_heatmap(tuned2_refit, metric = "mean_nll",
                                  normalize01 = TRUE, norm_scope = "facet",
                                  agg = TRUE, base_size = 10))
```

## Training results

::: panel-tabset

### Best spec (refit-LOSO)

```{r refit-best-spec}
#| eval: true
tuned2_refit <- readRDS("../data/stage2_tuning_refit.rds")
spec_refit_best <- stage2_spec_from_tuning(tuned2_refit)
spec_refit_best$w_floor <- 0.05

skip_slots <- c("best_row", "formula", "exclude_newseason")
spec_display_refit <- spec_refit_best[!names(spec_refit_best) %in% skip_slots]
data.frame(
  Specification = names(spec_display_refit),
  Value = sapply(spec_display_refit, function(x) paste(as.character(x), collapse = ", ")),
  row.names = NULL
) |> gt() |> tab_header(title = "Best M2 spec (refit-LOSO)")
```

### Fit by season

```{r refit-fit-plot}
#| eval: true
#| fig-height: 14
#| fig-width: 14
joint_out_refit <- train_stage2_joint(
  dat         = alignedD_prosp,
  template_df = template_df,
  spec        = spec_refit_best
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out_refit,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```

### Fit summary

```{r fit-summary}
#| eval: true
summary(joint_out_refit$fit)
```

:::

## Nested LOSO tuning (leakage-free)

The standard LOSO above uses a **single** reference curve (fitted on all seasons) as `template_df` for every fold. This leaks alignment information from the held-out season into the template covariate. **Nested LOSO** fixes this by re-estimating the reference curve per fold:

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold `template_df`.
3. Run M1 walk-forward on training seasons → M1 stacking features.
4. Train $M_2$ using the per-fold template and M1 features.
5. Run M1 walk-forward on $s^*$ → M1 test features.
6. Evaluate $M_2$ on $s^*$.

This is implemented by `nested_loso_cv()` and `nested_loso_grid_search()` (composable helpers in `nested_loso.R`).

### Smoke-test grid

A small grid for local validation — the full grid should be run on a server (see [Production grid](#production-grid) below).

```{r nested-loso-grid}
#| eval: true

# M0 detection params from stage1 tuning
params <- tuned$best_params

# Small smoke-test grid: 2 specs (local run uses skip_m1=TRUE)
smoke_specs <- list(
  base = stage2_make_spec(
    delta = 0L, K = 3L, T = "S", k_f = 6L,
    k_e = 6L, k_1 = 4L, k_2 = 0L, k_w = 0L, k_s = 4L,
    alpha_state = 0.25, lambda_w = 0, w_floor = 0.05
  ),
  smaller = stage2_make_spec(
    delta = 0L, K = 3L, T = "S", k_f = 4L,
    k_e = 4L, k_1 = 4L, k_2 = 0L, k_w = 0L, k_s = 4L,
    alpha_state = 0.30, lambda_w = 0, w_floor = 0.05
  )
)
```

```{r nested-loso-run}
#| eval: true

if (file.exists("../data/nested_loso_smoke.rds")) {
  nested_results <- readRDS("../data/nested_loso_smoke.rds")
} else {
  nested_results <- nested_loso_grid_search(
    allD          = allD,
    params        = params,
    specs         = smoke_specs,
    test_seasons  = c("2023-24", "2024-25"),
    skip_m1       = TRUE,
    eval_window   = 12L,
    k_deriv       = 10L,
    k_ref         = 8L,
    n_weeks       = 52L,
    manual_labels = manual_labels,
    n_cores       = 1L,
    verbose       = TRUE
  )
  saveRDS(nested_results, "../data/nested_loso_smoke.rds")
}
```

### Results

::: panel-tabset

#### Score summary

```{r nested-loso-summary}
#| eval: true
nested_results$summary |>
  gt() |>
  tab_header(title = "Nested LOSO: Score summary by spec") |>
  fmt_number(columns = c(mean_nll, brier, rmse_p), decimals = 4)
```

#### Scores by season

```{r nested-loso-scores-plot}
#| eval: true
#| fig-height: 5
#| fig-width: 10
best_cv <- nested_results$cv_results[[nested_results$best_spec_id]]
ggplotly(plot_nested_loso_scores(best_cv$scores, metric = "mean_nll"))
```

#### Per-season scores table

```{r nested-loso-season-table}
#| eval: true
nested_results$scores |>
  gt() |>
  tab_header(title = "Nested LOSO: Per-season scores") |>
  fmt_number(columns = c(mean_nll, brier, rmse_p), decimals = 4)
```

#### Out-of-sample predictions

```{r nested-loso-pred-plot}
#| eval: true
#| fig-height: 12
#| fig-width: 14
ggplotly(plot_nested_loso_predictions(best_cv, dat_raw = alignedD))
```

:::

### Fit-back with best spec

Refit $M_2$ on all historical data using the best nested-LOSO spec and the production reference curve. This mirrors the "Fit by season" plot above but uses the leakage-free tuned hyperparameters.

```{r nested-refit}
#| eval: true
best_spec_nested <- nested_results$best_spec

# Display best spec
skip_slots <- c("best_row", "formula", "exclude_newseason")
spec_display <- best_spec_nested[!names(best_spec_nested) %in% skip_slots]
data.frame(
  Specification = names(spec_display),
  Value = sapply(spec_display, function(x) paste(as.character(x), collapse = ", ")),
  row.names = NULL
) |> gt() |> tab_header(title = "Best nested-LOSO spec")
```

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
joint_out_nested <- nested_loso_refit_best(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec           = best_spec_nested
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out_nested,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```


### Production grid {#production-grid}

The smoke test above validates the pipeline with 2 specs on 2 seasons. For a proper hyperparameter search, the full grid below should be run on a server with `skip_m1 = FALSE` (includes M1 walk-forward stacking features). With 11 seasons and ~40 min per fold (M1 + M2), each spec takes ~7 hours, so the full grid below (~144 specs) will take several days on an 8-core machine.

**Axes being searched:**

| Parameter | Grid values | Rationale |
|-----------|-------------|-----------|
| `delta` | −2, −1, 0, 1, 2 | Template time-shift (weeks). Captures timing misalignment between the reference curve and the optimal forecast anchor. |
| `K` | 1, 3, 5 | Ramp width for phase transition. `K = 1` = abrupt switch; larger K = smoother blend from pre- to post-ignition template. |
| `T` | "S", "O" | Template entry mode. `S` = smooth spline on logit-template; `O` = offset (simpler, less flexible). |
| `k_f` | 4, 6, 8 | Basis dimension for template smooth (only when `T = "S"`). |
| `k_s` | 0, 4 | Season × week factor smooth dimension. `0` = off; `4` = season-specific weekly deviations. |
| `alpha_state` | 0.20, 0.25, 0.30 | EWMA smoothing for `z_ema`. Lower = smoother (more lag); higher = more responsive. |
| `lambda_w` | 0, 0.05 | Time-decay weight for older training seasons. `0` = equal weight; `0.05` = gentle down-weighting. |

```{r production-grid}
#| eval: false

# --- Production grid: ~144 specs ---
# Run on server with: Rscript -e 'source("run_nested_loso_full.R")'
prod_specs <- {
  grid <- tidyr::crossing(
    delta       = c(-2L, -1L, 0L, 1L, 2L),
    K           = c(1L, 3L, 5L),
    T           = c("S", "O"),
    k_s         = c(0L, 4L),
    alpha_state = c(0.20, 0.25, 0.30),
    lambda_w    = c(0, 0.05)
  )

  # k_f only matters for T="S"; for T="O" fix at 6
  grid_S <- grid |> dplyr::filter(T == "S") |>
    tidyr::crossing(k_f = c(4L, 6L, 8L))
  grid_O <- grid |> dplyr::filter(T == "O") |>
    dplyr::mutate(k_f = 6L)
  grid <- dplyr::bind_rows(grid_S, grid_O) |>
    dplyr::mutate(
      spec_id = dplyr::if_else(
        T == "S",
        sprintf("d%+d_K%d_S_kf%d_ks%d_a%.0f_lw%.0f",
                delta, K, k_f, k_s, alpha_state * 100, lambda_w * 100),
        sprintf("d%+d_K%d_O_ks%d_a%.0f_lw%.0f",
                delta, K, k_s, alpha_state * 100, lambda_w * 100)
      )
    )

  specs <- purrr::pmap(grid, function(delta, K, T, k_s, alpha_state,
                                       lambda_w, k_f, spec_id) {
    stage2_make_spec(
      delta = delta, K = K, T = T, k_f = k_f,
      k_e = 6L, k_1 = 4L, k_2 = 0L, k_w = 0L, k_s = k_s,
      alpha_state = alpha_state, lambda_w = lambda_w, w_floor = 0.05
    )
  })
  names(specs) <- grid$spec_id
  specs
}

cat("Production grid:", length(prod_specs), "specs\n")
cat("Estimated time: ~", round(length(prod_specs) * 11 * 40 / 60 / 8, 0),
    "hours on 8 cores\n")
```

To run on the server, save this script as `scripts/run_nested_loso_full.R`:

```{r server-script}
#| eval: false

# --- Server runner: saves results incrementally ---
# Usage: Rscript scripts/run_nested_loso_full.R

prod_results <- nested_loso_grid_search(
  allD          = allD,
  params        = params,
  specs         = prod_specs,
  test_seasons  = NULL,          # all 11 seasons
  skip_m1       = FALSE,         # full M1 walk-forward per fold
  eval_window   = 12L,
  k_deriv       = 10L,
  k_ref         = 8L,
  n_weeks       = 52L,
  manual_labels = manual_labels,
  n_cores       = min(parallel::detectCores() - 1L, 8L),
  verbose       = TRUE
)

saveRDS(prod_results, "data/nested_loso_production.rds")
cat("Done! Best spec:", prod_results$best_spec_id, "\n")
```

Once the server run completes, load results here:

```{r load-production-results}
#| eval: false

if (file.exists("../data/nested_loso_production.rds")) {
  prod_results <- readRDS("../data/nested_loso_production.rds")
  cat("Production grid:", nrow(prod_results$summary), "specs evaluated\n")
  cat("Best:", prod_results$best_spec_id,
      "| mean_nll =", round(prod_results$summary$mean_nll[1], 2), "\n")

  prod_results$summary |>
    head(20) |>
    gt() |>
    tab_header(title = "Top 20 specs (production nested LOSO)") |>
    fmt_number(columns = c(mean_nll, brier, rmse_p), decimals = 4)
}
```
