---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/retro_estimation.R")
source("../R/ignitionTraining.R")
source("../R/prospective_training.R")
source("../R/prospective_running.R")
source("../R/module_training.R")
source("../R/nested_loso.R")
source("../test/R/ignitionTraining.R")
# Pull internal helpers missing from R/ mirror
.plot_stage2_joint_fit_prosp <- flualign:::.plot_stage2_joint_fit_prosp

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned <- readRDS("../data/stage1_tuning.rds")
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

This document covers the $M_2$ positivity forecast model of the PAGe pipeline. $M_2$ is trained offline using all historical seasons and refitted online each week after ignition is detected. It takes alignment outputs from §1–6 of the [pipeline overview](pipeline_overview.html) as inputs.

**Prerequisites:**

- `alignedD` — historical data aligned using manual ignition labels (§1.2, reference curve estimation)
- `template_df` — reference curve from `estimateRef()` (§1.2)
- `det_all` — per-season ignition detections from tuned $M_0$ (§1.1)

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_w>0\}\;
  S_{2}^{h}\!\left(w^a_{s,w};\; k_w\right) \\[2pt]
&+ \mathbb{I}\{k_s>0\}\;
  S_{3,s}^{h}\!\left(w^a_{s,w};\;S, k_s\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_1>0\}\;
  S_{6}^{h}\!\left(d1_{s,w};\; k_1\right) \\[2pt]
&+ \mathbb{I}\{k_2>0\}\;
  S_{7}^{h}\!\left(d2_{s,w};\; k_2\right).
\end{aligned}
$$ {#eq-mod1}

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,Kr)$

- Template backbone covariate: injects the reference curve shape into M2's linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §M1 training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,Kr)=\omega(t_{s,w};Kr)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;Kr)=\min\!\left(1,\max\!\left(0,\dfrac{t}{Kr}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $Kr$ weeks (with $Kr=1$ meaning no ramp).

- **Notation warning — $\delta$ here is not M1's dilation $\delta_w$.** In M1, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time M2 sees it. The $\delta$ in M2 is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if M1's alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that M1's alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{2}^{h}(w^a_{s,w};k_w)$
- Global aligned-time correction, allows the model to "self-align" in real time with observed data
- **Hyperparams:** $k_w$ (basis dimension).
- **Toggle:** on if $k_w>0$; off if $k_w=0$.
- **R:** `s(newWeek, by=lead, bs="ts", k=k_w)`.

### $S^{h}_{3,s}(w^a_{s,w};S,k_s)$
- Season-specific deviation over aligned time, allows each season to deviate smoothly from the global aligned-time pattern.
- **Hyperparams:** $k_s$ (basis dimension).
- **Toggle:** on if $k_s>0$; off if $k_s=0$.
- **R:** `s(newWeek, by=interaction(season,lead), bs="fs", k=k_s, xt=...)`.
- **New-season prediction:** typically excluded because a brand-new season has an unseen factor level.

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay, tuned jointly with $\lambda_w$) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{6}^{h}(d1_{s,w};k_1)$
- Prospective derivative (slope), captures local growth/decline momentum on the logit scale, estimated in real time
- **Hyperparams:** $k_1$ (basis dimension).
- **Toggle:** on if $k_1>0$; off if $k_1=0$.
- **R (feature):** `d1_now`.
- **R (model):** `s(d1_now, by=lead, bs="ts", k=k_1)`.

### $S_{7}^{h}(d2_{s,w};k_2)$
- Prospective derivative (curvature), captures acceleration/curvature on the logit scale
- **Hyperparams:** $k_2$ (basis dimension).
- **Toggle:** on if $k_2>0$; off if $k_2=0$ (recommended if noisy).
- **R (feature):** `d2_now`.
- **R (model):** `s(d2_now, by=lead, bs="ts", k=k_2)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Tuning.** $\lambda_w$ is treated as a hyperparameter and selected jointly with $\alpha_{\text{state}}$ by LOSO. Evaluation during LOSO always uses **uniform weights** over the first `eval_window` post-ignition weeks — $\lambda_w$ governs what the model learns, not what it is judged on.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Online weekly refit

Let $s^*$ denote the current live season and $\mathcal{D}_{\text{hist}}$ the full historical dataset. At each week $w \geq w^{(\mathrm{ign})}_{s^*}$ after ignition detection, $M_2$ is refit on the augmented dataset

$$
\mathcal{D}_w = \mathcal{D}_{\text{hist}} \;\cup\; \bigl\{\,(s^*,\,w',\,x_{s^*,w'},\,n_{s^*,w'}) : w' \leq w\,\bigr\}.
$$

This allows the season random effect $b_{s^*}$ and the factor-smooth $S^h_{3,s^*}$ to be estimated from current-season observations, so both the level and shape of the forecast are corrected as data accumulate. All hyperparameters are fixed at their LOSO-tuned values; only the GAM coefficients are updated each week.

### Prospective derivatives

For each season $s$ and week $w$, derivatives are estimated on the **link scale** using only the observed history up to week $w$. We fit a **local quadratic** using only the most recent $k$ points up to week $w$, centred at $w$:

$$
z_{s,w} = \mathrm{logit}(\mathrm{clamp}(p_{s,w})) \approx a + b\,u + c\,u^2,
\qquad u = w' - w
$$

The prospective derivatives at week $w$ are:

$$
d1_{s,w} = b, \qquad d2_{s,w} = 2c.
$$

This estimator is **prospective-safe** — it only uses weeks $\le w$.

```{r compute-prospective-derivs}
#| fig-height: 10
#| fig-width: 12
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)

alignedD_long <- alignedD_prosp %>%
  select(season, weekF, d1_link, d2_link) %>%
  pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_line() +
    facet_wrap(~ season, scales = "free_y") +
    labs(x = "weekF", y = "Derivative on logit scale", color = "")
)
```

## Hyperparameter tuning

$M_2$ hyperparameters are selected by **nested leave-one-season-out (LOSO) cross-validation**. Standard LOSO leaks alignment information: it uses a single reference curve fitted on all seasons, including the held-out one. Nested LOSO fixes this by re-estimating the reference curve and all M1 predictions per fold.

**Protocol per fold:**

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold template.
3. Run M1 walk-forward on training seasons → M1 stacking features.
4. Train $M_2$ with the per-fold template and M1 features.
5. Run M1 walk-forward on $s^*$ → M1 test features.
6. Evaluate $M_2$ on $s^*$ using uniform-weight NLL over the first 12 post-ignition weeks.

Steps 1–3 (Phase 1) are pre-computed once per fold and cached; Phase 2 (steps 4–6) runs in minutes regardless of grid size.

### Tuning parameters

The table below lists every hyperparameter, its role in the model, and the values searched. Parameters marked *fixed* were held constant throughout based on prior analysis or implementation constraints.

```{r tuning-param-table}
#| echo: false
tibble::tribble(
  ~Parameter,  ~Description,                                                                                                                                                    ~Type,          ~`Grid / value`,
  "δ (delta)", "Integer-week nudge to the template covariate on the aligned axis. Compensates for residual timing offset after M1 alignment. δ=0 means no nudge.",             "Integer",      "Fixed: 0",
  "Kr",        "Template ramp length in weeks. Kr=1 means full template weight from week 1 of ignition; Kr>1 blends in gradually over Kr weeks.",                              "Integer ≥ 1",  "{1, 2, 3}",
  "k_f",       "Basis dimension for the template smooth $S_1^h$. Controls how nonlinearly the model maps the template covariate to the response. k_f=2 is near-linear.",       "Integer ≥ 2",  "{2, 3, 4}",
  "k_e",       "Basis dimension for the EMA-state smooth $S_4^h$. Controls nonlinearity of the current-level correction.",                                                     "Integer ≥ 0",  "{3, 4, 6}",
  "α_state",   "EWMA smoothing decay for the current-level covariate z_ema. Lower α = slower adaptation, longer effective memory (≈ 1/α weeks).",                             "Real ∈ (0,1)", "{0.03, 0.05, 0.07, 0.10, 0.15, 0.20, 0.25}",
  "k_s",       "Basis dimension for per-season deviation smooth $S_3^h$. k_s=0 disables season-specific shape correction.",                                                    "Integer ≥ 0",  "Fixed: 0",
  "λ_w",       "Time-decay weight exponent. Down-weights late-season training rows. λ_w=0 is standard uniform MLE.",                                                           "Real ≥ 0",     "Fixed: 0",
  "k_1",       "Basis dimension for slope smooth $S_6^h$ (prospective first derivative d1).",                                                                                  "Integer ≥ 0",  "Fixed: 4",
  "k_2",       "Basis dimension for curvature smooth $S_7^h$ (prospective second derivative d2). k_2=0 disables.",                                                             "Integer ≥ 0",  "Fixed: 0",
  "k_w",       "Basis dimension for global aligned-time smooth $S_2^h$. k_w=0 disables.",                                                                                      "Integer ≥ 0",  "Fixed: 0"
) |>
  gt() |>
  tab_header(title = "M2 tuning parameters") |>
  cols_width(Parameter ~ px(90), Description ~ px(460),
             Type ~ px(110), `Grid / value` ~ px(240)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#f0f4f8"),
            locations = cells_body(rows = c(2, 3, 4, 5)))
```

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the full approach to peak (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean binomial NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** regardless of $\lambda_w$.

### Tuning call

The nested LOSO is implemented as a two-phase script (`scripts/run_nested_loso_v4.R`). Phase 1 pre-computes M1 predictions once per fold; Phase 2 loops over all specs.

```{r tuning-call}
#| eval: false

# --- Phase 1: build fold + pre-compute M1 (one-time per held-out season) ---
fold <- nested_loso_build_fold(
  allD = allD, test_season = test_s, exclude_seasons = "2015-16",
  k_ref = 25L, ref_method = "fs", manual_labels = manual_labels
)
m1_train <- m1_walkforward_multi(
  allD = allD, ref = fold$ref, hyper = fold$hyper, params = params,
  seasons = fold$train_seasons,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L, parallel = TRUE
)
m1_test <- m1_walkforward_predictions(
  seasonD = allD[allD$season == test_s, ],
  ref = fold$ref, hyper = fold$hyper, params = params,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L
)

# --- Phase 2: for each candidate spec, train M2 and score held-out season ---
# v4 grid: Kr x k_f x k_e x alpha_state (189 specs, fixed delta=0, k_s=0, lambda_w=0)
grid <- tidyr::crossing(
  Kr          = c(1L, 2L, 3L),
  k_f         = c(2L, 3L, 4L),
  k_e         = c(3L, 4L, 6L),
  alpha_state = c(0.03, 0.05, 0.07, 0.10, 0.15, 0.20, 0.25)
)
spec <- stage2_make_spec(
  delta = 0L, Kr = Kr, T = "S", k_f = k_f, k_e = k_e,
  k_1 = 4L, k_2 = 0L, k_w = 0L, k_s = 0L,
  alpha_state = alpha_state, lambda_w = 0, w_floor = 0.05
)

m2      <- nested_loso_m2_train(fold = fold, m1_train_preds = m1_train, spec = spec)
eval_out <- nested_loso_m2_eval(
  allD = allD, fold = fold, m2_fit = m2, m1_test_preds = m1_test,
  spec = spec, eval_window = 12L, k_deriv = 10L,
  manual_labels = manual_labels, flag_args = flag_args
)
```

### Full grid results

```{r load-production-results}
#| eval: true

prod_results_v4 <- readRDS("../data/nested_loso_v4_production.rds")
cat("v4 grid:", nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
cat("Best:", prod_results_v4$best_spec_id,
    "| mean_nll =", round(prod_results_v4$summary$mean_nll[1], 4), "\n")
```

### Best spec {#sec-best-spec-interp}

```{r load-best-spec}
#| include: false
best_spec_nested <- prod_results_v4$best_spec
```

The tuned spec is `Kr=1, k_f=2, k_e=4, alpha_state=0.10` with `delta=0, k_s=0, lambda_w=0`.

**`delta = 0` — no template shift.**
M1's alignment already places the current season correctly on the reference curve. No residual systematic timing offset remains. This validates that multi-template alignment with `k_ref=25` is well-calibrated.

**`Kr = 1` — no ramp.**
The template covariate is applied at full weight from the first week after ignition. There is no benefit to a gradual blend — once ignition fires, the reference curve shape is immediately a reliable anchor. This reflects the high quality of the M1 alignment.

**`k_f = 2` — near-linear template smooth.**
The mapping from the template covariate $z_{s,w}$ to the logit-scale response is essentially linear (k_f=2 is the minimum non-trivial smooth in `mgcv`). The model trusts the reference curve shape and scales it linearly without bending — a tighter basis also reduces overfitting across 10 training seasons.

**`k_e = 4` — moderate EWMA spline.**
The EMA state `z_ema` enters with a 4-knot smooth — enough to capture a nonlinear current-level correction (different responses at high vs low baseline positivity) without overfitting. k_e=3 has slightly lower brier but higher NLL; k_e=6 overfits.

**`alpha_state = 0.10` — slow EMA, ≈10-week memory.**
The effective lookback is $1/\alpha = 10$ weeks. The model wants long-term memory of the positivity level — a slow average that tracks the season's overall trajectory rather than reacting to week-to-week noise. This parameter is genuinely interior: both lower (too sluggish) and higher (too reactive) values hurt. The grid has converged here.

**`k_s = 0` — no per-season deviation.**
Allowing each season its own weekly shape (factor-smooth) hurts out-of-sample. The shared reference curve captures enough cross-season variation — per-season wiggle room overfits on 10 training seasons.

**`lambda_w = 0` — uniform training weights.**
Down-weighting late-season observations provides no benefit. The model learns equally from all post-ignition weeks, including the declining phase. The late-season pattern is consistent enough across seasons to inform early-season accuracy.

**Summary.** The tuned $M_2$ is a **simple, pooled, near-linear GAM** that trusts the M1 template immediately at ignition, corrects for the current level via a 10-week EMA, and does not adapt per-season. With only 10 training seasons, parsimony generalises better than flexibility.

## Fit-back on historical data

Refit $M_2$ on all historical seasons using the best nested-LOSO spec and the production reference curve.

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
joint_out_nested <- nested_loso_refit_best(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec           = best_spec_nested
)

ggplotly(plot_stage2_joint_fit_by_season(
  joint_out   = joint_out_nested,
  dat_raw     = alignedD_prosp,
  ign_hat_df  = det_all$compare,
  template_df = template_df
))
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```
