---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/m0_retro.R")
source("../R/m0_training.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m2_spec_grid.R")
source("../R/m2_nested_loso.R")
# Pull internal helpers missing from R/ mirror
.plot_stage2_joint_fit_prosp <- flualign:::.plot_stage2_joint_fit_prosp

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
hyper       <- ref_cache$hyper
if (is.null(hyper)) stop("hyper not found in ref_production.rds — re-run docs/run.qmd first")
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned   <- readRDS("../data/stage1_tuning.rds")
params  <- tuned$best_params  # M0 detection params; used by m1_walkforward_multi
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Define the `$M_2$` model and its feature set.
- Explain the frozen GAM + online bias correction used after ignition.
- Tune the `$M_2$` hyperparameters with nested LOSO.
- Inspect the fit-back of the production model on historical seasons.

Inputs assumed here:

- `alignedD` — historical data aligned using manual ignition labels
- `template_df` — reference curve from `estimateRef()`
- `det_all` — per-season ignition detections from tuned `$M_0$`
:::

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_r>0\}\;
  S_{5}^{h}\!\left(z^{(\mathrm{resid})}_{s,w};\; k_r\right).
\end{aligned}
$$ {#eq-mod1}

At deployment time, a **trend-augmented bias correction** $\beta^{h}_w$ is added to $\eta^{h}_{s,w}$ before applying the inverse-logit (see @sec-bias-correction).

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,Kr)$

- Template backbone covariate: injects the reference curve shape into $M_2$'s linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §$M_1$ training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,Kr)=\omega(t_{s,w};Kr)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;Kr)=\min\!\left(1,\max\!\left(0,\dfrac{t}{Kr}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $Kr$ weeks (with $Kr=1$ meaning no ramp).

- **Notation warning — $\delta$ here is not $M_1$'s dilation $\delta_w$.** In $M_1$, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time $M_2$ sees it. The $\delta$ in $M_2$ is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if $M_1$'s alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that $M_1$'s alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{5}^{h}(z^{(\mathrm{resid})}_{s,w};k_r)$

Amplitude residual: captures the deviation of the current season's positivity level from what the template predicts, breaking the collinearity between `z_ema` and `logit_f_eff`.

$$
z^{(\mathrm{resid})}_{s,w} = z^{(\mathrm{ema})}_{s,w} - z_{s,w}(\delta, Kr).
$$

When $z^{(\mathrm{resid})}_{s,w} > 0$ the current season is running hotter than the template; when $< 0$ it is cooler. This signal lets $M_2$ adjust amplitude independently of the template's phase position.

- **Hyperparams:** $k_r$ (basis dimension).
- **Toggle:** on if $k_r > 0$; off if $k_r = 0$.
- **R (feature):** `z_resid = z_ema - logit_f_eff` computed in `prep_stage2_joint()`.
- **R (model):** `s(z_resid, by=lead, bs="ts", k=k_r)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Design choice.** $\lambda_w$ is fixed at 0 (uniform weighting). The pre-peak ignition-to-peak window is already the focus of evaluation and the model naturally emphasises it through the z_ema and z_resid covariates.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Frozen GAM with online bias correction {#sec-bias-correction}

$M_2$ is trained **once** on all historical seasons and deployed as a frozen GAM. Instead of weekly refit (which re-estimates the entire GAM each week with the current season appended), we use a **trend-augmented bias correction** applied on the logit scale at prediction time.

### Why frozen + bias replaces weekly refit

Weekly refit provided two benefits: (1) the season random effect $b_{s^*}$ absorbed inter-season amplitude differences, and (2) the smooth re-estimation captured current-season trajectory. Both are now handled more efficiently:

- **Season amplitude** → `z_resid` ($z^{(\mathrm{ema})} - z_{s,w}$) provides continuous, within-week amplitude deviation from the template — richer than a single-season random intercept.
- **Trajectory tracking** → the Holt EMA bias correction adapts per-horizon, per-week, anticipating trend direction — more responsive than re-estimating smooth basis functions from $\sim$12 new rows added to $\sim$120 historical.

The frozen approach is $\sim$12× faster for grid search (1 GAM fit per fold vs $\sim$12).

### Trend-augmented Holt EMA

At each evaluation week $w$, when a past prediction's target becomes observed, the bias correction is updated per horizon $h$:

$$
\begin{aligned}
r^h_w &= \mathrm{logit}(p^{(\mathrm{obs})}_w) - \mathrm{logit}(\hat{p}^{h}_{w}), \\[4pt]
\ell^h_w &= \alpha\, r^h_w + (1-\alpha)\,\bigl(\ell^h_{w-1} + \tau^h_{w-1}\bigr), \\[4pt]
\tau^h_w &= \beta\,\bigl(\ell^h_w - \ell^h_{w-1}\bigr) + (1-\beta)\,\tau^h_{w-1}, \\[4pt]
\beta^h_w &= \ell^h_w + \tau^h_w,
\end{aligned}
$$

where $\ell^h_w$ is the bias level, $\tau^h_w$ is the bias trend, and $\beta^h_w$ is the correction added to $\eta^h_{s,w}$ before applying $\mathrm{logit}^{-1}$. Parameters $\alpha = 0.3$ (level smoothing) and $\beta = 0.2$ (trend smoothing) are fixed design choices.

**Why trend-augmented over simple EMA.** A simple EMA always lags behind accelerating seasons — if bias is growing week-over-week (as during the rise phase), the correction arrives too late. The Holt trend term $\tau^h_w$ extrapolates one step ahead, anticipating growing bias and correcting proactively.

**Initialisation.** $\ell^h_0 = 0$, $\tau^h_0 = 0$ (no correction until the first target is observed).

### Prospective derivatives

For each season $s$ and week $w$, derivatives are estimated on the **link scale** using only the observed history up to week $w$. We fit a **local quadratic** using only the most recent $k$ points up to week $w$, centred at $w$:

$$
z_{s,w} = \mathrm{logit}(\mathrm{clamp}(p_{s,w})) \approx a + b\,u + c\,u^2,
\qquad u = w' - w
$$

The prospective derivatives at week $w$ are:

$$
d1_{s,w} = b, \qquad d2_{s,w} = 2c.
$$

This estimator is **prospective-safe** — it only uses weeks $\le w$.

```{r compute-prospective-derivs}
#| fig-height: 10
#| fig-width: 12
alignedD_prosp <- add_prospective_derivs_link(alignedD, k = 5L, eps = 1e-6, min_obs = 4L)

alignedD_long <- alignedD_prosp %>%
  select(season, weekF, d1_link, d2_link) %>%
  pivot_longer(c(d1_link, d2_link), names_to = "deriv", values_to = "value")

ggplotly(
  ggplot(alignedD_long, aes(x = weekF, y = value, color = deriv)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_line() +
    facet_wrap(~ season, scales = "free_y") +
    labs(x = "weekF", y = "Derivative on logit scale", color = "")
)
```

## Hyperparameter tuning

$M_2$ hyperparameters are selected by **nested leave-one-season-out (LOSO) cross-validation**. Standard LOSO leaks alignment information: it uses a single reference curve fitted on all seasons, including the held-out one. Nested LOSO fixes this by re-estimating the reference curve and all $M_1$ predictions per fold.

**Protocol per fold:**

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold template.
3. Run $M_1$ walk-forward on training seasons → $M_1$ stacking features.
4. Train $M_2$ with the per-fold template and $M_1$ features.
5. Run $M_1$ walk-forward on $s^*$ → $M_1$ test features.
6. Evaluate $M_2$ on $s^*$ using uniform-weight NLL over the first 12 post-ignition weeks.

Steps 1–3 (Phase 1) are pre-computed once per fold and cached; Phase 2 (steps 4–6) runs in minutes regardless of grid size.

### Tuning parameters

The table below lists every hyperparameter, its role in the model, and the values searched. Parameters marked *fixed* were held constant throughout based on prior analysis or implementation constraints.

```{r tuning-param-table}
#| echo: false
tibble::tribble(
  ~Parameter,  ~Description,                                                                                                                                                    ~Type,          ~`Grid / value`,
  "δ (delta)", "Integer-week nudge to the template covariate on the aligned axis. Compensates for residual timing offset after $M_1$ alignment. δ=0 means no nudge.",             "Integer",      "{-1, 0, 1}",
  "Kr",        "Template ramp length in weeks. Kr=1 means full template weight from week 1 of ignition; Kr>1 blends in gradually over Kr weeks.",                              "Integer ≥ 1",  "{1, 2, 3}",
  "k_f",       "Basis dimension for the template smooth $S_1^h$. Controls how nonlinearly the model maps the template covariate to the response.",                             "Integer ≥ 2",  "{3, 5, 7}",
  "k_e",       "Basis dimension for the EMA-state smooth $S_4^h$. Controls nonlinearity of the current-level correction.",                                                     "Integer ≥ 0",  "{2, 3, 4}",
  "α_state",   "EWMA smoothing decay for the current-level covariate z_ema. Lower α = slower adaptation, longer effective memory (≈ 1/α weeks).",                             "Real ∈ (0,1)", "{0.03, 0.05, 0.10}",
  "k_r",       "Basis dimension for amplitude-residual smooth $S_5^h$ (z_resid = z_ema − logit_f_eff). k_r=0 disables.",                                                      "Integer ≥ 0",  "{0, 3, 5}",
  "λ_w",       "Time-decay weight exponent. Down-weights late-season training rows. λ_w=0 is standard uniform MLE.",                                                           "Real ≥ 0",     "Fixed: 0",
  "k_1",       "Basis dimension for slope smooth (prospective first derivative d1). Dropped in v11 — derivative is noisy and redundant with z_ema + template.",                "Integer ≥ 0",  "Fixed: 0",
  "k_2",       "Basis dimension for curvature smooth (prospective second derivative d2). Dropped in v11.",                                                                     "Integer ≥ 0",  "Fixed: 0"
) |>
  gt() |>
  tab_header(title = "M_2 tuning parameters") |>
  cols_width(Parameter ~ px(90), Description ~ px(460),
             Type ~ px(110), `Grid / value` ~ px(240)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#f0f4f8"),
            locations = cells_body(rows = c(1, 2, 3, 4, 5, 6)))
```

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the full approach to peak (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean binomial NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** regardless of $\lambda_w$.

Predicted probabilities $\hat{p}_i$ pass through the **soft positivity ceiling** before NLL is computed, matching the transformation applied at deployment time. The ceiling is a tanh squash above the 95th-percentile of training positivity ($p_{\text{knee}}$), asymptoting toward $p_{\text{ceil}} = p_{\max}^{(\text{train})} + 0.5(p_{\max}^{(\text{train})} - p_{\text{knee}})$. Because historical seasons never exceed $\sim$34% positivity the cap is near-identity within the training distribution and has negligible effect on NLL or spec ranking. Its purpose is to ensure that the model evaluated during training is identical to the model deployed prospectively.

### Tuning call

The nested LOSO was run in three stages: v9 (1,296 specs), v10 (8 boundary-check specs), and v11 (729 specs with model simplification + z_resid + frozen GAM + bias correction). Results from v11 supersede earlier runs. The v11 entry-point script is `scripts/run_nested_loso_v11.R`.

**v11 changes from v9/v10:**

- **Dropped terms:** `k_n` (logN_now), `k_w` (aligned-week smooth), `k_s` (per-season factor-smooth), `k_1` (derivative d1) — all fixed at 0.
- **New covariate:** `z_resid = z_ema − logit_f_eff` (amplitude residual), controlled by `k_r`.
- **Frozen GAM evaluation** with trend-augmented Holt EMA bias correction (~12× faster than weekly refit).
- **Grid:** `delta × Kr × k_f × k_e × alpha_state × k_r` = 3×3×3×3×3×3 = 729 specs.

```{r tuning-call}
#| eval: false

# --- Phase 1: build fold + pre-compute $M_1$ (one-time per held-out season) ---
fold <- nested_loso_build_fold(
  allD = allD, test_season = test_s, exclude_seasons = "2015-16",
  k_ref = 25L, ref_method = "fs", manual_labels = manual_labels
)
m1_train <- m1_walkforward_multi(
  allD = allD, ref = fold$ref, hyper = fold$hyper, params = params,
  seasons = fold$train_seasons,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L, parallel = TRUE
)
m1_test <- m1_walkforward_predictions(
  seasonD = allD[allD$season == test_s, ],
  ref = fold$ref, hyper = fold$hyper, params = params,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L
)

# --- Phase 2: train frozen GAM once per spec, evaluate with bias correction ---
# v11 grid: delta × Kr × k_f × k_e × alpha_state × k_r (729 specs)
grid <- tidyr::crossing(
  delta       = c(-1L, 0L, 1L),
  Kr          = c(1L, 2L, 3L),
  k_f         = c(3L, 5L, 7L),
  k_e         = c(2L, 3L, 4L),
  alpha_state = c(0.03, 0.05, 0.10),
  k_r         = c(0L, 3L, 5L)
)
spec <- stage2_make_spec(
  delta = delta, Kr = Kr, T = "S", k_f = k_f, k_e = k_e,
  k_r = k_r, k_1 = 0L, k_2 = 0L, k_n = 0L, k_w = 0L, k_s = 0L,
  alpha_state = alpha_state, lambda_w = 0, w_floor = 0.05
)

m2      <- nested_loso_m2_train(fold = fold, m1_train_preds = m1_train, spec = spec)
eval_out <- nested_loso_m2_eval_frozen_bias(
  allD = allD, fold = fold, m2_fit = m2, m1_test_preds = m1_test,
  spec = spec, eval_window = 12L,
  manual_labels = manual_labels, flag_args = flag_args,
  bias_alpha = 0.3, bias_beta = 0.2
)
```

### Full grid results

```{r load-production-results}
#| eval: true

# Load most recent LOSO results: v11 > combined(v9+v10) > v9 > v8 > v4.
if (file.exists("../data/nested_loso_v11_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v11_production.rds")
  cat("v11 (frozen GAM + bias correction + z_resid) grid:",
      nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_combined_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_combined_production.rds")
  cat("Combined (v9+v10) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v10_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v10_production.rds")
  cat("v10 (alpha boundary check) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v9_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v9_production.rds")
  cat("v9 (full parameter search) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v8_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v8_production.rds")
  cat("v8 (weekly refit, k_f expanded) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else {
  prod_results_v4 <- readRDS("../data/nested_loso_v4_production.rds")
  cat("v4 grid (frozen-fit eval):", nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
}
cat("Best:", prod_results_v4$best_spec_id,
    "| mean_nll =", round(prod_results_v4$summary$mean_nll[1], 4), "\n")
```

### Best spec {#sec-best-spec-interp}

```{r load-best-spec}
#| include: false
best_spec_nested <- prod_results_v4$best_spec
```

The best spec is loaded from the most recent LOSO results file. Below is the interpretation of the v9/v10 production spec; after running the v11 grid search this section will be updated with the new best spec.

**Current production spec (v9/v10):** `Kr=1, k_f=5, k_e=3, alpha_state=0.05` with `delta=0, k_1=0, k_w=0, k_s=0, k_r=0, lambda_w=0`.

**v11 additions to interpret:** The v11 grid adds `k_r` (amplitude residual z_resid) and uses frozen GAM + trend-augmented bias correction instead of weekly refit. Key questions for v11:

- Does `k_r > 0` (z_resid enabled) improve over `k_r = 0`?
- Are `k_f` and `k_e` ranges still interior?
- Does `alpha_state` remain at 0.05?

**`delta = 0` — no template shift.**
$M_1$'s alignment already places the current season correctly on the reference curve. No residual systematic timing offset remains. All three delta values (−1, 0, +1) tie when the other parameters are fixed at the optimum — confirming the template is well-aligned.

**`Kr = 1` — no ramp.**
The template covariate is applied at full weight from the first week after ignition. Once ignition fires the reference curve shape is immediately a reliable anchor — a gradual blend adds no benefit. All `Kr` values (1, 2, 3) tie, confirming the ramp has no effect given the winning model structure.

**`k_f = 5` — moderately flexible template smooth.**
The logit-scale mapping from the template covariate $z_{s,w}$ uses 5 basis functions, capturing nonlinear relationships between the reference curve position and realised positivity. `k_f=5` is a clear interior optimum: `k_f=3` gives NLL+7.7 and `k_f=7` gives NLL+0.5.

**`k_e = 3` — minimal EWMA smooth.**
The EMA state `z_ema` enters with 3 basis functions (the mgcv minimum for thin-plate splines). In v9 `k_e=2` was specified and was silently promoted to 3 by mgcv — v10 confirmed `k_e=3` explicitly matches the v9 result identically. `k_e=4` is worse (NLL+3.9), confirming that only a near-linear EMA correction is needed.

**`alpha_state = 0.05` — slow EMA, ≈20-week memory.**
The effective lookback is $1/\alpha = 20$ weeks — a slow-decay tracker that approximates a running mean of within-season positivity. v9 found `alpha=0.05` best among {0.05, 0.10, 0.20}; v10 extended the search to {0.01, 0.02, 0.03} and confirmed `alpha=0.05` is the global optimum — going lower degrades performance sharply (alpha=0.03: NLL+2.7; alpha=0.01: NLL+40.8).

**`k_1 = 0`, `k_w = 0`, `k_s = 0` — derivative, week-trend and season-specific terms all disabled.**
Adding first-derivative (`k_1=4`: NLL+26.7), aligned-week trend (`k_w=6`: NLL+17.9), or per-season deviations (`k_s=4`) all hurt substantially. These terms were formally dropped from the model in v11.

**`lambda_w = 0` — uniform training weights.**
Down-weighting late-season observations provides no benefit. The model learns equally from all post-ignition weeks.

**Summary.** The v9/v10 tuned $M_2$ uses a **moderately flexible template smooth (`k_f=5`) with a minimal EMA correction (`k_e=3`, `alpha=0.05`)**. v11 builds on this by adding the amplitude residual `z_resid` and trend-augmented bias correction while dropping the derivative, aligned-week, and factor-spline terms.


## Fit-back on historical data

Refit $M_2$ on all historical seasons using the best nested-LOSO spec and the production reference curve.

**M1 walk-forward predictions for historical seasons** are computed here and saved with the production kit. They are passed to `nested_loso_refit_best()` (frozen GAM) and `run_m2_forecast()` at deployment so that `logit_f_eff` is M1-based in both contexts — ensuring no inconsistency. There is no data leakage: the predictions are walk-forward (each season uses only past data) and the current season is never in `m1_train_preds`.

```{r compute-m1-production-preds}
#| eval: true
#| message: false
# Production M1 predictions for ALL training seasons (no leakage — walk-forward).
# These replace template-based logit_f_eff in the frozen GAM.  Save them in
# m2_production.rds so load_prospective_kit() can thread them through
# run_m2_forecast() at deployment.
EXCLUDE_SEAS <- "2015-16"  # Same exclusion used throughout LOSO evaluation.
m1_preds_production <- m1_walkforward_multi(
  allD               = allD,
  ref                = ref,
  hyper              = hyper,
  params             = params,
  seasons            = sort(setdiff(unique(allD$season), EXCLUDE_SEAS)),
  temperature        = 0.25,
  rise_weight        = 1.0,
  trough_weight      = 0.1,
  peak_decay         = 0.3,
  slope_weight       = 0.5,
  slope_window       = 4L,
  dynamic_temp       = TRUE,
  dynamic_temp_pivot = 10L,
  parallel           = TRUE,
  verbose            = FALSE
)
message("m1_preds_production: ", nrow(m1_preds_production), " rows across ",
        length(unique(m1_preds_production$season)), " seasons")
```

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
joint_out_nested <- nested_loso_refit_best(
  alignedD_prosp = alignedD_prosp,
  template_df    = template_df,
  spec           = best_spec_nested,
  m1_preds       = m1_preds_production
)

p_fitback <- plot_stage2_joint_fit_by_season(
  out_m1     = joint_out_nested,
  feat_full  = joint_out_nested$train_data,
  dat_raw    = alignedD_prosp,
  ign_hat_df = det_all$compare
)

ggsave("figures/fit_back_by_season.png", p_fitback,
       width = 14, height = 14, dpi = 150)

ggplotly(p_fitback)
```

```{r save-m2-production}
#| eval: true
# Save the production M2 object for use by load_prospective_kit().
#
# Contents:
#   fit            — frozen production GAM (trained with M1-based logit_f_eff)
#   spec           — best nested-LOSO spec
#   template_df    — reference curve (newWeek, fit) for run_m2_forecast()
#   m1_train_preds — M1 walk-forward predictions for all historical training
#                    seasons.  load_prospective_kit() reads this and passes it
#                    to run_m2_forecast() so that logit_f_eff at deployment
#                    matches the feature space used to train the frozen GAM.
#                    No data leakage: predictions are walk-forward and the
#                    current season is never included.
m2_production <- list(
  fit            = joint_out_nested$fit,
  spec           = best_spec_nested,
  template_df    = template_df[, c("newWeek", "fit")],
  m1_train_preds = m1_preds_production
)
saveRDS(m2_production, "../data/m2_production.rds")
message("Saved m2_production.rds")
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```
