---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/m0_retro.R")
source("../R/m0_training.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m2_spec_grid.R")
source("../R/m2_nested_loso.R")
# Pull internal helpers missing from R/ mirror
.plot_stage2_joint_fit_prosp <- flualign:::.plot_stage2_joint_fit_prosp

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
hyper       <- ref_cache$hyper
if (is.null(hyper)) stop("hyper not found in ref_production.rds — re-run docs/run.qmd first")
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned   <- readRDS("../data/stage1_tuning.rds")
params  <- tuned$best_params  # M0 detection params; used by m1_walkforward_multi
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Define the `$M_2$` model and its feature set.
- Explain the frozen GAM + online bias correction used after ignition.
- Tune the `$M_2$` hyperparameters with nested LOSO.
- Inspect the fit-back of the production model on historical seasons.

Inputs assumed here:

- `alignedD` — historical data aligned using manual ignition labels
- `template_df` — reference curve from `estimateRef()`
- `det_all` — per-season ignition detections from tuned `$M_0$`
:::

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_r>0\}\;
  S_{5}^{h}\!\left(z^{(\mathrm{resid})}_{s,w};\; k_r\right) \\[2pt]
&+ \mathbb{I}\{k_{de}>0\}\;
  S_{6}^{h}\!\left(\Delta z^{(\mathrm{ema})}_{s,w};\; k_{de}\right).
\end{aligned}
$$ {#eq-mod1}

At deployment time, a **level-only Holt EMA bias correction** $\beta^{h}_w$ is added to $\eta^{h}_{s,w}$ before applying the inverse-logit (see @sec-bias-correction).

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,Kr)$

- Template backbone covariate: injects the reference curve shape into $M_2$'s linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §$M_1$ training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,Kr)=\omega(t_{s,w};Kr)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;Kr)=\min\!\left(1,\max\!\left(0,\dfrac{t}{Kr}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $Kr$ weeks (with $Kr=1$ meaning no ramp).

- **Notation warning — $\delta$ here is not $M_1$'s dilation $\delta_w$.** In $M_1$, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time $M_2$ sees it. The $\delta$ in $M_2$ is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if $M_1$'s alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that $M_1$'s alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{5}^{h}(z^{(\mathrm{resid})}_{s,w};k_r)$

Amplitude residual: captures the deviation of the current season's positivity level from what the template predicts, breaking the collinearity between `z_ema` and `logit_f_eff`.

$$
z^{(\mathrm{resid})}_{s,w} = z^{(\mathrm{ema})}_{s,w} - z_{s,w}(\delta, Kr).
$$

When $z^{(\mathrm{resid})}_{s,w} > 0$ the current season is running hotter than the template; when $< 0$ it is cooler. This signal lets $M_2$ adjust amplitude independently of the template's phase position.

- **Hyperparams:** $k_r$ (basis dimension).
- **Toggle:** on if $k_r > 0$; off if $k_r = 0$.
- **R (feature):** `z_resid = z_ema - logit_f_eff` computed in `prep_stage2_joint()`.
- **R (model):** `s(z_resid, by=lead, bs="ts", k=k_r)`.

### $S_{6}^{h}(\Delta z^{(\mathrm{ema})}_{s,w};k_{de})$

Rate-of-change of EMA state: provides the model with a purely observation-driven speed signal, enabling faster adaptation when the current season rises or falls unusually quickly.

$$
\Delta z^{(\mathrm{ema})}_{s,w} = z^{(\mathrm{ema})}_{s,w} - z^{(\mathrm{ema})}_{s,w-1},
$$

initialised to 0 at the first ignition week. Unlike $z^{(\mathrm{resid})}$, this covariate does not depend on M1 template quality — it is entirely M1-agnostic.

- **Hyperparams:** $k_{de}$ (basis dimension).
- **Toggle:** on if $k_{de} > 0$; off if $k_{de} = 0$.
- **R (feature):** `dz_ema = z_ema - lag(z_ema, default=first(z_ema))` per season in `prep_stage2_joint()`.
- **R (model):** `s(dz_ema, by=lead, bs="ts", k=k_de)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Design choice.** $\lambda_w$ is fixed at 0 (uniform weighting). The pre-peak ignition-to-peak window is already the focus of evaluation and the model naturally emphasises it through the z_ema and z_resid covariates.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Frozen GAM with online bias correction {#sec-bias-correction}

$M_2$ is trained **once** on all historical seasons and deployed as a frozen GAM. Instead of weekly refit (which re-estimates the entire GAM each week with the current season appended), we use a **level-only EMA bias correction** applied on the logit scale at prediction time.

### Why frozen + bias replaces weekly refit

Weekly refit provided two benefits: (1) the season random effect $b_{s^*}$ absorbed inter-season amplitude differences, and (2) the smooth re-estimation captured current-season trajectory. Both are now handled more efficiently:

- **Season amplitude** → `z_resid` ($z^{(\mathrm{ema})} - z_{s,w}$) provides continuous, within-week amplitude deviation from the template — richer than a single-season random intercept.
- **Trajectory tracking** → `dz_ema` ($\Delta z^{(\mathrm{ema})}$) informs the GAM about observation speed at training time, so the frozen model already encodes this signal. A level-only Holt EMA bias correction handles any residual season-specific offset — it is a safety net, not the primary trajectory mechanism.

The frozen approach is $\sim$12× faster for grid search (1 GAM fit per fold vs $\sim$12).

### Level-only Holt EMA

At each evaluation week $w$, when a past prediction's target becomes observed, the bias correction is updated per horizon $h$:

$$
\begin{aligned}
r^h_w &= \mathrm{logit}(p^{(\mathrm{obs})}_w) - \mathrm{logit}(\hat{p}^{h}_{w}), \\[4pt]
\ell^h_w &= \alpha\, r^h_w + (1-\alpha)\,\ell^h_{w-1}, \\[4pt]
\beta^h_w &= \ell^h_w,
\end{aligned}
$$

where $\ell^h_w$ is the bias level and $\beta^h_w$ is the correction added to $\eta^h_{s,w}$ before applying $\mathrm{logit}^{-1}$. Parameter $\alpha = 0.4$ (level smoothing) is a fixed design choice (~2.5-week half-life).

**Why level-only over trend-augmented.** The `dz_ema` covariate now handles structural trajectory trends within the GAM itself. The Holt trend term is redundant and was creating feedback: when structural bias (e.g. unusually fast rise) is captured by the GAM, the Holt trend term over-corrected. Level-only EMA is a cleaner safety net for distribution-shifted seasons.

**Initialisation.** $\ell^h_0 = 0$ (no correction until the first target is observed). Bias is reset to zero on the first post-peak week to avoid over-correcting the descent phase.

## Hyperparameter tuning

$M_2$ hyperparameters are selected by **nested leave-one-season-out (LOSO) cross-validation**. Standard LOSO leaks alignment information: it uses a single reference curve fitted on all seasons, including the held-out one. Nested LOSO fixes this by re-estimating the reference curve and all $M_1$ predictions per fold.

**Protocol per fold:**

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold template.
3. Run $M_1$ walk-forward on training seasons → $M_1$ stacking features.
4. Train $M_2$ with the per-fold template and $M_1$ features.
5. Run $M_1$ walk-forward on $s^*$ → $M_1$ test features.
6. Evaluate $M_2$ on $s^*$ using uniform-weight NLL over the first 12 post-ignition weeks.

Steps 1–3 (Phase 1) are pre-computed once per fold and cached; Phase 2 (steps 4–6) runs in minutes regardless of grid size.

### Tuning parameters

The table below lists every hyperparameter, its role in the model, and the values searched. Parameters marked *fixed* were held constant throughout based on prior analysis or implementation constraints.

```{r tuning-param-table}
#| echo: false
tibble::tribble(
  ~Parameter,  ~Description,                                                                                                                                                    ~Type,          ~`Grid / value`,
  "δ (delta)", "Integer-week nudge to the template covariate on the aligned axis. Compensates for residual timing offset after $M_1$ alignment. δ=0 means no nudge.",             "Integer",      "{-1, 0, 1}",
  "Kr",        "Template ramp length in weeks. Kr=1 means full template weight from week 1 of ignition; Kr>1 blends in gradually over Kr weeks.",                              "Integer ≥ 1",  "{1, 2, 3}",
  "k_f",       "Basis dimension for the template smooth $S_1^h$. Controls how nonlinearly the model maps the template covariate to the response.",                             "Integer ≥ 2",  "{3, 5, 7}",
  "k_e",       "Basis dimension for the EMA-state smooth $S_4^h$. Controls nonlinearity of the current-level correction.",                                                     "Integer ≥ 0",  "{2, 3, 4}",
  "α_state",   "EWMA smoothing decay for the current-level covariate z_ema. Lower α = slower adaptation, longer effective memory (≈ 1/α weeks).",                             "Real ∈ (0,1)", "{0.03, 0.05, 0.10}",
  "k_r",       "Basis dimension for amplitude-residual smooth $S_5^h$ (z_resid = z_ema − logit_f_eff). k_r=0 disables.",                                                      "Integer ≥ 0",  "{0, 2, 3}",
  "k_de",      "Basis dimension for rate-of-change smooth $S_6^h$ (dz_ema = z_ema[t] − z_ema[t−1]). k_de=0 disables. Added in v12 to capture fast-rise trajectories.",     "Integer ≥ 0",  "{0, 2, 3}",
  "λ_w",       "Time-decay weight exponent. Down-weights late-season training rows. λ_w=0 is standard uniform MLE.",                                                           "Real ≥ 0",     "Fixed: 0"
) |>
  gt() |>
  tab_header(title = "M_2 tuning parameters") |>
  cols_width(Parameter ~ px(90), Description ~ px(460),
             Type ~ px(110), `Grid / value` ~ px(240)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#f0f4f8"),
            locations = cells_body(rows = c(1, 2, 3, 4, 5, 6)))
```

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the full approach to peak (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean binomial NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** regardless of $\lambda_w$.

Predicted probabilities $\hat{p}_i$ pass through the **soft positivity ceiling** before NLL is computed, matching the transformation applied at deployment time. The ceiling is a tanh squash above the 95th-percentile of training positivity ($p_{\text{knee}}$), asymptoting toward $p_{\text{ceil}} = p_{\max}^{(\text{train})} + 0.5(p_{\max}^{(\text{train})} - p_{\text{knee}})$. Because historical seasons never exceed $\sim$34% positivity the cap is near-identity within the training distribution and has negligible effect on NLL or spec ranking. Its purpose is to ensure that the model evaluated during training is identical to the model deployed prospectively.

### Tuning call

The nested LOSO was run in stages: v9 (1,296 specs), v10 (8 boundary-check specs), v11 (729 specs with model simplification + z_resid + frozen GAM), and v12 (72 specs adding dz_ema). Results from v12 supersede earlier runs. The v12 entry-point script is `scripts/run_nested_loso_v12.R`.

**v12 changes from v11:**

- **New covariate:** `dz_ema = z_ema[t] − z_ema[t−1]` (observation speed), controlled by `k_de`. Purely M1-agnostic — enables faster self-alignment when seasons rise unusually quickly.
- **Simplified bias correction:** level-only Holt EMA (α=0.4), replacing trend-augmented Holt (α=0.3, β=0.2). The `dz_ema` GAM term now handles structural trajectory trends, making the Holt trend term redundant.

**v11 changes from v9/v10:**

- **Dropped terms:** `k_n` (logN_now), `k_w` (aligned-week smooth), `k_s` (per-season factor-smooth), `k_1`/`k_2` (derivatives d1, d2) — all fixed at 0.
- **New covariate:** `z_resid = z_ema − logit_f_eff` (amplitude residual), controlled by `k_r`.
- **Frozen GAM evaluation** with level-only Holt EMA bias correction (~12× faster than weekly refit).
- **Grid:** `delta × Kr × k_f × k_e × alpha_state × k_r` = 3×3×3×3×3×3 = 729 specs.

```{r tuning-call}
#| eval: false

# --- Phase 1: build fold + pre-compute $M_1$ (one-time per held-out season) ---
fold <- nested_loso_build_fold(
  allD = allD, test_season = test_s, exclude_seasons = "2015-16",
  k_ref = 25L, ref_method = "fs", manual_labels = manual_labels
)
m1_train <- m1_walkforward_multi(
  allD = allD, ref = fold$ref, hyper = fold$hyper, params = params,
  seasons = fold$train_seasons,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L, parallel = TRUE
)
m1_test <- m1_walkforward_predictions(
  seasonD = allD[allD$season == test_s, ],
  ref = fold$ref, hyper = fold$hyper, params = params,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
  dynamic_temp = TRUE, dynamic_temp_pivot = 10L
)

# --- Phase 2: train frozen GAM once per spec, evaluate with level-only bias correction ---
# v12 grid: delta × Kr × k_f × k_e × alpha_state × k_r × k_de (72 specs)
grid <- tidyr::crossing(
  delta       = 0L,
  Kr          = 1L,
  k_f         = c(3L, 5L),
  k_e         = c(2L, 3L),
  alpha_state = c(0.10, 0.15, 0.20),
  k_r         = c(2L, 3L),
  k_de        = c(0L, 2L, 3L)
)
spec <- stage2_make_spec(
  delta = delta, Kr = Kr, T = "S", k_f = k_f, k_e = k_e,
  k_r = k_r, k_de = k_de, k_n = 0L, k_w = 0L, k_s = 0L,
  alpha_state = alpha_state, lambda_w = 0, w_floor = 0.05
)

m2      <- nested_loso_m2_train(fold = fold, m1_train_preds = m1_train, spec = spec)
eval_out <- nested_loso_m2_eval_frozen_bias(
  allD = allD, fold = fold, m2_fit = m2, m1_test_preds = m1_test,
  spec = spec, eval_window = 12L,
  manual_labels = manual_labels, flag_args = flag_args,
  bias_alpha = 0.4
)
```

### Full grid results

```{r load-production-results}
#| eval: true

# Load most recent LOSO results: v12 > v11 > combined(v9+v10) > v9
if (file.exists("../data/nested_loso_v12_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v12_production.rds")
  cat("v12 (dz_ema + level-only Holt α=0.4) grid:",
      nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_v11_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_v11_production.rds")
  cat("v11 (frozen GAM + bias correction + z_resid) grid:",
      nrow(prod_results_v4$summary), "specs × 10 seasons nested LOSO\n")
} else if (file.exists("../data/nested_loso_combined_production.rds")) {
  prod_results_v4 <- readRDS("../data/nested_loso_combined_production.rds")
  cat("Combined (v9+v10) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
} else {
  prod_results_v4 <- readRDS("../data/nested_loso_v9_production.rds")
  cat("v9 (full parameter search) grid:", nrow(prod_results_v4$summary),
      "specs × 10 seasons nested LOSO\n")
}
cat("Best:", prod_results_v4$best_spec_id,
    "| mean_nll =", round(prod_results_v4$summary$mean_nll[1], 4), "\n")
```

### Best spec {#sec-best-spec-interp}

```{r load-best-spec}
#| include: false
best_spec_nested <- prod_results_v4$best_spec
```

**v12 production spec:** `k_f=3, k_e=2, alpha_state=0.2, k_r=2, k_de=0` with `delta=0, Kr=1, lambda_w=0`. Mean LOSO NLL = **33.49** (72-spec grid, 10 seasons).

v12 introduces two changes from v11: (1) `dz_ema` covariate ablation (`k_de ∈ {0,2,3}`) — k_de=0 won, confirming dz_ema adds no predictive signal in LOSO; (2) Holt EMA simplified to level-only with α raised 0.3→0.4. The result is NLL = 33.49, down from 34.03 in v11 — a further 1.6% improvement driven by the higher `alpha_state=0.2`.

**`delta = 0`, `Kr = 1` — template shift and ramp are inert.**
All delta and Kr values produce identical NLL. Fixed at simplest values.

**`k_f = 3` — minimal template smooth.**
Confirmed again in v12 (kf=5 gives NLL+1.0). $z_{\text{resid}}$ carries the nonlinear amplitude information.

**`k_e = 2` (mgcv promotes to 3) — minimal EMA smooth.**
Near-linear EMA correction suffices.

**`alpha_state = 0.2` — moderately responsive EMA, ≈5-week memory.**
Up from 0.15 in v11. With level-only Holt bias correction (no trend term), a higher α keeps the EMA signal fresh:

| alpha | NLL  |
|-------|------|
| 0.10  | 34.0 |
| 0.15  | 33.7 |
| **0.20** | **33.5** |

**`k_r = 2` — amplitude residual with minimal flexibility.**
$z_{\text{resid}} = z_{\text{ema}} - \text{logit}(f_{\text{eff}})$ with 2 basis functions. `k_r=2` and `k_r=3` tie (NLL identical to 3 d.p.); `k_r=2` chosen for parsimony.

**`k_de = 0` — dz_ema ablation: covariate not needed.**
v12 tested `dz_ema = z_ema[t] - z_ema[t-1]` as an additional rate-of-change covariate to help fast-rise seasons. In LOSO, k_de=0 outperformed k_de=2 and k_de=3 across all alpha/kr combinations. The GAM with level-only Holt bias (α=0.4) already tracks rapid changes adequately without an explicit derivative term.

**`lambda_w = 0` — uniform training weights.**
Down-weighting late-season observations provides no benefit.


## Fit-back on historical data

This section has two parts:

1. **In-sample GAM baseline** (`fit_back_by_season.png`) — the production GAM fitted on *all* historical seasons and predicted back on the same data. **No bias correction. Not deployment-equivalent.** Use this only to check whether the GAM formula is well-specified (i.e., does it capture the seasonal shape?).

2. **Out-of-sample LOSO predictions** (`loso_oos_by_season.png`) — each season held out, GAM trained on the remaining 9 seasons, Holt bias correction applied sequentially week-by-week as in deployment. **This is the honest, deployment-equivalent view.** The x-axis is weeks since ignition (ignition = 0); the evaluation window covers ignition through ignition + 12 weeks.

**M1 walk-forward predictions for historical seasons** are computed here and saved with the production kit. They are passed to `nested_loso_refit_best()` (frozen GAM) and `run_m2_forecast()` at deployment so that `logit_f_eff` is M1-based in both contexts. No data leakage: the predictions are walk-forward and the current season is never included.


```{r compute-m1-production-preds}
#| eval: true
#| message: false
# Production M1 predictions: load from saved m2_production.rds if available
# (avoids re-running the expensive walk-forward on each render).
# To force a recompute, delete data/m2_production.rds before rendering.
if (file.exists("../data/m2_production.rds")) {
  m1_preds_production <- readRDS("../data/m2_production.rds")$m1_train_preds
  message("m1_preds_production loaded from m2_production.rds: ",
          nrow(m1_preds_production), " rows across ",
          length(unique(m1_preds_production$season)), " seasons")
} else {
  EXCLUDE_SEAS <- "2015-16"
  m1_preds_production <- m1_walkforward_multi(
    allD = allD, ref = ref, hyper = hyper, params = params,
    seasons = sort(setdiff(unique(allD$season), EXCLUDE_SEAS)),
    temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
    peak_decay = 0.3, slope_weight = 0.5, slope_window = 4L,
    dynamic_temp = TRUE, dynamic_temp_pivot = 10L,
    parallel = TRUE, verbose = FALSE
  )
  message("m1_preds_production: ", nrow(m1_preds_production), " rows across ",
          length(unique(m1_preds_production$season)), " seasons")
}
```

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
# Load production GAM if already saved; otherwise refit.
if (file.exists("../data/m2_production.rds")) {
  m2_prod_obj      <- readRDS("../data/m2_production.rds")
  joint_out_nested <- list(fit = m2_prod_obj$fit, train_data = NULL)
  # Rebuild train_data for the in-sample plot by calling prep directly
  joint_out_nested <- nested_loso_refit_best(
    alignedD_prosp = alignedD,
    template_df    = template_df,
    spec           = best_spec_nested,
    m1_preds       = m1_preds_production
  )
} else {
  joint_out_nested <- nested_loso_refit_best(
    alignedD_prosp = alignedD,
    template_df    = template_df,
    spec           = best_spec_nested,
    m1_preds       = m1_preds_production
  )
}

p_fitback <- plot_stage2_joint_fit_by_season(
  out_m1     = joint_out_nested,
  feat_full  = joint_out_nested$train_data,
  dat_raw    = alignedD,
  ign_hat_df = det_all$compare
)

ggsave("figures/fit_back_by_season.png", p_fitback,
       width = 14, height = 14, dpi = 150)

# Note: this is the IN-SAMPLE GAM fit — no bias correction applied.
ggplotly(p_fitback + ggplot2::ggtitle(
  "In-sample GAM fit (training data, no bias correction)"))
```

```{r loso-oos-plot}
#| eval: true
#| fig-height: 14
#| fig-width: 10
#| message: false
# LOSO out-of-sample predictions: each season held out, GAM trained on remaining
# seasons, Holt bias correction applied sequentially week-by-week (= deployment).
# x-axis = weeks since ignition so all seasons start at 0 and the 13-week
# evaluation window is directly visible.  y-axis fixed at [0, 0.5].
best_cv_result <- prod_results_v4$cv_results[[prod_results_v4$best_spec_id]]

p_loso_oos <- plot_nested_loso_predictions(
  cv_result = best_cv_result,
  y_max     = 0.5,
  title     = "M2 LOSO out-of-sample (frozen GAM + Holt bias correction)"
)

ggsave("figures/loso_oos_by_season.png", p_loso_oos,
       width = 10, height = 14, dpi = 150)

ggplotly(p_loso_oos)
```

```{r loso-oos-scores}
#| eval: true
# Per-season NLL and RMSE from the best-spec LOSO run.
best_cv_result$scores |>
  dplyr::select(season, mean_nll, rmse_p, brier) |>
  dplyr::mutate(
    mean_nll = round(mean_nll, 2),
    rmse_p   = round(rmse_p, 4),
    brier    = round(brier, 5)
  ) |>
  knitr::kable(caption = "Per-season LOSO scores (out-of-sample, bias-corrected)")
```

```{r save-m2-production}
#| eval: true
# Save the production M2 object for use by load_prospective_kit().
# Skip if already saved with the current best spec (re-running the retrain
# is expensive; delete data/m2_production.rds to force a rebuild).
m2_prod_existing <- if (file.exists("../data/m2_production.rds"))
  readRDS("../data/m2_production.rds") else NULL
spec_changed <- is.null(m2_prod_existing) ||
  !identical(m2_prod_existing$spec$spec_id, best_spec_nested$spec_id)

if (spec_changed) {
  m2_production <- list(
    fit            = joint_out_nested$fit,
    spec           = best_spec_nested,
    template_df    = template_df[, c("newWeek", "fit")],
    m1_train_preds = m1_preds_production
  )
  saveRDS(m2_production, "../data/m2_production.rds")
  message("Saved m2_production.rds")
} else {
  message("m2_production.rds already up to date (spec_id unchanged)")
}
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```



