---
title: "$M_2$: Positivity Forecast Model"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(flualign)

source("../R/m0_retro.R")
source("../R/m0_training.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m2_spec_grid.R")
source("../R/m2_nested_loso.R")
# Pull internal helpers missing from R/ mirror
.plot_stage2_joint_fit_prosp <- flualign:::.plot_stage2_joint_fit_prosp

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Build alignedD using manual ignition labels
res  <- estimateDerivs(allD, k = 10)
outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))
alignedD <- alignIgnition(outs)

# Reference curve and template (from production cache or rebuild)
ref_cache <- if (file.exists("../data/ref_production.rds")) {
  readRDS("../data/ref_production.rds")
} else {
  train_allD <- filter(alignedD, !season %in% c("2015-16"))
  r <- estimateRef(alignedD = train_allD, exSeason = character(0), k = 8L, n_weeks = 52L)
  list(ref = r)
}
ref         <- ref_cache$ref
hyper       <- ref_cache$hyper
if (is.null(hyper)) stop("hyper not found in ref_production.rds — re-run docs/run.qmd first")
template_df <- ref$pred_df
g_ref_fun   <- ref$g_ref_fun

# Stage-1 tuning (for det_all used in training results plots)
tuned   <- readRDS("../data/stage1_tuning.rds")
params  <- tuned$best_params  # M0 detection params; used by m1_walkforward_multi
det_all <- detect_ignition_from_tuning(tuned, alignedD)
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Define the `$M_2$` model and its feature set.
- Explain the frozen GAM + online bias correction used after ignition.
- Tune the `$M_2$` hyperparameters with nested LOSO.
- Inspect the fit-back of the production model on historical seasons.

Inputs assumed here:

- `alignedD` — historical data aligned using manual ignition labels
- `template_df` — reference curve from `estimateRef()`
- `det_all` — per-season ignition detections from tuned `$M_0$`
:::

## Model specification

After retrospective ignition labelling and alignment, let $w^a_{s,w}$ denote the aligned week index (`newWeek`) corresponding to within-season week $w^f_{s,w}$ (`weekF`). For forecast horizon $h \in \{1,2\}$, we model the positives $x_{s,w+h}$ out of totals $n_{s,w+h}$ as

$$
x_{s,w+h} \mid n_{s,w+h} \sim \mathrm{Binomial}\!\left(n_{s,w+h},\; \pi_{s,w+h}\right).
$$

We fit a single **joint** model for $h=1,2$ by stacking the data across horizons and allowing all effects to vary by lead. The linear predictor is

$$
\begin{aligned}
\eta^{h}_{s,w} = \mathrm{logit}\!\left(\pi^{h}_{s,w}\right)
=&\;\alpha^{h} \\[2pt]
&+  b_s \\[2pt]
&+ \mathbb{I}\{\texttt{T="O"}\}\;
  z_{s,w}(\delta,Kr) \\[2pt]
&+ \mathbb{I}\{\texttt{T="S"}\}\;
  S_{1}^{h}\!\left(z_{s,w}(\delta,Kr);\; k_f\right) \\[2pt]
&+ \mathbb{I}\{k_e>0\}\;
  S_{4}^{h}\!\left(z^{(\mathrm{ema})}_{s,w};\; k_e\right) \\[2pt]
&+ \mathbb{I}\{k_r>0\}\;
  S_{5}^{h}\!\left(z^{(\mathrm{resid})}_{s,w};\; k_r\right) \\[2pt]
&+ \mathbb{I}\{k_{de}>0\}\;
  S_{6}^{h}\!\left(\Delta z^{(\mathrm{ema})}_{s,w};\; k_{de}\right).
\end{aligned}
$$ {#eq-mod1}

At deployment time, a **level-only Holt EMA bias correction** $\beta^{h}_w$ is added to $\eta^{h}_{s,w}$ before applying the inverse-logit (see @sec-bias-correction).

### Model terms

For each term in the model, the indicator functions $\mathbb{I}\{\cdot\}$ represent toggles that allow us to include/exclude the term from the model, and the hyperparameters ($k_*$) control the complexity of the corresponding smooth term.

Below, each term in Eq. (@eq-mod1) has (i) an interpretation, (ii) its hyperparameter(s), (iii) how to turn it on/off, and (iv) the corresponding `mgcv` implementation. All smooths are **lead-specific** (one smooth per horizon), implemented in `mgcv` using `by = lead`.

::: panel-tabset

### $\alpha^{h}$
- Lead-specific intercept, separate baseline for $h=1$ vs $h=2$ in joint model
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `-1 + lead`.

### $b_s$
- Season-level baseline shift shared across weeks/leads.
- **Hyperparams:** none.
- **Toggle:** always on.
- **R:** `s(season, bs="re")`.
- **New-season prediction:** typically excluded (see "Prediction note" below).

### $z_{s,w}(\delta,Kr)$

- Template backbone covariate: injects the reference curve shape into $M_2$'s linear predictor, ramped in gradually after ignition.

  The reference curve $g_{\mathrm{ref}}$ (from §$M_1$ training, on response scale) is evaluated at the aligned week `newWeek`, shifted by an integer $\delta$, then logit-transformed and ramped:

  $$
  f^{(\delta)}_{s,w} = g_{\mathrm{ref}}\!\left(w^a_{s,w} + \delta\right),
  \qquad
  z_{s,w}(\delta,Kr)=\omega(t_{s,w};Kr)\;\mathrm{logit}_{\varepsilon}\!\left(f^{(\delta)}_{s,w}\right),
  $$
  where $t_{s,w}=w^f_{s,w}-w^{(\mathrm{ign})}_s$ is weeks since ignition, and
  $$
  \omega(t;Kr)=\min\!\left(1,\max\!\left(0,\dfrac{t}{Kr}\right)\right)
  $$
  is a linear ramp from 0 at ignition to 1 after $Kr$ weeks (with $Kr=1$ meaning no ramp).

- **Notation warning — $\delta$ here is not $M_1$'s dilation $\delta_w$.** In $M_1$, $\tau_w$ (continuous shift) and $\delta_w$ (dilation/stretch) are fitted every week to align the current season to $g_{\mathrm{ref}}$. That alignment has already been absorbed into `newWeek` ($w^a_{s,w}$) by the time $M_2$ sees it. The $\delta$ in $M_2$ is a separate, fixed integer-week nudge applied to the template covariate, tuned once offline by LOSO grid search. It compensates for any systematic residual timing offset between the aligned time axis and the template peak — for example, if $M_1$'s alignment consistently centres slightly early. In practice, the tuned value is often $\delta = 0$, confirming that $M_1$'s alignment already handles the shift.

- **Hyperparams:**
  - shift $\delta$ (integer weeks; often 0 after tuning)
  - ramp length $Kr\ge 1$ (with $Kr=1$ meaning "no ramp"),
  - clamp $\varepsilon$ in $\mathrm{logit}_\varepsilon(\cdot)$ (fixed, e.g. $10^{-6}$),
  - smooth basis $k_f$ **only if** `T="S"`.
- **Toggle (model entry via `T`):**
  - `T="S"`: include a smooth of the template covariate;
  - `T="O"`: include the template covariate as an offset (no extra smooth);
  - `T="N"`: omit the template term entirely.
- **Toggle (feature-level off):** set `delta = NA` **or** `Kr = NA` to force `logit_f_eff = 0`.
- **R (feature):** `logit_f_eff` computed in `prep_stage2_joint()` from `delta` and `Kr`.
- **R (model):**
  - `T="S"`: `s(logit_f_eff, by=lead, bs="ts", k=k_f)`
  - `T="O"`: `offset(logit_f_eff)`
  - `T="N"`: (omit template term)

### $S_{4}^{h}(z^{(\mathrm{ema})}_{s,w};k_e)$

EMA state (current-level memory): provides the model with a smoothed summary of the current level of positivity, estimated causally in real time.

The covariate $z^{(\mathrm{ema})}_{s,w}$ is an exponentially weighted moving average of the logit-transformed positivity, computed recursively forward in time within each season:

$$
z^{(\mathrm{ema})}_{s,w} = \alpha_{\text{state}}\;\mathrm{logit}_\varepsilon(p_{s,w}) \;+\; (1-\alpha_{\text{state}})\; z^{(\mathrm{ema})}_{s,w-1},
$$

initialised at $z^{(\mathrm{ema})}_{s,1} = \mathrm{logit}_\varepsilon(p_{s,1})$. The smoothing parameter $\alpha_{\text{state}}\in(0,1)$ controls how quickly the memory adapts.

- **Hyperparams:** $\alpha_{\text{state}}\in(0,1)$ (EWMA decay) and $k_e$ (smooth basis dimension).
- **Toggle:** on if $k_e>0$; off if $k_e=0$.
- **R (feature):** `z_ema` computed via `ewma_recursive()` using `alpha_state`.
- **R (model):** `s(z_ema, by=lead, bs="ts", k=k_e)`.

### $S_{5}^{h}(z^{(\mathrm{resid})}_{s,w};k_r)$

Amplitude residual: captures the deviation of the current season's positivity level from what the template predicts, breaking the collinearity between `z_ema` and `logit_f_eff`.

$$
z^{(\mathrm{resid})}_{s,w} = z^{(\mathrm{ema})}_{s,w} - z_{s,w}(\delta, Kr).
$$

When $z^{(\mathrm{resid})}_{s,w} > 0$ the current season is running hotter than the template; when $< 0$ it is cooler. This signal lets $M_2$ adjust amplitude independently of the template's phase position.

- **Hyperparams:** $k_r$ (basis dimension).
- **Toggle:** on if $k_r > 0$; off if $k_r = 0$.
- **R (feature):** `z_resid = z_ema - logit_f_eff` computed in `prep_stage2_joint()`.
- **R (model):** `s(z_resid, by=lead, bs="ts", k=k_r)`.

### $S_{6}^{h}(\Delta z^{(\mathrm{ema})}_{s,w};k_{de})$

Rate-of-change of EMA state: provides the model with a purely observation-driven speed signal, enabling faster adaptation when the current season rises or falls unusually quickly.

$$
\Delta z^{(\mathrm{ema})}_{s,w} = z^{(\mathrm{ema})}_{s,w} - z^{(\mathrm{ema})}_{s,w-1},
$$

initialised to 0 at the first ignition week. Unlike $z^{(\mathrm{resid})}$, this covariate does not depend on M1 template quality — it is entirely M1-agnostic.

- **Hyperparams:** $k_{de}$ (basis dimension).
- **Toggle:** on if $k_{de} > 0$; off if $k_{de} = 0$.
- **R (feature):** `dz_ema = z_ema - lag(z_ema, default=first(z_ema))` per season in `prep_stage2_joint()`.
- **R (model):** `s(dz_ema, by=lead, bs="ts", k=k_de)`.

### $\lambda_w$ (time-decay weight)

To emphasise early post-ignition weeks during training — the period most critical for real-time forecasting — each observation is assigned a time-decay weight

$$
w_{s,w} = \frac{\exp(-\lambda_w\, t_{s,w})}{\bar{w}_s},
\qquad
\bar{w}_s = \operatorname{mean}_{w}\!\left(\exp(-\lambda_w\, t_{s,w})\right),
$$

where $t_{s,w} = w^f_{s,w} - w_s^{(\mathrm{ign})}$ is weeks since ignition. Setting $\lambda_w=0$ recovers standard (uniform) MLE.

The weighted binomial log-likelihood fitted by `mgcv::bam` is then

$$
\ell(\theta) = \sum_s\sum_{w}\sum_h w_{s,w}\;\log\Pr\!\left(x_{s,w+h}\mid n_{s,w+h},\,\pi^h_{s,w}\right).
$$

**Design choice.** $\lambda_w$ is fixed at 0 (uniform weighting). The pre-peak ignition-to-peak window is already the focus of evaluation and the model naturally emphasises it through the z_ema and z_resid covariates.

#### Late-season weight floor

An aggressive $\lambda_w$ reduces the effective weight on late-season observations to near zero. To prevent numerically unstable extrapolation beyond the peak while preserving the pre-peak fit, a **two-phase weight floor** is applied:

$$
w_{s,w} = \begin{cases}
\exp(-\lambda_w\, t_{s,w}) & t_{s,w} \le t_{\text{floor}} \\
\max\!\bigl(\exp(-\lambda_w\, t_{s,w}),\; w_{\text{floor}}\bigr) & t_{s,w} > t_{\text{floor}}
\end{cases}
$$

where $t_{\text{floor}}$ (`t_floor_start`, default 14) activates the floor beyond the pre-peak window, and $w_{\text{floor}}$ (`w_floor`, default 0.05) is the minimum relative weight. The floor is a fitting-time regularisation only — it does not affect the LOSO objective.

:::

## Frozen GAM with online bias correction {#sec-bias-correction}

$M_2$ is trained **once** on all historical seasons and deployed as a frozen GAM. Instead of weekly refit (which re-estimates the entire GAM each week with the current season appended), we use a **Holt EMA bias correction** applied on the logit scale at prediction time.

### Why frozen + bias replaces weekly refit

Weekly refit provided two benefits: (1) the season random effect $b_{s^*}$ absorbed inter-season amplitude differences, and (2) the smooth re-estimation captured current-season trajectory. Both are now handled more efficiently:

- **Season amplitude** → `z_resid` ($z^{(\mathrm{ema})} - z_{s,w}$) provides continuous, within-week amplitude deviation from the template — richer than a single-season random intercept.
- **Trajectory tracking** → `dz_ema` ($\Delta z^{(\mathrm{ema})}$) informs the GAM about observation speed at training time, so the frozen model already encodes this signal. A level-only Holt EMA bias correction handles any residual season-specific offset — it is a safety net, not the primary trajectory mechanism.

The frozen approach is $\sim$12× faster for grid search (1 GAM fit per fold vs $\sim$12).

### Holt EMA bias correction

At each evaluation week $w$, when a past prediction's target becomes observed, the bias correction is updated per horizon $h$ using Holt double exponential smoothing:

$$
\begin{aligned}
r^h_w &= \mathrm{logit}(p^{(\mathrm{obs})}_w) - \mathrm{logit}(\hat{p}^{h}_{w}), \\[4pt]
\ell^h_w &= \alpha\, r^h_w + (1-\alpha)\,(\ell^h_{w-1} + \tau^h_{w-1}), \\[4pt]
\tau^h_w &= \beta\,(\ell^h_w - \ell^h_{w-1}) + (1-\beta)\,\tau^h_{w-1}, \\[4pt]
\beta^h_{w,h'} &= \ell^h_w + h'\,\tau^h_w,
\end{aligned}
$$

where $\ell^h_w$ is the bias level, $\tau^h_w$ is the bias trend, and $\beta^h_{w,h'}$ is the correction added to $\eta^{h'}_{s,w}$ before applying $\mathrm{logit}^{-1}$ (with $h'$ the forecast horizon). Parameters $\alpha$ (level smoothing) and $\beta$ (trend smoothing) are jointly tuned by LOSO grid search.

**Tuning selected $\beta = 0$ (level-only).** The grid searched $\alpha \in \{0.2, 0.3, 0.4, 0.5\}$ and $\beta \in \{0.0, 0.1\}$. The trend term provided no benefit: $\beta=0$ won at every $\alpha$ level. With $\beta=0$ the formula collapses to standard level-only EMA with $\alpha=0.4$ (~2.5-week half-life). The implementation retains the full Holt structure for generality.

**Online season random effect.** Each week, a shrinkage estimate of the current season's amplitude offset is added to the bias correction:
$$
\hat{b}_{s^*,w} = \frac{\sum_{i \le w} \bigl[\mathrm{logit}(p_i) - \hat{\eta}_i^{(-s)}\bigr]}{n_w + \lambda_{re}},
$$
where $\hat{\eta}_i^{(-s)}$ is the GAM prediction excluding the season random effect and $\lambda_{re}=1$ is a shrinkage penalty. This replaces the fully-excluded new-season random effect from earlier versions and reduces systematic amplitude bias in the first few post-ignition weeks.

**Initialisation.** $\ell^h_0 = 0$, $\tau^h_0 = 0$, $\hat{b}_{s^*,0} = 0$. All three are reset to zero on the first post-peak week to avoid over-correcting the descent phase.

## Hyperparameter tuning

$M_2$ hyperparameters are selected by **nested leave-one-season-out (LOSO) cross-validation**. Standard LOSO leaks alignment information: it uses a single reference curve fitted on all seasons, including the held-out one. Nested LOSO fixes this by re-estimating the reference curve and all $M_1$ predictions per fold.

**Protocol per fold:**

1. Hold out season $s^*$.
2. Fit $g_{\mathrm{ref}}^{(-s^*)}$ on the remaining seasons → per-fold template.
3. Run $M_1$ walk-forward on training seasons → $M_1$ stacking features.
4. Train $M_2$ with the per-fold template and $M_1$ features.
5. Run $M_1$ walk-forward on $s^*$ → $M_1$ test features.
6. Evaluate $M_2$ on $s^*$ using uniform-weight NLL over the first 12 post-ignition weeks.

Steps 1–3 (Phase 1) are pre-computed once per fold and cached; Phase 2 (steps 4–6) runs in minutes regardless of grid size.

### Tuning parameters

The table below lists every hyperparameter, its role in the model, and the values searched. Parameters marked *fixed* were held constant throughout based on prior analysis or implementation constraints.

```{r tuning-param-table}
#| echo: false
tibble::tribble(
  ~Parameter,   ~Description,                                                                                                                                                    ~Type,          ~`Grid / value`,
  "δ (delta)",  "Integer-week nudge to the template covariate on the aligned axis. Compensates for residual timing offset after $M_1$ alignment. δ=0 means no nudge.",             "Integer",      "Fixed: 0",
  "Kr",         "Template ramp length in weeks. Kr=1 means full template weight from week 1 of ignition; Kr>1 blends in gradually over Kr weeks.",                              "Integer ≥ 1",  "Fixed: 1",
  "k_f",        "Basis dimension for the template smooth $S_1^h$. Controls how nonlinearly the model maps the template covariate to the response.",                             "Integer ≥ 2",  "{2, 3, 4, 5}",
  "k_e",        "Basis dimension for the EMA-state smooth $S_4^h$. Controls nonlinearity of the current-level correction.",                                                     "Integer ≥ 0",  "{2, 3}",
  "α_state",    "EWMA smoothing decay for the current-level covariate z_ema. Lower α = slower adaptation, longer effective memory (≈ 1/α weeks).",                             "Real ∈ (0,1)", "{0.10, 0.15, 0.20, 0.25, 0.30}",
  "k_r",        "Basis dimension for amplitude-residual smooth $S_5^h$ (z_resid = z_ema − logit_f_eff). k_r=0 disables.",                                                      "Integer ≥ 0",  "{0, 2, 3}",
  "k_de",       "Basis dimension for rate-of-change smooth $S_6^h$ (dz_ema). k_de>0 failed in v13/v14 due to poor dz_ema conditioning; standardized to unit-variance in v15, enabling tuning.",  "Integer ≥ 0",  "{0, 2} (v15)",
  "k_sp",       "Basis dimension for alignment-uncertainty smooth $S_7^h$ (logit_spread = weighted SD of per-template logit predictions from M1 ensemble). k_sp=0 disables.", "Integer ≥ 0",  "{0, 2} (v15)",
  "α (bias)",   "Holt EMA level-smoothing for the logit-scale bias corrector. Controls how quickly the bias adapts (~1/α week half-life).",                                     "Real ∈ (0,1)", "{0.1, 0.2, 0.3, 0.4}",
  "β (bias)",   "Holt EMA trend-smoothing for the bias corrector. β=0 collapses to level-only EMA.",                                                                           "Real ∈ [0,1)", "Fixed: 0.0",
  "λ_w",        "Time-decay weight exponent. Down-weights late-season training rows. λ_w=0 is standard uniform MLE.",                                                           "Real ≥ 0",     "Fixed: 0"
) |>
  gt() |>
  tab_header(title = "M_2 tuning parameters") |>
  cols_width(Parameter ~ px(90), Description ~ px(440),
             Type ~ px(110), `Grid / value` ~ px(240)) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#f0f4f8"),
            locations = cells_body(rows = c(3, 4, 5, 6, 7, 9, 10)))
```

### Evaluation metric

The evaluation window is fixed at $W_{\mathrm{eval}} = 12$ post-ignition weeks, covering the full approach to peak (median ignition-to-peak: 8.5 weeks; maximum: 12 weeks). The LOSO objective is the mean binomial NLL over all held-out seasons and evaluation rows:

$$
\overline{\mathrm{NLL}} = \frac{-1}{\sum_s|\mathcal{E}_s|}\sum_s\sum_{i\in\mathcal{E}_s} \log\Pr(x_i\mid n_i,\hat{p}_i),
\qquad
\mathcal{E}_s = \bigl\{(s,w,h) : 0 \le t_{s,w} \le W_{\mathrm{eval}}\bigr\}.
$$

Evaluation always uses **uniform weights** regardless of $\lambda_w$.

Predicted probabilities $\hat{p}_i$ pass through the **soft positivity ceiling** before NLL is computed, matching the transformation applied at deployment time. The ceiling is a tanh squash above the 95th-percentile of training positivity ($p_{\text{knee}}$), asymptoting toward $p_{\text{ceil}} = p_{\max}^{(\text{train})} + 0.5(p_{\max}^{(\text{train})} - p_{\text{knee}})$. Because historical seasons never exceed $\sim$34% positivity the cap is near-identity within the training distribution and has negligible effect on NLL or spec ranking. Its purpose is to ensure that the model evaluated during training is identical to the model deployed prospectively.

### Tuning call

Hyperparameters were selected by nested leave-one-season-out (LOSO) cross-validation. For each held-out season, M0 detection and M1 alignment were run on the remaining training seasons; M2 was trained on those aligned predictions and evaluated on the held-out season with week-by-week Holt EMA bias correction and online season RE — exactly replicating the prospective deployment loop.

Four rounds of search have been run:

- **v13** (960 specs = 4×k_f × 2×k_e × 5×alpha_state × 3×k_r × 4×bias_alpha × 2×bias_beta) confirmed `alpha_state=0.30` was at the grid boundary.
- **v13b** (15 specs) extended `alpha_state` to 0.50 and confirmed `0.30` is a genuine interior optimum under raw binomial NLL.
- **v14** re-scored v13 results using **Bernoulli NLL** (N-invariant; prevents post-COVID high-volume seasons from dominating). Best `bias_alpha` flipped 0.4 → 0.2; `alpha_state` optimum shifted to 0.40.
- **v14b** extended `alpha_state` under Bernoulli NLL and found the optimum is flat from 0.30–0.50 (interior).
- **v15** (1920 specs = 4×k_f × 2×k_e × 5×alpha_state × 3×k_r × 2×k_de × 2×k_sp × 4×bias_alpha) tested `k_de` (rate-of-change, after `dz_ema` standardization) and `k_sp` (alignment uncertainty `logit_spread`). Both new terms produced identical NLL to `k_de=0, k_sp=0`. **v14 spec confirmed as production model.** Entry-point: `scripts/run_nested_loso_v15.R`.

**Active hyperparameters (v15):**

- `k_f` ∈ {2, 3, 4, 5} — basis dimension for template-position smooth `s(logit_f_eff)`; best (v14) = 4.
- `k_e` ∈ {2, 3} — basis dimension for EMA state smooth `s(z_ema)`; k_e=2 and k_e=3 tie.
- `alpha_state` ∈ {0.30, 0.35, 0.40, 0.45, 0.50} — EMA decay for `z_ema`; best (v14b Bernoulli) = 0.40 (flat 0.30–0.50).
- `k_r` ∈ {0, 2, 3} — basis dimension for amplitude-residual smooth `s(z_resid)`; k_r=0 loses by Bernoulli NLL ≈ 0.013 nats.
- `k_de` ∈ {0, 2} — basis dimension for rate-of-change smooth `s(dz_ema)`; enabled after standardizing `dz_ema` to unit variance.
- `k_sp` ∈ {0, 2} — basis dimension for alignment uncertainty smooth `s(logit_spread)`, where `logit_spread` is the weighted SD of per-template logit predictions from M1.
- `bias_alpha` ∈ {0.1, 0.2, 0.3, 0.4} — Holt level smoothing; best (v14 Bernoulli) = 0.2.
- `bias_beta` fixed at 0.0 (level-only EMA; trend term provides no benefit).

**Fixed parameters:** `delta = 0`, `Kr = 1`, `lambda_w = 0`, `k_n = k_w = k_s = 0`.

```{r tuning-call}
#| eval: false

# --- Phase 1: build fold + pre-compute M1 (one-time per held-out season) ---
# M1 locked spec: k_ref=25, slope_weight=8, slope_window=6, dynamic_temp=FALSE
fold <- nested_loso_build_fold(
  allD = allD, test_season = test_s, exclude_seasons = "2015-16",
  k_ref = 25L, ref_method = "fs", manual_labels = manual_labels
)
m1_train <- m1_walkforward_multi(
  allD = allD, ref = fold$ref, hyper = fold$hyper, params = params,
  seasons = fold$train_seasons,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 8.0, slope_window = 6L,
  dynamic_temp = FALSE, dynamic_temp_pivot = 10L, parallel = TRUE
)
m1_test <- m1_walkforward_predictions(
  seasonD = allD[allD$season == test_s, ],
  ref = fold$ref, hyper = fold$hyper, params = params,
  temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
  peak_decay = 0.3, slope_weight = 8.0, slope_window = 6L,
  dynamic_temp = FALSE, dynamic_temp_pivot = 10L
)

# --- Phase 2: train frozen GAM once per spec, evaluate with Holt bias + online RE ---
# v15 grid: k_f × k_e × alpha_state × k_r × k_de × k_sp × bias_alpha = 960 specs
# k_de now feasible after dz_ema standardization; k_sp tests alignment uncertainty
grid <- tidyr::crossing(
  delta       = 0L, Kr = 1L,
  k_f         = c(2L, 3L, 4L, 5L),
  k_e         = c(2L, 3L),
  alpha_state = c(0.30, 0.35, 0.40, 0.45, 0.50),
  k_r         = c(0L, 2L, 3L),
  k_de        = c(0L, 2L),
  k_sp        = c(0L, 2L),
  bias_alpha  = c(0.1, 0.2, 0.3, 0.4),
  bias_beta   = 0.0
)
spec <- stage2_make_spec(
  delta = delta, Kr = Kr, T = "S", k_f = k_f, k_e = k_e,
  k_r = k_r, k_de = k_de, k_sp = k_sp, k_n = 0L, k_w = 0L, k_s = 0L,
  alpha_state = alpha_state, lambda_w = 0, w_floor = 0.05,
  bias_alpha = bias_alpha, bias_beta = bias_beta
)

m2       <- nested_loso_m2_train(fold = fold, m1_train_preds = m1_train, spec = spec)
eval_out <- nested_loso_m2_eval_frozen_bias(
  allD = allD, fold = fold, m2_fit = m2, m1_test_preds = m1_test,
  spec = spec, eval_window = 12L,
  manual_labels = manual_labels, flag_args = flag_args
)
```

### Full grid results

```{r load-production-results}
#| eval: true
v14_path <- "../data/nested_loso_v14_production.rds"
v15_path <- "../data/nested_loso_v15_production.rds"
if (file.exists(v15_path)) {
  prod_results <- readRDS(v15_path)
  cat("Grid:", nrow(prod_results$summary), "specs × 10 seasons nested LOSO (v15)\n")
} else {
  prod_results <- readRDS(v14_path)
  cat("Grid:", nrow(prod_results$summary), "specs × 10 seasons nested LOSO (v14, Bernoulli NLL)\n")
  cat("(v15 not yet available)\n")
}
cat("Best:", prod_results$best_spec_id,
    "| mean_bernoulli_nll =", round(min(prod_results$summary$mean_nll, na.rm = TRUE), 5), "\n")
```

### Best spec {#sec-best-spec-interp}

```{r load-best-spec}
#| include: false
best_spec_nested <- prod_results$best_spec
```

**Production spec (v14/v15):** `k_f=4, k_e=2, alpha_state=0.40, k_r=2, k_de=0, k_sp=0, bias_alpha=0.2, bias_beta=0.0` with `delta=0, Kr=1, lambda_w=0`. Mean LOSO Bernoulli NLL = **0.40628** (v14/v15, 1920 specs across 10 seasons). v15 (1920-spec grid adding `k_de ∈ {0,2}` and `k_sp ∈ {0,2}`) confirmed that neither rate-of-change nor alignment uncertainty improve over the v14 spec — Bernoulli NLL is identical at 0.406 for both `k_de=0` and `k_de=2`, and for both `k_sp=0` and `k_sp=2`. The v14 spec is therefore confirmed as the production model.

The model uses three effective covariates: template position (`s(logit_f_eff, k=4)`), EMA state (`s(z_ema, k=2)`, α=0.40), and amplitude residual (`s(z_resid, k=2)`). Holt EMA bias correction (α=0.2, β=0.0) plus online season random effect are applied at deployment and in LOSO evaluation.

**`delta = 0`, `Kr = 1` — template shift and ramp are inert.**
All delta and Kr values produce identical NLL. With `z_resid` and bias correction absorbing amplitude and timing adjustments, these parameters have no remaining signal.

**`k_f = 4` — moderate template smooth.**
`k_f=2` and `k_f=3` give higher NLL; `k_f=5` overfits. The template smooth needs modest flexibility to represent the bell-shaped reference curve.

**`k_e = 2` — minimal EMA smooth, k_e=3 ties.**
`k_e=2` and `k_e=3` produce identical NLL at all alpha_state levels; `k_e=2` chosen for parsimony.

**`alpha_state = 0.40` — ≈2.5-week EMA memory.**
The optimum is flat from 0.30–0.50 across both v14b and v15:

| alpha_state | best Bernoulli NLL |
|-------------|---------------------|
| 0.30 | 0.406 |
| 0.35 | 0.406 |
| **0.40** | **0.406** |
| 0.45 | 0.406 |
| 0.50 | 0.406 |

`alpha_state=0.40` chosen as the representative interior point. The model is insensitive to EMA speed across the full 0.30–0.50 range.

**`k_r = 2` — amplitude residual is load-bearing.**
`k_r=0` (drop `z_resid`) loses by Bernoulli NLL ≈ 0.013 nats. Under N-invariant scoring, `z_resid` still provides meaningful signal; `k_r=2` and `k_r=3` tie.

**`k_de = 0` — rate-of-change covariate provides no benefit.**
`k_de > 0` catastrophically failed in v13/v14 (NLL > 117) due to poor conditioning of unscaled `dz_ema`. In v15, `dz_ema` was standardized to unit variance (SD stored in `feature_ranges$dz_ema_sd`) — but even after rescaling, `k_de=2` and `k_de=0` produce identical Bernoulli NLL (0.406). The growth-rate signal is already captured by `z_resid` and the Holt bias corrector.

**`k_sp = 0` — alignment uncertainty provides no benefit.**
`logit_spread` (weighted SD of per-template logit predictions from M1's ensemble) was tested in v15 via `k_sp ∈ {0, 2}`. `k_sp=2` and `k_sp=0` produce identical Bernoulli NLL (0.406). The alignment uncertainty from M1 does not carry additional predictive signal beyond the existing covariates.

**`bias_alpha = 0.2`, `bias_beta = 0.0` — level-only Holt.**
`bias_beta = 0` (level-only) wins at every `bias_alpha` level. Under Bernoulli NLL, `bias_alpha ∈ {0.1, 0.2, 0.3}` all produce Bernoulli NLL ≈ 0.406 (flat); `bias_alpha=0.4` slightly degrades. `bias_alpha=0.2` is chosen as the representative interior point with a ~4.5-week half-life, preventing over-weighting of noisy early-season observations in low-N weeks.

**`lambda_w = 0` — uniform training weights.**
Down-weighting late-season observations provides no benefit.


## Fit-back on historical data

This section has two parts:

1. **In-sample GAM baseline** (`fit_back_by_season.png`) — the production GAM fitted on *all* historical seasons and predicted back on the same data. **No bias correction. Not deployment-equivalent.** Use this only to check whether the GAM formula is well-specified (i.e., does it capture the seasonal shape?).

2. **Out-of-sample LOSO predictions** (`loso_oos_by_season.png`) — each season held out, GAM trained on the remaining 9 seasons, Holt bias correction applied sequentially week-by-week as in deployment. **This is the honest, deployment-equivalent view.** The x-axis is weeks since ignition (ignition = 0); the evaluation window covers ignition through ignition + 12 weeks.

**M1 walk-forward predictions for historical seasons** are computed here and saved with the production kit. They are passed to `nested_loso_refit_best()` (frozen GAM) and `run_m2_forecast()` at deployment so that `logit_f_eff` is M1-based in both contexts. No data leakage: the predictions are walk-forward and the current season is never included.


```{r compute-m1-production-preds}
#| eval: true
#| message: false
# Production M1 predictions: load from saved m2_production.rds if available
# (avoids re-running the expensive walk-forward on each render).
# To force a recompute, delete data/m2_production.rds before rendering.
if (file.exists("../data/m2_production.rds")) {
  m1_preds_production <- readRDS("../data/m2_production.rds")$m1_train_preds
  message("m1_preds_production loaded from m2_production.rds: ",
          nrow(m1_preds_production), " rows across ",
          length(unique(m1_preds_production$season)), " seasons")
} else {
  EXCLUDE_SEAS <- "2015-16"
  # M1 locked spec: slope_weight=8, slope_window=6, dynamic_temp=FALSE
  m1_preds_production <- m1_walkforward_multi(
    allD = allD, ref = ref, hyper = hyper, params = params,
    seasons = sort(setdiff(unique(allD$season), EXCLUDE_SEAS)),
    temperature = 0.25, rise_weight = 1.0, trough_weight = 0.1,
    peak_decay = 0.3, slope_weight = 8.0, slope_window = 6L,
    dynamic_temp = FALSE, dynamic_temp_pivot = 10L,
    parallel = TRUE, verbose = FALSE
  )
  message("m1_preds_production: ", nrow(m1_preds_production), " rows across ",
          length(unique(m1_preds_production$season)), " seasons")
}
```

```{r nested-fit-by-season}
#| eval: true
#| fig-height: 14
#| fig-width: 14
# Load production GAM if already saved; otherwise refit.
if (file.exists("../data/m2_production.rds")) {
  m2_prod_obj      <- readRDS("../data/m2_production.rds")
  joint_out_nested <- list(fit = m2_prod_obj$fit, train_data = NULL)
  # Rebuild train_data for the in-sample plot by calling prep directly
  joint_out_nested <- nested_loso_refit_best(
    alignedD_prosp = alignedD,
    template_df    = template_df,
    spec           = best_spec_nested,
    m1_preds       = m1_preds_production
  )
} else {
  joint_out_nested <- nested_loso_refit_best(
    alignedD_prosp = alignedD,
    template_df    = template_df,
    spec           = best_spec_nested,
    m1_preds       = m1_preds_production
  )
}

p_fitback <- plot_stage2_joint_fit_by_season(
  out_m1     = joint_out_nested,
  feat_full  = joint_out_nested$train_data,
  dat_raw    = alignedD,
  ign_hat_df = det_all$compare
)

ggsave("figures/fit_back_by_season.png", p_fitback,
       width = 14, height = 14, dpi = 150)

# Note: this is the IN-SAMPLE GAM fit — no bias correction applied.
ggplotly(p_fitback + ggplot2::ggtitle(
  "In-sample GAM fit (training data, no bias correction)"))
```

```{r loso-oos-plot}
#| eval: true
#| fig-height: 14
#| fig-width: 10
#| message: false
# LOSO out-of-sample predictions: each season held out, GAM trained on remaining
# seasons, Holt bias correction applied sequentially week-by-week (= deployment).
# x-axis = weeks since ignition so all seasons start at 0 and the 13-week
# evaluation window is directly visible.  y-axis fixed at [0, 0.5].
best_cv_result <- prod_results$cv_results[[prod_results$best_spec_id]]

p_loso_oos <- plot_nested_loso_predictions(
  cv_result = best_cv_result,
  y_max     = 0.5,
  title     = "M2 LOSO out-of-sample (frozen GAM + Holt bias correction)"
)

ggsave("figures/loso_oos_by_season.png", p_loso_oos,
       width = 10, height = 14, dpi = 150)

ggplotly(p_loso_oos)
```

```{r loso-oos-scores}
#| eval: true
# Per-season NLL and RMSE from the best-spec LOSO run.
best_cv_result$scores |>
  dplyr::select(season, mean_nll, rmse_p, brier) |>
  dplyr::mutate(
    mean_nll = round(mean_nll, 2),
    rmse_p   = round(rmse_p, 4),
    brier    = round(brier, 5)
  ) |>
  knitr::kable(caption = "Per-season LOSO scores (out-of-sample, bias-corrected)")
```

```{r save-m2-production}
#| eval: true
# Save the production M2 object for use by load_prospective_kit().
# Skip if already saved with the current best spec (re-running the retrain
# is expensive; delete data/m2_production.rds to force a rebuild).
m2_prod_existing <- if (file.exists("../data/m2_production.rds"))
  readRDS("../data/m2_production.rds") else NULL
spec_changed <- is.null(m2_prod_existing) ||
  !identical(m2_prod_existing$spec$spec_id, best_spec_nested$spec_id)

if (spec_changed) {
  m2_production <- list(
    fit            = joint_out_nested$fit,
    spec           = best_spec_nested,
    template_df    = template_df[, c("newWeek", "fit")],
    m1_train_preds = m1_preds_production
  )
  saveRDS(m2_production, "../data/m2_production.rds")
  message("Saved m2_production.rds")
} else {
  message("m2_production.rds already up to date (spec_id unchanged)")
}
```

```{r nested-fit-summary}
#| eval: true
summary(joint_out_nested$fit)
```

## Function call graph

Functions used in this document for M2 model training, LOSO evaluation, and
the shared `m2_predict_one()` prediction core.

```{dot}
//| fig-width: 13
//| fig-height: 6
digraph m2_train_calls {
  graph [rankdir=LR, fontname="Helvetica", fontsize=11,
         nodesep=0.3, ranksep=0.9, concentrate=true];
  node  [fontname="Helvetica", fontsize=10, style=filled,
         shape=box, margin="0.12,0.05"];
  edge  [fontname="Helvetica", fontsize=9, arrowsize=0.6];

  // rank hint
  nested_loso_refit_best -> train_stage2_joint [style=invis];

  subgraph cluster_spec {
    label="Spec & Features (m2_spec_grid.R, m2_training.R)"; style=filled;
    color="#004085"; fillcolor="#cce5ff"; fontcolor="#004085";
    fontsize=11; fontname="Helvetica-Bold";
    stage2_make_spec            [label="stage2_make_spec"];
    stage2_build_joint_formula  [label="stage2_build\n_joint_formula"];
    stage2_exclude_newseason    [label="stage2_exclude\n_newseason"];
    prep_stage2_joint           [label="prep_stage2_joint"];
    add_prospective_derivs_link [label="add_prospective\n_derivs_link"];
    format_current_for_stage2   [label="format_current\n_for_stage2"];
  }
  subgraph cluster_loso {
    label="LOSO Grid Search (m2_nested_loso.R)"; style=filled;
    color="#003060"; fillcolor="#b8d4f0"; fontcolor="#003060";
    fontsize=11; fontname="Helvetica-Bold";
    nested_loso_grid_search          [label="nested_loso\n_grid_search"];
    nested_loso_m2_eval_weekly_refit [label="nested_loso_m2\n_eval_weekly_refit"];
    nested_loso_refit_best           [label="nested_loso\n_refit_best"];
    loso_m1_m2_joint                 [label="loso_m1_m2_joint"];
    score_stage2_metrics             [label="score_stage2_metrics"];
  }
  subgraph cluster_fit {
    label="Model Fitting (m2_training.R)"; style=filled;
    color="#002040"; fillcolor="#9ec5e8"; fontcolor="#002040";
    fontsize=11; fontname="Helvetica-Bold";
    train_stage2_joint         [label="train_stage2_joint"];
    train_stage2_joint_prepped [label="train_stage2_joint\n_prepped"];
    refit_stage2_weekly        [label="refit_stage2_weekly"];
  }
  subgraph cluster_m1 {
    label="M1 Walk-forward (pipeline_bridge.R)"; style=filled;
    color="#155724"; fillcolor="#c3e6cb"; fontcolor="#155724";
    fontsize=11; fontname="Helvetica-Bold";
    m1_walkforward_multi       [label="m1_walkforward_multi"];
    m1_walkforward_predictions [label="m1_walkforward\n_predictions"];
  }
  subgraph cluster_pred {
    label="Predict (m2_training.R)"; style=filled;
    color="#4a0080"; fillcolor="#e2ccff"; fontcolor="#4a0080";
    fontsize=11; fontname="Helvetica-Bold";
    m2_predict_one   [label="m2_predict_one", penwidth=2];
    make_soft_cap_fn [label="make_soft_cap_fn"];
  }

  stage2_make_spec           -> stage2_build_joint_formula;
  stage2_make_spec           -> stage2_exclude_newseason;
  prep_stage2_joint          -> add_prospective_derivs_link;
  format_current_for_stage2  -> add_prospective_derivs_link;
  format_current_for_stage2  -> train_stage2_joint;
  train_stage2_joint         -> train_stage2_joint_prepped;
  train_stage2_joint         -> prep_stage2_joint;
  train_stage2_joint         -> stage2_make_spec;
  train_stage2_joint_prepped -> stage2_build_joint_formula;
  refit_stage2_weekly        -> format_current_for_stage2;
  refit_stage2_weekly        -> prep_stage2_joint;
  refit_stage2_weekly        -> train_stage2_joint;
  refit_stage2_weekly        -> score_stage2_metrics;
  nested_loso_grid_search    -> m1_walkforward_multi;
  nested_loso_grid_search    -> refit_stage2_weekly;
  nested_loso_grid_search    -> train_stage2_joint;
  nested_loso_grid_search    -> stage2_make_spec;
  nested_loso_m2_eval_weekly_refit -> refit_stage2_weekly;
  nested_loso_m2_eval_weekly_refit -> m2_predict_one;
  nested_loso_refit_best     -> nested_loso_grid_search;
  nested_loso_refit_best     -> train_stage2_joint;
  loso_m1_m2_joint           -> m1_walkforward_predictions;
  loso_m1_m2_joint           -> prep_stage2_joint;
  loso_m1_m2_joint           -> train_stage2_joint;
  loso_m1_m2_joint           -> score_stage2_metrics;
  m1_walkforward_multi       -> m1_walkforward_predictions;
  prep_stage2_joint          -> m1_walkforward_predictions;
  m2_predict_one             -> make_soft_cap_fn;
}
```
