---
title: "$M_0$: Prospective Ignition Detector Training"
format:
  html:
    html-math-method: mathjax
    self-contained: true
    toc: true
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(mgcv)
library(plotly)
library(data.table)
library(MMWRweek)
library(gt)
library(PAGe)

source("../R/m0_retro.R")
source("../R/m0_training.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")

startWeek <- 27L

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(
    season, week, year,
    start_year = seasonstart,
    date = week_start_date,
    y = pos_flua,
    N = test_flu
  ) %>%
  mutate(
    neg   = N - y,
    date  = as.Date(date),
    mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
    nW_true   = n_weeks_in_start_year(start_year),
    weekS     = ((week - 35L)       %% nW_true) + 1L,
    weekF     = ((week - startWeek) %% nW_true) + 1L,
    p = y / N
  ) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Define the prospective ignition gates used by `$M_0$`.
- Tune the gate thresholds by leave-one-season-out evaluation.
- Inspect per-season detection behaviour and threshold tradeoffs.

The aligned historical data used here come from retrospective ignition labels and reference-curve preparation described in the [pipeline overview](pipeline_overview.html).
:::

```{r build-aligned}
#| message: false

# Retrospective ignition labels (manual) → aligned dataset for training
res  <- estimateDerivs(allD, k = 10)

outs <- res$data %>%
  group_by(season) %>%
  group_split(.keep = TRUE) %>%
  purrr::map(~ flagIgnition(
    df            = .x,
    p_thresh      = 0.01,
    k1            = 0.4,
    k_c           = 0.01,
    n_consec      = 2,
    min_window    = 10,
    w_min         = 21,
    w_max         = 21,
    d2_relax      = -0.01,
    manual_labels = manual_labels
  ))

alignedD <- alignIgnition(outs)
```

## $M_0$ — Prospective ignition detection

We pose ignition detection as a **prospective** decision problem. For each season $s$ and week $w$, the detector uses information available up to week $w$ to decide whether ignition has occurred, and returns the **earliest** ignition week.

## Ignition gates

We restrict ignition to an eligible within-season week window:

$$
g_{\mathrm{win}}(s,w)=\mathbf{1}\{w_{\min}\le w^f_{s,w}\le w_{\max}\}.
$$

We define four epidemiological gates used in voting. A fifth signal, the velocity gate (`cond_dp`), is computed and shown in the detection table for diagnostics but does not contribute to the vote.

::: panel-tabset

### Rolling-sum evidence gate

*Gate: `cond_sum` · Signal: `p_sumK` = $S_K(s,w)$ · Parameters: `K_sum` = $K_{\mathrm{sum}}$, `p_sum_thr` = $\tau_{\mathrm{sum}}$*

**Purpose.** A single elevated week could be a sampling artefact. Requiring the *sum* of positivity over several consecutive weeks to exceed a threshold filters out isolated spikes and demands that elevated signal be sustained across multiple observations before contributing a vote.

Let $K_{\mathrm{sum}}\ge 1$ and define a short rolling sum of weekly positivity:

$$
S_{K}(s,w)=\sum_{j=0}^{K_{\mathrm{sum}}-1} p_{s,w-j}.
$$

The rolling-sum gate is

$$
g_{\mathrm{sum}}(s,w)=
\mathbf{1}\left\{S_{K}(s,w)\ge \tau_{\mathrm{sum}}\right\}.
$$

### Smoothed positivity level gate

*Gate: `cond_p` · Signal: `p_sm` = $\tilde{p}_{s,w}$ · Parameters: `L` = $L$, `p_thr` = $\tau_p$*

**Purpose.** The raw weekly positivity $p_{s,w}$ is noisy. Smoothing over $L$ weeks suppresses week-to-week noise and confirms that the *current level* of positivity is meaningfully elevated rather than a transient fluctuation.

Let $\tilde p_{s,w}$ be a short rolling mean of weekly positivity with window length $L$:

$$
\tilde p_{s,w}=
\frac{1}{L}\sum_{j=0}^{L-1} p_{s,w-j},
\qquad L \in \{2,3\}.
$$

The level gate is

$$
g_{p}(s,w)=
\mathbf{1}\left\{\tilde p_{s,w}\ge \tau_{p}\right\}.
$$

### Cumulative prevalence gate

*Gate: `cond_prev` · Signal: `prev` = $\mathrm{Prv}_{s,w}$ · Parameter: `prev_thr` = $\tau_{\mathrm{prev}}$*

**Purpose.** A short burst of positive tests in a small sample can transiently elevate weekly positivity without representing a true population-level signal. The cumulative prevalence gate integrates evidence *across the entire season so far*, making it resistant to brief noise episodes: the season-wide test burden must reveal a meaningfully infected fraction of specimens.

Define cumulative prevalence evidence:

$$
\mathrm{Prv}_{s,w}=
\frac{\sum_{u \le w} y_{s,u}}{\sum_{u \le w} N_{s,u}}.
$$

The prevalence gate is

$$
g_{\mathrm{prev}}(s,w)=
\mathbf{1}\left\{\mathrm{Prv}_{s,w}\ge \tau_{\mathrm{prev}}\right\}.
$$

### Trend confirmation gate

*Gate: `cond_inc` · Signal: `dp` = $\Delta\tilde{p}_{s,w} = \tilde{p}_{s,w}-\tilde{p}_{s,w-1}$ · Parameters: `n_consec` = $n_{\mathrm{consec}}$, `eps` = $\epsilon$*

**Purpose.** Elevated positivity that is already declining is not ignition — it could be a receding false alarm. This gate requires that the smoothed signal has not been *sustainedly falling* over recent weeks, confirming that the elevated level is part of an ongoing rise rather than a peak already past. The noise tolerance $\epsilon$ allows one minor dip without disqualifying a genuine uptrend.

Let $k=n_{\mathrm{consec}}-1$ and $\epsilon \ge 0$ be a tolerance. We declare a sustained upward trend if at least $k-1$ of the last $k$ differences of `dp` exceed $-\epsilon$:

$$
g_{\mathrm{inc}}(s,w)=
\mathbf{1}\left\{
\sum_{j=0}^{k-1}\mathbf{1}\{\Delta \tilde p_{s,w-j} > -\epsilon\}\ \ge\ k-1
\right\}.
$$

Note that `cond_inc` checks *local direction* (week-to-week signs) but not magnitude. It passes as soon as positivity stops falling — even if it has been flat for weeks at a moderate level.

### Velocity gate (display only)

*Gate: `cond_dp` · Signal: `p_sm` lagged difference over `K_dp` weeks = $\tilde{p}_{s,w}-\tilde{p}_{s,w-K_{\mathrm{dp}}}$ · Parameters: `K_dp` = $K_{\mathrm{dp}}$, `dp_thr` = $\tau_{\mathrm{dp}}$*

**Note.** This gate is computed and included in the per-season detection table for diagnostic purposes, but it does **not** contribute to the $N_{\mathrm{req}}$-of-4 vote.

**Purpose.** The four voting gates can all be satisfied by a flat, moderately elevated signal. The velocity gate demands a *net displacement* of at least $\tau_{\mathrm{dp}}$ over the past $K_{\mathrm{dp}}$ weeks, distinguishing a genuinely rising epidemic from a flat signal sitting at a moderate level.

Let $K_{\mathrm{dp}}\ge 1$ be a look-back lag. The velocity gate is

$$
g_{\mathrm{dp}}(s,w)=
\mathbf{1}\left\{
\tilde p_{s,w}-\tilde p_{s,w-K_{\mathrm{dp}}}\ge \tau_{\mathrm{dp}}
\right\}.
$$

Unlike `cond_inc`, which checks whether each step is non-negative, `cond_dp` measures the *total rise* over the window.

:::

## Ignition rule: $N_{\mathrm{req}}$-of-4 voting

Define the vote count over the four epidemiological gates:

$$
H(s,w)=
g_{\mathrm{sum}}(s,w)+
g_{p}(s,w)+
g_{\mathrm{prev}}(s,w)+
g_{\mathrm{inc}}(s,w).
$$

Ignition is declared at the earliest week within the eligible window where all $N_{\mathrm{req}}$ gates agree:

$$
\hat w_s^{(\mathrm{ign})}=
\min\left\{
w:\
g_{\mathrm{win}}(s,w)=1,\
H(s,w)\ge N_{\mathrm{req}}
\right\},
$$

with $\hat w_s^{(\mathrm{ign})}=\mathrm{NA}$ if no week satisfies the rule. With $N_{\mathrm{req}}=4$, all four gates must agree simultaneously, ensuring that no single gate can trigger detection on its own.

---

## Leave One Season Out (LOSO) training of thresholds

To assess how well the ignition detector transfers to an unseen season, we use **leave-one-season-out (LOSO)** cross-validation. For each season $s$, we tune the rule thresholds using all seasons except $s$, then run the detector **prospectively** on the held-out season to obtain $\hat w_s^{(\mathrm{ign})}$. We summarize performance by aggregating the per-season ignition timing errors across folds, and select hyperparameters that minimize the chosen LOSO objective. This procedure reduces overfitting to idiosyncratic seasons and better reflects real deployment, where ignition must be detected for a new season without access to its future trajectory.

Let $w_s^{(\mathrm{ign})}$ be the ignition week for season $s$, $\theta$ denote the full set of thresholds and rule hyperparameters.

$$
\theta = (\tau_{\mathrm{sum}},K_{\mathrm{sum}},\tau_p,\tau_{\mathrm{prev}}, n_{\mathrm{consec}}, L, \epsilon, N_{\mathrm{req}}, w_{\min}, w_{\max})
$$

Detection error for season $s$ under parameter set $\theta$ is

$$
\mathrm{diff}_s(\theta)=\hat w_s^{(\mathrm{ign})}(\theta)-w_s^{(\mathrm{ign})}.
$$

If no week fires within $[w_{\min}, w_{\max}]$, the detector assigns $\hat w_s^{(\mathrm{ign})} = w_{\max}$ as a fallback, so misses never occur.

Detection one week early is treated as exact via an adjusted error with a $-1$ wiggle room:

$$
\mathrm{adj\_diff}_s = \max(\mathrm{diff}_s,\,0)+\max(-1-\mathrm{diff}_s,\,0),
$$

so $\mathrm{adj\_diff}=0$ when $\mathrm{diff}\in\{-1,0\}$, and errors grow symmetrically beyond that window.

Per-season loss:

$$
\ell(\mathrm{adj\_diff}) = \mathrm{adj\_diff}
+ \gamma\max(\mathrm{adj\_diff}-2,\,0),
\qquad \gamma \ge 0,
$$

where the $\gamma$ term penalizes adjusted errors exceeding 2 weeks. Late and missing detections are not penalized separately.

A global tuning objective is

$$
\mathrm{Score}(\theta)=
\sum_s \ell(\mathrm{adj\_diff}_s(\theta))
+\lambda\max_s \mathrm{adj\_diff}_s(\theta).
$$

**Parameter selection.** Among all grid candidates, the best $\hat\theta$ is selected by the following lexicographic criterion (in priority order):

1. Minimize $n_{>2}(\theta)$: number of seasons with $\mathrm{adj\_diff}>2$.
2. Minimize $\max_s\,\mathrm{adj\_diff}_s(\theta)$.
3. Minimize $\mathrm{Score}(\theta)$ (final tiebreaker).

The tuned $\hat\theta$ is then applied prospectively in real time to determine ignition timing and trigger downstream forecasting.

```{r loso-tuning}
if (file.exists("../data/stage1_tuning.rds")) {
  tuned <- readRDS("../data/stage1_tuning.rds")
} else {
  grid_loso <- data.table::CJ(
    cls_thr   = 0.26,
    use_cls   = FALSE,
    p_thr     = c(0.002, 0.003, 0.004, 0.005),
    prev_thr  = c(0.001, 0.002, 0.003),
    n_consec  = 5L,
    L         = 2L,
    eps       = 0,
    K_sum     = 5L,
    p_sum_thr = c(0.050, 0.055, 0.060),
    N_req     = 4L,
    w_min     = 13L,
    w_max     = 26L,
    K_dp      = 3L,
    dp_thr    = 0.01,
    sorted    = FALSE
  )

  tuned <- loso_M0v2(
    dat  = alignedD,
    grid = as.data.frame(grid_loso),
    score_col = "p_cls_p",
    drop_seasons = c("2015-16"),
    exSeason_tune = NULL,
    fit_args = list(
      fit_base = TRUE, fit_slope = FALSE, fit_fs = FALSE,
      event_k = 1L, lead = 1L, A_pre = 6L, B_post = 6L,
      k_week = 6L, k_p = 8L, k_fs = 4L,
      select = FALSE, verbose = FALSE
    ),
    tune_args = list(
      miss_penalty = 0, lambda = 20, kappa = 0,
      gamma = 25, gamma_late = 0,
      iWeek = TRUE, ncores = 10L,
      verbose = FALSE, progress_every = 200L
    ),
    verbose = TRUE
  )

  saveRDS(tuned, "../data/stage1_tuning.rds")
}

tuned$compare %>% gt()
```

::: panel-tabset

### LOSO results

```{r loso-results}
tuned$runtime

keep_params <- c("p_thr", "prev_thr", "p_sum_thr", "eps", "dp_thr",
                 "n_consec", "L", "K_sum", "N_req", "w_min", "w_max", "K_dp")
bp_display  <- tuned$best_params[names(tuned$best_params) %in% keep_params]

data.frame(Parameter = names(unlist(bp_display)),
           Value     = unlist(bp_display), row.names = NULL) |>
  gt() |>
  tab_header(title = "Best M_0 parameters (LOSO tuned)")

det_all <- detect_ignition_from_tuning(tuned, alignedD)
det_all$compare |>
  gt() |>
  tab_header(title = "Ignition detection: predicted vs truth") |>
  cols_label(iWeek_hat = "Detected week", iWeek_true = "True week", diff = "Diff") |>
  tab_style(style = cell_fill(color = "#fff3cd"),
            locations = cells_body(rows = abs(diff) > 2)) |>
  tab_footnote(
    footnote = "True week is the retrospective manual label. It is expert-defined, not a formally validated ground truth.",
    locations = cells_column_labels(columns = iWeek_true)
  )
```

```{r season-tabs}
#| echo: false
seasons_in_data <- sort(unique(det_all$data$season))
season_tables <- lapply(
  seasons_in_data,
  function(s) {
    htmltools::tagList(
      htmltools::tags$h3(s),
      plot_season_detection_table(det_all, s)
    )
  }
)
do.call(htmltools::tagList, season_tables)
```

:::

## Function call graph

Functions used in this document, showing how `fitIgnition()` and the M0
detection chain relate to each other.

```{dot}
//| fig-width: 10
//| fig-height: 4
digraph m0_calls {
  graph [rankdir=LR, fontname="Helvetica", fontsize=11,
         nodesep=0.4, ranksep=0.9];
  node  [fontname="Helvetica", fontsize=10, style=filled,
         shape=box, margin="0.12,0.05"];
  edge  [fontname="Helvetica", fontsize=9, arrowsize=0.6];

  subgraph cluster_m0train {
    label="M0: Training (m0_training.R)"; style=filled;
    color="#856404"; fillcolor="#fff3cd"; fontcolor="#856404";
    fontsize=11; fontname="Helvetica-Bold";
    fitIgnition      [label="fitIgnition"];
    scoreIgnition    [label="scoreIgnition"];
    loso_M0v2        [label="loso_M0v2"];
  }
  subgraph cluster_m0flag {
    label="M0: Detection (flagIgnition.R, m0_runtime.R)"; style=filled;
    color="#6d4c00"; fillcolor="#ffe8a1"; fontcolor="#6d4c00";
    fontsize=11; fontname="Helvetica-Bold";
    flag_ignition_rules     [label="flag_ignition_rules"];
    detect_ignition_from_tuning [label="detect_ignition\n_from_tuning"];
    run_ignition_weekly     [label="run_ignition_weekly"];
    run_m0_detection        [label="run_m0_detection"];
  }
  subgraph cluster_m0retro {
    label="M0: Retrospective (m0_retro.R)"; style=filled;
    color="#5a3e00"; fillcolor="#ffd970"; fontcolor="#5a3e00";
    fontsize=11; fontname="Helvetica-Bold";
    m0_retro_run   [label="m0_retro_run"];
    m0_retro_score [label="m0_retro_score"];
  }

  fitIgnition             -> scoreIgnition;
  fitIgnition             -> loso_M0v2;
  detect_ignition_from_tuning -> flag_ignition_rules;
  run_ignition_weekly     -> detect_ignition_from_tuning;
  run_m0_detection        -> flag_ignition_rules;
  m0_retro_run            -> fitIgnition;
  m0_retro_run            -> m0_retro_score;
}
```
