---
title: "LOSO Walk-forward Alignment Evaluation"
format:
  html:
    self-contained: true
    toc: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(gamm4)
library(gratia)
library(MMWRweek)
library(plotly)
library(gt)
library(flualign)
library(future)
library(furrr)

source("../R/m1_reference.R")
source("../R/m1_loso.R")
source("../R/utils.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m1_runtime.R")
source("../R/m1_multi_template.R")
source("../R/m1_fit.R")
startWeek <- 27

tuned  <- readRDS("../data/stage1_tuning.rds")
params <- tuned$best_params

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(season, week, year, start_year = seasonstart, date = week_start_date,
         y = pos_flua, N = test_flu) %>%
  mutate(neg = N - y, date = as.Date(date),
         mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
         nW_true   = n_weeks_in_start_year(start_year),
         weekF     = ((week - startWeek) %% nW_true) + 1L,
         p = y / N) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

manual_labels_orig <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)
manual_labels <- manual_labels_orig - 1L

EXCLUDE_SEAS <- "2015-16"

# Helper: convert newWeek -> flu week (epidemiological week)
# weekF = newWeek - anchorWeek + iWeek_hat
# flu_week = ((weekF - 1 + startWeek) %% nW) + 1
# For display we use nW=52 as default
newWeek_to_fluWeek <- function(nw, anchor, iWeek_hat, nW = 52L) {
  weekF <- nw - anchor + iWeek_hat
  ((weekF - 1L + startWeek) %% nW) + 1L
}
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Covers

- Evaluate walk-forward `$M_1$` alignment under leave-one-season-out validation.
- Compare the tuned alignment settings on peak-week accuracy.
- Inspect parameter trajectories and season-specific behaviour.
:::

## Reference curve and alignment

All alignment uses the **FS median reference curve**: `estimateRef(method = "fs", agg = "median")`. This fits a factor-smooth GAM with per-season random curves, then takes the pointwise median across training seasons on the logit scale. The resulting population reference is robust to outlier seasons but retains week-level shape features — including jagged transitions where the median switches between seasons.

### Week numbering

Three week indices appear in this analysis:

1. **Flu week** (epidemiological week): the CDC MMWR week. A flu season starts around week 35 (late August).
2. **weekF** (season-indexed): `((flu_week - 27) %% nW) + 1`. Week 27 maps to weekF=1. This is the internal season index used for modeling.
3. **newWeek** (alignment-indexed): after alignment, each season is anchored so that ignition falls at a common `anchorWeek` (median ignition across training seasons, typically ~19). The alignment model operates in newWeek space.

All plots below use **newWeek** on the x-axis. A secondary flu-week annotation line at y=0.05 maps newWeek back to epidemiological weeks for clinical context.

## Parameter tuning via LOSO grid search

The alignment hyperparameters were optimized using leave-one-season-out (LOSO) cross-validation across three sequential grid searches (v5–v7), totalling **67 configurations**. All searches fix `multi_temperature = 0.25` and `align_rise_weight = 1.0` (established in an earlier 153-spec sweep). The headline metric is Weibull-weighted peak MAE: $w(t) = \exp(-(0.1\,t)^2)$, $t$ = weeks since ignition.

| Search | Varied parameters | Specs | Baseline MAE |
|--------|------------------|-------|-------------|
| v5 | `slope_weight` ∈ {0, 0.2, 0.5, 1, 2}, `slope_window` ∈ {4, 6, 8} | 15 | 1.929 |
| v6 | `slope_weight` ∈ {2, 3, 5, 8}, `slope_window` ∈ {6, 8, 10, 12}, `template_shift` ∈ {0, 1} | 32 | 1.629 → 1.275 |
| v7 | `slope_weight` ∈ {8, 12, 16, 20, 30}, `k_ref` ∈ {25, 30, 40, 50} | 20 | confirmed 1.275 |

`slope_weight` controls the slope-similarity softmax bonus: $\exp(-\text{slope\_weight} \times |\text{obs\_slope} - \text{tmpl\_slope}|)$. Higher values concentrate ensemble weight on templates whose current growth rate matches the observed season.

```{r tuning-results}
#| echo: false
#| message: false
#| warning: false

combined <- readRDS("../data/m1_alignment_tuning_combined.rds")

top_results <- combined %>%
  arrange(mae_weibull) %>%
  head(20) %>%
  select(version, k_ref, slope_window, slope_weight, template_shift, mae_weibull) %>%
  mutate(mae_weibull = round(mae_weibull, 4))

gt(top_results) %>%
  cols_label(
    version       = "Grid",
    k_ref         = "k_ref",
    slope_window  = "Slope Window",
    slope_weight  = "Slope Weight",
    template_shift = "Shift",
    mae_weibull   = "MAE (Weibull) ★"
  ) %>%
  cols_align(align = "center") %>%
  tab_style(
    style = cell_fill(color = "#ffffcc", alpha = 0.8),
    locations = cells_body(rows = 1)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = mae_weibull, rows = 1)
  ) %>%
  tab_options(table.font.size = "13px", table.width = pct(80))
```

**Locked specification**: k_ref=25, slope_window=6, slope_weight=8, template_shift=0, temperature=0.25 → **Peak MAE = 1.275 weeks**

Both grid boundaries are interior: `slope_weight` peaks at 8 then degrades monotonically through 30; `k_ref=25` beats all larger values tested up to 50.

### Tuning insights

1. **slope_weight=8 optimal** — slope similarity improves alignment substantially. The full range traced from 0 (no slope bonus, MAE 1.929) through 8 (optimal, MAE 1.275) then degrades to 30 (MAE 1.516), confirming an interior optimum.

2. **slope_window=6 optimal** — 6-week slope window outperforms 4, 8, 10, 12 at all tested slope weights. Narrower windows are noisy; wider windows smooth away the current-week signal.

3. **template_shift=0** is optimal — no systematic lag correction needed. shift=1 is competitive only at low slope weights but falls behind once slope matching is active.

4. **k_ref=25** beats larger values (30, 40, 50) — higher basis dimensions overfit the reference GAM and degrade generalisation.

5. **Weibull-weighted MAE** is the headline metric — upweights early prospective weeks when uncertainty is highest and clinical decisions are most time-sensitive.

::: {.callout-note}
**Median scoring.** From this point forward, `tune_m1_alignment()` also reports `mae_med_weibull` (Weibull-weighted MAE using the weighted-median peak rather than the weighted-mean peak). The v5–v7 checkpoints pre-date this addition; both metrics will be reported for v8+ grid searches.
:::

---

## Multi-template alignment

**Multi-template alignment** aligns to each training season's individual curve separately, then ensembles the results:

1. `estimateRef(method="fs")` returns `eta_mat`: a 52 $\times$ $n_\text{train}$ matrix of per-season logit-scale curves from the factor-smooth GAM. Each column is a smooth GAM prediction for one training season.
2. For each template (training season), build a `splinefun()` reference curve and run `align_forecast_pipeline_dilate()` to estimate ($\tau$, $\delta$, $a$, $b$) and produce a forecast.
3. Compute per-template NLL (negative log-likelihood normalized per observation). Templates that better match the observed data get lower NLL.
4. Ensemble via **softmax weights** combining NLL and slope similarity:
$$w_s \propto \exp\!\bigl(-\mathrm{NLL}_s / T\bigr) \cdot \exp\!\bigl(-\lambda_s \,|\hat{\sigma}_{\mathrm{obs}} - \hat{\sigma}_s|\bigr)$$
where $T=0.25$ and $\lambda_s = 8$ (slope weight). Lower temperature concentrates weight on the best-fitting template; slope similarity rewards templates whose current growth rate matches the observed season.
5. **Logit-scale ensemble**: predictions are averaged in log-odds space, then back-transformed. This respects binomial geometry and prevents probability-scale averaging distortion near 0 or 1.
6. **Dual peak estimates**: weighted mean and weighted median of per-template peak positions, both in the logit-aligned time axis.
7. **Alignment uncertainty** (`logit_spread`): weighted SD of per-template logit-scale predictions at each future week — propagated as a covariate to $M_2$.

This allows the model to "recognize" which historical season the current one most resembles and weight forecasts accordingly — without committing to a single reference shape.

```{r run-multi-template}
#| message: false
#| warning: false

MULTI_CACHE <- "../data/align_multi_cache.rds"
CKPT_FILE   <- "../data/align_ckpt/ckpt_multi.rds"
if (!dir.exists(dirname(CKPT_FILE))) dir.create(dirname(CKPT_FILE), recursive = TRUE)

if (file.exists(MULTI_CACHE)) {
  wf <- readRDS(MULTI_CACHE)
  message("Loaded multi-template result from cache: ", MULTI_CACHE)
} else {
  # LOCKED PARAMETERS from LOSO grid search (67 configs across v5-v7)
  # Peak MAE = 1.275 weeks (Weibull-weighted)
  wf <- loso_walkforward(
    allD              = allD,
    params            = params,
    manual_labels     = manual_labels,
    exclude_seasons   = EXCLUDE_SEAS,
    k_deriv           = 20L,
    k_ref             = 25L,           # ← TUNED: interior optimum (tested 15–50)
    n_weeks           = 52L,
    buffer_weeks      = 5L,
    curvature_ratio   = 1.0,
    template_shift    = 0L,            # ← TUNED: no offset needed
    slope_window      = 6L,            # ← TUNED: 6-week slope window optimal
    slope_weight      = 8.0,           # ← TUNED: interior optimum (tested 0–30)
    peak_weight_boost = 3,
    peak_weight_decay = 0.3,
    align_rise_weight = 1.0,           # ← TUNED: uniform loss weights optimal
    use_multi_template = TRUE,
    ref_method        = "fs",
    multi_temperature = 0.25,          # ← TUNED: concentration on best template
    n_cores           = parallel::detectCores() - 1L,
    checkpoint_file   = CKPT_FILE,
    verbose           = TRUE
  )
  saveRDS(wf, MULTI_CACHE)
  message("Done. Saved to ", MULTI_CACHE)
}
```

## Tuning summary

$M_1$ alignment hyperparameters were optimized over 67 configurations across three sequential LOSO grid searches (v5–v7). The locked specification is:

| Parameter | Value | Tuned? |
|-----------|-------|--------|
| `k_ref` | 25 | ✓ interior (tested 15–50) |
| `multi_temperature` | 0.25 | ✓ (fixed from earlier sweep) |
| `template_shift` | 0 | ✓ interior |
| `slope_window` | 6 | ✓ optimal |
| `slope_weight` | 8 | ✓ interior optimum |
| `align_rise_weight` | 1.0 | ✓ (fixed from earlier sweep) |

**Best Weibull-weighted peak MAE (LOSO) = 1.275 weeks** (improvement from 1.929 baseline via slope-similarity weighting).

## Alignment parameters over time

One trace per season. Each point = one prospective evaluation at that `eval_week`.

```{r param-plots-setup}
#| echo: false
#| warning: false

param_vars <- c("tau", "delta", "a", "b", "t_peak")
param_labels <- c(
  tau    = "tau - time shift",
  delta  = "delta - scale",
  a      = "a - lower asymptote",
  b      = "b - upper asymptote",
  t_peak = "t_peak - predicted peak (newWeek)"
)

df_params <- wf$params_df %>%
  filter(!is.na(iWeek_hat)) %>%
  select(season, eval_week, all_of(param_vars))

seasons_pal <- setNames(
  RColorBrewer::brewer.pal(min(11, length(unique(df_params$season))), "Set3"),
  sort(unique(df_params$season))
)

make_param_plot <- function(pvar) {
  df <- df_params %>% select(season, eval_week, value = all_of(pvar))
  plot_ly(df, x = ~eval_week, y = ~value, color = ~season,
          colors = seasons_pal,
          type = "scatter", mode = "lines+markers",
          marker = list(size = 6), line = list(width = 1),
          legendgroup = ~season, showlegend = TRUE) %>%
    layout(
      xaxis  = list(title = "Eval week (weekF)"),
      yaxis  = list(title = param_labels[[pvar]]),
      legend = list(title = list(text = "Season"))
    )
}
```

::: {.panel-tabset}

```{r param-tabs}
#| echo: false
#| warning: false
#| results: asis

param_tab_chunks <- lapply(param_vars, function(pvar) {
  e <- new.env(parent = globalenv())
  e$p <- make_param_plot(pvar)
  knitr::knit_child(
    text = c(sprintf("### %s", param_labels[[pvar]]), "",
             "```{r echo=FALSE, warning=FALSE}", "p", "```", ""),
    envir = e, quiet = TRUE
  )
})
cat(unlist(param_tab_chunks), sep = "\n")
```

:::

## Results by season

Per-season walkforward alignment using the multi-template ensemble. Grey = FS median reference curve. Blue = fitted past, red = forecast. Dashed grey = ignition, solid black = current newWeek, dashed red = estimated peak, solid blue = observed peak. Orange markers at y=0.05 show corresponding flu weeks.

```{r season-setup}
#| echo: false
#| warning: false

all_seasons_wf <- sort(unique(wf$params_df$season))

true_peaks_obs <- allD %>%
  mutate(p_obs = y / N) %>%
  filter(!is.na(p_obs), is.finite(p_obs), N > 0) %>%
  group_by(season) %>%
  slice_max(p_obs, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(season, true_peak_weekF = weekF)

make_season_plot <- function(s, ew, s_params, ref_curve, obs_s,
                              true_peak_weekF, nW_s) {
  pw           <- wf$params_df %>% filter(season == s, eval_week == ew)
  iWeek_hat_ew <- pw$iWeek_hat[[1]]
  anchor_ew    <- pw$anchorWeek[[1]]
  t_peak_ew    <- pw$t_peak[[1]]
  peak_nW      <- round(t_peak_ew)
  current_newW <- ew - iWeek_hat_ew + anchor_ew
  true_peak_newW <- if (!is.na(true_peak_weekF))
    true_peak_weekF - iWeek_hat_ew + anchor_ew else NA_real_

  fc_smooth <- wf$forecast_df %>%
    filter(season == s, eval_week == ew, kind == "forecast") %>%
    arrange(newWeek)
  if (nrow(fc_smooth) == 0) return(NULL)
  junction  <- fc_smooth$newWeek[which.min(abs(fc_smooth$newWeek - current_newW))]
  fc_past   <- fc_smooth %>% filter(newWeek <= junction)
  fc_future <- fc_smooth %>% filter(newWeek >= junction)

  obs_aligned <- obs_s %>%
    filter(weekF <= ew) %>%
    mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

  y_max <- max(c(fc_smooth$p_hi, obs_aligned$p_obs), na.rm = TRUE) * 1.1
  y_max <- max(y_max, 0.15)

  nw_range <- range(c(fc_smooth$newWeek, obs_aligned$newWeek), na.rm = TRUE)
  nw_ticks <- seq(max(1L, floor(nw_range[1])), min(52L, ceiling(nw_range[2])), by = 2L)
  flu_ticks <- tibble(
    newWeek  = as.numeric(nw_ticks),
    flu_week = newWeek_to_fluWeek(nw_ticks, anchor_ew, iWeek_hat_ew, nW_s),
    y        = 0.05
  )
  current_fluWk <- newWeek_to_fluWeek(current_newW, anchor_ew, iWeek_hat_ew, nW_s)

  plt <- plot_ly() %>%
    add_lines(data = ref_curve, x = ~newWeek, y = ~p_ref,
              line = list(color = "rgba(160,160,160,0.5)", width = 1.2),
              name = "FS ref") %>%
    add_ribbons(data = fc_smooth, x = ~newWeek, ymin = ~p_lo, ymax = ~p_hi,
                fillcolor = "rgba(150,150,150,0.15)",
                line = list(color = "transparent"), name = "95% CI") %>%
    add_lines(data = fc_past, x = ~newWeek, y = ~p_hat,
              line = list(color = "steelblue", width = 2.5), name = "Fitted") %>%
    add_lines(data = fc_future, x = ~newWeek, y = ~p_hat,
              line = list(color = "firebrick", width = 2.5), name = "Forecast") %>%
    add_markers(data = obs_aligned, x = ~newWeek, y = ~p_obs,
                marker = list(color = "black", size = 5, opacity = 0.75),
                name = "Observed") %>%
    add_text(data = flu_ticks, x = ~newWeek, y = ~y,
             text = ~as.character(flu_week),
             textfont = list(size = 9, color = "darkorange"),
             name = "Flu wk", showlegend = FALSE) %>%
    add_segments(x = anchor_ew, xend = anchor_ew, y = 0, yend = y_max,
                 line = list(color = "grey50", dash = "dash", width = 1.5),
                 name = paste0("Ignition (nW", anchor_ew, ")")) %>%
    add_segments(x = current_newW, xend = current_newW, y = 0, yend = y_max,
                 line = list(color = "black", dash = "solid", width = 2),
                 name = paste0("Now (nW", current_newW, ")")) %>%
    add_segments(x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
                 line = list(color = "firebrick", dash = "dash", width = 1.5),
                 name = paste0("Est. peak (nW", peak_nW, ")"))

  if (!is.na(true_peak_newW)) {
    plt <- plt %>%
      add_segments(x = true_peak_newW, xend = true_peak_newW, y = 0, yend = y_max,
                   line = list(color = "steelblue", dash = "solid", width = 1.5),
                   name = paste0("Obs. peak (nW", round(true_peak_newW), ")"))
  }

  plt %>% layout(
    title = paste0(s, " | nW ", current_newW, " (flu wk ", current_fluWk,
                   ") | est peak nW ", peak_nW),
    xaxis = list(title = "newWeek (aligned)"),
    yaxis = list(title = "Positivity", range = c(0, y_max)),
    legend = list(orientation = "h", y = -0.20)
  )
}
```

::: {.panel-tabset}

```{r season-tabs}
#| echo: false
#| warning: false
#| results: asis

season_chunks <- lapply(all_seasons_wf, function(s) {
  s_all <- wf$params_df %>% filter(season == s, !is.na(iWeek_hat))
  if (nrow(s_all) == 0) {
    return(knitr::knit_child(
      text = c(sprintf("## %s", s), "", "*No forecasts — ignition not detected.*", ""),
      quiet = TRUE
    ))
  }

  if (!"peak_passed" %in% names(s_all))
    s_all <- s_all %>% mutate(peak_weekF = round(t_peak - anchorWeek + iWeek_hat),
                               peak_passed = FALSE)

  first_peak_ew <- s_all %>% filter(peak_passed) %>%
    slice_min(eval_week, n = 1, with_ties = FALSE) %>% pull(eval_week)
  cutoff_ew <- if (length(first_peak_ew) > 0) first_peak_ew + 2L else max(s_all$eval_week)
  s_params  <- s_all %>% filter(eval_week <= cutoff_ew)

  # Build data table
  s_disp <- s_params %>%
    mutate(current_nW = eval_week - iWeek_hat + anchorWeek,
           peak_nW    = round(t_peak)) %>%
    select(current_nW, peak_nW, tau, delta, t_peak, t_peak_lo, t_peak_hi, peak_passed)
  tbl <- gt(s_disp) %>%
    cols_label(current_nW = "nW", peak_nW = "Est. peak nW",
               tau = "τ", delta = "δ", t_peak = "t_peak",
               t_peak_lo = "lo", t_peak_hi = "hi", peak_passed = "Peak?") %>%
    fmt_number(columns = c(tau, delta, t_peak, t_peak_lo, t_peak_hi), decimals = 2) %>%
    fmt_integer(columns = c(current_nW, peak_nW)) %>%
    tab_style(style = cell_fill(color = "#fff0f0"),
              locations = cells_body(rows = peak_passed == TRUE)) %>%
    tab_options(table.width = pct(100), table.font.size = "13px")

  obs_s <- allD %>% filter(season == s) %>% mutate(p_obs = y / N)
  true_peak_weekF_s <- true_peaks_obs %>% filter(season == s) %>%
    pull(true_peak_weekF)
  true_peak_weekF_s <- if (length(true_peak_weekF_s) > 0) true_peak_weekF_s[[1]] else NA_integer_
  nW_s   <- if (!is.null(obs_s$nW_true[1]) && !is.na(obs_s$nW_true[1])) obs_s$nW_true[1] else 52L
  ref_s  <- wf$ref_list[[s]]
  ref_curve_s <- tibble(newWeek = seq(1, 52, 0.5),
                         p_ref   = plogis(ref_s$g_ref_fun(seq(1, 52, 0.5))))

  # Per-eval-week sub-tabs
  week_chunks <- lapply(s_params$eval_week, function(ew) {
    e <- new.env(parent = globalenv())
    e$plt <- make_season_plot(s, ew, s_params, ref_curve_s, obs_s,
                               true_peak_weekF_s, nW_s)
    if (is.null(e$plt)) return(NULL)
    anchor_ew <- wf$params_df %>% filter(season == s, eval_week == ew) %>%
      pull(anchorWeek) %>% `[[`(1)
    iWeek_hat_ew <- wf$params_df %>% filter(season == s, eval_week == ew) %>%
      pull(iWeek_hat) %>% `[[`(1)
    current_nW <- ew - iWeek_hat_ew + anchor_ew
    knitr::knit_child(
      text = c(sprintf("#### nW %d (wkF %d)", as.integer(current_nW), ew), "",
               "```{r echo=FALSE, warning=FALSE}", "plt", "```", ""),
      envir = e, quiet = TRUE
    )
  })
  week_chunks <- Filter(Negate(is.null), week_chunks)

  # Data tab
  e2 <- new.env(parent = globalenv())
  e2$tbl_obj <- tbl
  data_chunk <- knitr::knit_child(
    text = c("#### Data", "",
             "```{r echo=FALSE}", "gt::as_raw_html(tbl_obj) |> htmltools::HTML() |> print()", "```", ""),
    envir = e2, quiet = TRUE
  )

  knitr::knit_child(
    text = c(sprintf("## %s", s), "",
             "::: {.panel-tabset}", "",
             unlist(week_chunks), "",
             data_chunk,
             ":::", ""),
    quiet = TRUE
  )
})

cat(unlist(season_chunks), sep = "\n")
```

:::
