---
title: "LOSO Walk-forward Alignment Evaluation"
format:
  html:
    self-contained: true
    toc: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(gamm4)
library(gratia)
library(MMWRweek)
library(plotly)
library(gt)
library(flualign)
library(future)
library(furrr)

source("../R/estimateRef.R")
source("../R/loso_alignment.R")
source("../R/utils.R")
source("../R/prospective_running.R")
source("../R/prospective_alignment.R")
source("../R/align_multi_template.R")
source("../R/fit.R")
startWeek <- 27

tuned  <- readRDS("../data/stage1_tuning.rds")
params <- tuned$best_params

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(season, week, year, start_year = seasonstart, date = week_start_date,
         y = pos_flua, N = test_flu) %>%
  mutate(neg = N - y, date = as.Date(date),
         mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
         nW_true   = n_weeks_in_start_year(start_year),
         weekF     = ((week - startWeek) %% nW_true) + 1L,
         p = y / N) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

manual_labels_orig <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)
manual_labels <- manual_labels_orig - 1L

EXCLUDE_SEAS <- "2015-16"

# Helper: convert newWeek -> flu week (epidemiological week)
# weekF = newWeek - anchorWeek + iWeek_hat
# flu_week = ((weekF - 1 + startWeek) %% nW) + 1
# For display we use nW=52 as default
newWeek_to_fluWeek <- function(nw, anchor, iWeek_hat, nW = 52L) {
  weekF <- nw - anchor + iWeek_hat
  ((weekF - 1L + startWeek) %% nW) + 1L
}
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

## Reference curve and alignment

All alignment uses the **FS median reference curve**: `estimateRef(method = "fs", agg = "median")`. This fits a factor-smooth GAM with per-season random curves, then takes the pointwise median across training seasons on the logit scale. The resulting population reference is robust to outlier seasons but retains week-level shape features — including jagged transitions where the median switches between seasons.

### Week numbering

Three week indices appear in this analysis:

1. **Flu week** (epidemiological week): the CDC MMWR week. A flu season starts around week 35 (late August).
2. **weekF** (season-indexed): `((flu_week - 27) %% nW) + 1`. Week 27 maps to weekF=1. This is the internal season index used for modeling.
3. **newWeek** (alignment-indexed): after alignment, each season is anchored so that ignition falls at a common `anchorWeek` (median ignition across training seasons, typically ~19). The alignment model operates in newWeek space.

All plots below use **newWeek** on the x-axis. A secondary flu-week annotation line at y=0.05 maps newWeek back to epidemiological weeks for clinical context.

## Parameter Tuning via LOSO Grid Search

The alignment hyperparameters were optimized using leave-one-season-out (LOSO) cross-validation over an extended grid search:

- **`k_ref`**: 15, 20, 25, 30, 35 (basis dimension for FS reference GAM)
- **`multi_temperature`**: 0.05, 0.10, 0.15, 0.20, 0.25, 0.5, 1.0, 2.0 (softmax temperature; lower = peakier template weights)
- **`template_shift`**: -2, -1, 0, +1 (offset alignment templates by weeks)
- **`align_rise_weight`**: 1.0 (uniform time weighting — no emphasis on rising limb)

**153 configurations** evaluated on Weibull-weighted peak MAE.

```{r tuning-results}
#| echo: false
#| message: false
#| warning: false

tune_m1 <- readRDS("../data/m1_alignment_tuning_v3.rds")

# Top 15 by mae_weibull
top_results <- tune_m1$scores %>%
  arrange(mae_weibull) %>%
  head(15) %>%
  select(k_ref, multi_temperature, template_shift,
         mae_uniform, mae_exp, mae_weibull) %>%
  mutate(across(mae_uniform:mae_weibull, ~round(., 4)))

gt(top_results) %>%
  cols_label(
    k_ref = "k_ref",
    multi_temperature = "Temperature",
    template_shift = "Template Shift",
    mae_uniform = "MAE (uniform weight)",
    mae_exp = "MAE (exp decay)",
    mae_weibull = "MAE (Weibull) ★"
  ) %>%
  cols_align(align = "center") %>%
  tab_style(
    style = cell_fill(color = "#ffffcc", alpha = 0.8),
    locations = cells_body(rows = 1)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = mae_weibull, rows = 1)
  ) %>%
  tab_options(table.font.size = "13px", table.width = pct(80))
```

**Best configuration (★)**: k_ref=25, temperature=0.25, template_shift=0, **Peak MAE = 1.169 weeks**

### Tuning insights

1. **Temperature=0.25 optimal** across the full range tested (0.05–2.0). Lower temperature concentrates softmax weight on the best-matching template, but going too low (≤0.20) overcommits to a single template and degrades performance (MAE 1.47–3.02). The sweet spot at 0.25 balances commitment to the best analog with hedging across templates.

2. **Template_shift=0** is optimal — no systematic detection lag correction needed. The previous shift=−1 finding was an artifact of the smaller initial grid.

3. **k_ref=25** (mid-range basis dimension) balances flexibility and smoothness of the population reference curve.

4. **align_rise_weight=1.0** (uniform loss weights) is optimal — no time-based emphasis needed. The multi-template method adapts locally without requiring pre-ignition down-weighting.

5. **Weibull-weighted MAE** is the headline metric — it accounts for peak uncertainty near the season peak and is most relevant for decision-making (e.g., hospital resource planning).

---

## Multi-template alignment

**Multi-template alignment** aligns to each training season's individual curve separately, then ensembles the results:

1. `estimateRef(method="fs")` returns `eta_mat`: a 52 $\times$ $n_\text{train}$ matrix of per-season logit-scale curves from the factor-smooth GAM. Each column is a smooth GAM prediction for one training season.
2. For each template (training season), build a `splinefun()` reference curve and run `align_forecast_pipeline_dilate()` to estimate ($\tau$, $\delta$, $a$, $b$) and produce a forecast.
3. Compute per-template NLL (negative log-likelihood normalized per observation). Templates that better match the observed data get lower NLL.
4. Ensemble via **softmax weights**: $w_s = \exp(-\text{NLL}_s / T) \big/ \sum_s \exp(-\text{NLL}_s / T)$ where $T$ is the temperature parameter (default 1.0). Lower temperature concentrates weight on the best-fitting template.
5. Final forecast: weighted average of per-template predictions. Peak estimate: weighted average of per-template peaks.

This allows the model to "recognize" which historical season the current one most resembles, and weight forecasts accordingly — without committing to a single reference shape.

```{r run-multi-template}
#| message: false
#| warning: false

MULTI_CACHE <- "../data/align_multi_cache.rds"
CKPT_FILE   <- "../data/align_ckpt/ckpt_multi.rds"
if (!dir.exists(dirname(CKPT_FILE))) dir.create(dirname(CKPT_FILE), recursive = TRUE)

if (file.exists(MULTI_CACHE)) {
  wf <- readRDS(MULTI_CACHE)
  message("Loaded multi-template result from cache: ", MULTI_CACHE)
} else {
  # BEST TUNED PARAMETERS from LOSO grid search (153 configs evaluated)
  # Peak MAE = 1.169 weeks (Weibull-weighted)
  wf <- loso_walkforward(
    allD              = allD,
    params            = params,
    manual_labels     = manual_labels,
    exclude_seasons   = EXCLUDE_SEAS,
    k_deriv           = 20L,
    k_ref             = 25L,           # ← TUNED: mid-range basis dimension
    n_weeks           = 52L,
    buffer_weeks      = 5L,
    curvature_ratio   = 1.0,
    template_shift    = 0L,            # ← TUNED: no offset needed
    peak_weight_boost = 3,
    peak_weight_decay = 0.3,
    align_rise_weight = 1.0,           # ← TUNED: uniform loss weights optimal
    use_multi_template = TRUE,
    ref_method        = "fs",
    multi_temperature = 0.25,          # ← TUNED: concentration on best template
    n_cores           = parallel::detectCores() - 1L,
    checkpoint_file   = CKPT_FILE,
    verbose           = TRUE
  )
  saveRDS(wf, MULTI_CACHE)
  message("Done. Saved to ", MULTI_CACHE)
}
```

## Tuning summary

The M1 alignment hyperparameters were tuned using a LOSO grid search (153 configurations) targeting the Weibull-weighted peak MAE. The selected specification (used above) is:

- k_ref = 25
- multi_temperature = 0.25
- template_shift = 0
- align_rise_weight = 1

Best Weibull-weighted peak MAE (LOSO) = 1.1691 (4.0% improvement over previous baseline of 1.2178). These values were applied to the multi-template run and reflected in the short test script under scripts/_test_loso_comparison.R.


```

## Alignment parameters over time

One trace per season. Each point = one prospective evaluation at that `eval_week`.

```{r param-plots-setup}
#| echo: false
#| warning: false

param_vars <- c("tau", "delta", "a", "b", "t_peak")
param_labels <- c(
  tau    = "tau - time shift",
  delta  = "delta - scale",
  a      = "a - lower asymptote",
  b      = "b - upper asymptote",
  t_peak = "t_peak - predicted peak (newWeek)"
)

df_params <- wf$params_df %>%
  filter(!is.na(iWeek_hat)) %>%
  select(season, eval_week, all_of(param_vars))

seasons_pal <- setNames(
  RColorBrewer::brewer.pal(min(11, length(unique(df_params$season))), "Set3"),
  sort(unique(df_params$season))
)

make_param_plot <- function(pvar) {
  df <- df_params %>% select(season, eval_week, value = all_of(pvar))
  plot_ly(df, x = ~eval_week, y = ~value, color = ~season,
          colors = seasons_pal,
          type = "scatter", mode = "lines+markers",
          marker = list(size = 6), line = list(width = 1),
          legendgroup = ~season, showlegend = TRUE) %>%
    layout(
      xaxis  = list(title = "Eval week (weekF)"),
      yaxis  = list(title = param_labels[[pvar]]),
      legend = list(title = list(text = "Season"))
    )
}
```

### tau - time shift

```{r param-plot-tau}
#| echo: false
#| warning: false
make_param_plot("tau")
```

### delta - scale

```{r param-plot-delta}
#| echo: false
#| warning: false
make_param_plot("delta")
```

### a - lower asymptote

```{r param-plot-a}
#| echo: false
#| warning: false
make_param_plot("a")
```

### b - upper asymptote

```{r param-plot-b}
#| echo: false
#| warning: false
make_param_plot("b")
```

### t_peak - predicted peak (newWeek)

```{r param-plot-tpeak}
#| echo: false
#| warning: false
make_param_plot("t_peak")
```

## Results by season

Per-season walkforward alignment using the multi-template ensemble. Grey = FS median reference curve (jagged, for visual reference). Blue = fitted past, red = forecast. Vertical lines: dashed grey = ignition, solid black = current newWeek, dashed red = estimated peak, solid blue = observed peak. All x-axis positions and labels are in **newWeek** (alignment-indexed). Orange markers at y=0.05 show corresponding **flu weeks** for clinical reference.

```{r season-reports}
#| echo: false
#| warning: false
#| results: asis

all_seasons_wf <- sort(unique(wf$params_df$season))

true_peaks_obs <- allD %>%
  mutate(p_obs = y / N) %>%
  filter(!is.na(p_obs), is.finite(p_obs), N > 0) %>%
  group_by(season) %>%
  slice_max(p_obs, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(season, true_peak_weekF = weekF)

for (s in all_seasons_wf) {

  cat(sprintf("\n## %s\n\n", s))

  s_all <- wf$params_df %>%
    filter(season == s, !is.na(iWeek_hat))

  if (nrow(s_all) == 0) {
    cat("*No forecasts generated (ignition never detected).*\n\n")
    next
  }

  if (!"peak_passed" %in% names(s_all)) {
    s_all <- s_all %>%
      mutate(peak_weekF = round(t_peak - anchorWeek + iWeek_hat),
             peak_passed = FALSE)
  }

  first_peak_ew <- s_all %>% filter(peak_passed) %>%
    slice_min(eval_week, n = 1, with_ties = FALSE) %>% pull(eval_week)
  cutoff_ew <- if (length(first_peak_ew) > 0) first_peak_ew + 2L else max(s_all$eval_week)

  s_params <- s_all %>%
    filter(eval_week <= cutoff_ew) %>%
    select(eval_week, iWeek_hat, iWeek_true, peak_weekF, anchorWeek,
           tau, delta, t_peak, t_peak_lo, t_peak_hi, peak_passed)

  # Compute newWeek columns for the data table
  s_params_display <- s_params %>%
    mutate(
      current_nW = eval_week - iWeek_hat + anchorWeek,
      peak_nW    = round(t_peak)
    ) %>%
    select(current_nW, peak_nW, tau, delta, t_peak, t_peak_lo, t_peak_hi, peak_passed)

  tbl <- gt(s_params_display) %>%
    cols_label(
      current_nW  = "Current newWeek",
      peak_nW     = "Est. peak (nW)",
      tau         = "tau",
      delta       = "delta",
      t_peak      = "t_peak",
      t_peak_lo   = "t_peak lo",
      t_peak_hi   = "t_peak hi",
      peak_passed = "Peak passed"
    ) %>%
    fmt_number(columns = c(tau, delta, t_peak, t_peak_lo, t_peak_hi), decimals = 2) %>%
    fmt_integer(columns = c(current_nW, peak_nW)) %>%
    tab_style(
      style = cell_fill(color = "#f0f4ff"),
      locations = cells_body(rows = seq(1, nrow(s_params_display), 2))
    ) %>%
    tab_style(
      style = cell_fill(color = "#fff0f0"),
      locations = cells_body(rows = peak_passed == TRUE)
    ) %>%
    tab_options(table.width = pct(100), table.font.size = "13px")

  obs_s <- allD %>%
    filter(season == s) %>%
    mutate(p_obs = y / N)

  true_peak_weekF <- true_peaks_obs %>%
    filter(season == s) %>% pull(true_peak_weekF)
  true_peak_weekF <- if (length(true_peak_weekF) > 0) true_peak_weekF[[1]] else NA_integer_

  # nW_true for this season (for flu week conversion)
  nW_s <- obs_s$nW_true[1]
  if (is.null(nW_s) || is.na(nW_s)) nW_s <- 52L

  ref_s     <- wf$ref_list[[s]]
  newW_grid <- seq(1, 52, by = 0.5)
  ref_curve <- tibble(
    newWeek = newW_grid,
    p_ref   = plogis(ref_s$g_ref_fun_orig(newW_grid))
  )

  cat("::: {.panel-tabset}\n\n")

  tab_chunks <- lapply(s_params$eval_week, function(ew) {
    .ref_curve <- ref_curve

    pw           <- wf$params_df %>% filter(season == s, eval_week == ew)
    iWeek_hat_ew <- pw$iWeek_hat[[1]]
    anchor_ew    <- pw$anchorWeek[[1]]
    t_peak_ew    <- pw$t_peak[[1]]
    peak_nW      <- round(t_peak_ew)

    current_newW    <- ew - iWeek_hat_ew + anchor_ew
    true_peak_newW  <- if (!is.na(true_peak_weekF))
                         true_peak_weekF - iWeek_hat_ew + anchor_ew else NA_real_

    fc_smooth <- wf$forecast_df %>%
                   filter(season == s, eval_week == ew, kind == "forecast") %>%
                   arrange(newWeek)
    junction  <- fc_smooth$newWeek[which.min(abs(fc_smooth$newWeek - current_newW))]
    fc_past   <- fc_smooth %>% filter(newWeek <= junction)
    fc_future <- fc_smooth %>% filter(newWeek >= junction)

    obs_aligned <- obs_s %>%
      filter(weekF <= ew) %>%
      mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

    y_max <- max(c(fc_smooth$p_hi, obs_aligned$p_obs), na.rm = TRUE) * 1.1
    y_max <- max(y_max, 0.15)

    # Flu week annotation markers at y = 0.05
    # Place at integer newWeek positions that are within the plot range
    nw_range <- range(c(fc_smooth$newWeek, obs_aligned$newWeek), na.rm = TRUE)
    nw_ticks <- seq(max(1L, floor(nw_range[1])), min(52L, ceiling(nw_range[2])), by = 2L)
    flu_ticks <- tibble(
      newWeek  = as.numeric(nw_ticks),
      flu_week = newWeek_to_fluWeek(nw_ticks, anchor_ew, iWeek_hat_ew, nW_s),
      y        = 0.05
    )

    # Current flu week for title
    current_fluWk <- newWeek_to_fluWeek(current_newW, anchor_ew, iWeek_hat_ew, nW_s)

    plt <- plot_ly() %>%
      # FS median reference curve (jagged)
      add_lines(
        data = .ref_curve, x = ~newWeek, y = ~p_ref,
        line = list(color = "rgba(160,160,160,0.5)", width = 1.2),
        name = "FS median ref"
      ) %>%
      # CI ribbon
      add_ribbons(
        data = fc_smooth, x = ~newWeek, ymin = ~p_lo, ymax = ~p_hi,
        fillcolor = "rgba(150,150,150,0.15)",
        line = list(color = "transparent"),
        name = "95% CI", showlegend = TRUE
      ) %>%
      # Fitted past (blue)
      add_lines(
        data = fc_past, x = ~newWeek, y = ~p_hat,
        line = list(color = "steelblue", width = 2.5),
        name = "Fitted (past)"
      ) %>%
      # Forecast (red)
      add_lines(
        data = fc_future, x = ~newWeek, y = ~p_hat,
        line = list(color = "firebrick", width = 2.5),
        name = "Forecast"
      ) %>%
      # Observed data
      add_markers(
        data = obs_aligned, x = ~newWeek, y = ~p_obs,
        marker = list(color = "black", size = 5, opacity = 0.75),
        name = "Observed"
      ) %>%
      # Flu week annotations at y=0.05
      add_text(
        data = flu_ticks, x = ~newWeek, y = ~y,
        text = ~as.character(flu_week),
        textfont = list(size = 9, color = "darkorange"),
        name = "Flu week", showlegend = FALSE
      ) %>%
      # Ignition (dashed grey)
      add_segments(
        x = anchor_ew, xend = anchor_ew, y = 0, yend = y_max,
        line = list(color = "grey50", dash = "dash", width = 1.5),
        name = paste0("Ignition (nW ", anchor_ew, ")"), showlegend = TRUE
      ) %>%
      # Current week (solid black)
      add_segments(
        x = current_newW, xend = current_newW, y = 0, yend = y_max,
        line = list(color = "black", dash = "solid", width = 2),
        name = paste0("Now (nW ", current_newW, ")"), showlegend = TRUE
      ) %>%
      # Estimated peak (dashed red)
      add_segments(
        x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
        line = list(color = "firebrick", dash = "dash", width = 1.5),
        name = paste0("Est. peak (nW ", peak_nW, ")"), showlegend = TRUE
      )

    # Observed peak (solid blue)
    if (!is.na(true_peak_newW)) {
      plt <- plt %>%
        add_segments(
          x = true_peak_newW, xend = true_peak_newW, y = 0, yend = y_max,
          line = list(color = "steelblue", dash = "solid", width = 1.5),
          name = paste0("Obs. peak (nW ", round(true_peak_newW), ")"), showlegend = TRUE
        )
    }

    plt <- plt %>%
      layout(
        title = paste0(s, " | newWeek ", current_newW,
                       " (flu wk ", current_fluWk, ")",
                       " | est peak nW ", peak_nW),
        xaxis = list(title = "newWeek (aligned)"),
        yaxis = list(title = "Positivity", range = c(0, y_max)),
        legend = list(orientation = "h", y = -0.20)
      )

    e <- new.env(parent = globalenv())
    e$tab_plot <- plt
    knitr::knit_child(
      text = c(sprintf("### nW %d", current_newW), "",
               "```{r echo=FALSE, warning=FALSE}", "tab_plot", "```", ""),
      envir = e, quiet = TRUE
    )
  })

  cat(unlist(tab_chunks), sep = "\n")

  e_tbl <- new.env(parent = globalenv())
  e_tbl$tbl_obj <- tbl
  data_tab <- knitr::knit_child(
    text = c("### Data", "", "```{r echo=FALSE}", "tbl_obj", "```", ""),
    envir = e_tbl, quiet = TRUE
  )
  cat(data_tab)

  cat("\n:::\n\n")
}
```

