---
title: "Post-Peak Flag Tuning"
format:
  html:
    self-contained: true
    toc: true
    page-layout: full
    grid:
      body-width: 1000px
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(gt)
library(MMWRweek)
library(PAGe)

source("../R/utils.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/pipeline_runtime.R")
source("../R/m1_runtime.R")

allD <- read.csv("../data/flu_testing_data.csv") %>%
  select(season, week, year, start_year = seasonstart, date = week_start_date,
         y = pos_flua, N = test_flu) %>%
  mutate(neg = N - y, date = as.Date(date),
         mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
         nW_true   = n_weeks_in_start_year(start_year),
         weekF     = ((week - 27L) %% nW_true) + 1L,
         p = y / N) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))

# Load LOSO walk-forward results (required — run loso_walkforward.qmd first)
wf <- readRDS("../data/align_multi_cache.rds")

# Backfill peak_passed / peak_weekF if loaded from a pre-refactor cache
if (!"peak_passed" %in% names(wf$params_df)) {
  wf$params_df <- wf$params_df %>%
    mutate(
      peak_weekF    = round(t_peak - anchorWeek + iWeek_hat),
      last_obs_newW = eval_week - iWeek_hat + anchorWeek,
      peak_passed   = !is.na(last_obs_newW) & !is.na(t_peak_hi) &
                      last_obs_newW >= t_peak_hi
    ) %>%
    select(-last_obs_newW)
}
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

## Overview

`tune_peak_detection()` evaluates combinations of two parameters that control
when the aligner declares the epidemic peak has passed:

| Parameter | Role |
|-----------|------|
| `use_ci` | If `TRUE`, threshold = upper CI bound of estimated peak (`t_peak_hi`). If `FALSE`, threshold = point estimate (`t_peak`). |
| `buffer_weeks` | Additional weeks added to the threshold before declaring peak passed. Negative = early trigger. |

The evaluation runs directly on the pre-computed `params_df` from
[`loso_walkforward.qmd`](loso_walkforward.html) — **no LOSO rerun needed**.

For each `(season, eval_week)` row, the algorithm checks whether
`last_obs_newWeek >= threshold`. The first `eval_week` per season where this
fires is the **detection week**. Detection delay = `detection_weekF − true_peak_weekF`.

**Tuning goal:** minimise mean detection delay among seasons with no false
positive (`delay >= 0`). Pick the smallest `buffer_weeks` with `fp_rate == 0`.

## Run tuning

Results cached to `../data/peak_det_tuning.rds`.

```{r peak-det-tuning}
#| message: false
#| warning: false

PK_DET_CACHE <- "../data/peak_det_tuning.rds"

if (file.exists(PK_DET_CACHE)) {
  pk_tuning <- readRDS(PK_DET_CACHE)
  message("Loaded from cache: ", PK_DET_CACHE)
} else {
  pk_tuning <- tune_peak_detection(
    params_df         = wf$params_df,
    allD              = allD,
    use_ci_grid       = c(TRUE, FALSE),
    buffer_weeks_grid = -2:5
  )
  saveRDS(pk_tuning, PK_DET_CACHE)
  message("Saved to ", PK_DET_CACHE)
}
```

## Heatmap

Mean detection delay (weeks after true peak). Cells with `fp_rate > 0` show
the false-positive percentage instead.

```{r peak-det-heatmap}
#| echo: false
#| warning: false

pk_heat <- pk_tuning %>%
  mutate(
    use_ci_label = ifelse(use_ci, "CI upper bound", "Point estimate"),
    label        = ifelse(fp_rate == 0,
                          sprintf("delay %.1f", mean_delay),
                          sprintf("FP %.0f%%", fp_rate * 100))
  )

plot_ly(pk_heat,
        x = ~buffer_weeks, y = ~use_ci_label,
        z = ~mean_delay, text = ~label,
        type = "heatmap",
        colorscale = "Blues",
        showscale = TRUE,
        hovertemplate = paste(
          "use_ci: %{y}<br>buffer: %{x}<br>",
          "mean delay: %{z:.1f} wks<br>",
          "fp_rate: %{customdata:.1%}<extra></extra>"
        ),
        customdata = ~fp_rate) %>%
  add_annotations(
    x = ~buffer_weeks, y = ~use_ci_label,
    text = ~label,
    showarrow = FALSE,
    font = list(size = 11, color = "black")
  ) %>%
  layout(
    title  = "Mean detection delay — blank/NA cells have fp_rate > 0",
    xaxis  = list(title = "buffer_weeks", dtick = 1),
    yaxis  = list(title = "Threshold type")
  )
```

## Summary table

```{r peak-det-table}
#| echo: false
#| warning: false

best_no_fp <- pk_tuning %>%
  filter(fp_rate == 0) %>%
  arrange(mean_delay) %>%
  slice(1)

gt(pk_tuning %>% arrange(use_ci, buffer_weeks)) %>%
  cols_label(
    use_ci       = "use_ci",
    buffer_weeks = "buffer_weeks",
    fp_rate      = "FP rate",
    mean_delay   = "Mean delay (wks)",
    median_delay = "Median delay",
    max_delay    = "Max delay",
    n_seasons    = "N seasons"
  ) %>%
  fmt_number(columns = c(fp_rate, mean_delay, median_delay, max_delay),
             decimals = 2) %>%
  tab_style(
    style     = list(cell_fill(color = "#d4edda"), cell_text(weight = "bold")),
    locations = cells_body(
      rows = use_ci == best_no_fp$use_ci & buffer_weeks == best_no_fp$buffer_weeks
    )
  ) %>%
  tab_footnote(
    "Green = recommended setting (fp_rate == 0, lowest mean delay)."
  ) %>%
  tab_options(table.width = pct(75), table.font.size = "13px")
```

## Recommendation

```{r recommendation}
#| echo: false

cat(sprintf(
  "Recommended: use_ci = %s, buffer_weeks = %d\n  fp_rate = %.0f%%  |  mean delay = %.1f wks  |  max delay = %.0f wks",
  best_no_fp$use_ci, best_no_fp$buffer_weeks,
  best_no_fp$fp_rate * 100,
  best_no_fp$mean_delay,
  best_no_fp$max_delay
))
```
