---
title: "PAGe: A Practical Guide"
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
execute:
  eval: false
---

PAGe training is an offline workflow. The examples below are not executed
during documentation builds because full fitting and nested LOSO are
computationally substantial.

## Prepare authorized surveillance data

PAGe does not ship surveillance observations. Input must contain one row per
season and `weekF`, positive tests `y`, and either total tests `N` or negative
tests `neg`.

```{r}
library(PAGe)

allD <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()
```

## Refresh the locked production kit

Use refresh when the model structure is already approved. It fits the deployed
v16 specification (`k_f=4`, `k_e=2`, `alpha_state=0.15`, `k_sp=6`, `k_r=0`,
`k_de=0`, `delta=0`, `Kr=1`, `bias_alpha=0.05`, `bias_beta=0`). The 2025-26
season remains excluded as a prospective holdout by default.

```{r}
training <- train_pipeline(allD, mode = "refresh")
kit <- training$kit
saveRDS(kit, "page_kit.rds")
```

## Replay the holdout and apply promotion gates

Both candidate and incumbent kits must have been trained without 2025-26.

```{r}
candidate <- replay_season_holdout(kit, allD, season = "2025-26")
incumbent <- replay_season_holdout(incumbent_kit, allD, season = "2025-26")
promotion <- check_promotion(candidate$metrics, incumbent$metrics)

# A passing report releases 2025-26 into the refresh for 2026-27.
next_training <- train_pipeline(allD, mode = "refresh", promotion = promotion)
```

Promotion requires at least 2% Bernoulli-NLL improvement, at most 5% MAE
degradation at any horizon, and at most 10% MAE degradation in any phase. A
failed or malformed report keeps the holdout excluded. The versioned report
schema validates the locked thresholds and internal gate consistency; it does
not provide cryptographic proof of who created the report.

## Retune when model selection is required

```{r}
retuned <- train_pipeline(
  allD,
  mode = "retune",
  previous_results = prior_m2_results,
  selection_method = "min_nll", # alternatives: "one_se", "pareto"
  racing = FALSE,
  checkpoint_dir = "results/checkpoints"
)
```

The adaptive grid retains the v16 incumbent, uses compatible prior results,
keeps diverse finalists, adds local neighbors, and expands reached boundaries.
Optional `racing=TRUE` requires a user-supplied partial-fold evaluator. Racing
only removes clear losers; every survivor receives full nested-LOSO evaluation
before selection.

## Produce a frozen prospective forecast

```{r}
kit <- readRDS("page_kit.rds")
current <- prepare_surveillance_data(current_csv, season = "2026-27")
forecast <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(forecast, history = allD)
```

Frozen-GAM deployment is the default validated path. Archive the input
snapshot, issue date, kit version, manual overrides, and returned M0/M1/M2 state
for reproducibility. Weekly refitting is available only as an explicit legacy
compatibility option.
