---
title: "PAGe: A Practical Guide"
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
execute:
  eval: false
---

PAGe training is an offline workflow. The examples below are not executed
during documentation builds because full fitting and nested LOSO are
computationally substantial.

## Prepare authorized surveillance data

PAGe does not ship surveillance observations. Input must contain one row per
season and `weekF`, positive tests `y`, and either total tests `N` or negative
tests `neg`.

```{r}
library(PAGe)

allD <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()
```

## Declare the governed season sets

The guarded stage API never infers training membership from the rows that
happen to be present. Declare training, permanent exclusions, untouched
holdouts, and prospective application seasons explicitly:

```{r}
selection <- validate_season_selection(
  allD,
  training_seasons = development_seasons,
  exclude_seasons = c("2011-12", "2015-16", "2020-21", "2021-22"),
  holdout_seasons = "2025-26",
  application_seasons = "2026-27"
)
```

The four sets must be mutually disjoint and every identifier must exist in the
canonical data. `season_selection(x)` retrieves the recorded selection from a
governed tuning result, stage artifact, or assembled kit.

## Train and freeze each stage

Each stage has separate tuning, tuning-validation, fitting, and freezing
operations. A downstream stage accepts only frozen upstream artifacts with the
same season selection and expected artifact identities.

```{r}
# M0: ignition
m0_tuning <- tune_m0(allD, selection = selection)
validate_m0_tuning(m0_tuning)
m0 <- fit_m0(allD, selection, config = m0_tuning$best_params) |>
  freeze_m0(tuning = m0_tuning)

# M1: alignment
m1_tuning <- tune_m1(allD, m0 = m0, selection = selection)
validate_m1_tuning(m1_tuning)
approved_m1_config <- as.list(m1_tuning$best[1, , drop = FALSE])
m1 <- fit_m1(allD, selection, m0, config = approved_m1_config) |>
  freeze_m1(tuning = m1_tuning)

# M2: forecast
m2_tuning <- tune_m2(allD, selection, m0, m1, grid = m2_grid)
validate_m2_tuning(m2_tuning)
m2 <- fit_m2(allD, selection, m0, m1, config = m2_tuning$best_spec) |>
  freeze_m2(tuning = m2_tuning)

kit <- assemble_kit(m0, m1, m2)
validate_page_kit(kit)
```

The example uses the validated one-row `m1_tuning$best` selection as the fixed
M1 configuration. A tuning result is not itself a fitted stage, and a draft fit
is not a deployment input. `assemble_kit()` fails if a governed chain contains
a draft stage, mismatched season selection, or broken upstream identity.

`tune_m1()` normally inherits M0's manual labels and records the shifted M1
label vector in its result. To reproduce a historical M1 search that used a
different target-label vector, pass that vector explicitly with
`manual_labels=` in the unshifted M0 coordinate system. This override changes
the M1 tuning target only; it should not silently replace the labels used to
fit M0 or the production reference.

M2 tuning constructs labels separately for every LOSO fold. The held-out
season is excluded from `manual_labels_train`, while `manual_labels_test` is
`NULL`, so held-out ignition timing is detected prospectively. Any checkpoint
created with the held-out label supplied to the evaluation fold is leaked and
must be recomputed rather than merely re-ranked.

## Compatibility refresh of the locked specification

Use refresh when the model structure is already approved. It fits the coded
locked v16 specification (`k_f=4`, `k_e=2`, `alpha_state=0.15`, `k_sp=6`, `k_r=0`,
`k_de=0`, `delta=0`, `Kr=1`, `bias_alpha=0.05`, `bias_beta=0`). The 2025-26
season remains excluded as a prospective holdout by default.

```{r}
training <- train_pipeline(allD, mode = "refresh")
kit <- training$kit
saveRDS(kit, "page_kit.rds")
```

`train_pipeline()` is the high-level compatibility orchestrator and composes
the guarded stage lifecycle above for both refresh and retune. Use the explicit
calls when a workflow needs to inspect or gate each stage independently.

## Replay the holdout and apply promotion gates

Both candidate and incumbent kits must have been trained without 2025-26.
Governed evaluation runs through `scripts/acceptance/replay_2025_26.R`, which
saves the private replay bundle and disclosure-safe manifest before returning
a promotion decision. Do not pass a bare `check_promotion()` report to
`train_pipeline()`; it has no artifact provenance and cannot release the
holdout.

After a passing acceptance run, use
`season2526/run_retrain_venkata.R` with the saved decision bundle, manifest,
authorized data, candidate kit, and incumbent kit. That script verifies every
bound SHA-256 hash, constructs short-lived verified promotion evidence, and
performs the fixed post-promotion refresh.

Promotion requires at least 2% Bernoulli-NLL improvement, at most 5% MAE
degradation at any horizon, and at most 10% MAE degradation in any phase. A
failed or malformed report keeps the holdout excluded. The versioned report
schema validates the locked thresholds and internal gate consistency. The
retained artifacts and their verified hashes—not the forgeable R class—are the
production safety boundary.

## Retune when model selection is required

```{r}
retuned <- train_pipeline(
  allD,
  mode = "retune",
  previous_results = prior_m2_results,
  selection_method = "min_nll", # alternatives: "one_se", "pareto"
  racing = FALSE,
  checkpoint_dir = "results/checkpoints"
)
```

The adaptive grid retains the v16 incumbent, uses compatible prior results,
keeps diverse finalists, adds local neighbors, and expands reached boundaries.
Optional `racing=TRUE` requires a user-supplied partial-fold evaluator. Racing
only removes clear losers; every survivor receives full nested-LOSO evaluation
before selection.

### Diagnose and expand the grid

A selected value should normally have tested values on both sides. If the
winner is at a lower or upper edge, retain the existing candidates and add one
valid value beyond that edge using the adjacent observed spacing. Then rerun
the same complete folds and selection rule. Prefer one-factor local neighbors
before adding targeted interactions; do not expand every axis into an
unbounded Cartesian grid.

A boundary can be legitimate when it is a null or hard constraint. For
example, an M2 smooth dimension of zero means that feature is off and should
not be expanded to an invalid negative value. Record whether each boundary is
expanded, accepted as a constraint/null, or left unresolved because the
predeclared grid cap was reached.

For the default M1 grid, a selected `k_ref=25` or `slope_weight=8` is on the
lower edge. If that selection survives the complete governed run, adjacent
spacing suggests testing `k_ref=20` and `slope_weight=4` in the next
pre-holdout development round while retaining the original candidates.

For M2, prefer:

```{r}
next_grid <- plan_m2_grid(
  previous_results = prior_m2_results,
  max_finalists = 6L,
  max_specs = 64L
)
```

The planner retains the coded incumbent and prior finalists, adds nearest local
neighbors, expands valid winning boundaries by one observed step, and preserves
stable specification identities. Stop expanding when the winner is bracketed,
a boundary is a documented null/constraint, improvement is flat relative to
fold uncertainty, or the declared computational cap is reached. Never alter
the grid after viewing the prospective holdout.

The [tuning playbook](tuning-playbook.md) gives parameter-specific M0, M1, and
M2 advice plus a boundary-report schema.

## Produce a frozen prospective forecast

```{r}
kit <- readRDS("page_kit.rds")
current <- prepare_surveillance_data(current_csv, season = "2026-27")
forecast <- run_pipeline(kit, current, mode = "frozen")
plot_forecast(forecast, history = allD)
```

Frozen-GAM deployment is the default validated path. Archive the input
snapshot, issue date, kit version, manual overrides, and returned M0/M1/M2 state
for reproducibility. Weekly refitting is available only as an explicit legacy
compatibility option.
