---
title: "PAGe: A Practical Guide"
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
execute:
  eval: false
---

PAGe separates its supported user API from its internal training and reproduction machinery. This guide uses only exported functions. For implementation details, see [`stage-api-map.md`](stage-api-map.md) and [`tuning-playbook.md`](tuning-playbook.md).

::: {.callout-tip title="Related walkthroughs"}
For a training-focused walkthrough, see [`training_walkthrough.qmd`](training_walkthrough.qmd). For current ORVT ingestion, real-time forecasting, the walk-forward report, and HTTP deployment, see [`realtime_deployment_walkthrough.qmd`](realtime_deployment_walkthrough.qmd). The complete exported namespace is listed in [`public-api.qmd`](public-api.qmd).
:::

## Prepare authorized surveillance data

PAGe does not ship surveillance observations. Load an authorized local source or supported official source, then validate the canonical contract.

```{r}
library(PAGe)

allD <- page_load_surveillance(
  source = "/authorized/path/flu_history.csv",
  season = NULL
)

allD <- prepare_surveillance_data(allD)
validate_surveillance_data(allD)
```

## Declare the governed season sets

Training membership is explicit rather than inferred from whatever rows happen to be present.

```{r}
selection <- validate_season_selection(
  allD,
  training_seasons = setdiff(
    sort(unique(allD$season)),
    c("2011-12", "2015-16", "2020-21", "2021-22", "2025-26")
  ),
  exclude_seasons = c("2011-12", "2015-16", "2020-21", "2021-22"),
  holdout_seasons = "2025-26",
  application_seasons = character()
)
```

The four season sets must be mutually disjoint. `season_selection(x)` retrieves the recorded selection from a governed result or kit.

## Train PAGe

`page_train()` owns the governed tuning, validation, fitting, freezing, and kit-assembly boundaries.

For a fixed-specification refresh:

```{r}
trained <- page_train(
  data = allD,
  selection = selection,
  mode = "refresh"
)
```

For governed retuning and model selection:

```{r}
retuned <- page_train(
  data = allD,
  selection = selection,
  mode = "retune",
  prospective_holdout = "2025-26",
  checkpoint_dir = "results/checkpoints",
  selection_method = "min_nll",
  verbose = TRUE
)
```

Alternative supported selection methods include `"one_se"` and `"pareto"` where the underlying training workflow permits them.

Low-level tuning grids, boundary expansion, stage freezing, fold runners, cache helpers, and artifact constructors are intentionally internal. Repository-owned reproduction scripts may use them explicitly, but external users should train through `page_train()`.

## Save and validate a kit

```{r}
page_save_kit(
  trained$kit,
  "results/kits/page-kit.rds"
)

kit <- page_load_kit("results/kits/page-kit.rds")
page_validate_kit(kit, mode = "frozen")
```

`page_save_kit()` and `page_load_kit()` validate the governed artifact rather than acting as unguarded `saveRDS()`/`readRDS()` wrappers.

## Advanced component-level interface

Researchers can work directly with the scientific components when they already have explicit fixed configurations.

```{r}
m0 <- m0_fit(allD, selection, config = m0_config)
m0_state <- m0_detect(m0_kit, current_data)

m1 <- m1_fit(allD, selection, m0, config = m1_config)
m1_state <- m1_predict(m1_kit, current_data, m0_state)
peak <- m1_peak_posterior(
  library = m1_library,
  current_data = current_data,
  activation_week = ignition_week,
  origin_week = origin_week
)
passage <- m1_passage_posterior(
  library = m1_library,
  current_data = current_data,
  activation_week = ignition_week,
  origin_week = origin_week
)

m2 <- m2_fit(allD, selection, m0, m1, config = m2_config)
m2_state <- m2_predict(m2_kit, current_data, m1_state)
```

These functions expose scientific model operations; they do not expose the development-generation names of the implementations underneath them.

## Evaluate and replay

```{r}
evaluation <- evaluate_forecasts(
  allD,
  holdouts = sort(unique(allD$season))
)

replay <- replay_holdout(
  kit,
  allD
)
```

Promotion decisions remain separate from model fitting:

```{r}
gate <- check_promotion(candidate_metrics, incumbent_metrics)
evidence <- verify_promotion(promotion_bundle)
```

The production safety boundary is the retained artifact evidence and verified hashes, not merely the R class of a decision object.

## Forecast from the current production release

For the packaged current production release:

```{r}
forecast <- page_forecast(
  current_data,
  season = "2026-27"
)

print(forecast)
```

For a previously saved kit in the general real-time workflow:

```{r}
forecast_now <- page_forecast_now(
  kit = kit,
  data = current_data
)
```

## Generate the walk-forward report

```{r}
page_walkforward_report(
  current_data,
  season = "2026-27",
  output_file = "PAGe_2026-27_walkforward.html",
  self_contained = TRUE
)
```

The report is generated from R; the HTML artifact is not a hand-maintained source file.

## Aggregate stratified forecasts

The same aggregation API can combine pathogen types, age groups, regions, sites, or other strata.

```{r}
ab <- aggregate_strata(
  estimate = c(A = 0.05, B = 0.02),
  se = c(A = 0.004, B = 0.002),
  method = "sum",
  dependence = "shared_denominator",
  bounds = c(0, 1)
)
```

For age groups or other denominator partitions:

```{r}
all_age <- aggregate_strata(
  estimate = c(`0-17` = 0.10, `18-64` = 0.20, `65+` = 0.30),
  se = c(`0-17` = 0.01, `18-64` = 0.02, `65+` = 0.03),
  method = "weighted_mean",
  weights = c(`0-17` = 100, `18-64` = 300, `65+` = 100),
  dependence = "independent",
  bounds = c(0, 1)
)
```

When joint simulation or posterior draws exist, prefer `aggregate_strata_draws()` because it propagates the full dependence structure directly.

## Provenance and manuscript work

Keep input snapshots, release identifiers, hashes, training selections, and generated reports with each governed run. Manuscript-specific protocols, drafts, publication analyses, and manuscript runners live under the repository's separate `manuscript/` tree; production package code remains under `PAGe/`.
