---
title: "PAGe Pipeline: End-to-End Walkthrough"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    page-layout: full
    grid:
      body-width: 1200px
---

```{=html}
<style>
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
.callout-tip  { border-left: 4px solid #28a745; background: #f0fff4; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
.callout-note { border-left: 4px solid #4a90d9; background: #f4f8fd; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
</style>
```

This document walks through the complete **M0 → M1 → M2** forecasting pipeline
in two parts:

1. **Training** (offline): tune M0 ignition thresholds, build M1 reference curve,
   tune and train M2 forecast model, assemble the deployment kit.
2. **Deployment** (online, every new data week): load the kit and run
   `run_m0()` → `run_m1()` → `run_m2()` (or the single `run_pipeline()` call).

For deep dives into individual steps see:
[`ignition_training.html`](ignition_training.html) |
[`estimateRef.html`](estimateRef.html) |
[`loso_walkforward.html`](loso_walkforward.html) |
[`forecast_training.html`](forecast_training.html)

---

## Quick Start

Train everything and deploy in ~10 lines. Uses `loso_seasons = "alternating"` for
M2 (every other season as LOSO test fold) so the first run completes in roughly
30 minutes instead of 2–3 hours.

```{r quick-start}
#| eval: false

library(PAGe)

allD <- load_flu_data()                             # load surveillance history

m0       <- tune_m0(allD)                          # M0: LOSO grid search
m1       <- build_m1(allD, m0)                     # M1: reference curve + hyperparams
m2_loso  <- build_m2(allD, m0, m1,                # M2: nested LOSO tuning (~30 min)
                     loso_seasons = "alternating")
m2_model <- train_m2(allD, m0, m1,                # M2: production GAM fit
                     m2_loso$best_spec)
kit      <- assemble_kit(m0, m1, m2_model,        # bundle artifacts for deployment
                         save_ref_path = "data/ref_production.rds",
                         save_m2_path  = "data/m2_production.rds")

# --- Deployment (every new week) ---
current  <- getCurrentD() |> dplyr::filter(season == "2025-26")
wf       <- run_pipeline(kit, current)             # full M0 → M1 → M2 in one call
```

Switch `loso_seasons = "all"` for a full production-quality LOSO (all seasons as
test folds, ~2 hours). See [Customization](#customization) for other options.

---

## Setup

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(MMWRweek)
library(plotly)
library(gt)

r_dir <- normalizePath("../R", mustWork = TRUE)
for (f in c(
  "utils.R", "m0_retro.R", "m0_training.R", "flagIgnition.R",
  "m1_reference_helpers.R", "m1_reference.R",
  "m1_hyperparams.R", "m1_fit.R", "check_scale_identifiability.R",
  "m1_runtime.R", "m1_multi_template.R",
  "m1_peak_status.R", "m1_peak_summary.R", "m1_loso.R",
  "m2_training.R", "m2_spec_grid.R", "m2_nested_loso.R",
  "pipeline_runtime_helpers.R", "pipeline_bridge.R",
  "m0_runtime.R", "m2_runtime.R", "pipeline_runtime.R",
  "pipeline_training.R", "getCurrentD.R"
)) {
  fp <- file.path(r_dir, f)
  if (file.exists(fp)) source(fp)
}

`%||%` <- function(x, y) if (is.null(x)) y else x
startWeek <- 27L
DATA_DIR  <- normalizePath("../data", mustWork = TRUE)
```

---

## Part 1 — Training (offline)

Run once per season (or whenever surveillance history grows). Each function
handles one stage and returns an object that the next stage consumes.

### 1.1 Load data

```{r load-data}
#| message: false
#| warning: false

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(MMWRweek::MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L)
}

allD <- read.csv(file.path(DATA_DIR, "flu_testing_data.csv")) |>
  dplyr::select(season, week, year, start_year = seasonstart,
                date = week_start_date, y = pos_flua, N = test_flu) |>
  dplyr::mutate(
    neg     = N - y,
    date    = as.Date(date),
    nW_true = n_weeks_in_start_year(start_year),
    weekF   = ((week - startWeek) %% nW_true) + 1L,
    p       = y / N
  ) |>
  dplyr::filter(!season %in% c("2011-12", "2020-21", "2021-22"))

cat(sprintf("allD: %d rows | %d seasons (%s to %s)\n",
            nrow(allD), length(unique(allD$season)),
            min(allD$season), max(allD$season)))
```

### 1.2 M0 — Ignition Detection

`tune_m0()` runs a LOSO grid search over ignition threshold parameters.
Each held-out season is scored against manually-verified ignition weeks.

```{r train-m0}
#| eval: false

m0 <- tune_m0(allD, loso_seasons = "all")

cat("Best M0 params:\n")
print(m0$best_params)
```

```{r load-m0}
#| message: false

m0_raw  <- readRDS(file.path(DATA_DIR, "stage1_tuning.rds"))
manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)
flag_args <- list(
  p_thresh = 0.01, k1 = 0.4, k_c = 0.01, n_consec = 2L,
  min_window = 10L, w_min = 21L, w_max = 21L, d2_relax = -0.01
)
# Reconstruct an m0 object from the saved tuning result
m0 <- list(
  best_params   = m0_raw$best_params,
  tuning        = m0_raw,
  manual_labels = manual_labels,
  flag_args     = flag_args
)
cat("M0 best params:\n"); print(m0$best_params)
```

::: {.callout-note}
**Default grid**: 36 parameter combinations (4 `p_thr` × 3 `prev_thr` × 3
`p_sum_thr`). Full LOSO over all seasons takes ~2 minutes. See
[`ignition_training.html`](ignition_training.html) for scoring details.
:::

### 1.3 M1 — Reference Curve

`build_m1()` aligns all training seasons to their ignition weeks, fits the
factor-smooth reference curve (`estimateRef`), and learns per-season alignment
search bounds (`learn_alignment_hyperparams`).

```{r train-m1}
#| eval: false

m1 <- build_m1(allD, m0)

cat(sprintf("M1 ref: anchorWeek=%d | %d training seasons\n",
            m1$ref$anchorWeek, length(m1$seasons_used)))

# Optional: tune alignment hyperparameters (k_ref × slope_weight, 20-spec grid)
# m1_tune <- tune_m1(allD, m0, m1)
```

```{r load-m1}
#| message: false

ref_cache <- readRDS(file.path(DATA_DIR, "ref_production.rds"))
m1 <- list(
  ref           = ref_cache$ref,
  hyper         = ref_cache$hyper,
  aligned_train = ref_cache$hist_data,
  m1_params     = ref_cache$M1_PARAMS %||% list(
    k_ref = 25L, ref_method = "fs", temperature = 0.25,
    rise_weight = 1.0, trough_weight = 0.1, peak_decay = 0.3,
    slope_weight = 8.0, slope_window = 6L,
    dynamic_temp = FALSE, dynamic_temp_pivot = 10L
  ),
  seasons_used  = sort(unique(ref_cache$hist_data$season))
)
cat(sprintf("M1 ref loaded: anchorWeek=%d, k_ref=%d, slope_weight=%.1f\n",
            m1$ref$anchorWeek,
            m1$m1_params$k_ref,
            m1$m1_params$slope_weight))
```

```{r plot-ref-curve}
#| echo: false
#| warning: false
#| fig-height: 4

nw <- seq(1, 52)
ref_df <- data.frame(
  newWeek = nw,
  p_ref   = plogis(m1$ref$g_ref_fun(nw))
)
ggplot(ref_df, aes(newWeek, p_ref)) +
  geom_line(linewidth = 1.2, colour = "steelblue") +
  labs(title = "M1 Reference Curve",
       x = "Aligned week (newWeek)", y = "Positivity") +
  theme_minimal()
```

::: {.callout-note}
**Locked production spec**: `k_ref=25, temperature=0.25, slope_weight=8.0,
slope_window=6, dynamic_temp=FALSE` — Weibull-weighted peak MAE = 1.275 weeks
(v5–v7 grid). See [`loso_walkforward.html`](loso_walkforward.html).
:::

### 1.4 M2 — Forecast Model

`build_m2()` runs a three-phase nested LOSO to find the best M2 spec.
`train_m2()` then fits the production GAM on all training seasons.

```{r train-m2}
#| eval: false

# Phase 1 + 2: nested LOSO grid search
# loso_seasons = "alternating" → ~30 min; "all" → ~2 hours (production)
m2_loso <- build_m2(allD, m0, m1, loso_seasons = "alternating")

cat(sprintf("\nTop 5 specs:\n"))
print(head(m2_loso$summary[, c("spec_id", "bernoulli_nll")], 5))

# Phase 3: production GAM fit
m2_model <- train_m2(allD, m0, m1, m2_loso$best_spec)
cat(sprintf("GAM EDF: %.2f\n", sum(m2_model$fit$edf)))
```

```{r load-m2}
#| message: false

m2_prod <- readRDS(file.path(DATA_DIR, "m2_production.rds"))
cat(sprintf(
  "M2 loaded: EDF=%.1f | spec_version=%s\n  spec: k_f=%d, k_e=%d, alpha_state=%.2f, k_r=%d, k_de=%d, k_sp=%d\n",
  sum(m2_prod$fit$edf),
  m2_prod$spec_version %||% "unknown",
  m2_prod$spec$k_f, m2_prod$spec$k_e, m2_prod$spec$alpha_state,
  m2_prod$spec$k_r, m2_prod$spec$k_de %||% 0L, m2_prod$spec$k_sp %||% 0L
))
```

::: {.callout-note}
**Production spec (v15-postfix)**: `k_f=5, k_e=2, alpha_state=0.40, k_r=0, k_de=0, k_sp=2, bias_alpha=0.5, bias_beta=0`
— mean Bernoulli NLL = 0.5959 pre-L2 / 0.5796 post-L2 (7200-spec grid, 10 seasons).
`bias_alpha` is now tuned inside LOSO (was deployment-only pre-fix).
See [`forecast_training.html`](forecast_training.html).
:::

### 1.5 Assemble the Deployment Kit

`assemble_kit()` bundles all trained artifacts into the format expected by
`run_pipeline()`. It also optionally saves to disk.

```{r assemble-kit}
#| eval: false

kit <- assemble_kit(
  m0, m1,
  m2_model        = list(
    fit              = m2_prod$fit,
    feature_ranges   = m2_prod$feature_ranges,
    m1_train_preds   = m2_prod$m1_train_preds,
    spec             = m2_prod$spec,
    training_seasons = m2_prod$training_seasons,
    spec_version     = m2_prod$spec_version %||% "v15-postfix"
  ),
  best_spec_id  = m2_prod$best_spec_id,
  save_ref_path = "data/ref_production.rds",
  save_m2_path  = "data/m2_production.rds"
)
```

```{r build-kit}
kit <- load_prospective_kit(DATA_DIR)

cat(sprintf(paste0(
  "Kit contents:\n",
  "  M1: k_ref=%d, temperature=%.2f, slope_weight=%.1f, anchorWeek=%d\n",
  "  M0: %d params loaded\n",
  "  M2: EDF=%.1f | spec_version=%s\n"),
  kit$M1_PARAMS$k_ref,
  kit$M1_PARAMS$temperature,
  kit$M1_PARAMS$slope_weight %||% NA_real_,
  kit$ref$anchorWeek,
  length(kit$m0_params),
  sum(kit$m2_production$fit$edf),
  kit$m2_production$spec_version %||% "unknown"
))
```

---

## Part 2 — Deployment (online, every new data week)

Load the kit once and call `run_pipeline()` each week, or call the three
step functions individually for finer control.

### 2.1 Load current season data

```{r get-current}
#| message: false
#| warning: false

currentD_full <- getCurrentD(startWeek = startWeek)
currentD_full <- dplyr::filter(currentD_full, season == "2025-26")

cat(sprintf("Current season: 2025-26 | Weeks observed: %d | weekF range: %d–%d\n",
            nrow(currentD_full),
            min(currentD_full$weekF), max(currentD_full$weekF)))
```

### 2.2 Run the full pipeline

`run_pipeline()` calls M0 → M1 → M2 in sequence and returns all results.
Use the individual step functions when you need to inspect intermediate results
or override between stages.

```{r run-all}
#| eval: false

# Option A: one call
wf <- run_pipeline(kit, currentD_full)

# Option B: step by step (same result, allows inspection/override)
r0 <- run_m0(kit, currentD_full)
r1 <- run_m1(kit, currentD_full, r0)
r2 <- run_m2(kit, currentD_full, r1)
```

### 2.3 M0 — Ignition Detection

```{r run-m0}
#| message: false

manual_ign_week <- NA_integer_   # set to an integer to override

m0_res <- run_m0(kit, currentD_full,
                 manual_ign_week = manual_ign_week, verbose = TRUE)

cat(sprintf("\nM0 result: iWeek_locked = %s | overridden = %s\n",
            ifelse(is.na(m0_res$iWeek_locked), "not detected",
                   as.character(m0_res$iWeek_locked)),
            m0_res$overridden))
```

::: {.callout-tip}
**When to override**: If surveillance data are delayed or M0 fires a week
early/late, set `manual_ign_week` before calling `run_m1()`. The corrected
`iWeek_locked` propagates to all downstream covariates.
:::

### 2.4 M1 — Alignment Walk-Forward

```{r run-m1}
#| message: false

m1_res <- run_m1(kit, currentD_full, m0_res, walk_start = 5L, verbose = TRUE)

active <- dplyr::filter(m1_res$params_df, !is.na(iWeek_hat))
if (nrow(active) > 0) {
  last <- dplyr::slice_max(active, eval_week, n = 1, with_ties = FALSE)
  cat(sprintf(paste0(
    "M1 walk-forward: %d eval weeks\n",
    "Latest (weekF %d): state=%s | τ=%.2f | δ=%.2f\n",
    "  Est. peak weekF: %s | Peak passed: %s\n"),
    nrow(active),
    last$eval_week, last$state, last$tau, last$delta_m1,
    ifelse(is.na(last$peak_weekF), "?", as.character(last$peak_weekF)),
    ifelse(last$peak_passed, "YES", "no")
  ))
} else {
  cat("M1: alignment not yet active (ignition not detected).\n")
}
```

```{r m1-table}
#| echo: false
#| warning: false

if (nrow(active) > 0) {
  gt(active |> dplyr::select(eval_week, state, iWeek_hat, tau, delta_m1,
                               peak_weekF, peak_passed)) |>
    fmt_number(columns = where(is.numeric), decimals = 3) |>
    fmt_integer(columns = c(eval_week, iWeek_hat, peak_weekF)) |>
    tab_style(
      style     = cell_fill(color = "#fff0f0"),
      locations = cells_body(rows = peak_passed == TRUE)
    ) |>
    tab_header(title = "M1 alignment parameters — all eval weeks") |>
    tab_options(table.width = pct(100), table.font.size = "13px")
}
```

### 2.5 M2 — 1- and 2-week-ahead Forecast

```{r run-m2}
#| message: false

m2_res <- run_m2(kit, currentD_full, m1_res, verbose = TRUE)

if (nrow(m2_res$m2_preds) > 0) {
  latest_ew <- max(m2_res$m2_preds$eval_week)
  latest_m2 <- dplyr::filter(m2_res$m2_preds, eval_week == latest_ew)
  cat(sprintf("\nM2 forecasts at weekF=%d:\n", latest_ew))
  for (i in seq_len(nrow(latest_m2))) {
    r <- latest_m2[i, ]
    cat(sprintf("  h=%d → target weekF %d: M2=%.1f%% [%.1f–%.1f]  |  M1=%.1f%%\n",
                r$h, r$target_weekF,
                r$m2_p * 100, r$m2_lo * 100, r$m2_hi * 100,
                r$m1_p * 100))
  }
}
```

### 2.6 Alignment and forecast plot

```{r plot-current}
#| echo: false
#| warning: false
#| fig-height: 5

if (nrow(active) > 0 && nrow(m2_res$m2_preds) > 0) {
  ew           <- max(active$eval_week)
  pw           <- dplyr::filter(m1_res$params_df, eval_week == ew)
  iWeek_hat_ew <- pw$iWeek_hat[[1]]
  anchor_ew    <- kit$ref$anchorWeek
  peak_wF      <- pw$peak_weekF[[1]]
  t_peak_ew    <- pw$t_peak[[1]]

  current_newW  <- ew - iWeek_hat_ew + anchor_ew
  ignition_newW <- anchor_ew

  fc <- dplyr::filter(m1_res$m1_curves, eval_week == ew, kind == "forecast") |>
    dplyr::arrange(newWeek)
  junction   <- fc$newWeek[which.min(abs(fc$newWeek - current_newW))]
  fc_past    <- dplyr::filter(fc, newWeek <= junction)
  fc_future  <- dplyr::filter(fc, newWeek >= junction)

  obs_s      <- currentD_full |> dplyr::mutate(p_obs = y / N)
  obs_aligned <- obs_s |>
    dplyr::filter(weekF <= ew) |>
    dplyr::mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)
  obs_future  <- obs_s |>
    dplyr::filter(weekF > ew) |>
    dplyr::mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

  m2_ew      <- dplyr::filter(m2_res$m2_preds, eval_week == ew)
  m2_aligned <- m2_ew |>
    dplyr::mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) |>
    dplyr::arrange(newWeek)

  m2_retro <- m2_res$m2_preds |>
    dplyr::filter(eval_week < ew, h == 1) |>
    dplyr::mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) |>
    dplyr::filter(newWeek >= ignition_newW) |>
    dplyr::select(newWeek, m2_p) |>
    dplyr::arrange(newWeek)

  jv            <- obs_aligned$p_obs[obs_aligned$weekF == ew]
  junction_val  <- if (length(jv) > 0 && !is.na(jv[1])) jv[1] else NA_real_
  junction_pt   <- tibble::tibble(newWeek = current_newW, m2_p = junction_val)

  y_max <- max(c(fc$p_hat, obs_s$p_obs, m2_ew$m2_hi), na.rm = TRUE) * 1.15
  y_max <- max(y_max, 0.15)

  nw_grid    <- seq(1, 52, by = 0.5)
  ref_curve  <- tibble::tibble(newWeek = nw_grid,
                                p_ref = plogis(kit$ref$g_ref_fun(nw_grid)))

  plt <- plot_ly() |>
    add_lines(data = ref_curve, x = ~newWeek, y = ~p_ref,
              line = list(color = "rgba(160,160,160,0.4)", width = 1),
              name = "Reference") |>
    add_lines(data = fc_past, x = ~newWeek, y = ~p_hat,
              line = list(color = "steelblue", width = 2), name = "M1 fitted") |>
    add_lines(data = fc_future, x = ~newWeek, y = ~p_hat,
              line = list(color = "steelblue", width = 2, dash = "dash"),
              name = "M1 forecast") |>
    add_markers(data = obs_aligned, x = ~newWeek, y = ~p_obs,
                marker = list(color = "black", size = 5), name = "Observed") |>
    add_markers(data = obs_future, x = ~newWeek, y = ~p_obs,
                marker = list(color = "black", size = 5, opacity = 0.35,
                              symbol = "circle-open"), name = "Future obs") |>
    add_segments(x = anchor_ew,    xend = anchor_ew,    y = 0, yend = y_max,
                 line = list(color = "grey50", dash = "dash", width = 1),
                 name = paste0("Ign wk ", iWeek_hat_ew)) |>
    add_segments(x = current_newW, xend = current_newW, y = 0, yend = y_max,
                 line = list(color = "black", width = 2),
                 name = paste0("Now wk ", ew))

  m2_past_line <- dplyr::bind_rows(m2_retro, junction_pt) |> dplyr::arrange(newWeek)
  if (nrow(m2_past_line) > 1) {
    plt <- plt |>
      add_lines(data = m2_past_line, x = ~newWeek, y = ~m2_p,
                line = list(color = "firebrick", width = 2), name = "M2 fitted")
  }
  m2_future_line <- dplyr::bind_rows(junction_pt,
                                      m2_aligned |> dplyr::select(newWeek, m2_p))
  plt <- plt |>
    add_lines(data = m2_future_line, x = ~newWeek, y = ~m2_p,
              line = list(color = "firebrick", width = 2, dash = "dash"),
              name = "M2 forecast") |>
    add_markers(data = m2_aligned, x = ~newWeek, y = ~m2_p,
                marker = list(color = "firebrick", size = 7, symbol = "diamond"),
                name = "M2 pts", showlegend = FALSE)

  if (!is.na(t_peak_ew)) {
    plt <- plt |>
      add_segments(x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
                   line = list(color = "green", dash = "dash", width = 1.5),
                   name = paste0("Est. peak wk ", peak_wF))
  }

  plt |> layout(
    title = paste0("Current state: weekF=", ew,
                   " | state=", pw$state[[1]],
                   if (!is.na(peak_wF)) paste0(" | peak wk ", peak_wF) else ""),
    xaxis = list(title = "Aligned week (newWeek)"),
    yaxis = list(title = "Positivity", range = c(0, y_max)),
    legend = list(orientation = "h", y = -0.20)
  )
}
```

### 2.7 Forecast accuracy

```{r accuracy}
#| echo: false
#| warning: false

if (nrow(m2_res$m2_preds) > 0) {
  obs_lookup <- currentD_full |>
    dplyr::transmute(target_weekF = as.integer(weekF), p_obs = y / N)

  acc <- m2_res$m2_preds |>
    dplyr::filter(!is.na(m2_p)) |>
    dplyr::left_join(obs_lookup, by = "target_weekF") |>
    dplyr::filter(!is.na(p_obs)) |>
    dplyr::mutate(
      m2_ae     = abs(m2_p - p_obs),
      m1_ae     = abs(m1_p - p_obs),
      m2_covers = (p_obs >= m2_lo) & (p_obs <= m2_hi)
    )

  if (nrow(acc) > 0) {
    acc |>
      dplyr::group_by(h) |>
      dplyr::summarise(n = n(), MAE_M2 = mean(m2_ae), MAE_M1 = mean(m1_ae),
                       Coverage = mean(m2_covers), .groups = "drop") |>
      gt() |>
      tab_header(title = "Forecast accuracy: M2 vs M1 (realised weeks only)") |>
      fmt_number(columns = c(MAE_M2, MAE_M1), decimals = 4) |>
      fmt_percent(columns = Coverage, decimals = 0) |>
      tab_options(table.width = pct(60), table.font.size = "14px")
  } else {
    cat("No realised observations at target weeks yet.\n")
  }
}
```

---

## Customization {#customization}

All high-level functions accept custom season sets and grids.

```{r customization}
#| eval: false

# Custom training seasons (skip some outlier seasons)
m0 <- tune_m0(allD, exclude = c("2011-12", "2015-16", "2020-21", "2021-22", "2022-23"))

# Full LOSO for production-quality M2 tuning
m2_loso <- build_m2(allD, m0, m1, loso_seasons = "all")

# Test only specific seasons as LOSO folds
m2_loso <- build_m2(allD, m0, m1,
                     loso_seasons = c("2016-17", "2018-19", "2022-23", "2024-25"))

# Subset the M2 grid (e.g. fix k_de=0, vary other dims)
my_grid <- default_m2_grid() |> dplyr::filter(k_de == 0L)
m2_loso <- build_m2(allD, m0, m1, grid = my_grid)

# Use a checkpoint directory for resumable M2 tuning
m2_loso <- build_m2(allD, m0, m1,
                     checkpoint_dir = "data/m2_loso_ckpt")

# Tune M1 alignment hyperparameters (optional — defaults are well-tuned)
m1_tune <- tune_m1(allD, m0, m1, loso_seasons = "all")
# Then rebuild m1 with the tuned k_ref and slope_weight if they differ from defaults

# Change bias_alpha (deployment parameter, not tuned via LOSO)
m2_loso <- build_m2(allD, m0, m1, bias_alpha = 0.3)

# Manual ignition override at deployment time
r0 <- run_m0(kit, current_data)
r0$ign_out$iWeek_hat_locked <- 20L   # force week 20
r0$iWeek_locked             <- 20L
r1 <- run_m1(kit, current_data, r0)
r2 <- run_m2(kit, current_data, r1)
```

---

## Function reference

| Function | Stage | Purpose |
|---|---|---|
| `build_m0()` | M0 | Align seasons to ignition weeks (no tuning) |
| `tune_m0()` | M0 | LOSO grid search for ignition thresholds |
| `build_m1()` | M1 | Reference curve + alignment hyperparameters |
| `tune_m1()` | M1 | LOSO grid search for alignment hyperparameters |
| `build_m2()` | M2 | Nested LOSO grid search for forecast spec |
| `train_m2()` | M2 | Production GAM fit on all training seasons |
| `assemble_kit()` | — | Bundle artifacts for deployment |
| `run_m0()` / `run_m0_detection()` | Deploy | Ignition detection for current season |
| `run_m1()` / `run_m1_alignment()` | Deploy | Alignment walk-forward |
| `run_m2()` / `run_m2_forecast()` | Deploy | 1- and 2-week-ahead forecasts |
| `run_pipeline()` / `run_prospective_pipeline()` | Deploy | Full M0→M1→M2 in one call |
| `load_prospective_kit()` | Deploy | Load saved artifacts from disk |

: {tbl-colwidths="[25,10,65]"}
