---
title: "PAGe Pipeline: End-to-End Walkthrough"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    page-layout: full
    grid:
      body-width: 1200px
---

```{=html}
<style>
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
.callout-tip { border-left: 4px solid #28a745; background: #f0fff4; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
.callout-note { border-left: 4px solid #4a90d9; background: #f4f8fd; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
</style>
```

This document walks through the complete **M0 → M1 → M2** forecasting pipeline in two parts:

1. **Training** (offline, once per season or when surveillance data accumulate): fit M0 ignition thresholds, M1 reference curve and alignment hyperparameters, and M2 forecast model.
2. **Deployment** (online, every new data week): load pre-built artifacts and run `run_m0_detection()` → `run_m1_alignment()` → `run_m2_forecast()`.

For deep dives into each training step see:
[`ignition_training.html`](ignition_training.html) |
[`estimateRef.html`](estimateRef.html) |
[`loso_walkforward.html`](loso_walkforward.html) |
[`forecast_training.html`](forecast_training.html)

## Setup

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(MMWRweek)
library(plotly)
library(gt)

# Source all pipeline functions
r_dir  <- normalizePath("../R",   mustWork = TRUE)
r_files <- c(
  "utils.R", "m0_retro.R", "m1_reference_helpers.R", "m1_reference.R",
  "m1_hyperparams.R", "m1_fit.R", "identifiability.R",
  "check_scale_identifiability.R", "m1_runtime.R",
  "align_forecast_pipeline_dilate.R", "m1_multi_template.R",
  "m1_peak_status.R",
  "m1_peak_flags.R", "m1_peak_summary.R", "m0_training.R", "m1_loso.R",
  "m2_training.R", "pipeline_runtime_helpers.R",
  "m0_runtime.R", "m2_runtime.R", "pipeline_runtime.R",
  "forecast_post_peak_gam.R", "m2_spec_grid.R",
  "pipeline_bridge.R", "m2_nested_loso.R", "getCurrentD.R"
)
for (f in r_files) {
  fp <- file.path(r_dir, f)
  if (file.exists(fp)) source(fp)
}
source(file.path(normalizePath("../flualign/R", mustWork = TRUE), "flagIgnition.R"))

startWeek <- 27L
DATA_DIR  <- normalizePath("../data", mustWork = TRUE)
```

## Historical data

The pipeline is trained on all historical influenza seasons. Seasons `2020-21`
and `2021-22` are excluded due to COVID disruption. Season `2015-16` is used for
M1 reference curve estimation but excluded from LOSO alignment folds (ignition
outlier).

```{r load-hist}
#| message: false
#| warning: false

flu_hist_path <- file.path(DATA_DIR, "flu_testing_data.csv")
flu_hist <- read.csv(flu_hist_path) |>
  transmute(
    season,
    week,
    year,
    start_year = seasonstart,
    date = as.Date(week_start_date),
    y = pos_flua,
    N = test_flu
  ) |>
  filter(!season %in% c("2011-12", "2020-21", "2021-22")) |>
  mutate(
    neg = N - y,
    p = y / N,
    weekF = as.integer(((week - 27L) %% (52L + as.integer(MMWRweek::MMWRweek(
      as.Date(paste0(start_year, "-12-31"))
    )$MMWRweek == 53L))) + 1L),
    season = as.character(season)
  )

seasons_available <- sort(unique(flu_hist$season))
cat(sprintf("Seasons: %s\n(%d seasons, %d rows)\n",
            paste(seasons_available, collapse = ", "),
            length(seasons_available), nrow(flu_hist)))
```

---

## Part 1 — Training (offline)

These steps run once, before the current season begins. The outputs are saved
as RDS files in `data/` and loaded at deployment time.

### Step 1 — Train M0: ignition detection

M0 fits threshold gates (`cond_sum`, `cond_p`, `cond_prev`, `cond_inc`) by
LOSO grid search. Each fold holds out one season and tunes thresholds on the
remaining seasons. The best configuration is selected by minimising detection
timing error against manually labelled ignition weeks.

For full details see [`ignition_training.html`](ignition_training.html).

```{r train-m0}
#| eval: false

# Manual ignition labels (weekF, season-relative)
manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

# LOSO tuning grid (see ignition_training.html for full grid)
stage1_result <- tuneIgnitionGrid_M0v2(
  flu_hist     = flu_hist,
  manual_labels = manual_labels,
  startWeek    = startWeek
)

saveRDS(stage1_result, file.path(DATA_DIR, "stage1_tuning.rds"))
cat("Best M0 params:\n")
print(stage1_result$best_params)
```

```{r load-m0}
#| message: false

stage1 <- readRDS(file.path(DATA_DIR, "stage1_tuning.rds"))
cat("M0 best params loaded:\n")
print(stage1$best_params)
```

### Step 2 — Train M1: reference curve and alignment hyperparameters

M1 training has two substeps:

**2a — Reference curve** (`estimateRef()`): aligns historical seasons to their
manually-labelled ignition weeks. The factor-smooth (`fs`) method fits a GAM
with `s(newWeek, season, bs="fs")`, producing per-season curves `eta_mat` used
for multi-template alignment. Full analysis: [`estimateRef.html`](estimateRef.html).

**2b — Alignment hyperparameters** (`tune_m1_alignment()` / `loso_walkforward()`):
a 153-config grid is evaluated by LOSO walk-forward to select `k_ref`,
`temperature`, and `shift` minimising Weibull-weighted peak MAE.
Full analysis: [`loso_walkforward.html`](loso_walkforward.html).

```{r train-m1}
#| eval: false

# 2a — Estimate reference curve (FS method)
ref_result <- estimateRef(
  flu_hist      = flu_hist,
  manual_labels = manual_labels,
  method        = "fs",
  k_ref         = 25L,
  startWeek     = startWeek
)

# 2b — M1 LOSO walk-forward hyperparameter tuning
loso_result <- loso_walkforward(
  flu_hist      = flu_hist,
  manual_labels = manual_labels,
  startWeek     = startWeek,
  grid = expand.grid(
    k_ref       = c(10L, 25L, 40L, 60L),
    temperature = c(0.10, 0.25, 0.50),
    shift       = c(-2L, -1L, 0L, 1L)
  )
)

best_m1 <- loso_result$best_config
# k_ref=25, temperature=0.25, shift=0 → Weibull peak MAE = 1.169

# Re-estimate ref with best k_ref, save with M1_PARAMS attached
ref_production <- estimateRef(
  flu_hist      = flu_hist,
  manual_labels = manual_labels,
  method        = "fs",
  k_ref         = best_m1$k_ref,
  startWeek     = startWeek
)
ref_production$M1_PARAMS <- list(
  k_ref              = best_m1$k_ref,
  temperature        = best_m1$temperature,
  rise_weight        = 1.0,
  trough_weight      = 0.1,
  peak_decay         = 0.3,
  slope_weight       = 0.5,
  slope_window       = 4L,
  dynamic_temp       = TRUE,
  dynamic_temp_pivot = 10L
)

saveRDS(ref_production, file.path(DATA_DIR, "ref_production.rds"))
cat("Saved ref_production.rds — anchorWeek:", ref_production$anchorWeek, "\n")
```

```{r load-m1}
#| message: false

ref_cache <- readRDS(file.path(DATA_DIR, "ref_production.rds"))
ref <- ref_cache$ref
cat(sprintf("M1 ref loaded: anchorWeek=%d, k_ref=%d, temperature=%.2f\n",
            ref$anchorWeek,
            ref_cache$M1_PARAMS$k_ref,
            ref_cache$M1_PARAMS$temperature))
```

### Step 3 — Train M2: nested LOSO hyperparameter tuning and production fit

M2 is a joint binomial GAM for h∈{1,2} in aligned-week space. Hyperparameters
are selected by **nested LOSO**: for each held-out season, the reference curve
and M1 predictions are re-estimated on the remaining seasons to prevent leakage.
Full analysis: [`forecast_training.html`](forecast_training.html).

```{r train-m2}
#| eval: false

# Phase 1: pre-compute M1 walk-forward predictions per LOSO fold
m1_folds <- nested_loso_build_folds(
  flu_hist      = flu_hist,
  manual_labels = manual_labels,
  startWeek     = startWeek,
  k_ref         = 25L,
  temperature   = 0.25
)

# Phase 2: sweep M2 spec grid (189 specs)
spec_grid <- make_stage2_spec_grid()   # built-in 189-spec grid
nested_result <- nested_loso_m2_eval(
  folds     = m1_folds,
  spec_grid = spec_grid,
  flu_hist  = flu_hist
)
# Best spec: delta=0, Kr=1, k_f=2, k_e=4, alpha_state=0.10, lambda_w=0

# Phase 3: refit on full historical data with best spec
best_spec <- nested_result$best_spec
m2_production_fit <- nested_loso_refit_best(
  flu_hist      = flu_hist,
  manual_labels = manual_labels,
  startWeek     = startWeek,
  best_spec     = best_spec,
  k_ref         = 25L,
  temperature   = 0.25
)

m2_production <- list(fit = m2_production_fit, spec = best_spec)
saveRDS(m2_production, file.path(DATA_DIR, "m2_production.rds"))
cat("Saved m2_production.rds\n")
```

```{r load-m2}
#| message: false

m2_prod <- readRDS(file.path(DATA_DIR, "m2_production.rds"))
cat(sprintf("M2 model loaded: %d coefficients | lead levels: %s\n",
            length(coef(m2_prod$fit)),
            paste(levels(m2_prod$fit$model$lead), collapse = ", ")))
```

### Step 4 — Build and verify the deployment kit

`load_prospective_kit()` bundles all offline-trained artifacts into a single
named list, verifying that required files exist.

```{r build-kit}
kit <- load_prospective_kit(DATA_DIR)

cat(sprintf(paste0(
  "Kit contents:\n",
  "  M1: k_ref=%d, temperature=%.2f, anchorWeek=%d\n",
  "  M0: %d params loaded\n",
  "  M2: %d coefs | best spec: delta=%s, Kr=%s\n"),
  kit$M1_PARAMS$k_ref,
  kit$M1_PARAMS$temperature,
  kit$ref$anchorWeek,
  length(kit$m0_params),
  length(coef(kit$m2_production$fit)),
  kit$best_spec$delta,
  kit$best_spec$Kr
))
```

---

## Part 2 — Deployment (online, every new data week)

Each week, load the current season data and call the three modular pipeline
functions. You can stop at any stage:

- Stop after **M0** to inspect ignition detection only.
- Stop after **M1** for peak timing estimates without forecasts.
- Run all three for full 1- and 2-week-ahead positivity forecasts.

### Get current season data

```{r get-current}
#| message: false
#| warning: false

currentD_full <- getCurrentD(startWeek = startWeek)
currentD_full <- filter(currentD_full, season == "2025-26")

cat(sprintf("Current season: 2025-26 | Weeks observed: %d | weekF range: %d–%d\n",
            nrow(currentD_full),
            min(currentD_full$weekF), max(currentD_full$weekF)))
```

### Step A — M0: ignition detection

`run_m0_detection()` applies the trained M0 thresholds to the current season.
If ignition has not fired automatically, you can pass `manual_ign_week` to
override.

```{r run-m0}
#| message: false

# Optional: set manual_ign_week to an integer to override automatic detection
# e.g. manual_ign_week <- 20L
manual_ign_week <- NA_integer_

m0 <- run_m0_detection(
  kit             = kit,
  current_data    = currentD_full,
  manual_ign_week = manual_ign_week,
  verbose         = TRUE
)

cat(sprintf("\nM0 result: iWeek_locked = %s | overridden = %s\n",
            ifelse(is.na(m0$iWeek_locked), "not detected",
                   as.character(m0$iWeek_locked)),
            m0$overridden))
```

::: {.callout-tip}
**When to override**: If surveillance data are delayed or M0 fires a week
early/late due to a noisy season, set `manual_ign_week` to the known ignition
week before calling `run_m1_alignment()`. The corrected `iWeek_locked` will
propagate to all downstream alignment and forecast covariates.
:::

```{r m0-detection-table}
#| echo: false
#| warning: false

ign_tbl <- m0$ign_out$weekly_flags
if (!is.null(ign_tbl) && nrow(ign_tbl) > 0) {
  ign_show <- ign_tbl |>
    filter(weekF >= max(1, m0$iWeek_locked - 5, na.rm = TRUE)) |>
    select(any_of(c("weekF", "p", "cond_sum", "cond_p", "cond_prev",
                    "cond_inc", "ignited"))) |>
    head(10)
  if (nrow(ign_show) > 0) {
    gt(ign_show) |>
      tab_header(title = "M0 gate flags near ignition") |>
      fmt_number(columns = where(is.numeric), decimals = 3) |>
      tab_options(table.width = pct(70), table.font.size = "13px")
  }
}
```

### Step B — M1: alignment walk-forward

`run_m1_alignment()` takes the M0 result and runs the multi-template ensemble
alignment from ignition week to the last observed week. Each week's partial
curve is aligned to each training season's template; an ensemble peak estimate
is updated as data accumulate.

You may stop here if you only need peak timing (no M2 forecast required).

```{r run-m1}
#| message: false

m1 <- run_m1_alignment(
  kit          = kit,
  current_data = currentD_full,
  m0_result    = m0,
  walk_start   = 5L,
  verbose      = TRUE
)

# Summary
active <- filter(m1$params_df, !is.na(iWeek_hat))
if (nrow(active) > 0) {
  last <- slice_max(active, eval_week, n = 1, with_ties = FALSE)
  cat(sprintf(paste0(
    "M1 walk-forward: %d eval weeks\n",
    "Latest (weekF %d): state=%s | τ=%.2f | δ=%.2f\n",
    "  Est. peak weekF: %s | Peak passed: %s\n"),
    nrow(active),
    last$eval_week,
    last$state,
    last$tau,
    last$delta_m1,
    ifelse(is.na(last$peak_weekF), "?", as.character(last$peak_weekF)),
    ifelse(last$peak_passed, "YES", "no")
  ))
} else {
  cat("M1: alignment not yet active (ignition not detected).\n")
}
```

::: {.callout-tip}
**Using M1 results for peak detection only**: `m1$params_df` contains
`peak_weekF` and `peak_passed` at every evaluation week. If `peak_passed == TRUE`,
the epidemic has passed its peak. No M2 call is needed for this use case.
:::

```{r m1-params-table}
#| echo: false
#| warning: false

if (nrow(active) > 0) {
  gt(active |> select(eval_week, state, iWeek_hat, tau, delta_m1,
                       peak_weekF, t_peak, peak_passed)) |>
    fmt_number(columns = where(is.numeric), decimals = 3) |>
    fmt_integer(columns = c(eval_week, iWeek_hat, peak_weekF)) |>
    tab_style(
      style = cell_fill(color = "#fff0f0"),
      locations = cells_body(rows = peak_passed == TRUE)
    ) |>
    tab_header(title = "M1 alignment parameters — all eval weeks") |>
    tab_options(table.width = pct(100), table.font.size = "13px")
}
```

### Step C — M2: 1- and 2-week-ahead forecast

`run_m2_forecast()` takes the M1 result and predicts h=1 and h=2 positivity
using the production GAM. Extrapolation guards are applied automatically:
- `logN_now` is capped at the training range maximum to prevent spline extrapolation.
- Predicted positivity is soft-capped using a tanh squash above the 95th percentile
  of training positivity.

```{r run-m2}
#| message: false

m2 <- run_m2_forecast(
  kit          = kit,
  current_data = currentD_full,
  m1_result    = m1,
  verbose      = TRUE
)

# Latest forecasts
if (nrow(m2$m2_preds) > 0) {
  latest_ew <- max(m2$m2_preds$eval_week)
  latest_m2 <- filter(m2$m2_preds, eval_week == latest_ew)

  cat(sprintf("\nM2 forecasts at weekF=%d:\n", latest_ew))
  for (i in seq_len(nrow(latest_m2))) {
    r <- latest_m2[i, ]
    cat(sprintf("  h=%d → target weekF %d: M2=%.1f%% [%.1f–%.1f]  |  M1=%.1f%%\n",
                r$h, r$target_weekF,
                r$m2_p * 100, r$m2_lo * 100, r$m2_hi * 100,
                r$m1_p * 100))
  }
}
```

### Current season: alignment + forecast plot

```{r plot-current}
#| echo: false
#| warning: false
#| fig-height: 5

if (nrow(active) > 0 && nrow(m2$m2_preds) > 0) {
  ew           <- max(active$eval_week)
  pw           <- filter(m1$params_df, eval_week == ew)
  iWeek_hat_ew <- pw$iWeek_hat[[1]]
  anchor_ew    <- kit$ref$anchorWeek
  peak_wF      <- pw$peak_weekF[[1]]
  t_peak_ew    <- pw$t_peak[[1]]

  current_newW  <- ew - iWeek_hat_ew + anchor_ew
  ignition_newW <- anchor_ew

  fc <- filter(m1$m1_curves, eval_week == ew, kind == "forecast") |>
    arrange(newWeek)
  junction     <- fc$newWeek[which.min(abs(fc$newWeek - current_newW))]
  fc_past      <- filter(fc, newWeek <= junction)
  fc_future    <- filter(fc, newWeek >= junction)

  obs_s <- currentD_full |> mutate(p_obs = y / N)
  obs_aligned <- obs_s |>
    filter(weekF <= ew) |>
    mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)
  obs_future  <- obs_s |>
    filter(weekF > ew) |>
    mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

  m2_ew <- filter(m2$m2_preds, eval_week == ew)
  m2_aligned <- m2_ew |>
    mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) |>
    arrange(newWeek)

  m2_retro <- m2$m2_preds |>
    filter(eval_week < ew, h == 1) |>
    mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) |>
    filter(newWeek >= ignition_newW) |>
    select(newWeek, m2_p) |>
    arrange(newWeek)

  jv <- obs_aligned$p_obs[obs_aligned$weekF == ew]
  junction_val <- if (length(jv) > 0 && !is.na(jv[1])) jv[1] else NA_real_
  junction_pt  <- tibble(newWeek = current_newW, m2_p = junction_val)

  y_max <- max(c(fc$p_hat, obs_s$p_obs, m2_ew$m2_hi), na.rm = TRUE) * 1.15
  y_max <- max(y_max, 0.15)

  newW_grid <- seq(1, 52, by = 0.5)
  ref_curve_df <- tibble(
    newWeek = newW_grid,
    p_ref   = plogis(kit$ref$g_ref_fun(newW_grid))
  )

  plt <- plot_ly() |>
    add_lines(data = ref_curve_df, x = ~newWeek, y = ~p_ref,
              line = list(color = "rgba(160,160,160,0.4)", width = 1),
              name = "Reference") |>
    add_lines(data = fc_past, x = ~newWeek, y = ~p_hat,
              line = list(color = "steelblue", width = 2),
              name = "M1 fitted") |>
    add_lines(data = fc_future, x = ~newWeek, y = ~p_hat,
              line = list(color = "steelblue", width = 2, dash = "dash"),
              name = "M1 forecast") |>
    add_markers(data = obs_aligned, x = ~newWeek, y = ~p_obs,
                marker = list(color = "black", size = 5),
                name = "Observed") |>
    add_markers(data = obs_future, x = ~newWeek, y = ~p_obs,
                marker = list(color = "black", size = 5, opacity = 0.35,
                              symbol = "circle-open"),
                name = "Future obs") |>
    add_segments(x = anchor_ew, xend = anchor_ew, y = 0, yend = y_max,
                 line = list(color = "grey50", dash = "dash", width = 1),
                 name = paste0("Ignition wk ", iWeek_hat_ew)) |>
    add_segments(x = current_newW, xend = current_newW, y = 0, yend = y_max,
                 line = list(color = "black", width = 2),
                 name = paste0("Now wk ", ew))

  m2_past_line <- bind_rows(m2_retro, junction_pt) |> arrange(newWeek)
  if (nrow(m2_past_line) > 1) {
    plt <- plt |>
      add_lines(data = m2_past_line, x = ~newWeek, y = ~m2_p,
                line = list(color = "firebrick", width = 2),
                name = "M2 fitted")
  }
  m2_future_line <- bind_rows(junction_pt,
                               m2_aligned |> select(newWeek, m2_p))
  plt <- plt |>
    add_lines(data = m2_future_line, x = ~newWeek, y = ~m2_p,
              line = list(color = "firebrick", width = 2, dash = "dash"),
              name = "M2 forecast") |>
    add_markers(data = m2_aligned, x = ~newWeek, y = ~m2_p,
                marker = list(color = "firebrick", size = 7, symbol = "diamond"),
                name = "M2 pts", showlegend = FALSE)

  if (!is.na(t_peak_ew)) {
    plt <- plt |>
      add_segments(x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
                   line = list(color = "green", dash = "dash", width = 1.5),
                   name = paste0("Est. peak wk ", peak_wF))
  }

  plt |> layout(
    title = paste0("Current state: weekF=", ew,
                   " | state=", pw$state[[1]],
                   if (!is.na(peak_wF)) paste0(" | peak wk ", peak_wF) else ""),
    xaxis = list(title = "Aligned week (newWeek)"),
    yaxis = list(title = "Positivity", range = c(0, y_max)),
    legend = list(orientation = "h", y = -0.20)
  )
}
```

---

## Full pipeline in three lines

Once trained artifacts exist in `data/`, the entire deployment for a new week
reduces to:

```{r full-pipeline-example}
#| eval: false

kit <- load_prospective_kit(DATA_DIR)
current_data <- getCurrentD(startWeek = 27L) |> filter(season == "2025-26")

# --- Option A: run everything at once ---
wf <- run_prospective_pipeline(kit, current_data)

# --- Option B: step by step (allows inspection/override between stages) ---
m0 <- run_m0_detection(kit, current_data)
# Optionally: m0$ign_out$ign_week_locked <- 20L  # manual override

m1 <- run_m1_alignment(kit, current_data, m0_result = m0)
# Optionally stop here — m1$params_df has peak estimates

m2 <- run_m2_forecast(kit, current_data, m1_result = m1)
# m2$m2_preds has h=1 and h=2 forecasts for every eval week
```

---

## Forecast accuracy

```{r accuracy}
#| echo: false
#| warning: false

if (nrow(m2$m2_preds) > 0) {
  obs_lookup <- currentD_full |>
    transmute(target_weekF = as.integer(weekF), p_obs = y / N)

  acc <- m2$m2_preds |>
    filter(!is.na(m2_p)) |>
    left_join(obs_lookup, by = "target_weekF") |>
    filter(!is.na(p_obs)) |>
    mutate(
      m2_ae = abs(m2_p - p_obs),
      m1_ae = abs(m1_p - p_obs),
      m2_covers = (p_obs >= m2_lo) & (p_obs <= m2_hi)
    )

  if (nrow(acc) > 0) {
    acc |>
      group_by(h) |>
      summarise(
        n        = n(),
        MAE_M2   = mean(m2_ae),
        MAE_M1   = mean(m1_ae),
        Coverage = mean(m2_covers),
        .groups  = "drop"
      ) |>
      gt() |>
      tab_header(title = "Forecast accuracy: M2 vs M1 (realised weeks only)") |>
      fmt_number(columns = c(MAE_M2, MAE_M1), decimals = 4) |>
      fmt_percent(columns = Coverage, decimals = 0) |>
      tab_options(table.width = pct(60), table.font.size = "14px")
  } else {
    cat("No realised observations at target weeks yet.\n")
  }
}
```
