---
title: "PAGe: A Practical Guide"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    page-layout: full
    grid:
      body-width: 1100px
---

```{=html}
<style>
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 1.6em;
  padding: 0.35em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
.callout-note    { border-left: 4px solid #4a90d9; background: #f4f8fd; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
.callout-tip     { border-left: 4px solid #28a745; background: #f0fff4; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
.callout-warning { border-left: 4px solid #ffc107; background: #fffbf0; padding: 0.8em 1em; margin: 1em 0; border-radius: 0 4px 4px 0; }
table { border-collapse: collapse; width: 100%; }
th, td { padding: 0.45em 0.75em; border: 1px solid #ddd; }
th { background: #f0f4f8; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

PAGe (**Phase-Aligned Gated Epidemic Forecasting**) is an R package for
producing 1–2 week ahead forecasts of seasonal respiratory virus positivity.
This guide walks through the complete workflow using Canadian influenza
surveillance data.

The workflow has two stages:

- **Stage 1 — Training** (run once): fit and tune three model components
  (M0, M1, M2) on historical seasons and assemble a deployment kit.
- **Stage 2 — Deployment** (run each week): load the kit and call one function
  to produce forecasts from current-season data.

<div class="callout-warning">
**Stage 1 takes 1–3 hours on first run** because it fits M1 alignment
for every LOSO fold. Run `scripts/train_walkthrough_demo.R` in the background
before rendering this document. Once the cache exists at
`data/walkthrough/`, this document renders instantly.
</div>

---

## Data format

The pipeline needs one row per (season, MMWR week) with these columns:

| Column   | Description |
|----------|-------------|
| `season` | Season label, e.g. `"2019-20"` |
| `weekF`  | Within-season week index (1 = first week after the season start) |
| `y`      | Positive tests |
| `N`      | Total tests |
| `neg`    | Negative tests (`N − y`) |
| `p`      | Positivity fraction (`y / N`) |

The helper below converts the repository's raw surveillance CSV to that
format. For your own data, write an equivalent function and drop it in at the
top of `scripts/train_walkthrough_demo.R`.

```{r load_fn, eval=FALSE}
library(PAGe)
library(MMWRweek)
library(dplyr)

START_WEEK <- 27L   # MMWR week that begins each season

n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(
    MMWRweek::MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L
  )
}

load_flu_data <- function(path = "data/flu_testing_data.csv") {
  read.csv(path) |>
    dplyr::select(
      season, week, start_year = seasonstart,
      y = pos_flua, N = test_flu
    ) |>
    dplyr::mutate(
      neg   = N - y,
      nW    = n_weeks_in_start_year(start_year),
      weekF = ((week - START_WEEK) %% nW) + 1L,
      p     = y / N
    )
}
```

---

## Stage 1 — Training

Stage 1 code is shown below **but marked `eval=FALSE`** because the computations
take 1–3 hours and are meant to run once via
`scripts/train_walkthrough_demo.R`.  The output of each step is cached;
Stage 2 below loads those cached artifacts.

### Season roles

We use every other season for training and hold out 2025–26 for deployment.

```{r season_roles, eval=FALSE}
TRAIN_SEASONS <- c("2012-13", "2014-15", "2017-18", "2019-20", "2023-24")
PROSP_SEASON  <- "2025-26"

allD       <- load_flu_data()
train_allD <- allD |> dplyr::filter(season %in% TRAIN_SEASONS)
prosp_data <- allD |> dplyr::filter(season == PROSP_SEASON)
```

### M0 — Ignition detection

```{r m0, eval=FALSE}
CACHE_DIR <- "data/walkthrough"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

# Tunes ignition-detection thresholds via leave-one-season-out grid search.
# Saves ~28 KB; runs in < 5 minutes.
m0 <- tune_m0(
  allD         = train_allD,
  loso_seasons = "all",
  exclude      = character(0)
)
saveRDS(m0, file.path(CACHE_DIR, "m0.rds"))
```

### M1 — Phase alignment

```{r m1, eval=FALSE}
# Fits the epidemic reference curve and alignment hyperparameters.
# Saves ~2.4 MB; runs in < 2 minutes.
m1 <- build_m1(
  allD    = train_allD,
  m0      = m0,
  exclude = character(0)
)
saveRDS(m1, file.path(CACHE_DIR, "m1.rds"))
```

### M2 — Leave-one-season-out evaluation

```{r m2_loso, eval=FALSE}
# LOSO grid search over M2 hyperparameters (k_f, k_e, alpha_state, k_sp, …).
# Phase 1 builds M1 fold predictions (~20 min per fold × 5 folds = ~100 min).
# Phase 2 fits M2 GAMs across the grid (~5 min for a small grid).
# Use default_m2_grid() for a full 480-spec production search; the demo grid
# below finishes in ~30 minutes and is sufficient to illustrate the API.
demo_grid <- tidyr::crossing(
  delta       = 0,
  Kr          = 1,
  k_f         = c(4L, 6L),
  k_e         = 2L,
  alpha_state = 0.15,
  k_r         = 0L,
  k_de        = 0L,
  k_sp        = 6L
)

loso <- build_m2(
  allD         = train_allD,
  m0           = m0,
  m1           = m1,
  loso_seasons = "all",
  exclude_seas = character(0),
  grid         = demo_grid
)
saveRDS(loso, file.path(CACHE_DIR, "m2_loso.rds"))
```

<div class="callout-note">
The M1 fold-building phase (Phase 1 inside `build_m2`) is the bottleneck: it
runs M1 walk-forward predictions for each LOSO fold.  On the first run this
takes ~20 minutes per fold.  Subsequent renders load `m2_loso.rds` instantly.
</div>

### M2 — Final model training

```{r m2_train, eval=FALSE}
m2_model <- train_m2(
  allD      = train_allD,
  m0        = m0,
  m1        = m1,
  best_spec = loso$best_spec,
  exclude   = character(0),
  verbose   = TRUE
)
saveRDS(m2_model, file.path(CACHE_DIR, "m2_model.rds"))
```

### Assemble the deployment kit

```{r kit_assemble, eval=FALSE}
kit <- assemble_kit(
  m0       = m0,
  m1       = m1,
  m2_model = m2_model
)
saveRDS(kit, file.path(CACHE_DIR, "kit.rds"))
```

<div class="callout-tip">
**Stage 1 complete.** The kit at `data/walkthrough/kit.rds` contains
everything needed for deployment. Hand this single file to Stage 2 — no
training data needs to travel with it.
</div>

---

## Stage 2 — Deployment

Stage 2 is what runs every week during the active season. It loads the kit and
processes new surveillance data through M0 → M1 → M2 in one call.

Here we demonstrate on **2025–26** (held out from training throughout Stage 1).

```{r stage2_setup}
library(PAGe)
library(MMWRweek)
library(dplyr)
library(ggplot2)

START_WEEK <- 27L
n_weeks_in_start_year <- function(start_year) {
  52L + as.integer(
    MMWRweek::MMWRweek(as.Date(paste0(start_year, "-12-31")))$MMWRweek == 53L
  )
}
load_flu_data <- function(path = "data/flu_testing_data.csv") {
  read.csv(path) |>
    dplyr::select(season, week, start_year = seasonstart,
                  y = pos_flua, N = test_flu) |>
    dplyr::mutate(neg = N - y, nW = n_weeks_in_start_year(start_year),
                  weekF = ((week - START_WEEK) %% nW) + 1L, p = y / N)
}
```

### Load kit and deployment data

```{r load_kit}
# In a real workflow: kit <- readRDS("data/walkthrough/kit.rds")
# For this demo we load the production kit (same structure as assemble_kit() output).
kit <- load_prospective_kit("data/")

allD       <- load_flu_data()
prosp_data <- allD |> dplyr::filter(season == "2025-26")

cat("Training seasons  :", paste(kit$m2_production$training_seasons, collapse = ", "), "\n")
cat("Deployment season : 2025-26\n")
cat("Data rows         :", nrow(prosp_data), "\n")
```

### Walk-forward simulation

`run_prospective_pipeline()` simulates week-by-week deployment: at each week
it uses only data available up to that point, then forecasts 1 and 2 weeks
ahead.

```{r deployment}
CACHE_DIR <- "data/walkthrough"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)
wf_cache  <- file.path(CACHE_DIR, "wf.rds")

if (file.exists(wf_cache)) {
  wf <- readRDS(wf_cache)
} else {
  wf <- run_prospective_pipeline(
    kit          = kit,
    current_data = prosp_data,
    mode         = "frozen",
    verbose      = FALSE
  )
  saveRDS(wf, wf_cache)
}

cat("Forecast rows :", nrow(wf$m2_preds), "\n")
cat("Horizons      :", paste(sort(unique(wf$m2_preds$h)), collapse = ", "), "weeks\n")
```

### Forecast plot

```{r forecast_plot, fig.width=11, fig.height=5}
fc1 <- wf$m2_preds |> dplyr::filter(h == 1L)
obs <- prosp_data |> dplyr::mutate(pct = y / N)

ggplot() +
  geom_ribbon(
    data = fc1,
    aes(x = target_weekF, ymin = m2_lo, ymax = m2_hi),
    fill = "#4a90d9", alpha = 0.25
  ) +
  geom_line(
    data = fc1,
    aes(x = target_weekF, y = m2_p, colour = "Forecast (h=1)"),
    linewidth = 1
  ) +
  geom_point(
    data = obs,
    aes(x = weekF, y = pct, colour = "Observed"),
    size = 2
  ) +
  scale_colour_manual(
    values = c("Forecast (h=1)" = "#1e5f8e", "Observed" = "#333333")
  ) +
  labs(
    title  = "Stage 2 walk-forward deployment — 2025-26",
    x      = "Season week",
    y      = "Positivity",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Accuracy summary

```{r accuracy}
obs_join <- prosp_data |> dplyr::select(weekF, obs = p)

wf$m2_preds |>
  dplyr::inner_join(obs_join, by = c("target_weekF" = "weekF")) |>
  dplyr::group_by(h) |>
  dplyr::summarise(
    n    = dplyr::n(),
    mae  = mean(abs(m2_p - obs), na.rm = TRUE),
    rmse = sqrt(mean((m2_p - obs)^2, na.rm = TRUE)),
    bias = mean(m2_p - obs, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 4, caption = "Walk-forward accuracy — 2025-26")
```

<div class="callout-tip">
**That's it.**  Each week append the new observation row to `prosp_data` and
call `run_prospective_pipeline()` again.  The kit never changes during the
season.
</div>

---

## Using your own data

To apply PAGe to a different virus or geography:

1. **Prepare your data** in the format shown at the top of this document.
2. **Edit** `TRAIN_SEASONS` in `scripts/train_walkthrough_demo.R` to list the
   seasons you want for training, and set `PROSP_SEASON` to the current season.
3. **Run** `Rscript scripts/train_walkthrough_demo.R` once to build Stage 1
   artifacts. This produces `data/walkthrough/kit.rds`.
4. **Deploy** — load `kit.rds` and call `run_prospective_pipeline(kit, new_data)`.

| Column   | Type      | Note |
|----------|-----------|------|
| `season` | character | e.g. `"2022-23"` |
| `weekF`  | integer   | 1 = first week after season start |
| `y`      | integer   | positive tests |
| `N`      | integer   | total tests |
| `neg`    | integer   | `N − y` |
| `p`      | numeric   | `y / N` |

<div class="callout-warning">
**Minimum seasons.** M1 needs at least 5 seasons to build a stable reference
curve; M2 LOSO requires at least 4 distinct folds. With fewer seasons, results
will be noisier.
</div>
