---
title: "Prospective Deployment Pipeline"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    page-layout: full
    grid:
      body-width: 1200px
---

```{=html}
<style>
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

::: {.callout-important}
## What This Page Shows

- The live walk-forward pipeline for the current season.
- The current state of ignition, alignment, and forecast outputs.
- How weekly `$M_2$` forecasts evolve as more 2025-26 data arrive.

This page is the runtime view. Training details live in
[`estimateRef.html`](estimateRef.html),
[`loso_walkforward.html`](loso_walkforward.html), and
[`forecast_training.html`](forecast_training.html).
:::

## Runtime Setup

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(MMWRweek)
library(plotly)
library(gt)

# Source all pipeline functions
r_dir <- normalizePath("../R", mustWork = TRUE)
r_files <- c(
  "utils.R", "m0_retro.R", "m1_reference_helpers.R", "m1_reference.R",
  "m1_hyperparams.R", "m1_fit.R", "identifiability.R",
  "check_scale_identifiability.R", "m1_runtime.R",
  "align_forecast_pipeline_dilate.R", "m1_multi_template.R",
  "m1_peak_status.R",
  "m1_peak_flags.R", "m1_peak_summary.R", "m0_training.R", "m1_loso.R",
  "m2_training.R", "pipeline_runtime_helpers.R",
  "m0_runtime.R", "m2_runtime.R", "pipeline_runtime.R",
  "forecast_post_peak_gam.R", "m2_spec_grid.R",
  "pipeline_bridge.R", "m2_nested_loso.R", "getCurrentD.R"
)
for (f in r_files) {
  fp <- file.path(r_dir, f)
  if (file.exists(fp)) source(fp)
}
source(file.path(normalizePath("../flualign/R", mustWork = TRUE), "flagIgnition.R"))

startWeek  <- 27L
WALK_START <- 5L
DATA_DIR   <- normalizePath("../data", mustWork = TRUE)
WF_CACHE   <- file.path(DATA_DIR, "deploy_wf_cache.rds")

# ---- Cache control ----
# Set FORCE_REFRESH = TRUE to rerun the walk-forward (e.g., after new data
# arrive, changing MANUAL_IGN_WEEK, or updating pipeline code).
# The cache does NOT auto-invalidate.
FORCE_REFRESH <- FALSE

# ---- Manual ignition override ----
# Set to an integer weekF to override M0 detection (e.g., 20L).
# Set to NA_integer_ to use automatic M0 detection.
MANUAL_IGN_WEEK <- NA_integer_
```

## Inputs

All offline-trained components (reference curve, $M_1$ hyperparameters, $M_0$ thresholds,
$M_2$ model) are loaded from `data/` via `load_prospective_kit()`.

```{r load-kit}
#| message: false

kit <- load_prospective_kit(DATA_DIR)

cat(sprintf(
  "Kit loaded:\n  k_ref=%d | temperature=%.2f\n  M_2 coefficients: %d\n  anchorWeek: %d\n",
  kit$M1_PARAMS$k_ref,
  kit$M1_PARAMS$temperature,
  length(coef(kit$m2_production$fit)),
  kit$ref$anchorWeek
))
```

### Current season data

```{r get-current}
#| message: false
#| warning: false

currentD_full <- getCurrentD(startWeek = startWeek)
currentD_full <- filter(currentD_full, season == "2025-26")

cat(sprintf("Season: 2025-26 | Weeks observed: %d | weekF range: %d–%d\n",
            nrow(currentD_full),
            min(currentD_full$weekF), max(currentD_full$weekF)))
```

## Walk-forward Results: $M_0$ → $M_1$ → $M_2$

For each evaluation week $w$, the pipeline runs sequentially:

1. **$M_0$ — Ignition detection**: `run_ignition_weekly()` determines whether the season has ignited. Once locked, $\hat{w}^{(\text{ign})}$ is fixed.
2. **$M_1$ — Alignment**: `run_alignment_prospective_multi()` fits $\tau$, $\delta$, $a$, $b$ to align the partial curve to reference templates, producing a peak estimate and forecast curve.
3. **$M_2$ — Forecast**: The production GAM predicts $\hat{p}_{w+1}$ and $\hat{p}_{w+2}$ using $M_1$'s aligned prediction as a stacking covariate.

```{r walk-forward}
#| message: false
#| warning: false

if (file.exists(WF_CACHE) && !isTRUE(FORCE_REFRESH)) {
  wf  <- readRDS(WF_CACHE)
  ref <- kit$ref
  message("Loaded walk-forward from cache. Set FORCE_REFRESH = TRUE to rerun.")
} else {
  wf <- run_prospective_pipeline(
    kit             = kit,
    current_data    = currentD_full,
    walk_start      = WALK_START,
    manual_ign_week = MANUAL_IGN_WEEK,
    verbose         = TRUE
  )
  ref <- kit$ref
  saveRDS(wf, WF_CACHE)
  message("Walk-forward complete. Saved to ", WF_CACHE)
}
```


## Current Snapshot

```{r current-state}
#| echo: false

last_row <- wf$params_df %>%
  filter(!is.na(iWeek_hat)) %>%
  slice_max(eval_week, n = 1, with_ties = FALSE)

if (nrow(last_row) > 0) {
  latest_m2 <- wf$m2_preds %>%
    filter(eval_week == last_row$eval_week)

  cat(sprintf(paste0(
    "State: %s | Ignition: weekF %d | Latest eval: weekF %d\n",
    "Est. peak: weekF %s | Peak passed: %s\n"),
    last_row$state, last_row$iWeek_hat, last_row$eval_week,
    ifelse(is.na(last_row$peak_weekF), "?", as.character(last_row$peak_weekF)),
    ifelse(last_row$peak_passed, "YES", "no")
  ))

  if (nrow(latest_m2) > 0) {
    cat("\nLatest M_2 forecasts:\n")
    for (i in seq_len(nrow(latest_m2))) {
      r <- latest_m2[i, ]
      cat(sprintf("  h=%d → target weekF %d: M_2=%.1f%% [%.1f, %.1f] | M_1=%.1f%%\n",
                  r$h, r$target_weekF,
                  r$m2_p * 100, r$m2_lo * 100, r$m2_hi * 100,
                  r$m1_p * 100))
    }
  }
} else {
  cat("No alignment yet — ignition not detected.\n")
}
```


## $M_1$ alignment parameters over time

```{r param-evolution}
#| echo: false
#| warning: false

df_params <- wf$params_df %>% filter(!is.na(iWeek_hat))

if (nrow(df_params) > 0) {
  param_vars <- c("tau", "delta_m1", "a", "b", "t_peak")
  param_labels <- c(
    tau      = "τ — time shift",
    delta_m1 = "δ — dilation",
    a        = "a — intercept",
    b        = "b — slope",
    t_peak   = "t_peak — est. peak (aligned)"
  )

  make_param_plot <- function(pvar) {
    df <- df_params %>% select(eval_week, value = all_of(pvar))
    plot_ly(df, x = ~eval_week, y = ~value,
            type = "scatter", mode = "lines+markers",
            line = list(color = "steelblue", width = 1.5),
            marker = list(size = 5, color = "steelblue")) %>%
      layout(xaxis = list(title = "Eval weekF"),
             yaxis = list(title = param_labels[[pvar]]))
  }
}
```

::: {.panel-tabset}

```{r param-tabs}
#| echo: false
#| warning: false
#| results: asis

if (nrow(df_params) > 0) {
  tabs <- lapply(param_vars, function(pvar) {
    e <- new.env(parent = globalenv())
    e$p <- make_param_plot(pvar)
    knitr::knit_child(
      text = c(sprintf("### %s", param_labels[[pvar]]), "",
               "```{r echo=FALSE, warning=FALSE}", "p", "```", ""),
      envir = e, quiet = TRUE
    )
  })
  cat(unlist(tabs), sep = "\n")
} else {
  cat("### No data\n\nIgnition not yet detected.\n")
}
```

:::


## $M_2$ forecast evolution

Each tab shows the $M_2$ 2-week-ahead forecast at a given evaluation week, overlaid
with $M_1$ alignment and observed data. Blue = $M_1$ (solid: fitted, dashed: forecast).
Red = $M_2$ (solid: past h=1 predictions, dashed: current forecast). Green dashed =
estimated peak. Vertical black line = current evaluation week.

```{r m2-forecast-plots}
#| echo: false
#| warning: false
#| results: asis

df_active <- wf$params_df %>% filter(!is.na(iWeek_hat))

if (nrow(df_active) == 0) {
  cat("*No forecasts yet — ignition not detected.*\n\n")
} else {
  obs_s <- currentD_full %>% mutate(p_obs = y / N)

  newW_grid <- seq(1, 52, by = 0.5)
  ref_curve <- tibble(
    newWeek = newW_grid,
    p_ref   = plogis(ref$g_ref_fun(newW_grid))
  )

  cat("::: {.panel-tabset}\n\n")

  tabs <- lapply(df_active$eval_week, function(ew) {
    pw <- wf$params_df %>% filter(eval_week == ew)
    iWeek_hat_ew <- pw$iWeek_hat[[1]]
    anchor_ew    <- ref$anchorWeek
    state_ew     <- pw$state[[1]]
    peak_wF      <- pw$peak_weekF[[1]]
    t_peak_ew    <- pw$t_peak[[1]]

    current_newW  <- ew - iWeek_hat_ew + anchor_ew
    ignition_newW <- anchor_ew  # ignition always maps to anchorWeek

    fc <- wf$m1_curves %>%
      filter(eval_week == ew, kind == "forecast") %>%
      arrange(newWeek)

    if (nrow(fc) == 0) return(NULL)

    junction  <- fc$newWeek[which.min(abs(fc$newWeek - current_newW))]
    fc_past   <- fc %>% filter(newWeek <= junction)
    fc_future <- fc %>% filter(newWeek >= junction)

    obs_aligned <- obs_s %>%
      filter(weekF <= ew) %>%
      mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

    m2_ew <- wf$m2_preds %>% filter(eval_week == ew)

    y_max <- max(c(fc$p_hat, obs_aligned$p_obs, m2_ew$m2_hi, obs_s$p_obs), na.rm = TRUE) * 1.15
    y_max <- max(y_max, 0.15)

    obs_future <- obs_s %>%
      filter(weekF > ew) %>%
      mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

    plt <- plot_ly() %>%
      add_lines(
        data = ref_curve, x = ~newWeek, y = ~p_ref,
        line = list(color = "rgba(160,160,160,0.4)", width = 1),
        name = "Reference"
      ) %>%
      add_lines(
        data = fc_past, x = ~newWeek, y = ~p_hat,
        line = list(color = "steelblue", width = 2),
        name = "M_1 fitted"
      ) %>%
      add_lines(
        data = fc_future, x = ~newWeek, y = ~p_hat,
        line = list(color = "steelblue", width = 2, dash = "dash"),
        name = "M_1 forecast"
      ) %>%
      add_markers(
        data = obs_aligned, x = ~newWeek, y = ~p_obs,
        marker = list(color = "black", size = 5, opacity = 0.75),
        name = "Observed"
      ) %>%
      add_markers(
        data = obs_future, x = ~newWeek, y = ~p_obs,
        marker = list(color = "black", size = 5, opacity = 0.35,
                      symbol = "circle-open"),
        name = "Future obs"
      ) %>%
      add_segments(
        x = anchor_ew, xend = anchor_ew, y = 0, yend = y_max,
        line = list(color = "grey50", dash = "dash", width = 1),
        name = paste0("Ignition (wk ", iWeek_hat_ew, ")")
      ) %>%
      add_segments(
        x = current_newW, xend = current_newW, y = 0, yend = y_max,
        line = list(color = "black", width = 2),
        name = paste0("Now (wk ", ew, ")")
      )

    # M2 line: solid (past h=1 realized) + dashed (current forecast)
    # Both start at ignition week to mirror M1
    if (nrow(m2_ew) > 0) {
      m2_aligned <- m2_ew %>%
        mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) %>%
        arrange(newWeek)

      # Retrospective M2: h=1 predictions from past eval_weeks, starting at ignition
      m2_retro <- wf$m2_preds %>%
        filter(eval_week < ew, h == 1) %>%
        mutate(newWeek = target_weekF - iWeek_hat_ew + anchor_ew) %>%
        filter(newWeek >= ignition_newW) %>%
        select(newWeek, m2_p) %>%
        arrange(newWeek)

      junction_val <- obs_aligned$p_obs[obs_aligned$weekF == ew]
      junction_val <- if (length(junction_val) > 0 && !is.na(junction_val[1]))
        junction_val[1] else NA_real_
      junction_pt  <- tibble(newWeek = current_newW, m2_p = junction_val)

      # Solid red: ignition → past realized predictions → "now"
      m2_past_line <- bind_rows(m2_retro, junction_pt) %>% arrange(newWeek)
      if (nrow(m2_past_line) > 1) {
        plt <- plt %>%
          add_lines(
            data = m2_past_line, x = ~newWeek, y = ~m2_p,
            line = list(color = "firebrick", width = 2, dash = "solid"),
            name = "M_2 fitted"
          )
      }

      # Dashed red: "now" → h=1, h=2 forecast
      m2_future_line <- bind_rows(junction_pt, m2_aligned %>% select(newWeek, m2_p))
      plt <- plt %>%
        add_lines(
          data = m2_future_line, x = ~newWeek, y = ~m2_p,
          line = list(color = "firebrick", width = 2, dash = "dash"),
          name = "M_2 forecast"
        ) %>%
        add_markers(
          data = m2_aligned, x = ~newWeek, y = ~m2_p,
          marker = list(color = "firebrick", size = 7, symbol = "diamond"),
          name = "M_2 forecast pts",
          showlegend = FALSE
        )
    }

    if (!is.na(t_peak_ew)) {
      plt <- plt %>%
        add_segments(
          x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
          line = list(color = "green", dash = "dash", width = 1.5),
          name = paste0("Est. peak (wk ", peak_wF, ")")
        )
    }

    plt <- plt %>%
      layout(
        title = paste0("weekF=", ew, " | state=", state_ew,
                       if (!is.na(peak_wF)) paste0(" | peak wk ", peak_wF) else ""),
        xaxis = list(title = "Aligned week (newWeek)"),
        yaxis = list(title = "Positivity", range = c(0, y_max)),
        legend = list(orientation = "h", y = -0.20)
      )

    e <- new.env(parent = globalenv())
    e$plt <- plt
    knitr::knit_child(
      text = c(sprintf("### wk %d", ew), "",
               "```{r echo=FALSE, warning=FALSE}", "plt", "```", ""),
      envir = e, quiet = TRUE
    )
  })

  valid_tabs <- Filter(Negate(is.null), tabs)
  cat(unlist(valid_tabs), sep = "\n")
  cat("\n:::\n\n")
}
```


## $M_2$ forecast accuracy

Comparing $M_2$ forecasts against realised observations at target weeks (when available).

```{r forecast-accuracy}
#| echo: false
#| warning: false

if (nrow(wf$m2_preds) > 0) {
  obs_lookup <- currentD_full %>%
    transmute(target_weekF = as.integer(weekF), p_obs = y / N)

  acc <- wf$m2_preds %>%
    filter(!is.na(m2_p)) %>%
    left_join(obs_lookup, by = "target_weekF") %>%
    filter(!is.na(p_obs)) %>%
    mutate(
      m2_err = m2_p - p_obs,
      m1_err = m1_p - p_obs,
      m2_ae  = abs(m2_err),
      m1_ae  = abs(m1_err),
      m2_covers = (p_obs >= m2_lo) & (p_obs <= m2_hi)
    )

  if (nrow(acc) > 0) {
    summary_acc <- acc %>%
      group_by(h) %>%
      summarise(
        n        = n(),
        MAE_M2   = mean(m2_ae),
        MAE_M1   = mean(m1_ae),
        RMSE_M2  = sqrt(mean(m2_err^2)),
        RMSE_M1  = sqrt(mean(m1_err^2)),
        Coverage = mean(m2_covers),
        .groups  = "drop"
      )

    gt(summary_acc) %>%
      tab_header(title = "Forecast accuracy: M_2 vs M_1 (realised weeks only)") %>%
      fmt_number(columns = c(MAE_M2, MAE_M1, RMSE_M2, RMSE_M1), decimals = 4) %>%
      fmt_percent(columns = Coverage, decimals = 0) %>%
      tab_options(table.width = pct(80), table.font.size = "14px")
  } else {
    cat("No realised observations at target weeks yet.\n")
  }
} else {
  cat("No M_2 predictions available.\n")
}
```

### Forecast time series

```{r forecast-timeseries}
#| echo: false
#| warning: false

if (nrow(wf$m2_preds) > 0 && any(!is.na(wf$m2_preds$m2_p))) {
  obs_lookup <- currentD_full %>%
    transmute(target_weekF = as.integer(weekF), p_obs = y / N)

  m2_ts <- wf$m2_preds %>%
    filter(!is.na(m2_p)) %>%
    left_join(obs_lookup, by = "target_weekF")

  m2_h1 <- filter(m2_ts, h == 1)
  m2_h2 <- filter(m2_ts, h == 2)

  plt_ts <- plot_ly() %>%
    add_markers(
      data = obs_lookup, x = ~target_weekF, y = ~p_obs,
      marker = list(color = "black", size = 5),
      name = "Observed"
    )

  if (nrow(m2_h1) > 0) {
    plt_ts <- plt_ts %>%
      add_ribbons(
        data = m2_h1, x = ~target_weekF, ymin = ~m2_lo, ymax = ~m2_hi,
        fillcolor = "rgba(70,130,180,0.15)",
        line = list(color = "transparent"),
        name = "h=1 CI", showlegend = FALSE
      ) %>%
      add_lines(
        data = m2_h1, x = ~target_weekF, y = ~m2_p,
        line = list(color = "steelblue", width = 2),
        name = "M_2 h=1"
      )
  }

  if (nrow(m2_h2) > 0) {
    plt_ts <- plt_ts %>%
      add_ribbons(
        data = m2_h2, x = ~target_weekF, ymin = ~m2_lo, ymax = ~m2_hi,
        fillcolor = "rgba(178,34,34,0.12)",
        line = list(color = "transparent"),
        name = "h=2 CI", showlegend = FALSE
      ) %>%
      add_lines(
        data = m2_h2, x = ~target_weekF, y = ~m2_p,
        line = list(color = "firebrick", width = 2),
        name = "M_2 h=2"
      )
  }

  plt_ts <- plt_ts %>%
    layout(
      title = "M_2 forecast time series vs observed",
      xaxis = list(title = "Target weekF"),
      yaxis = list(title = "Positivity"),
      legend = list(orientation = "h", y = -0.15)
    )

  plt_ts
}
```


## Walk-forward Data Table

```{r data-table}
#| echo: false
#| warning: false

if (nrow(wf$params_df) > 0) {
  tbl_data <- wf$params_df %>%
    filter(!is.na(iWeek_hat)) %>%
    select(eval_week, state, iWeek_hat, tau, delta_m1,
           peak_weekF, t_peak, peak_passed, fallback)

  if (nrow(wf$m2_preds) > 0) {
    m2_wide <- wf$m2_preds %>%
      filter(!is.na(m2_p)) %>%
      select(eval_week, h, m2_p) %>%
      mutate(h_label = paste0("m2_h", h)) %>%
      select(-h) %>%
      pivot_wider(names_from = h_label, values_from = m2_p)

    tbl_data <- left_join(tbl_data, m2_wide, by = "eval_week")
  }

  gt(tbl_data) %>%
    fmt_number(columns = where(is.numeric), decimals = 3) %>%
    fmt_integer(columns = c(eval_week, iWeek_hat, peak_weekF)) %>%
    tab_style(
      style = cell_fill(color = "#fff0f0"),
      locations = cells_body(rows = peak_passed == TRUE)
    ) %>%
    tab_header(title = "Walk-forward alignment + forecast parameters") %>%
    tab_options(table.width = pct(100), table.font.size = "13px")
}
```
