---
title: "PAGe Real-Time Deployment Walkthrough"
subtitle: "Official ORVT input, reproducible forecasting, and the weekly HTTP API"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-copy: true
execute:
  eval: false
---

This guide covers the operational path after a PAGe model kit has been frozen or promoted. It presents two supported user patterns:

1. **direct R forecasting** with a validated kit and official surveillance input;
2. **service/API use** through the weekly v4 HTTP API.

::: {.callout-important title="Current deployment status"}
The checked-in weekly HTTP API is a governed **shadow-only** service. Its deployment is pinned to an immutable forecast release. Do not substitute a different release ID in an API request; discover the release from `/v1/release` and use that exact value.
:::

## 1. Direct R: load a frozen kit

For a normal frozen kit saved by the package:

```{r}
library(PAGe)

kit <- page_load_kit("/secure/PAGe/page_kit.rds")
validate_page_kit(kit, mode = "frozen")
```

For a kit promoted through the immutable registry, load both the private kit and its disclosure-safe deployment manifest:

```{r}
kit <- load_promoted_kit(
  kit_path = "/secure/PAGe/deployment-registry/<deployment-id>/promoted_kit.rds",
  deployment_manifest_path =
    "results/deployment-audit/<deployment-id>/deployment_manifest.json"
)
```

## 2. Direct R: ingest current official surveillance

### Live PHO ORVT feed

`getCurrentD()` is the lowest-level ORVT reader. For 2026–27 it resolves the supported PHO filename convention, validates the PHO schema, aggregates the provincial counts, converts surveillance week to PAGe `weekF`, and records source provenance.

```{r}
current_a <- getCurrentD(
  virus = "Influenza A",
  season = "2026-27",
  include_predecessor = FALSE,
  cache_dir = "results/orvt-cache"
)

attr(current_a, "source_url_or_path")
attr(current_a, "sha256")
attr(current_a, "pho_layout")

tail(current_a[, c(
  "season", "weekF", "week_start_date", "week_end_date",
  "N", "y", "p"
)])
```

### Pinned downloaded ORVT CSV

For reproducibility, an operator can save the official CSV first and then pass the exact file to the same resolver:

```{r}
orvt_file <- "/secure/PAGe/input/ORVT_Lab_Testing_Data_2025-26_2026-27.csv"

current_a <- getCurrentD(
  data = orvt_file,
  virus = "Influenza A",
  season = "2026-27",
  include_predecessor = FALSE
)

current_b <- getCurrentD(
  data = orvt_file,
  virus = "Influenza B",
  season = "2026-27",
  include_predecessor = FALSE
)

stopifnot(identical(current_a$weekF, current_b$weekF))
```

Hash the raw file before modeling:

```{r}
raw_sha256 <- digest::digest(
  file = orvt_file,
  algo = "sha256",
  serialize = FALSE
)
raw_sha256
```

## 3. Direct R: use the high-level surveillance loader

`page_load_surveillance()` is the normal package-level resolver. It accepts the live ORVT feed (`source = NULL`), a downloaded ORVT CSV, a historical PAGe CSV, a daily `.RData` extract, or an already-canonical data frame.

```{r}
current <- page_load_surveillance(
  source = orvt_file,
  season = "2026-27",
  include_predecessor = FALSE
)

source_info <- attr(current, "page_source")
source_info
```

The `page_source` record includes the source kind/path, SHA-256 when available, retrieval time, resolved season, number of weeks, latest `weekF`, and latest week end.

For the live feed instead:

```{r}
current_live <- page_load_surveillance(
  source = NULL,
  season = "2026-27",
  cache_dir = "results/orvt-cache",
  include_predecessor = FALSE
)
```

## 4. Direct R: produce the current forecast

`page_forecast_now()` is the simplest real-time R API. It validates the frozen kit, resolves the current surveillance source, runs the frozen prospective pipeline, and returns the latest forecast rows plus ignition and source provenance.

::: {.callout-note title="Direct R versus the governed weekly A+B product"}
`page_forecast_now()` is the package-level frozen-kit convenience path. The current governed weekly A+B operational product is the strict weekly transaction exposed by the v4 service later in this guide. Use the transaction/API path when reproducing the official four A/B × +1/+2 weekly outputs.
:::

```{r}
run <- page_forecast_now(
  kit = kit,
  source = orvt_file,
  season = "2026-27",
  timing_mode = "fractional",
  verbose = TRUE
)

run$forecast
run$ignition
run$source
```

For the live feed, set `source = NULL`:

```{r}
run_live <- page_forecast_now(
  kit = kit,
  source = NULL,
  season = "2026-27",
  cache_dir = "results/orvt-cache",
  timing_mode = "fractional"
)
```

Pre-ignition seasons can legitimately return zero M2 forecast rows. That is a valid model state, not automatically an error.

## 6. Generate the walk-forward HTML report from R

The package can render the same interactive walk-forward report used for the current 2026-27 weekly review. Supply either a canonical typed A+B panel as a data frame or the official ORVT CSV path accepted by `page_v3_forecast()`.

```{r}
report_file <- page_v3_walkforward_report(
  data = orvt_file,
  season = "2026-27",
  output_file = "PAGe_2026-27_week12_walkforward.html",
  self_contained = TRUE
)

report_file
```

For an already-canonical typed panel:

```{r}
typed_panel <- read.csv(
  "/secure/PAGe/input/typed_ab_weekly.csv",
  stringsAsFactors = FALSE
)

page_v3_walkforward_report(
  data = typed_panel,
  season = "2026-27",
  output_file = "PAGe_2026-27_week12_walkforward.html"
)
```

The generated report includes the A and B hero rows, individual A/B +1/+2 forecasts with 95% confidence intervals, retrospective prediction intervals where the target-week denominator is already known, a single summed A+B forecast line, the Flu A peak CDF/hazard views, and the peak-normalized M1 full-season shape view with observed 2026-27 points. With `self_contained = TRUE`, Plotly JavaScript is embedded and the HTML has no network dependency. Historical Week 8-11 diagnostics and report-only uncertainty/shape support are packaged separately from the frozen forecast-model identity.

## 6. Strict reproducible ORVT transaction

For governed weekly replay, use one explicit official ORVT CSV and one frozen content-addressed release. The helper below intentionally disables automatic source fallback.

```{bash}
cd /home/yeli/repos/PAGe-m1-v2

ORVT=/secure/PAGe/input/ORVT_Lab_Testing_Data_2025-26_2026-27.csv
RELEASE_DIR=/secure/PAGe/releases/<release-id>
RAW_SHA="$(sha256sum "$ORVT" | awk '{print $1}')"

bash scripts/reproduce_weekly_orvt_v1.sh \
  --season=2026-27 \
  --input="$ORVT" \
  --release-dir="$RELEASE_DIR" \
  --output-root=results/weekly-shadow-release-v5 \
  --expected-sha256="$RAW_SHA"
```

The transaction archives the raw source, binds its SHA-256, records the release identity, constructs the typed A+B panel, runs the governed weekly stack, and publishes only after the transaction seals complete.

To prove a later replay is equivalent:

```{bash}
python3 scripts/verify_weekly_reproducibility_v1.py \
  <first-sealed-transaction-dir> \
  <second-sealed-transaction-dir>
```

Accept only:

```text
REPRODUCIBILITY_CHECK=PASS
```

The verifier checks the season/origin, strict `source_mode=orvt`, raw-source SHA-256, release ID, supplied typed-panel SHA-256, effective-panel SHA-256, and all four A/B × +1/+2 forecasts.

## 7. Weekly HTTP API: user setup

The v4 service is normally loopback-only behind an approved same-host proxy. Users need:

- an API base URL;
- a bearer token supplied outside source code;
- the current immutable release ID, discovered from the API;
- a unique idempotency key for each publication cycle/request option.

In R, keep the token in an environment variable rather than embedding it in a script:

```{r}
Sys.setenv(PAGE_API_TOKEN = "<set outside shared code>")

api_base <- "http://127.0.0.1:8088"
token <- Sys.getenv("PAGE_API_TOKEN")
stopifnot(nzchar(token))
```

The examples below use `httr2`.

```{r}
library(httr2)
```

## 8. Check API liveness and readiness from R

`/healthz` checks process liveness and does not require a bearer token:

```{r}
health <- request(paste0(api_base, "/healthz")) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

health
```

`/readyz` reports whether the deployment/runtime preflight is currently ready:

```{r}
ready_resp <- request(paste0(api_base, "/readyz")) |>
  req_perform()

resp_status(ready_resp)
resp_body_json(ready_resp, simplifyVector = TRUE)
```

Do not submit a weekly run until readiness is `200` / `ready = TRUE`.

## 9. Discover the pinned release from R

Do not hard-code a release from an old runbook. Ask the running service which immutable release it is serving:

```{r}
release <- request(paste0(api_base, "/v1/release")) |>
  req_auth_bearer_token(token) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

release$release_id
release$api_contract_version
release$api_deployment_id
```

Use `release$release_id` in the trigger body.

## 10. Trigger a weekly run from R

The request must include a stable `Idempotency-Key`. Reusing the same key with a different request is rejected.

```{r}
publication_id <- "2026-10-02-week12-official-orvt"

trigger <- request(paste0(api_base, "/v1/weekly-runs")) |>
  req_auth_bearer_token(token) |>
  req_headers(`Idempotency-Key` = publication_id) |>
  req_body_json(list(
    season = "2026-27",
    expected_release_id = release$release_id
  )) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

trigger
run_id <- trigger$run_id
stopifnot(grepl("^wr_[0-9a-f]{32}$", run_id))
```

The API returns `202` for a newly accepted/active job. An idempotent replay of an already terminal job can return `200`.

The optional `a_shadow_option = "exp050_h2"` requests the experimental A+2 test-volume challenger without changing the four canonical forecasts:

```{r}
trigger_shadow <- request(paste0(api_base, "/v1/weekly-runs")) |>
  req_auth_bearer_token(token) |>
  req_headers(`Idempotency-Key` = paste0(publication_id, "-exp050")) |>
  req_body_json(list(
    season = "2026-27",
    expected_release_id = release$release_id,
    a_shadow_option = "exp050_h2"
  )) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)
```

## 11. Poll run status from R

```{r}
get_run_status <- function(run_id) {
  request(paste0(api_base, "/v1/weekly-runs/", run_id)) |>
    req_auth_bearer_token(token) |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

status <- get_run_status(run_id)
status
```

For an interactive client, poll at a modest interval until the job becomes terminal:

```{r}
repeat {
  status <- get_run_status(run_id)
  message("status = ", status$status)

  if (status$status %in% c("succeeded", "failed")) break
  Sys.sleep(5)
}

stopifnot(identical(status$status, "succeeded"))
```

## 12. Retrieve forecasts and monitoring from R

```{r}
result <- request(
  paste0(api_base, "/v1/weekly-runs/", run_id, "/result")
) |>
  req_auth_bearer_token(token) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

result$season
result$origin_weekF
result$release_id
result$effective_panel_sha256
result$forecasts
result$monitoring
```

A successful canonical weekly result contains the four A/B × +1/+2 forecast rows. Monitoring is a disclosure-safe projection from validated child artifacts; for example, A can expose M0 ignition state and M1 peak-timing fields when available.

## 13. Retrieve provenance from R

Always pair a forecast with its provenance projection.

```{r}
provenance <- request(
  paste0(api_base, "/v1/weekly-runs/", run_id, "/provenance")
) |>
  req_auth_bearer_token(token) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

provenance
```

At minimum, verify that the result and provenance agree on the release and effective panel identities supported by the response schema/configuration.

```{r}
stopifnot(identical(result$release_id, release$release_id))
stopifnot(identical(
  result$effective_panel_sha256,
  provenance$effective_panel_sha256
))
```

For official ORVT runs, operator-side transaction provenance should additionally show `source_mode = "orvt"` and the archived raw-source SHA-256. The public HTTP projection intentionally does not expose arbitrary raw artifacts.

## 14. Experimental probability queries from R

Probability endpoints are immutable queries against the run-bound probability snapshot. They do not recompute the model.

### Positivity exceedance

Example: probability that Flu A positivity at +1 week exceeds 2%:

```{r}
p_a_gt_2 <- request(
  paste0(
    api_base,
    "/v1/weekly-runs/", run_id,
    "/probability/positivity"
  )
) |>
  req_auth_bearer_token(token) |>
  req_url_query(
    type = "A",
    horizon = 1,
    threshold = 0.02
  ) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

p_a_gt_2
```

The positivity threshold is a **proportion**, so `0.02` means 2%.

### Peak timing

Example: probability the Flu A peak occurred before Week 21:

```{r}
p_peak_before_21 <- request(
  paste0(
    api_base,
    "/v1/weekly-runs/", run_id,
    "/probability/peak"
  )
) |>
  req_auth_bearer_token(token) |>
  req_url_query(
    type = "A",
    week = 21
  ) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)

p_peak_before_21
```

If the timing/probability snapshot is unavailable for a type or origin, the endpoint returns the documented unavailable/error state rather than fabricating a probability.

## 15. Minimal user-facing API helper in R

A small wrapper can keep authentication and endpoint handling consistent:

```{r}
page_api_get <- function(path, query = NULL) {
  req <- request(paste0(api_base, path)) |>
    req_auth_bearer_token(token)

  if (!is.null(query)) {
    req <- do.call(req_url_query, c(list(req), query))
  }

  req |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

page_api_get("/v1/release")
page_api_get(paste0("/v1/weekly-runs/", run_id, "/result"))
page_api_get(paste0("/v1/weekly-runs/", run_id, "/provenance"))
```

## 16. Operator deployment/preflight path

Service deployment is an operator task rather than an end-user R call. The current v4 builder validates the pinned release and produces a content-addressed API deployment directory.

```{bash}
Rscript --vanilla scripts/build_v3_weekly_api_deployment_v4.R \
  --deployment-root=<ARTIFACT_MOUNT>/api-deployments \
  --artifact-mount=<ARTIFACT_MOUNT> \
  --artifact-fs-type=nfs4 \
  --artifact-mount-source=<MOUNT_SOURCE> \
  --job-root=<ARTIFACT_MOUNT>/weekly-api \
  --output-root=<ARTIFACT_MOUNT>/weekly-shadow \
  --source-mode=orvt \
  --season=2026-27 \
  --bind-host=127.0.0.1 \
  --port=8088 \
  --rscript=/usr/bin/Rscript \
  --forecast-release-dir=<REPO>/artifacts/v3-shadow-release-v3/<release-id> \
  --max-runtime-seconds=1800 \
  --mode=production
```

For an official PHO publication, preflight the source before triggering:

```{bash}
Rscript --vanilla 2026/run_page_orvt_source_preflight_v1.R \
  --season=2026-27 \
  --min-weekF=<expected-origin> \
  --result-path=/tmp/page-orvt-source.json
```

Review the selected official URL, source SHA-256, latest week/end date, and A/B counts before the weekly transaction is started.

## 17. Operational acceptance checklist

Before treating a weekly result as reproducible, confirm:

1. the kit/release identity is immutable and validated;
2. the source is the intended official ORVT vintage;
3. the raw input SHA-256 has been recorded;
4. the latest week/end date is the expected publication origin;
5. the transaction sealed successfully;
6. result and provenance panel hashes agree;
7. there are four finite canonical A/B × +1/+2 forecasts when the model is eligible to issue them;
8. any unavailable M0/M1 timing state is displayed as unavailable/not detected rather than synthesized;
9. replaying the same source and release reproduces the same semantic outputs.

## Related documentation

- [`training_walkthrough.qmd`](training_walkthrough.qmd) — offline training and frozen-kit creation.
- [`deployment-workflow.qmd`](deployment-workflow.qmd) — acceptance, fixed-spec post-holdout refresh, and immutable promotion.
- [`v3-weekly-deployment-api-operations-2026-09-27.md`](v3-weekly-deployment-api-operations-2026-09-27.md) — operator details for API v4.
- [`v3-weekly-deployment-api-openapi-v1.yaml`](v3-weekly-deployment-api-openapi-v1.yaml) — HTTP contract.
