---
title: "2025-26 Season: Prospective Alignment"
format:
  html:
    self-contained: true
    toc: true
    page-layout: full
    grid:
      body-width: 1200px
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(mgcv)
library(gamm4)
library(gratia)
library(MMWRweek)
library(plotly)
library(gt)
library(flualign)
library(future)
library(furrr)

source("../R/m1_reference.R")
source("../R/m1_loso.R")
source("../R/utils.R")
source("../R/pipeline_runtime_helpers.R")
source("../R/m0_runtime.R")
source("../R/m2_runtime.R")
source("../R/m2_spec_grid.R")
source("../R/m2_training.R")
source("../R/pipeline_runtime.R")
source("../R/m1_runtime.R")
source("../R/m1_multi_template.R")
source("../R/getCurrentD.R")

startWeek  <- 27L
WALK_START <- 5L
EXCLUDE_SEAS <- "2015-16"

tuned  <- readRDS("../data/stage1_tuning.rds")
params <- tuned$best_params

manual_labels <- c(
  "2012-13" = 18L, "2013-14" = 20L, "2014-15" = 20L,
  "2015-16" = 24L, "2016-17" = 19L, "2017-18" = 20L,
  "2018-19" = 19L, "2019-20" = 22L, "2022-23" = 15L,
  "2023-24" = 20L, "2024-25" = 23L
)

# Historical data (all seasons, excluding current + bad seasons)
allD_hist <- read.csv("../data/flu_testing_data.csv") %>%
  select(season, week, year, start_year = seasonstart, date = week_start_date,
         y = pos_flua, N = test_flu) %>%
  mutate(neg = N - y, date = as.Date(date),
         mmwr_year = ifelse(week >= 35L, start_year, start_year + 1L),
         nW_true   = n_weeks_in_start_year(start_year),
         weekF     = ((week - startWeek) %% nW_true) + 1L,
         p = y / N) %>%
  filter(!season %in% c("2011-12", "2020-21", "2021-22", "2025-26"))
```

```{=html}
<style>
h1 {
  margin-top: 1.8em;
  padding: 0.45em 1em;
  background: #1a3a5c;
  color: #ffffff;
  border-radius: 5px;
}
h2 {
  margin-top: 1.8em;
  padding: 0.4em 0.9em;
  border-left: 5px solid #1e5f8e;
  background: #e8f1f8;
  color: #1a3a5c;
  border-radius: 0 4px 4px 0;
}
h3 {
  margin-top: 2em;
  padding: 0.4em 0.8em;
  border-left: 4px solid #4a90d9;
  background: #f4f8fd;
  border-radius: 0 4px 4px 0;
}
h4 {
  margin-top: 1.5em;
  padding: 0.3em 0.7em;
  border-left: 3px solid #a0bfe0;
  background: #f9fbfe;
  border-radius: 0 3px 3px 0;
}
</style>
```

## Production reference curve

Trained on all non-excluded historical seasons (`r paste(sort(unique(filter(allD_hist, !season %in% EXCLUDE_SEAS)$season)), collapse = ", ")`). Cached in `../data/ref_production.rds`.

```{r build-ref}
#| message: false
#| warning: false

REF_CACHE <- "../data/ref_production.rds"

if (file.exists(REF_CACHE)) {
  ref_cache <- readRDS(REF_CACHE)
  message("Loaded production ref from cache: ", REF_CACHE)
} else {
  train_allD <- filter(allD_hist, !season %in% EXCLUDE_SEAS)
  flag_args  <- list(p_thresh = 0.01, k1 = 0.4, k_c = 0.01, n_consec = 2L,
                     min_window = 10L, w_min = 21L, w_max = 21L, d2_relax = -0.01)

  res_deriv  <- estimateDerivs(train_allD, k = 10L)
  train_outs <- res_deriv$data %>%
    group_by(season) %>%
    group_split(.keep = TRUE) %>%
    map(~ do.call(flagIgnition, c(list(df = .x, manual_labels = manual_labels), flag_args)))

  aligned_train <- alignIgnition(train_outs)
  ref  <- estimateRef(alignedD = aligned_train, exSeason = character(0),
                      k = 25L, n_weeks = 52L, method = "fs")
  ref$g_ref_fun_orig <- ref$g_ref_fun   # store unshifted for plotting
  hyper <- learn_alignment_hyperparams(ref$dat, ref$g_ref_fun)

  # Prospective-safe derivatives for hist_data (needed for weekly M2 refit)
  alignedD_prosp <- add_prospective_derivs_link(aligned_train, k = 5L, eps = 1e-6, min_obs = 4L)
  if (!"N" %in% names(alignedD_prosp))
    alignedD_prosp$N <- alignedD_prosp$y + alignedD_prosp$neg

  ref_cache <- list(ref = ref, hyper = hyper, hist_data = alignedD_prosp)
  saveRDS(ref_cache, REF_CACHE)
  message("Production ref built and cached: ", REF_CACHE)
}

ref   <- ref_cache$ref
hyper <- ref_cache$hyper
```

```{r ref-plot}
#| echo: false
#| warning: false

newW_grid <- seq(1, 52, by = 0.5)
ref_df_plot <- tibble(
  newWeek = newW_grid,
  p_ref   = plogis(ref$g_ref_fun(newW_grid))
)

plot_ly() %>%
  add_lines(data = ref_df_plot, x = ~newWeek, y = ~p_ref,
            line = list(color = "steelblue", width = 2),
            name = "Reference curve") %>%
  layout(xaxis = list(title = "newWeek (aligned)"),
         yaxis = list(title = "Positivity"),
         title = "Production reference curve (k_ref = 25, FS method, all training seasons)")
```

## Current season data

```{r get-current}
#| message: false
#| warning: false

currentD_full <- getCurrentD(startWeek = startWeek)
# Keep only 2025-26
currentD_full <- filter(currentD_full, season == "2025-26")

cat(sprintf("Season: 2025-26 | Weeks observed: %d | weekF range: %d – %d\n",
            nrow(currentD_full),
            min(currentD_full$weekF), max(currentD_full$weekF)))
```

## Walk-forward alignment

Starting at `weekF = WALK_START` (`r WALK_START`). Results cached in `../data/run_wf_cache.rds`.

```{r run-wf}
#| message: false
#| warning: false

WF_CACHE <- "../data/run_wf_cache.rds"

if (file.exists(WF_CACHE)) {
  wf <- readRDS(WF_CACHE)
  message("Loaded walk-forward from cache.")
} else {
  walk_end_s <- max(currentD_full$weekF, na.rm = TRUE)

  # Run ignition detection once on all available data
  ign_out <- run_ignition_weekly(
    currentSeason  = currentD_full,
    ign_fit_or_gam = NULL,
    params         = params,
    start_week     = 1L
  )

  walk_start_s <- if (!is.na(ign_out$ign_week_locked)) {
    max(WALK_START, as.integer(ign_out$ign_week_locked))
  } else {
    WALK_START
  }
  eval_weeks <- seq(walk_start_s, walk_end_s)

  iWeek_true <- NA_integer_   # unknown for current season

  week_results <- lapply(eval_weeks, function(ew) {
    season_data_to_ew <- filter(currentD_full, weekF <= ew)
    n_obs <- nrow(season_data_to_ew)

    ap <- run_alignment_prospective_multi(
      currentSeason   = season_data_to_ew,
      ref             = ref,
      hyper           = hyper,
      ign_out         = ign_out,
      use_ci          = TRUE,
      buffer_weeks    = 0L,
      allow_scale     = NULL,
      level           = 0.95,
      min_obs         = 4L,
      curvature_ratio = 1.0,
      temperature     = 0.25,   # ← TUNED: v5-v7 LOSO grid
      rise_weight     = 1.0,    # ← TUNED
      slope_weight    = 8.0,    # ← TUNED: interior optimum (0→30 tested)
      slope_window    = 6L      # ← TUNED: optimal window
    )

    if (ap$state == "pre_ignition") {
      ign_locked_w <- ign_out$ign_week_locked
      reason <- if (is.na(ign_locked_w) || ign_locked_w > ew) "no_ignition"
                else if (n_obs < 4L) "too_few_obs"
                else "alignment_error"
      iWeek_hat_ew <- if (!is.na(ign_locked_w) && ign_locked_w <= ew)
                        as.integer(ign_out$iWeek_hat_locked) else NA_integer_

      params_row <- tibble(
        eval_week = ew, n_obs = n_obs, iWeek_hat = iWeek_hat_ew,
        iWeek_true = iWeek_true, state = ap$state,
        tau = NA_real_, delta = NA_real_, a = NA_real_, b = NA_real_,
        allow_scale = NA, delta_on = NA,
        t_peak = NA_real_, t_peak_lo = NA_real_, t_peak_hi = NA_real_,
        peak_weekF = NA_integer_, peak_passed = FALSE,
        fallback_reason = reason, anchorWeek = ref$anchorWeek
      )
      return(list(params = params_row, forecast = NULL))
    }

    params_row <- tibble(
      eval_week       = ew,
      n_obs           = n_obs,
      iWeek_hat       = ap$iWeek_hat,
      iWeek_true      = iWeek_true,
      state           = ap$state,
      tau             = ap$tau,
      delta           = ap$delta,
      a               = ap$a,
      b               = ap$b,
      allow_scale     = ap$allow_scale,
      delta_on        = ap$delta_on,
      t_peak          = ap$t_peak,
      t_peak_lo       = ap$t_peak_ci[1],
      t_peak_hi       = ap$t_peak_ci[2],
      peak_weekF      = ap$peak_weekF,
      peak_passed     = ap$peak_passed,
      fallback_reason = ap$fallback_reason,
      anchorWeek      = ref$anchorWeek
    )

    forecast_row <- ap$forecast_df %>%
      mutate(eval_week = ew,
             newWeek = as.numeric(newWeek),
             p_hat   = as.numeric(p_hat),
             p_lo    = as.numeric(p_lo),
             p_hi    = as.numeric(p_hi)) %>%
      select(eval_week, newWeek, p_hat, p_lo, p_hi, kind)

    list(params = params_row, forecast = forecast_row)
  })

  params_df   <- bind_rows(map(week_results, "params"))
  forecast_df <- bind_rows(map(week_results, "forecast"))

  wf <- list(params_df = params_df, forecast_df = forecast_df, ign_out = ign_out)
  saveRDS(wf, WF_CACHE)
  message("Walk-forward complete. Saved to ", WF_CACHE)
}
```

## Current state

```{r current-state}
#| echo: false

last_row <- wf$params_df %>%
  filter(!is.na(iWeek_hat)) %>%
  slice_max(eval_week, n = 1, with_ties = FALSE)

if (nrow(last_row) > 0) {
  cat(sprintf(
    "State: %s | Ignition: weekF %d | Est. peak: weekF %d [%d, %d] | Latest eval: weekF %d\n",
    last_row$state,
    last_row$iWeek_hat,
    last_row$peak_weekF,
    wf$params_df %>% filter(!is.na(iWeek_hat)) %>% slice_max(eval_week, n=1) %>% pull(t_peak_lo) %>% {round(. - ref$anchorWeek + last_row$iWeek_hat)},
    wf$params_df %>% filter(!is.na(iWeek_hat)) %>% slice_max(eval_week, n=1) %>% pull(t_peak_hi) %>% {round(. - ref$anchorWeek + last_row$iWeek_hat)},
    last_row$eval_week
  ))
} else {
  cat("No alignment yet — ignition not detected.\n")
}
```

## Alignment parameters over time

```{r param-plots-setup}
#| echo: false
#| warning: false

param_vars <- c("tau", "delta", "a", "b", "t_peak")
param_labels <- c(
  tau    = "τ — time shift",
  delta  = "δ — scale",
  a      = "a — lower asymptote",
  b      = "b — upper asymptote",
  t_peak = "t_peak — predicted peak (aligned)"
)

df_params <- wf$params_df %>%
  filter(!is.na(iWeek_hat)) %>%
  select(eval_week, all_of(param_vars))

make_param_plot <- function(pvar) {
  df <- df_params %>% select(eval_week, value = all_of(pvar))
  plot_ly(df, x = ~eval_week, y = ~value,
          type = "scatter", mode = "lines+markers",
          line = list(color = "steelblue", width = 1.5),
          marker = list(size = 6, color = "steelblue")) %>%
    layout(xaxis = list(title = "Eval weekF"),
           yaxis = list(title = param_labels[[pvar]]))
}
```

::: {.panel-tabset}

```{r param-tabs}
#| echo: false
#| warning: false
#| results: asis

param_tab_chunks <- lapply(param_vars, function(pvar) {
  e <- new.env(parent = globalenv())
  e$p <- make_param_plot(pvar)
  knitr::knit_child(
    text = c(sprintf("### %s", param_labels[[pvar]]), "",
             "```{r echo=FALSE, warning=FALSE}", "p", "```", ""),
    envir = e, quiet = TRUE
  )
})
cat(unlist(param_tab_chunks), sep = "\n")
```

:::

## Results by eval week

Grey line = reference curve (background). Blue = fitted past, red = forecast.
Dashed grey = ignition, solid black = current eval week, dashed red = est. peak.

```{r week-reports}
#| echo: false
#| warning: false
#| results: asis

s_all <- wf$params_df %>% filter(!is.na(iWeek_hat))

if (nrow(s_all) == 0) {
  cat("*No forecasts yet — ignition not detected.*\n\n")
} else {
  # Extend 2 weeks past first peak-detection week
  first_peak_ew <- s_all %>% filter(peak_passed) %>%
    slice_min(eval_week, n = 1, with_ties = FALSE) %>% pull(eval_week)
  cutoff_ew <- if (length(first_peak_ew) > 0) first_peak_ew + 2L else max(s_all$eval_week)

  s_params <- s_all %>% filter(eval_week <= cutoff_ew)

  # gt parameter table
  tbl <- gt(s_params %>% select(eval_week, iWeek_hat, state, peak_weekF,
                                 tau, delta, t_peak, t_peak_lo, t_peak_hi, peak_passed)) %>%
    cols_label(eval_week = "Eval wkF", iWeek_hat = "Ign (hat)", state = "State",
               peak_weekF = "Est. peak", tau = "τ", delta = "δ",
               t_peak = "t_peak", t_peak_lo = "lo", t_peak_hi = "hi",
               peak_passed = "Peak passed") %>%
    fmt_number(columns = c(tau, delta, t_peak, t_peak_lo, t_peak_hi), decimals = 2) %>%
    fmt_integer(columns = c(eval_week, iWeek_hat, peak_weekF)) %>%
    tab_style(style = cell_fill(color = "#fff0f0"),
              locations = cells_body(rows = peak_passed == TRUE)) %>%
    tab_options(table.width = pct(100), table.font.size = "13px")

  # Reference curve for background
  newW_grid <- seq(1, 52, by = 0.5)
  ref_curve <- tibble(
    newWeek = newW_grid,
    p_ref   = plogis(ref$g_ref_fun(newW_grid))
  )

  obs_s <- currentD_full %>% mutate(p_obs = y / N)

  cat("::: {.panel-tabset}\n\n")

  tab_chunks <- lapply(s_params$eval_week, function(ew) {
    .ref_curve <- ref_curve

    pw           <- wf$params_df %>% filter(eval_week == ew)
    iWeek_hat_ew <- pw$iWeek_hat[[1]]
    anchor_ew    <- pw$anchorWeek[[1]]
    t_peak_ew    <- pw$t_peak[[1]]
    peak_wF      <- pw$peak_weekF[[1]]
    state_ew     <- pw$state[[1]]

    current_newW <- ew - iWeek_hat_ew + anchor_ew

    fc_smooth <- wf$forecast_df %>%
      filter(eval_week == ew, kind == "forecast") %>%
      arrange(newWeek)

    if (nrow(fc_smooth) == 0) return(NULL)

    junction  <- fc_smooth$newWeek[which.min(abs(fc_smooth$newWeek - current_newW))]
    fc_past   <- fc_smooth %>% filter(newWeek <= junction)
    fc_future <- fc_smooth %>% filter(newWeek >= junction)

    obs_aligned <- obs_s %>%
      filter(weekF <= ew) %>%
      mutate(newWeek = weekF - iWeek_hat_ew + anchor_ew)

    y_max <- max(c(fc_smooth$p_hi, obs_aligned$p_obs), na.rm = TRUE) * 1.1
    y_max <- max(y_max, 0.15)

    plt <- plot_ly() %>%
      add_lines(
        data = .ref_curve, x = ~newWeek, y = ~p_ref,
        line = list(color = "rgba(160,160,160,0.5)", width = 1.2),
        name = "Template (orig)"
      ) %>%
      add_ribbons(
        data = fc_smooth, x = ~newWeek, ymin = ~p_lo, ymax = ~p_hi,
        fillcolor = "rgba(150,150,150,0.15)",
        line = list(color = "transparent"),
        name = "95% CI", showlegend = TRUE
      ) %>%
      add_lines(
        data = fc_past, x = ~newWeek, y = ~p_hat,
        line = list(color = "steelblue", width = 2.5),
        name = "Fitted (past)"
      ) %>%
      add_lines(
        data = fc_future, x = ~newWeek, y = ~p_hat,
        line = list(color = "firebrick", width = 2.5),
        name = "Forecast"
      ) %>%
      add_markers(
        data = obs_aligned, x = ~newWeek, y = ~p_obs,
        marker = list(color = "black", size = 5, opacity = 0.75),
        name = "Observed"
      ) %>%
      add_segments(
        x = anchor_ew, xend = anchor_ew, y = 0, yend = y_max,
        line = list(color = "grey50", dash = "dash", width = 1.5),
        name = paste0("Ignition (wk ", iWeek_hat_ew, ")"), showlegend = TRUE
      ) %>%
      add_segments(
        x = current_newW, xend = current_newW, y = 0, yend = y_max,
        line = list(color = "black", dash = "solid", width = 2),
        name = paste0("Now (wk ", ew, ")"), showlegend = TRUE
      )

    if (!is.na(t_peak_ew)) {
      plt <- plt %>%
        add_segments(
          x = t_peak_ew, xend = t_peak_ew, y = 0, yend = y_max,
          line = list(color = "firebrick", dash = "dash", width = 1.5),
          name = paste0("Est. peak (wk ", peak_wF, ")"), showlegend = TRUE
        )
    }

    plt <- plt %>%
      layout(
        title = paste0("2025-26 | weekF=", ew, " | ign=", iWeek_hat_ew,
                       " | state=", state_ew,
                       if (!is.na(peak_wF)) paste0(" | est peak wk ", peak_wF) else ""),
        xaxis = list(title = "newWeek (aligned)"),
        yaxis = list(title = "Positivity", range = c(0, y_max)),
        legend = list(orientation = "h", y = -0.20)
      )

    e <- new.env(parent = globalenv())
    e$tab_plot <- plt
    knitr::knit_child(
      text = c(sprintf("### nW %d", current_newW), "",
               "```{r echo=FALSE, warning=FALSE}", "tab_plot", "```", ""),
      envir = e, quiet = TRUE
    )
  })

  valid_tabs <- Filter(Negate(is.null), tab_chunks)
  cat(unlist(valid_tabs), sep = "\n")

  # Data tab
  e_tbl <- new.env(parent = globalenv())
  e_tbl$tbl_obj <- tbl
  data_tab <- knitr::knit_child(
    text = c("### Data", "", "```{r echo=FALSE}", "tbl_obj", "```", ""),
    envir = e_tbl, quiet = TRUE
  )
  cat(data_tab)

  cat("\n:::\n\n")
}
```
