---
title: "R Source Map"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    page-layout: full
    grid:
      body-width: 1200px
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note}
This page is a source map for the root-level `R/` directory, which is mirrored into `flualign/R/` for the installable package. It is meant to answer two questions quickly: which file owns a piece of the pipeline, and which functions live in that file.
:::

## At a glance

```{r}
file_meta <- data.frame(
  file = c(
    "m0_training.R", "m0_retro.R", "m0_runtime.R", "flagIgnition.R",
    "m1_reference.R", "m1_reference_helpers.R", "m1_reference_loader.R",
    "m1_fit.R", "m1_hyperparams.R", "m1_loso.R", "m1_runtime.R",
    "m1_multi_template.R", "m1_peak_status.R", "m1_peak_flags.R",
    "m1_peak_summary.R", "m1_peak_distribution.R", "m1_alignment_plots.R",
    "m2_training.R", "m2_spec_grid.R", "m2_nested_loso.R", "m2_runtime.R",
    "pipeline_bridge.R", "pipeline_runtime_helpers.R", "pipeline_runtime.R",
    "align_forecast_pipeline_dilate.R", "forecast_post_peak_gam.R",
    "checkSeasonLength.R", "check_scale_identifiability.R", "identifiability.R",
    "getCurrentD.R", "score_params.R", "simulate.R",
    "plot.R", "plotRes.R", "plotSeasonCurves.R",
    "utils.R", "migrate_r_packages.R",
    "retired.R"
  ),
  area = c(
    rep("M_0: ignition", 4),
    rep("M_1: reference and alignment", 13),
    rep("M_2: forecast training and runtime", 4),
    rep("Pipeline orchestration", 3),
    rep("Supporting analyses and utilities", 13),
    "Retired"
  ),
  purpose = c(
    "Train and tune the ignition detector, including LOSO utilities and ignition decision helpers.",
    "Run retrospective ignition grid searches and helper plots for historical alignment to ignition anchors.",
    "Run weekly M_0 detection on a current season and visualize the weekly snapshots.",
    "Apply the four-rule threshold detector to flag the ignition week for a single season.",
    "Estimate the reference epidemic curve and derivative summaries from historical data.",
    "Define reusable reference-curve closures and helpers that package the fitted GAM as accessors.",
    "Load example reference objects and built-in package data used in demos and documentation.",
    "Fit tau and delta by matching a season to the reference curve.",
    "Learn high-level M_1 alignment hyperparameters from retrospective tuning results.",
    "Run LOSO and walk-forward evaluation for alignment and hyperparameter tuning.",
    "Run prospective M_1 alignment, peak detection, and peak calibration.",
    "Combine multiple reference templates into the multi-template alignment ensemble.",
    "Turn alignment outputs into a coarse peak-state label such as pre-peak or post-peak.",
    "Flag whether the season has likely passed its peak in real time.",
    "Summarize peak timing implied by an alignment fit.",
    "Estimate the peak-week distribution and candidate peak-rule settings across seasons.",
    "Plot how alignment trajectories evolve as more weekly data arrive.",
    "Core M_2 prediction functions: m2_predict_one() (shared prediction core), soft-cap, feature derivation, model fitting, and weekly refit.",
    "M_2 model specification, formula construction, feature engineering helpers, and the grid-search spec builder.",
    "Nested LOSO evaluation for the full M_1 -> M_2 forecasting stack, including weekly-refit and frozen-fit paths.",
    "Build pseudo-prospective M_2 snapshots and turn them into forecast series.",
    "M_1 walk-forward predictions and joint M_1 -> M_2 LOSO evaluation bridge.",
    "Provide small shared helpers used across runtime code paths.",
    "End-to-end prospective pipeline entry points: run_prospective_pipeline() and run_m2_forecast().",
    "Legacy exploratory helper for dilation-based alignment plus forecasting in one function.",
    "Legacy post-peak forecasting helper based on a GAM without further alignment updates.",
    "Check whether a season has enough observed weeks to support downstream analysis.",
    "Check whether scale is identifiable early enough for alignment-based forecasting.",
    "Older identifiability helper retained for comparison with the current rule.",
    "Extract or derive the current scale parameter used in older alignment workflows.",
    "Score ignition-threshold parameter settings on historical data.",
    "Simulate flu seasons and load the built-in historical influenza example data.",
    "Generic forecast plotting helper.",
    "Legacy plotting helper for summarizing result objects.",
    "Plot season-level observed and fitted curves with ignition metadata.",
    "General mathematical and table-building utilities shared across the codebase.",
    "Local helper for migrating installed R packages between R versions.",
    "Archive of legacy and deprecated functions extracted from active source files. Not sourced by the pipeline."
  ),
  stringsAsFactors = FALSE
)

files <- sort(list.files("../R", pattern = "\\.[Rr]$", full.names = TRUE))

extract_desc <- function(lines, i) {
  j <- i - 1
  block <- character()
  while (j >= 1 && grepl("^#'", lines[j])) {
    block <- c(lines[j], block)
    j <- j - 1
  }
  block <- sub("^#' ?", "", block)
  block <- trimws(block)
  block <- block[nzchar(block)]
  block <- block[!grepl("^@", block)]
  if (!length(block)) {
    return(NA_character_)
  }
  block[1]
}

inventory_one <- function(path) {
  lines <- readLines(path, warn = FALSE)
  idx <- grep("^[A-Za-z.][A-Za-z0-9._]*\\s*<-\\s*function\\s*\\(", lines)
  if (!length(idx)) {
    return(data.frame(
      file = basename(path),
      fun = character(),
      description = character(),
      stringsAsFactors = FALSE
    ))
  }

  out <- data.frame(
    file = basename(path),
    fun = sub("^([A-Za-z.][A-Za-z0-9._]*)\\s*<-.*$", "\\1", lines[idx]),
    description = vapply(idx, function(i) extract_desc(lines, i), character(1)),
    stringsAsFactors = FALSE
  )

  out[!duplicated(out[c("file", "fun")]), , drop = FALSE]
}

fun_inventory <- do.call(rbind, lapply(files, inventory_one))

desc_overrides <- c(
  checkSeasonLength = "Check whether each season has enough usable observations for the intended analysis window.",
  getCurrentD = "Extract the current delta or scale estimate from older alignment result objects.",
  plot_det_facet = "Plot ignition-detection behavior across parameter settings and seasons.",
  run_ignition_weekly = "Run the weekly M_0 ignition detector on one current-season data stream.",
  plot_ignition_weekly_snapshots = "Visualize weekly M_0 detection snapshots over a season.",
  detect_ignition_from_tuning = "Apply tuned ignition parameters to produce weekly ignition decisions.",
  loso_M0v2 = "Run leave-one-season-out evaluation for the M_0 v2 ignition detector.",
  negloglik_tau_delta = "Compute the alignment objective used to estimate tau and delta.",
  tau_profile_se = "Approximate uncertainty for tau from the profile objective.",
  fit_tau_delta_old = "Older tau-delta fitting routine retained for comparison.",
  cov_tau_delta_from_profile = "Estimate a covariance matrix for tau and delta from the profile surface.",
  safe_obj = "Safely evaluate an objective function and return a large penalty on failure.",
  learn_alignment_hyperparams = "Summarize retrospective tuning results into a chosen M_1 hyperparameter setting.",
  peak_summary_from_fit = "Summarize predicted peak timing and related quantities from an alignment fit.",
  stage2_extract_hyperparams = "Extract the Stage-2 hyperparameters needed for runtime prediction from a fitted object.",
  score_stage2_metrics = "Compute NLL, Brier, and RMSE forecast error metrics for Stage-2 model comparison.",
  train_stage2_joint_prepped = "Fit a Stage-2 GAM after the training data have already been prepared by prep_stage2_joint().",
  m2_predict_one = "Shared prediction core for M_2: builds newdata, aligns factor levels, applies season RE inclusion rule, soft cap, and optional CI.",
  make_soft_cap_fn = "Build the tanh soft-cap function that compresses predictions above the 95th-percentile knee toward a ceiling.",
  stage2_make_spec = "Construct a fully-specified Stage-2 model spec including formula and exclude terms.",
  stage2_build_joint_formula = "Build the mgcv GAM formula from a Stage-2 spec object.",
  stage2_exclude_newseason = "Return the smooth terms to exclude when predicting for a season not in the training data.",
  nested_loso_grid_search = "Parallel nested LOSO grid search over Stage-2 specs, supporting both weekly-refit and frozen-fit modes.",
  nested_loso_m2_eval_weekly_refit = "Evaluate one Stage-2 spec on one LOSO fold using weekly refit; calls m2_predict_one().",
  nested_loso_refit_best = "Refit the production Stage-2 model on all historical data using the best spec from grid search.",
  m1_walkforward_predictions = "Compute M_1 walk-forward alignment predictions at each evaluation week across a season.",
  m1_walkforward_multi = "Run m1_walkforward_predictions() for all LOSO folds in parallel.",
  loso_m1_m2_joint = "Joint M_1 -> M_2 LOSO evaluation: runs walk-forward M_1 per fold then fits and scores Stage-2.",
  run_prospective_pipeline = "Top-level entry point: run the full M_0 -> M_1 -> M_2 prospective pipeline for one season.",
  run_m2_forecast = "Run the M_2 weekly refit and produce forecasts for the current season; calls m2_predict_one().",
  .plot_stage2_joint_fit_prosp = "Internal plotting helper for prospective Stage-2 fit diagnostics.",
  load_prospective_kit = "Load the data bundle needed to run the prospective weekly pipeline.",
  plot_forecast = "Plot observed data alongside forecast trajectories.",
  plotRes = "Plot legacy result objects in a compact summary view.",
  simulate_flu_seasons = "Simulate synthetic seasonal flu trajectories for experimentation.",
  load_flu_hist = "Load the bundled historical influenza dataset used in examples.",
  num_deriv = "Compute a numerical derivative for a univariate function.",
  template_peak_u = "Return the reference-curve location of the peak on the template time scale.",
  is_cov_ok = "Check whether a covariance estimate is numerically usable.",
  wilson_ci = "Compute a Wilson confidence interval for a binomial proportion.",
  n_weeks_in_start_year = "Return the number of epidemiological weeks in the starting calendar year of a season.",
  logit = "Apply the logit transform.",
  g_ref_safe = "Safely evaluate the reference curve on the link scale.",
  make_g_ref_fun = "Build a callable reference function from a fitted GAM.",
  make_g_ref_mu_se = "Build a callable reference mean-and-SE accessor from a fitted GAM.",
  makeTable = "Create a formatted summary table from vector inputs.",
  mark_season_weeks = "Mark which observations belong to the active in-season portion of each season.",
  get_newWeek_from_week = "Map surveillance week numbers into within-season week indices.",
  remove_global_functions = "Remove globally assigned helper functions created during exploratory work."
)

missing_desc <- is.na(fun_inventory$description) | !nzchar(fun_inventory$description)
fun_inventory$description[missing_desc] <- desc_overrides[fun_inventory$fun[missing_desc]]
fun_inventory$description[is.na(fun_inventory$description) | !nzchar(fun_inventory$description)] <-
  "No inline description is available; see the source for details."

summary_tbl <- merge(file_meta, aggregate(fun ~ file, fun_inventory, length), by = "file", all.x = TRUE)
summary_tbl$fun[is.na(summary_tbl$fun)] <- 0L
names(summary_tbl)[names(summary_tbl) == "fun"] <- "Functions"
summary_tbl <- summary_tbl[match(file_meta$file, summary_tbl$file), ]

knitr::kable(
  summary_tbl[, c("file", "area", "Functions", "purpose")],
  col.names = c("File", "Area", "Functions", "Role"),
  align = c("l", "l", "r", "l")
)
```

## Call graph

The diagram below shows which active pipeline functions call which others,
colour-coded by stage. **Solid boxes** are defined in `R/`; **ellipses** are
referenced but not found in active source (package internals or implicit).
Arrows point from caller → callee.

```{dot}
//| fig-width: 16
//| fig-height: 7
digraph PAGe_pipeline {
  graph [rankdir=LR, fontname="Helvetica", fontsize=11,
         nodesep=0.3, ranksep=1.0, concentrate=true];
  node  [fontname="Helvetica", fontsize=10, style=filled,
         shape=box, margin="0.12,0.05"];
  edge  [fontname="Helvetica", fontsize=9, arrowsize=0.6];

  // Force left-to-right pipeline order via invisible rank edges
  fitIgnition -> run_alignment_prospective [style=invis];
  run_alignment_prospective -> train_stage2_joint [style=invis];
  train_stage2_joint -> run_prospective_pipeline [style=invis];

  // ── M0: Ignition ──────────────────────────────────────────
  subgraph cluster_m0 {
    label="M0: Ignition"; style=filled;
    color="#856404"; fillcolor="#fff3cd"; fontcolor="#856404";
    fontsize=11; fontname="Helvetica-Bold";
    fitIgnition      [label="fitIgnition"];
    run_m0_detection [label="run_m0_detection"];
  }

  // ── M1: Alignment ─────────────────────────────────────────
  subgraph cluster_m1 {
    label="M1: Alignment"; style=filled;
    color="#155724"; fillcolor="#c3e6cb"; fontcolor="#155724";
    fontsize=11; fontname="Helvetica-Bold";
    run_alignment_prospective  [label="run_alignment\n_prospective"];
    m1_walkforward_predictions [label="m1_walkforward\n_predictions"];
    m1_walkforward_multi       [label="m1_walkforward\n_multi"];
    inject_m1_into_snapshots   [label="inject_m1_into\n_snapshots"];
  }

  // ── M2: Spec & Features ───────────────────────────────────
  subgraph cluster_m2spec {
    label="M2: Spec & Features"; style=filled;
    color="#004085"; fillcolor="#cce5ff"; fontcolor="#004085";
    fontsize=11; fontname="Helvetica-Bold";
    stage2_make_spec            [label="stage2_make_spec"];
    stage2_build_joint_formula  [label="stage2_build\n_joint_formula"];
    stage2_exclude_newseason    [label="stage2_exclude\n_newseason"];
    prep_stage2_joint           [label="prep_stage2_joint"];
    add_prospective_derivs_link [label="add_prospective\n_derivs_link"];
    format_current_for_stage2   [label="format_current\n_for_stage2"];
  }

  // ── M2: Training / LOSO ───────────────────────────────────
  subgraph cluster_m2train {
    label="M2: Training / LOSO"; style=filled;
    color="#003060"; fillcolor="#b8d4f0"; fontcolor="#003060";
    fontsize=11; fontname="Helvetica-Bold";
    train_stage2_joint               [label="train_stage2_joint"];
    train_stage2_joint_prepped       [label="train_stage2_joint\n_prepped"];
    refit_stage2_weekly              [label="refit_stage2\n_weekly"];
    score_stage2_metrics             [label="score_stage2\n_metrics"];
    nested_loso_grid_search          [label="nested_loso\n_grid_search"];
    nested_loso_m2_eval_weekly_refit [label="nested_loso_m2\n_eval_weekly_refit"];
    nested_loso_refit_best           [label="nested_loso\n_refit_best"];
    loso_m1_m2_joint                 [label="loso_m1_m2_joint"];
  }

  // ── M2: Predict (shared core) ─────────────────────────────
  subgraph cluster_m2pred {
    label="M2: Predict"; style=filled;
    color="#4a0080"; fillcolor="#e2ccff"; fontcolor="#4a0080";
    fontsize=11; fontname="Helvetica-Bold";
    m2_predict_one   [label="m2_predict_one", penwidth=2];
    make_soft_cap_fn [label="make_soft_cap_fn"];
  }

  // ── Deployment ────────────────────────────────────────────
  subgraph cluster_deploy {
    label="Deployment"; style=filled;
    color="#721c24"; fillcolor="#f8d7da"; fontcolor="#721c24";
    fontsize=11; fontname="Helvetica-Bold";
    run_prospective_pipeline [label="run_prospective\n_pipeline"];
    run_m2_forecast          [label="run_m2_forecast"];
  }

  // ── Edges ─────────────────────────────────────────────────
  // M0
  run_m0_detection -> fitIgnition;

  // M1
  m1_walkforward_multi     -> m1_walkforward_predictions;
  m1_walkforward_multi     -> run_alignment_prospective;
  inject_m1_into_snapshots -> run_alignment_prospective;
  prep_stage2_joint        -> m1_walkforward_predictions;

  // M2 spec
  stage2_make_spec          -> stage2_build_joint_formula;
  stage2_make_spec          -> stage2_exclude_newseason;
  format_current_for_stage2 -> add_prospective_derivs_link;
  format_current_for_stage2 -> train_stage2_joint;
  prep_stage2_joint         -> add_prospective_derivs_link;

  // M2 training
  train_stage2_joint         -> train_stage2_joint_prepped;
  train_stage2_joint         -> prep_stage2_joint;
  train_stage2_joint         -> stage2_make_spec;
  train_stage2_joint_prepped -> stage2_build_joint_formula;
  refit_stage2_weekly        -> format_current_for_stage2;
  refit_stage2_weekly        -> prep_stage2_joint;
  refit_stage2_weekly        -> train_stage2_joint;
  refit_stage2_weekly        -> score_stage2_metrics;
  nested_loso_grid_search    -> m1_walkforward_multi;
  nested_loso_grid_search    -> refit_stage2_weekly;
  nested_loso_grid_search    -> train_stage2_joint;
  nested_loso_grid_search    -> stage2_make_spec;
  nested_loso_m2_eval_weekly_refit -> refit_stage2_weekly;
  nested_loso_m2_eval_weekly_refit -> m2_predict_one;
  nested_loso_m2_eval_weekly_refit -> make_soft_cap_fn;
  nested_loso_refit_best     -> nested_loso_grid_search;
  nested_loso_refit_best     -> train_stage2_joint;
  loso_m1_m2_joint           -> m1_walkforward_predictions;
  loso_m1_m2_joint           -> prep_stage2_joint;
  loso_m1_m2_joint           -> train_stage2_joint;
  loso_m1_m2_joint           -> score_stage2_metrics;

  // M2 predict
  m2_predict_one -> make_soft_cap_fn;

  // Deployment
  run_prospective_pipeline -> run_m0_detection;
  run_prospective_pipeline -> run_m2_forecast;
  run_m2_forecast          -> run_m0_detection;
  run_m2_forecast          -> refit_stage2_weekly;
  run_m2_forecast          -> m2_predict_one;
  run_m2_forecast          -> make_soft_cap_fn;
  run_m2_forecast          -> stage2_exclude_newseason;
  run_m2_forecast          -> add_prospective_derivs_link;
}
```

## By module

```{r results='asis'}
areas <- unique(file_meta$area)

for (area in areas) {
  cat("## ", area, "\n\n", sep = "")
  area_rows <- file_meta[file_meta$area == area, , drop = FALSE]

  for (i in seq_len(nrow(area_rows))) {
    file_i <- area_rows$file[i]
    cat("### `", file_i, "`\n\n", sep = "")
    cat(area_rows$purpose[i], "\n\n")
    cat("Mirror: `flualign/R/", file_i, "`.\n\n", sep = "")

    funcs_i <- fun_inventory[fun_inventory$file == file_i, c("fun", "description"), drop = FALSE]
    if (!nrow(funcs_i)) {
      cat("This file does not define top-level functions.\n\n")
      next
    }

    rownames(funcs_i) <- NULL
    names(funcs_i) <- c("Function", "Description")
    print(knitr::kable(funcs_i, align = c("l", "l")))
    cat("\n")
  }
}
```
