---
title: "PAGe Training Walkthrough"
subtitle: "From historical surveillance data to a frozen, auditable model kit"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-copy: true
execute:
  eval: false
---

This guide is the user-facing training path for PAGe. It shows the public R API first, then the lower-level governed stage API used when a model developer needs to inspect M0, M1, and M2 separately.

::: {.callout-important title="Training is offline"}
Training and model selection are offline operations. Do not retune a model from the current surveillance week. A prospective holdout must remain untouched until the candidate is frozen and the acceptance workflow is run.
:::

## 1. Install/load PAGe

Run from an R environment that satisfies the repository runtime contract.

```{r}
library(PAGe)
```

For a source checkout rather than an installed package:

```{r}
pkgload::load_all("PAGe")
```

## 2. Prepare historical surveillance data

PAGe expects one observation per season/week with positive tests `y`, total tests `N`, and a season-relative `weekF` coordinate. The package normalization layer should be used rather than hand-building downstream model matrices.

```{r}
history <- load_flu_hist("/authorized/path/flu_history.csv") |>
  prepare_surveillance_data()

validate_surveillance_data(history)

history |>
  subset(select = c(season, weekF, y, N, p)) |>
  head()
```

A useful first audit is the available season set:

```{r}
sort(unique(history$season))
```

## 3. Record expert ignition labels

The convenience workflow can prompt interactively, but reproducible training should use a named decimal-week vector or a previously saved expert annotation object.

```{r}
ignition_weeks <- c(
  "2012-13" = 18.0,
  "2013-14" = 20.0,
  "2014-15" = 20.0,
  "2016-17" = 19.0,
  "2017-18" = 20.0,
  "2018-19" = 19.0,
  "2019-20" = 22.0,
  "2022-23" = 15.0,
  "2023-24" = 20.0,
  "2024-25" = 23.0
)

labels <- page_label_ignitions(
  history,
  ignition_weeks = ignition_weeks,
  annotator = "model-development-team",
  interactive = FALSE
)

labels
labels$table
```

The decimal expert annotation is retained for provenance. The current legacy M0/M1 training contract receives the integer projection used by the training workflow.

## 4. Declare training, exclusion, and holdout seasons

Never infer the governed season set from whichever rows happen to be present in a file.

```{r}
selection <- validate_season_selection(
  history,
  training_seasons = setdiff(
    sort(unique(history$season)),
    c("2011-12", "2015-16", "2020-21", "2021-22", "2025-26")
  ),
  exclude_seasons = c("2011-12", "2015-16", "2020-21", "2021-22"),
  holdout_seasons = "2025-26",
  application_seasons = character()
)

selection
```

The four season sets are checked for consistency and are carried through the governed stage artifacts.

## 5. Recommended high-level training API

For most users, `page_train()` is the simplest public entry point. It combines expert ignition annotations with the governed `train_pipeline()` lifecycle and returns a frozen kit.

### Refresh a locked specification

Use `mode = "refresh"` when the model specification has already been selected and only a fixed-spec refit is intended.

```{r}
trained <- page_train(
  data = history,
  labels = labels,
  mode = "refresh",
  prospective_holdout = "2025-26",
  exclude = c("2011-12", "2015-16", "2020-21", "2021-22"),
  n_cores = max(1L, parallel::detectCores() - 1L),
  checkpoint_dir = "results/checkpoints/refresh",
  verbose = TRUE
)

trained
page_validate_kit(trained$kit, mode = "frozen")
```

### Retune when model selection is intended

Retuning is a development operation and should be completed without looking at the prospective holdout.

```{r}
trained <- page_train(
  data = history,
  labels = labels,
  mode = "retune",
  prospective_holdout = "2025-26",
  exclude = c("2011-12", "2015-16", "2020-21", "2021-22"),
  n_cores = max(1L, parallel::detectCores() - 1L),
  checkpoint_dir = "results/checkpoints/retune",
  selection_method = "min_nll",
  verbose = TRUE
)
```

Alternative final selection methods supported by `train_pipeline()` are `"one_se"` and `"pareto"`.

## 6. Save and reload the frozen kit

Use the package helpers rather than an unvalidated bare `saveRDS()`/`readRDS()` pair.

```{r}
page_save_kit(
  trained,
  "results/kits/candidate_pre_holdout.rds"
)

kit <- page_load_kit("results/kits/candidate_pre_holdout.rds")
page_validate_kit(kit, mode = "frozen")
```

`page_save_kit()` validates the frozen kit before atomically publishing it. `page_load_kit()` validates again on load.

## 7. Advanced model-level API

The supported public training workflow is `page_train()`. Use `mode = "retune"` when PAGe should run the governed tuning and model-selection lifecycle; low-level tuning, boundary, freeze, and artifact-assembly helpers are package internals.

Researchers who need direct component access can work with fixed, explicitly chosen configurations through the stable model-level API:

```{r}
# Advanced scientific interfaces; config objects are explicit inputs.
m0 <- m0_fit(history, selection, config = m0_config)
m0_state <- m0_detect(m0_kit, current_data)

m1 <- m1_fit(history, selection, m0, config = m1_config)
m1_state <- m1_predict(m1_kit, current_data, m0_state)
peak_posterior <- m1_peak_posterior(
  library = m1_library,
  current_data = current_data,
  activation_week = ignition_week,
  origin_week = origin_week
)

m2 <- m2_fit(history, selection, m0, m1, config = m2_config)
m2_state <- m2_predict(m2_kit, current_data, m1_state)
```

For ordinary training, do not manually reproduce the internal tune/validate/freeze chain. `page_train()` owns those governance boundaries and returns the validated training result and kit. Repository-owned historical reproductions may still call internal helpers explicitly, but those helpers are not part of the supported external API.

## 8. Inspect training provenance

Keep the data identity, labels, training result, and kit identity together.

```{r}
trained$provenance
trained$labels$provenance
season_selection(trained$kit)
```

For an additional data-file checksum outside R:

```{r}
digest::digest(
  file = "/authorized/path/flu_history.csv",
  algo = "sha256",
  serialize = FALSE
)
```

## 9. Holdout acceptance and promotion are separate

A frozen pre-holdout kit is a candidate, not yet a promoted deployment kit. The governed lifecycle after training is:

```text
frozen candidate
  -> untouched holdout replay
  -> artifact-bound promotion decision
  -> fixed-spec post-acceptance refresh
  -> immutable promotion registry
  -> verified load
```

The operator commands for these phases are documented in [`deployment-workflow.qmd`](deployment-workflow.qmd). In particular, do not use a hand-constructed `check_promotion()` object to release the holdout.

A registered deployment kit is loaded with both the immutable kit path and its deployment manifest:

```{r}
promoted_kit <- load_promoted_kit(
  kit_path = "/secure/PAGe/deployment-registry/<deployment-id>/promoted_kit.rds",
  deployment_manifest_path =
    "results/deployment-audit/<deployment-id>/deployment_manifest.json"
)
```

## 10. Minimal reproducible training record

For every governed training run, retain at least:

1. the exact historical input file and SHA-256;
2. the expert ignition annotation table and annotation provenance;
3. the explicit season selection;
4. M0/M1/M2 tuning results or fixed-spec configuration;
5. checkpoints and boundary-review evidence when retuning;
6. the frozen candidate kit;
7. the package/runtime version information;
8. the later acceptance/refit/promotion manifests, if the candidate is promoted.

Useful session capture:

```{r}
sessionInfo()
packageVersion("PAGe")
```

## Next step

Once a kit is frozen or promoted, continue with [`realtime_deployment_walkthrough.qmd`](realtime_deployment_walkthrough.qmd) for current surveillance ingestion, direct R forecasting, and the v4 weekly HTTP API.
