---
title: "PAGe v3 2026–27 Walk-Forward Review: weekF8–11"
subtitle: "M0 ignition, M1 timing, and M2 forecast diagnostics"
author: "PAGe"
date: "2026-09-28"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 120)

root <- normalizePath("..", winslash = "/", mustWork = TRUE)
data_dir <- file.path(root, "artifacts", "v3-week8-11-walkforward-report-v1")

state <- read.csv(file.path(data_dir, "m0_m1_state_week8_11.csv"), stringsAsFactors = FALSE, check.names = FALSE)
m2 <- read.csv(file.path(data_dir, "m2_walkforward_week8_11.csv"), stringsAsFactors = FALSE, check.names = FALSE)
score <- read.csv(file.path(data_dir, "m2_score_to_date_week8_11.csv"), stringsAsFactors = FALSE, check.names = FALSE)
meta_raw <- read.delim(file.path(data_dir, "report_metadata.tsv"), sep = "\t", stringsAsFactors = FALSE, check.names = FALSE)
meta <- setNames(meta_raw$value, meta_raw$key)

fmt <- function(x, digits = 3) ifelse(is.na(x), "—", formatC(x, format = "f", digits = digits))
```

## Executive summary

This report shows the 2026–27 PAGe v3 walk-forward state from **weekF8 through weekF11**, the latest available week in the current OLIS panel.

> **Governance note:** weekF8–11 are **pre-issuance diagnostics**. The canonical v3 lower bound is now **weekF12**. Numerical M2 values shown here are retrospective walk-forward diagnostics generated from information available at each origin; they were **not governed issued forecasts**.

As of weekF11 (week ending `r state$week_end_date[state$origin_weekF == 11]`):

- Flu A positivity was **`r fmt(state$A_observed_pct[state$origin_weekF == 11])`%**.
- Flu B positivity was **`r fmt(state$B_observed_pct[state$origin_weekF == 11], 4)`%**.
- **M0-A had not ignited** at any origin from weekF8–11.
- **M1-A remained pre-ignition**; no A peak-timing state was active.
- **M1-B remained outside its weekF12 timing window**; no B timing event was active.
- M2 state forecasts were nevertheless computable for diagnostic replay. Through targets already observed by weekF11, the to-date MAE was **`r fmt(score$MAE_pp[score$type == "A" & score$horizon == 1])` pp for A+1** and **`r fmt(score$MAE_pp[score$type == "B" & score$horizon == 1])` pp for B+1**.

## Observed surveillance signal

```{r}
#| label: observed-table
obs <- state[, c("origin_weekF", "week_start_date", "week_end_date", "A_observed_pct", "B_observed_pct")]
names(obs) <- c("weekF", "week start", "week end", "A positivity (%)", "B positivity (%)")
obs[["A positivity (%)"]] <- round(obs[["A positivity (%)"]], 3)
obs[["B positivity (%)"]] <- round(obs[["B positivity (%)"]], 4)
knitr::kable(obs, align = c("r", "l", "l", "r", "r"))
```

```{r}
#| label: observed-plot
#| fig-cap: "Observed A and B positivity through weekF11."
matplot(
  state$origin_weekF,
  cbind(state$A_observed_pct, state$B_observed_pct),
  type = "o", pch = c(16, 17), lty = 1, col = c("#D55E00", "#0072B2"),
  xlab = "weekF", ylab = "positivity (%)", xaxt = "n"
)
axis(1, at = state$origin_weekF)
legend("topleft", legend = c("Flu A", "Flu B"), col = c("#D55E00", "#0072B2"), pch = c(16, 17), lty = 1, bty = "n")
grid(col = "grey90")
```

## M0: A ignition walk-forward

M0 was evaluated prospectively at each origin using only data available through that week. Across weekF8–11, the detector did **not** fire.

```{r}
#| label: m0-table
m0 <- state[, c("origin_weekF", "M0_A_ignited", "M0_A_ignition_weekF")]
names(m0) <- c("origin weekF", "A ignited?", "estimated ignition weekF")
m0[["estimated ignition weekF"]] <- ifelse(is.na(m0[["estimated ignition weekF"]]), "—", format(m0[["estimated ignition weekF"]]))
knitr::kable(m0, align = c("r", "c", "r"))
```

Interpretation: the increase in A positivity by weekF9–10 was not sufficient to satisfy the frozen M0 ignition rules. There was therefore no M0-to-M1 ignition handoff during these origins.

## M1: timing state

### Influenza A

Because M0 had not ignited, M1-A stayed inactive/pre-ignition throughout the review window.

### Influenza B

The canonical B timing/activity lower bound is now weekF12. Therefore weekF8–11 are explicitly **pre-window**, with no B timing event or passage posterior available.

```{r}
#| label: m1-table
m1 <- state[, c(
  "origin_weekF",
  "M1_A_state", "M1_A_peak_weekF", "M1_A_prob_peak_passed",
  "M1_B_state", "M1_B_activity_weekF", "M1_B_reason"
)]
names(m1) <- c(
  "origin weekF",
  "M1-A state", "A peak weekF", "A P(peak passed)",
  "M1-B state", "B activity weekF", "B reason"
)
m1[["A peak weekF"]] <- fmt(m1[["A peak weekF"]], 2)
m1[["A P(peak passed)"]] <- fmt(m1[["A P(peak passed)"]], 3)
m1[["B activity weekF"]] <- fmt(m1[["B activity weekF"]], 2)
knitr::kable(m1, align = c("r", "l", "r", "r", "l", "r", "l"))
```

## M2: walk-forward forecast diagnostics

The table below reports what the frozen state models would have predicted using only information available at each origin. These are **diagnostic replay values**, not issued forecasts, because all origins are before weekF12.

- A +1/+2 route: exact A1 state model.
- B +1 route: exact B1 state model.
- B +2 route: exact B1 fallback because B timing is unavailable before weekF12.

```{r}
#| label: m2-table
m2_show <- m2[, c(
  "origin_weekF", "origin_week_end", "type", "horizon", "target_weekF",
  "forecast_pct", "observed_target_pct", "error_pp", "abs_error_pp", "route"
)]
names(m2_show) <- c(
  "origin", "origin week end", "type", "horizon", "target",
  "forecast (%)", "observed (%)", "error (pp)", "abs error (pp)", "route"
)
for (nm in c("forecast (%)", "observed (%)", "error (pp)", "abs error (pp)")) {
  m2_show[[nm]] <- fmt(m2_show[[nm]], 4)
}
knitr::kable(m2_show, align = c("r", "l", "c", "r", "r", "r", "r", "r", "r", "l"))
```

### Score to date

Only targets already observed by weekF11 can be scored. Therefore:

- weekF8 contributes +1 and +2 targets;
- weekF9 contributes +1 and +2 targets;
- weekF10 contributes +1 only;
- weekF11 forecasts are not yet scoreable.

```{r}
#| label: score-table
score_show <- score
names(score_show) <- c("type", "horizon", "MAE (pp)", "n scored")
score_show[["MAE (pp)"]] <- round(score_show[["MAE (pp)"]], 4)
knitr::kable(score_show, align = c("c", "r", "r", "r"))
```

```{r}
#| label: score-bars
#| fig-cap: "Walk-forward MAE to date for targets realized by weekF11."
labels <- paste0(score$type, "+", score$horizon)
barplot(
  score$MAE_pp,
  names.arg = labels,
  col = ifelse(score$type == "A", "#D55E00", "#0072B2"),
  ylab = "MAE (percentage points)", xlab = "target",
  ylim = c(0, max(score$MAE_pp) * 1.2)
)
grid(nx = NA, ny = NULL, col = "grey90")
```

## Origin-by-origin notes

```{r}
#| label: origin-notes
notes <- lapply(8:11, function(w) {
  st <- state[state$origin_weekF == w, , drop = FALSE]
  mm <- m2[m2$origin_weekF == w, , drop = FALSE]
  data.frame(
    weekF = w,
    A_pct = st$A_observed_pct,
    B_pct = st$B_observed_pct,
    M0_A = ifelse(st$M0_A_ignited, paste0("ignited @ ", st$M0_A_ignition_weekF), "not ignited"),
    M1_A = st$M1_A_state,
    M1_B = paste(st$M1_B_state, st$M1_B_reason, sep = " / "),
    A_h1 = mm$forecast_pct[mm$type == "A" & mm$horizon == 1],
    A_h2 = mm$forecast_pct[mm$type == "A" & mm$horizon == 2],
    B_h1 = mm$forecast_pct[mm$type == "B" & mm$horizon == 1],
    B_h2 = mm$forecast_pct[mm$type == "B" & mm$horizon == 2],
    stringsAsFactors = FALSE
  )
})
notes <- do.call(rbind, notes)
for (nm in c("A_pct", "B_pct", "A_h1", "A_h2", "B_h1", "B_h2")) notes[[nm]] <- round(notes[[nm]], 4)
names(notes) <- c("weekF", "A obs %", "B obs %", "M0-A", "M1-A", "M1-B", "A+1 %", "A+2 %", "B+1 %", "B+2 %")
knitr::kable(notes, align = c("r", "r", "r", "l", "l", "l", "r", "r", "r", "r"))
```

## Legacy / v2 / v3 context

These weekF8–11 rows are **v3 pre-issuance diagnostics**, not issued forecasts. Historical model comparisons therefore remain separate from this prospective report.

- On the matched M1 common ledger in `artifacts/m1-v2-lowrank-posterior-v10-release-consistent/common_ledger_vs_legacy_v16_strict_future_release.csv`, v2 mean absolute peak error is about **1.093 weeks** versus **1.375 weeks** for legacy across 81 common origin/truth rows in 10 seasons. This is an M1 timing comparison only.
- A direct matched-target-ledger legacy/v2 M2-A benchmark is available under `artifacts/m2-v2-legacy-matched-a-v1/`; its primary 10-season comparison uses identical target counts but not perfectly identical training vintages.
- Canonical v2/v3 operational comparisons begin at weekF12 and must come from `2026/run_weekly_shadow_release_v5.R`, which runs both systems from the same archived source and effective typed panel. Transaction deltas describe model-output differences, not forecast accuracy until the target is observed.
- No verified common-ledger legacy/v2 M0 performance comparison is claimed here.

### WeekF12 A-ignition diagnostic

The frozen M0-A detector uses `w_min=12` and a 1-SE sampling-noise tolerance for recent raw-positivity drops. The weekF10→11 decrease is therefore still classified as raw-step stable. A direct diagnostic that appends the current M2 weekF12 A projection (~1.677%) to the observed weekF1–11 history returns **M0 ignition at weekF12**. This is a counterfactual model-state check, not an observed weekF12 result.

See `docs/v3-legacy-v2-v3-comparison-plan-2026-09-28.md` for the comparison contract.

## Governance and provenance

```{r}
#| label: provenance-table
prov <- data.frame(
  item = c(
    "Season",
    "Walk-forward origins",
    "Latest available weekF",
    "Latest available week end",
    "Canonical forecast release",
    "Typed-panel SHA-256",
    "Effective-panel SHA-256",
    "Governed minimum forecast origin",
    "Report status"
  ),
  value = c(
    meta[["season"]], meta[["origins"]], meta[["latest_weekF"]], meta[["latest_week_end"]],
    meta[["forecast_release_id"]], meta[["panel_sha256"]], meta[["effective_panel_sha256"]],
    meta[["governed_min_origin_week"]], meta[["report_status"]]
  ),
  stringsAsFactors = FALSE
)
knitr::kable(prov, align = c("l", "l"))
```

The model-state and M2 diagnostic CSVs are generated by:

`../scripts/v3_generate_week8_11_walkforward_report_data.R`

The underlying data products are stored under:

`../artifacts/v3-week8-11-walkforward-report-v1/`

This report deliberately does not reinterpret pre-weekF12 model values as prospective issued forecasts. Its purpose is to show how M0, M1, and M2 evolved walk-forward before the governed issuance window opened.
